### 2018-09-18 doc/network.tex
new repo..

\documentclass{article}
\title{Inferring generation-interval distributions from contact tracing data}
\author{Sang Woo Park, David Champredon and Jonathan Dushoff}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\usepackage{hyperref}
\usepackage{natbib}
\usepackage{hyperref}
\bibliographystyle{chicago}
\date{\today}

\usepackage{xspace}
\newcommand*{\ie}{i.e.\@\xspace}

\usepackage{color}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\begin{document}
\maketitle

\section{Introduction}

An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic [CITE].
However, directly measuring the reproductive number often requires knowledge of entire disease history and may not be feasible early in an epidemic [CITE].
Instead, the reproductive number can be \emph{indirectly} estimated from exponential grwoth rate, which can easily be estimated from incidence data [CITE].
These two quantities are linked by generation interval distributions \citep{wallinga2007generation}.

Generation interval is defined as the time between when a person becomes infected and when that person infects another person.
As each individual experiences different course of infection, \emph{individual} generation interval distribution varies among infectors [CITE sven].
Hence, the \emph{intrinsic} generation interval distribution, which provides link between $r$ and \RR, is a pooled distribution across all potential infections.

Due to individual variation in infection time, the observed generation interval distribution can change depending on when and how it is measured \citep{champredon2015intrinsic}. \jd{Other cites first. Maybe cite us later when talking about qualitative effects. We need to talk clearly also about censored intervals}
There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals are measured by evaluating the infectiousness of infected people,
%% without regard to actual infection events,
while \emph{observed} generation intervals refer to the time between actual infection events.
Observed generation intervals in turn can be \emph{aggregated} across time, measured \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others, or measured \emph{backward in time} by look at a cohort and asking when their infectors were infected.
When aggregated generation intervals are observed until a given time in an ongoing epidemic, we refer to these as \emph{censored} (or right-censored) intervals.

Early in the epidemic when depletion of susceptible is negligible, we expect the forward generation interval distribution to be similar to the intrinsic generation interval distribution.
As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
\swp{While this statement is true, it relies on the assumption that forward generation will be measured after all infected individuals that are infected early have recovered. We should note (and maybe explore the idea) that forward GI may be affected by censoring.}

Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution.
When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long.

In practice, generation intervals are measured by contact tracing throughout the course of an epidemic, usually aggregated to increase the amount of information available. 
\jd{I kind of feel like, if you go on forever, aggregated GIs approach intrinsic GIs. This obviously isn't true for a single epidemic (observed GIs are shorter overall). On the other hand, it obviously \emph{is} true if you add births and get to a stable equilibrium. So I guess the remaining question is, do we get convergence if we have stable cycles? What we think about this will affect how we write the rest of this P.} 
If observation goes on throughout the disease cycle, then (something about something).
On the other hand, if calculations are being done based on early-outbreak data, the available censored data is best thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions.

\swp{Need a paragraph about spatial effect?}
\jd{Yes, we do!}

In this study, we explore the temporal and spatial variation in the observed generation interval obtained from contact tracing.
We show that using the observed generation interval distribution directly will always underestimate the reproductive number \swp{need to confirm this statement when we're done with the ms}. 
We provide a statistical framework of recovering the intrinsic generation interval distribution from the observed generation interval distribution.

\swp{Going to rename the sections... but I think I like the current order...}
\section{Measuring generation intervals - Theory}

Following [CITE], let $K(t)$ be the infection kernel. 
The reproduction number is defined as
$$
\RR = \int_0^\infty K(t).
$$
Then, the intrinsic generation interval distributions is defined as
$$
g(t) = \frac{K(t)}{\RR}.
$$
Intrinsic generation-interval distribution can be considered as an intrinsic characteristic of a single average infector in a fully susceptible population.

\swp{Insert renewal equation approach}

\subsection{Right-censored interval}

Assume that contact tracing is performed from the beginning of an epidemic to time $t$.
The number of infection occuring at time $s$ caused by infectors who were themselves infected at time $s-\tau$ is given by
\begin{equation}
i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
\end{equation}
Then, total number of secondary infections that are $\tau$ time steps apart and occur before time $t$:
\begin{equation}
\RR \int_\tau^t i(s-\tau) g(\tau) S(s) ds.
\end{equation}
Then, the censored interval at time $t$ is given by
\begin{equation}
c_t(\tau)= \frac{\RR \int_\tau^t i(s-\tau) g(\tau) S(s) ds}{\RR \int_0^t \int_x^t i(s-x) g(x) S(s) ds dx}.
\end{equation}
We note that the expression in the denominater is equivalent to cumulative incidence at time $t$.
The intuition behind this is that we are normalizing acrosss all incidence before time $t$.
Then, we have
\begin{equation}
c_t(\tau) = \frac{\RR \int_\tau^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t i(s) ds}.
\end{equation}
For convenience, we ignore normalizing constants and write
\begin{equation}\label{eq:obsg}
c_t(\tau) \propto g(\tau) \int_{0}^t i(s-\tau) S(s) ds.
\end{equation}

For a single epidemic, the observed mean generation interval through contact tracing will always be shorter than intrinsic mean generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, as contact tracing is performed up to time $t$, any infection events that occur after time $t$ are not observed.
Any individuals that are infected before time $t$ can only complete infection events that are shorter than time $t$ and the distribution of all infection events that occur before time $t$ will be concentrated on shorter intervals.
In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), the censored generation interval distribution is equivalent to the inverse exponentially weighted intrisnc generation interval distribution:
\begin{equation}
\tsub{c}{exp}(\tau) \propto g(\tau) \exp(-r\tau),
\end{equation}
Second, number of susceptibles decrease over the course of an epidemic and any infector is less likely to infect susceptible individuals through long generation intervals than it would have in a fully susceptible population \citep{champredon2015intrinsic}.
As a result, even if contact tracing is performed through an entire epidemic, mean generation interval will be underestimated.

\begin{figure}
% \includegraphics[width=\textwidth]{../contact_trace/temporal_effect.pdf}
\caption{Fill this out and re-do simulation with realistic Ebola parameters for consistency.}
\label{fig:censor}
\end{figure}

When an epidemic is in an equilibrium state (i.e., incidence $i(t)$ and the number of susceptibles $S(t)$ is constant over time), the observed generation interval distribution is equivalent to the intrinsic generation interval distribution. \swp{TODO: make a figure}
On the other hand, when an epidemic is in a cycle, we get ???

\subsection{Local depletion of susceptibles}

\swp{cite trapman somewhere}
The intrinsic generation-interval distribution is often defined as the distribution of times at which secondary infections occur [CITE].
This definition implicitly assumes that all infectious contacts made give rise to infection. 
Given limited contact network, depletion of susceptibles occur locally and the same person can be contacted multiple times.
Infectious contacts give rise to an infection only when the susceptible individual is contacted for the first time.
Therefore, the intrinsic generation-interval distribution is the distribution of times at which infectious contacts are made.

The \emph{effective} generation-interval distribution -- the distribution of times at which secondary infections occur -- depends on the probability that a susceptible individual has not been contacted yet and the intrinsic generation-interval distribution.
Given a constant per pair contact rate, $\beta(t)$, the effective generation-interval distribution is given by \swp{Need to double check this and/or make it more general.}
\begin{equation}
\tsub{g}{eff}(\tau) \propto g(\tau) \exp\left(-\int_0^{\tau} \beta(t) dt\right).
\end{equation}
Typically, when homogeneous contact and constant contact rates are assumed,
the per pair contact rate is sufficiently small ($\RR/N$ where $N$ is the population size) that the effect of local depletion of susceptibles is negligible.

The difference between effective and intrinsic generation-interval distribution can be best demonstrated by using a tree network: a single infector connected to small number of susceptible individuals (Figure ???).

\section{Statistical approach to estimating generation-interval distribution}

The observed generation-interval distribution is a weighted intrinsic generation-interval distribution (equation~\ref{eq:obsg}),
Then, the intrinsic generation interval can be recovered by taking the inverse weights:
\begin{equation}
g(\tau) \propto g_t(\tau) \frac{1}{\int_{0}^t i(s-\tau) S(s) ds}
\end{equation}
However, this method requires a knowledge of susceptible dynamics and may not be feasible in practice.

During exponential growth period, we can write $i(\tau) \propto \exp(r t)$, where $r$ is the exponential growth rate.
Assuming that $S(t) \approx 1$, the observed generation-interval distribution during growth period can be written as follows:
\begin{equation}
\tsub{g}{exp}(\tau) \propto g(\tau) \exp(-r\tau),
\end{equation}
Taking the inverse weight, we obtain the following expression for the intrinsic generation-interval distribution:
\begin{equation}
g(\tau) \propto \tsub{g}{exp}(\tau) \exp(r\tau)
\end{equation}
and the reproductive number:
\begin{equation}
\RR = \int_0^\infty \tsub{g}{exp}(\tau) \exp(r\tau) d\tau.
\end{equation}
Therefore, applying the Lotka-Euler equation using the observed generation interval via contact tracing results in underestimation of reproductive number:
\begin{equation}
\int_0^\infty \tsub{g}{exp}(\tau) \exp(r\tau) d\tau > \left(\int_0^\infty \tsub{g}{exp}(\tau) \exp(-r\tau) d\tau\right)^{-1}
\end{equation}
This method provides a non-parametric approach for inferring the intrinsic generation-interval distribution and the reproductive number from contact tracing data.

\swp{Need to rewrite this section. It's merely a place holder for now:}
While the non-parametric method is simple, it does not use all available information from contact tracing data.
In particular, it does not take into account who infected whom.
Assuming a poisson process, we can obtain a likelihood for observing infections:
\begin{equation}
\RR^{n_e} \cdot \prod g(\tau_e) \cdot \exp \left(- \RR \int_0^{c - t_{\tiny\textrm{inf}}} g(s) ds \right)
\end{equation}
This method requires us to make an assumption about the generation-interval distributions... See example...

\section{Methods}

\bibliography{network}
\end{document}

### 2018-09-18 analysis/temporal_effect.R

@@ -1,0 +1,106 @@
+source("ebola_param_seminr.R")
+library(deSolve)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+intrinsic_fun <- function(tau) {
+    sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+## ode
+
+seir.ode <- function(t, y, parms) {
+    with(as.list(c(y, parms)), {
+        dS <- -beta/N*S*I
+        dE1 <- beta/N*S*I - m.sigma*sigma*E1
+        dE2 <- m.sigma*sigma*E1 - m.sigma*sigma*E2
+        dI <- m.sigma*sigma*E2 - gamma*I
+        list(c(dS, dE1, dE2, dI))
+    })
+}
+
+parms <- c(beta=beta, sigma=sigma, m.sigma=m.sigma, gamma=gamma, N=N)
+y <- c(S = N-1, E1=0, E2=0, I=10)
+
+dt <- 1
+tmax <- 300
+t <- seq(0, tmax, dt)
+out <- as.data.frame(ode(y, t, seir.ode, parms))
+names(out)[1] <- "t"
+
+incidence <- beta*out$S*out$I/N
+c.incidence <- cumsum(incidence)
+
+out$incidence <- incidence
+
+genden <- hist(gen, breaks=t, plot=FALSE)$density
+
+censor <- function(tau, tmax) {
+    m <- which(t==tmax)
+    total <- c.incidence[m]
+    
+    i <- which(t==tau)
+    
+    beta*sum(incidence[1:(m-i+1)] * out$S[i:m]/N)/total*genden[i]
+}
+
+mean.censor <- function(tmax) {
+    if (tmax<=0){
+        return(0)
+    } else{
+        t <- t[t <= tmax]
+        
+        dc <- sapply(t, censor, tmax=tmax)
+        return(sum(dc*t)/sum(dc))
+    }
+}
+
+tau <- seq(0, tmax, 5)
+cm <- sapply(tau, mean.censor)
+
+censor.df <- data.frame(
+    t=tau
+    , mean=cm
+)
+
+exp.df <- data.frame(
+    t=t[which(t==100)],
+    incidence=incidence[which(t==100)],
+    mean=cm[which(tau==100)]
+)
+
+out.shade <- rbind(out[out$t <= 100, c("t", "incidence")], c(100, 0), c(0,0))
+colnames(out.shade)[1] <- "t"
+
[…]

### 2018-09-18 analysis/ebola_param_seminr.R
update

@@ -1,0 +1,22 @@
+set.seed(101)
+beta <- 2/5
+sigma <- 1/11.4
+m.sigma <- round(1.75)
+gamma <- 1/5
+n.gamma <- round((5/4.7)^2)
+N <- 60000
+
+true.mean <- 1/sigma + 1/gamma
+
+nsamp <- 1e6
+
+lat <- rgamma(nsamp, m.sigma, sigma*m.sigma)
+inf <- rgamma(nsamp, n.gamma, gamma*n.gamma)
+
+gen <- lat + runif(nsamp, 0, sample(inf, prob=inf, replace=TRUE)) 
+
+true.var <- var(gen)
+
+true.shape <- true.mean^2/true.var
+
+true.R <- beta/gamma

### 2018-09-18 analysis/temporal_effect.R

@@ -1,12 +1,9 @@
-source("ebola_param_seminr.R")
 library(deSolve)
 library(dplyr)
 library(ggplot2); theme_set(theme_bw())
 library(gridExtra)
 
-intrinsic_fun <- function(tau) {
-    sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
-}
+source("ebola_param_seminr.R")
 
 ## ode
 
@@ -20,85 +17,110 @@ seir.ode <- function(t, y, parms) {
     })
 }
 
+## we want to simulate equilibrium state!
+seirs.ode <- function(t, y, parms) {
+	with(as.list(c(y, parms)), {
+		dS <- gamma * I -beta/N*S*I
+		dE1 <- beta/N*S*I - m.sigma*sigma*E1
+		dE2 <- m.sigma*sigma*E1 - m.sigma*sigma*E2
+		dI <- m.sigma*sigma*E2 - gamma*I
+		list(c(dS, dE1, dE2, dI))
+	})
+}
+
+sumfun <- function(out,
+				   tmax,
+				   t,
+				   tauby=5) {
+	incidence <- beta*out$S*out$I/N
+	c.incidence <- cumsum(incidence)
+	
+	out$incidence <- incidence
+	
+	genden <- hist(gen, breaks=t, plot=FALSE)$density
+	
+	censor <- function(tau, tmax) {
+		m <- which(t==tmax)
+		total <- c.incidence[m]
+		
+		i <- which(t==tau)
+		
+		beta*sum(incidence[1:(m-i+1)] * out$S[i:m]/N)/total*genden[i]
+	}
+	
+	mean.censor <- function(tmax) {
+		if (tmax<=0){
+			return(0)
+		} else{
+			t <- t[t <= tmax]
+			
+			dc <- sapply(t, censor, tmax=tmax)
+			return(sum(dc*t)/sum(dc))
+		}
+	}
+	
+	tau <- seq(0, tmax, tauby)
+	cm <- sapply(tau, mean.censor)
+	
+	censor.df <- data.frame(
+		t=tau
+		, mean=cm
+	)
+}
+
 parms <- c(beta=beta, sigma=sigma, m.sigma=m.sigma, gamma=gamma, N=N)
-y <- c(S = N-1, E1=0, E2=0, I=10)
+y <- c(S = N-10, E1=0, E2=0, I=10)
 
-dt <- 1
[…]

### 2018-09-19 doc/network.tex
update
need to sleep

@@
 \author{Sang Woo Park, David Champredon and Jonathan Dushoff}
+
+\usepackage{graphics}
+\usepackage{adjustbox}
 
@@
 \begin{figure}
-% \includegraphics[width=\textwidth]{../contact_trace/temporal_effect.pdf}
-\caption{Fill this out and re-do simulation with realistic Ebola parameters for consistency.}
+\includegraphics[width=\textwidth]{../analysis/temporal_effect.pdf}
+\caption{Fill this out.}
 \label{fig:censor}
@@
 This definition implicitly assumes that all infectious contacts made give rise to infection. 
-Given limited contact network, depletion of susceptibles occur locally and the same person can be contacted multiple times.
+Given limited contact network, a susceptible individual can be contacted multiple times and the depletion of susceptibles occur locally.
 Infectious contacts give rise to an infection only when the susceptible individual is contacted for the first time.
-Therefore, the intrinsic generation-interval distribution is the distribution of times at which infectious contacts are made.
+Therefore, the intrinsic generation-interval distribution is actually the distribution of times at which infectious contacts are made.
 
-The \emph{effective} generation-interval distribution -- the distribution of times at which secondary infections occur -- depends on the probability that a susceptible individual has not been contacted yet and the intrinsic generation-interval distribution.
-Given a constant per pair contact rate, $\beta(t)$, the effective generation-interval distribution is given by \swp{Need to double check this and/or make it more general.}
+The \emph{effective} generation-interval distribution -- the distribution of times at which secondary infections occur from a single primary infector -- depends on the probability that a susceptible individual has not been contacted yet and the intrinsic generation-interval distribution.
+Given a per pair contact rate $\beta(t)$ that does not vary across pairs for a single primary infector, the effective generation-interval distribution is given by \swp{Need to double check this and/or make it more general.}
 \begin{equation}
@@
 Typically, when homogeneous contact and constant contact rates are assumed,
-the per pair contact rate is sufficiently small ($\RR/N$ where $N$ is the population size) that the effect of local depletion of susceptibles is negligible.
+the per pair contact rate is sufficiently small ($\beta \approx \RR/N$ where $N$ is the population size) that the effect of multiple contacts is negligible.
 
-The difference between effective and intrinsic generation-interval distribution can be best demonstrated by using a tree network: a single infector connected to small number of susceptible individuals (Figure ???).
+The difference between effective and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a tree network with artificially high $\RR$ (Figure ???
+to be made).
+We observe that the distribution of generation times matches with the effective generation-interval distribution, whereas the distribution of contact times matches with the intrinsic generation-interval distribution.
+On the other hand, if a stochastic infection processes are simulated on a small homogeneous network (e.g., a household) with all else being equal, the distribution of generation times is even shorter than the effective generation intervals because a susceptible individual may be contacted by multiple infectors (and only the first contact gives rise to infection).
+Hence, we expect there to be varying degrees of local depletion of susceptibles in a heterogeneous contact structure and the accurate measurement of generation interval will be more difficult.
+
+\subsection{Spatial spread of an epidemic}
 

### 2018-09-19 analysis/local_effect.R
working on it

@@ -1,0 +1,55 @@
+library(igraph)
+source("ebola_param_seminr.R")
+source("../R/seminr.R")
+source("../R/generation.R")
+
+beta <- 1
+
+N <- 5
+
+nsim <- 1000
+
+set.seed(101)
+reslist_full <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+	rr <- seminr.full(N, beta/(N-1), sigma, m.sigma, gamma, n.gamma, I0 = 1, imax=10,
+					  keep.intrinsic = TRUE)
+	reslist_full[[i]] <- rr
+	i <- i+1
+}
+
+g <- graph.tree(5, children=4)
+
+set.seed(101)
+reslist_local <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+	rr <- seminr(g, beta/(N-1), sigma, m.sigma, gamma, n.gamma, initial_infected=1, imax=10,
+				 keep.intrinsic = TRUE)
+	
+	if (nrow(rr$data) > 1) {
+		reslist_local[[i]] <- rr
+		i <- i+1
+	}
+}
+
+gendata_full <- lapply(reslist_full, network.generation, type="backward", plot=FALSE)
+gendata_full <- unlist(gendata_full)
+gendata_local <- lapply(reslist_local, network.generation, type="backward", plot=FALSE)
+gendata_local <- unlist(gendata_local)
+mean(gendata_full)
+mean(gendata_local)
+
+t <- seq(0, 110, 1)
+genden <- hist(gen, breaks=t, plot=FALSE)$density
+
+
+hist(gendata_local, freq=FALSE, breaks=30)
+lines(head(t, -1), genden)
+
+true.mean <- 1/sigma + 1/gamma
+
+
+
+

### 2018-09-19 analysis/measles.R

@@ -1,0 +1,147 @@
+library(deSolve)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+intrinsic_fun <- function(tau, parm) {
+	with(as.list(parm), {
+		sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+	})
+}
+
+## ode
+
+measles.ode <- function(t, y, parms) {
+	with(as.list(c(y, parms)), {
+		
+		beta <- b0 * (1 + b1 * cos(2 * pi * t/period))
+		
+		incidence <- beta * S * I/N
+		
+		dS <- mu * (N-S) - incidence
+		dE <- incidence - (sigma+mu)*E
+		dI <- sigma*E - (gamma+mu)*I
+		dR <- (gamma+mu)*I - mu*R
+		list(c(dS, dE, dI, dR))
+	})
+}
+
+sumfun <- function(out,
+				   tmax,
+				   t,
+				   parm,
+				   tauby=5) {
+	incidence <- out$incidence
+	c.incidence <- cumsum(incidence)
+	
+	genden <- intrinsic_fun(t, parm)
+	
+	censor <- function(tau, tmax) {
+		m <- which(t==tmax)
+		total <- c.incidence[m]
+		
+		i <- which(t==tau)
+		
+		sum(incidence[1:(m-i+1)] * out$S[i:m]/parm[["N"]])*genden[i]
+	}
+	
+	mean.censor <- function(tmax) {
+		if (tmax<=0){
+			return(0)
+		} else{
+			t <- t[t <= tmax]
+			
+			dc <- sapply(t, censor, tmax=tmax)
+			return(sum(dc*t)/sum(dc))
+		}
+	}
+	
+	tau <- seq(0, tmax, tauby)
+	cm <- sapply(tau, mean.censor)
+	
+	censor.df <- data.frame(
+		t=tau
+		, mean=cm
+	)
+	
+	censor.df
+}
+
+parms <- c(
+	b0=1250/365,
+	b1=0.15,
+	sigma=1/8,
+	gamma=1/5,
+	mu=1/50/365,
+	period=365,
+	N=5e7
+)
+
+y <- c(S=0.05, E=0.0001, I=0.0001, R=1-0.05-0.0001-0.0001) * 5e7
+
+dt <- 1
+dt2 <- 0.2
[…]

### 2018-09-19 analysis/measles_param.R

@@ -1,0 +1,23 @@
+set.seed(101)
+beta <- 2/5
+mu <- 1/50
+sigma <- 1/11.4
+m.sigma <- round(1.75)
+gamma <- 1/5
+n.gamma <- round((5/4.7)^2)
+N <- 5e7
+
+true.mean <- 1/sigma + 1/gamma
+
+nsamp <- 1e6
+
+lat <- rgamma(nsamp, m.sigma, sigma*m.sigma)
+inf <- rgamma(nsamp, n.gamma, gamma*n.gamma)
+
+gen <- lat + runif(nsamp, 0, sample(inf, prob=inf, replace=TRUE)) 
+
+true.var <- var(gen)
+
+true.shape <- true.mean^2/true.var
+
+true.R <- beta/gamma

### 2018-09-19 doc/network.tex
Reorganizing to be more make-y (Sorry!)

@@
 Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic [CITE].
-However, directly measuring the reproductive number often requires knowledge of entire disease history and may not be feasible early in an epidemic [CITE].
-Instead, the reproductive number can be \emph{indirectly} estimated from exponential grwoth rate, which can easily be estimated from incidence data [CITE].
-These two quantities are linked by generation interval distributions \citep{wallinga2007generation}.
+However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
+Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can easily be estimated from incidence data [CITE].
+These two quantities are linked by generation-interval distributions \citep{wallinga2007generation}.
 
 Generation interval is defined as the time between when a person becomes infected and when that person infects another person.
-As each individual experiences different course of infection, \emph{individual} generation interval distribution varies among infectors [CITE sven].
-Hence, the \emph{intrinsic} generation interval distribution, which provides link between $r$ and \RR, is a pooled distribution across all potential infections.
+As each individual experiences different course of infection, \emph{individual} generation-interval distribution varies among infectors [CITE sven].
+The \emph{intrinsic} generation-interval distribution, which provides link between $r$ and \RR, is a pooled distribution across all potential infections.
+\jd{I'd like a more mechanistic definition -- it reflects intrinsic infectiousness and is often used by modelers. What do we say in the other MS?}
+\jd{Maybe we don't need this s. What is it trying to do? See below.}
 
-Due to individual variation in infection time, the observed generation interval distribution can change depending on when and how it is measured \citep{champredon2015intrinsic}.
+Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{champredon2015intrinsic}.
 \jd{Other cites first.
@@
 
-Early in the epidemic when depletion of susceptible is negligible, we expect the forward generation interval distribution to be similar to the intrinsic generation interval distribution.
+Early in the epidemic when depletion of susceptible is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
 As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
+\jd{See how this relates to where you talk about similar shortening below.}
 \swp{While this statement is true, it relies on the assumption that forward generation will be measured after all infected individuals that are infected early have recovered.
@@
 On the other hand, if calculations are being done based on early-outbreak data, the available censored data is best thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions.
+\jd{See how this relates to where you talk about similar shortening above and below.}
 
@@
 In this study, we explore the temporal and spatial variation in the observed generation interval obtained from contact tracing.
-We show that using the observed generation interval distribution directly will always underestimate the reproductive number \swp{need to confirm this statement when we're done with the ms}. 
-We provide a statistical framework of recovering the intrinsic generation interval distribution from the observed generation interval distribution.
+We show that using the observed generation-interval distribution directly will always underestimate the reproductive number \swp{need to confirm this statement when we're done with the ms}. 
+We provide a statistical framework of recovering the intrinsic generation-interval distribution from the observed generation-interval distribution.
 
@@
 $$
-Then, the intrinsic generation interval distributions is defined as
+Then, the intrinsic generation-interval distributions is defined as
 $$
@@
 For a single epidemic, the observed mean generation interval through contact tracing will always be shorter than intrinsic mean generation interval (Figure~\ref{fig:censor}).
+\jd{See how this relates to where you talk about similar shortening above.}
 There are two reasons for this phenomenon.
@@
 Any individuals that are infected before time $t$ can only complete infection events that are shorter than time $t$ and the distribution of all infection events that occur before time $t$ will be concentrated on shorter intervals.
-In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), the censored generation interval distribution is equivalent to the inverse exponentially weighted intrisnc generation interval distribution:
+In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrisnc generation-interval distribution:
 \begin{equation}
@@
 
-When an epidemic is in an equilibrium state (i.e., incidence $i(t)$ and the number of susceptibles $S(t)$ is constant over time), the observed generation interval distribution is equivalent to the intrinsic generation interval distribution.
-\swp{TODO: make a figure}
-On the other hand, when an epidemic is in a cycle, we get ???
+When an epidemic is in an equilibrium state (i.e., incidence $i(t)$ and the number of susceptibles $S(t)$ is constant over time), the observed generation-interval distribution is equivalent to the intrinsic generation-interval distribution.
+On the other hand, when an epidemic is in a cycle, we get ???  \swp{TODO: Continue this exciting research!}
 
@@
 \end{equation}
-Typically, when homogeneous contact and constant contact rates are assumed,
+Typically, when homogeneous mixing is assumed,
 the per pair contact rate is sufficiently small ($\beta \approx \RR/N$ where $N$ is the population size) that the effect of multiple contacts is negligible.
@@
 On the other hand, if a stochastic infection processes are simulated on a small homogeneous network (e.g., a household) with all else being equal, the distribution of generation times is even shorter than the effective generation intervals because a susceptible individual may be contacted by multiple infectors (and only the first contact gives rise to infection).
+\jd{Let's think about how to explain this. In particular, the first one is where we have our correction that works \ldots}.
 Hence, we expect there to be varying degrees of local depletion of susceptibles in a heterogeneous contact structure and the accurate measurement of generation interval will be more difficult.

### 2018-09-21 doc/network.tex
I'm supposed to commit these to avoid conflicts or whatever
working on it

@@
 Generation interval is defined as the time between when a person becomes infected and when that person infects another person.
-As each individual experiences different course of infection, \emph{individual} generation-interval distribution varies among infectors [CITE sven].
-The \emph{intrinsic} generation-interval distribution, which provides link between $r$ and \RR, is a pooled distribution across all potential infections.
-\jd{I'd like a more mechanistic definition -- it reflects intrinsic infectiousness and is often used by modelers. What do we say in the other MS?}
-\jd{Maybe we don't need this s. What is it trying to do? See below.}
-
-Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{champredon2015intrinsic}.
-\jd{Other cites first.
-Maybe cite us later when talking about qualitative effects.
-We need to talk clearly also about censored intervals}
+Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time}.
 There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals are measured by evaluating the infectiousness of infected people,
@@
 while \emph{observed} generation intervals refer to the time between actual infection events.
-Observed generation intervals in turn can be \emph{aggregated} across time, measured \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others, or measured \emph{backward in time} by look at a cohort and asking when their infectors were infected.
-When aggregated generation intervals are observed until a given time in an ongoing epidemic, we refer to these as \emph{censored} (or right-censored) intervals.
+Observed generation intervals in turn can be \emph{aggregated} across time, measured \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others, or measured \emph{backward in time} by look at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
+% When aggregated generation intervals are observed until a given time in an ongoing epidemic (often via contact tracing), we refer to these as \emph{censored} (or right-censored) intervals.
 
@@
 \jd{See how this relates to where you talk about similar shortening below.}
-\swp{While this statement is true, it relies on the assumption that forward generation will be measured after all infected individuals that are infected early have recovered.
-We should note (and maybe explore the idea) that forward GI may be affected by censoring.}
-
 Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
@@
 When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long.
+\swp{Where can I cite champredon? First sentence? Last sentence? Or don't cite DC because I already did in the previous paragraph?}
 
-In practice, generation intervals are measured by contact tracing throughout the course of an epidemic, usually aggregated to increase the amount of information available. 
+In practice, generation intervals are measured by contact tracing throughout the course of an epidemic, usually aggregated to increase the amount of information available; we refer to these as \emph{censored} (or right-censored) intervals.
 \jd{I kind of feel like, if you go on forever, aggregated GIs approach intrinsic GIs.
@@
 So I guess the remaining question is, do we get convergence if we have stable cycles?
-What we think about this will affect how we write the rest of this P.}
+What we think about this will affect how we write the rest of this P.} \swp{Going to leave this comment here so that we can come back later.}
 If observation goes on throughout the disease cycle, then (something about something).
@@
 
-\swp{Need a paragraph about spatial effect?}
-\jd{Yes, we do!}
+Aggregating generation intervals across heterogeneous environment also affects the observed generation intervals.
+Infection is more likely to spread quicker than on average among closely related inidividuals, and the local decrease in susceptibles result in shorter realized generation intervals.
+\cite{trapman2016inferring} gave a similar argument to explain the difference in the the reproductive numbers between a homogeneously mixing population and a heterogeneously (network structured) mixing population.
+Their result can be interpreted as a consequence of changes in generation-interval distributions: shorter generation intervals lead to smaller reproductive number [CITE].
 
-In this study, we explore the temporal and spatial variation in the observed generation interval obtained from contact tracing.
+\swp{Will rewrite this paragraph after writing the rest of it:}
+In this study, we explore the temporal and spatial variation in the observed generation-interval distributions obtained from contact tracing.
 We show that using the observed generation-interval distribution directly will always underestimate the reproductive number \swp{need to confirm this statement when we're done with the ms}. 
@@
 \section{Measuring generation intervals - Theory}
+
+We begin by introducing an analytical framework.
+Let $K_a(\tau)$ be an individual level kernel, i.e., the rate at which secondary infections are caused by an individual $a$ \citep{svensson2015influence}.
+\swp{Need to say something more about Sven; come back later}
+Then, the population-level kernel is given by integrating over individual variations:
+\begin{equation}
+K(\tau) = \int K_a (\tau) dA.
+\end{equation}
 

### 2018-09-21 doc/network.tex
more

@@
 As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
-\jd{See how this relates to where you talk about similar shortening below.}
 Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
@@
 On the other hand, if calculations are being done based on early-outbreak data, the available censored data is best thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions.
-\jd{See how this relates to where you talk about similar shortening above and below.}
 
@@
 \swp{Going to rename the sections... but I think I like the current order...}
-\section{Measuring generation intervals - Theory}
+\section{Generation-interval distributions across time}
 
@@
 \swp{Need to say something more about Sven; come back later}
-Then, the population-level kernel is given by integrating over individual variations:
+Then, the unconditional, population-level kernel is given by integrating over individual variations:
 \begin{equation}
@@
 \end{equation}
+The basic reproductive number -- average number of secondary cases caused by an average primary case in a fully susceptible population -- is defined as: 
+\begin{equation}
+\RR_0 = \int K(t).
+\end{equation}
+Then, the intrinsic generation-interval is simply the population-level kernel normalized by the basic reproductive number:
+\begin{equation}
+g(\tau) = \frac{K(t)}{\RR_0}.
+\end{equation}
+Finally, the renewal equation model can be used to model incidence over time:
+\begin{equation}
+i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR S(\tau) \int g(s) i(\tau-s) ds
+\end{equation}
+where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
 
-Following [CITE], let $K(t)$ be the infection kernel. 
-The reproduction number is defined as
-$$
-\RR = \int_0^\infty K(t).
-$$
-Then, the intrinsic generation-interval distributions is defined as
-$$
-g(t) = \frac{K(t)}{\RR}.
-$$
-Intrinsic generation-interval distribution can be considered as an intrinsic characteristic of a single average infector in a fully susceptible population.
-
-\swp{Insert renewal equation approach}
-
-\subsection{Right-censored interval}
+\subsection*{Right-censored interval}
 
@@
 For a single epidemic, the observed mean generation interval through contact tracing will always be shorter than intrinsic mean generation interval (Figure~\ref{fig:censor}).
-\jd{See how this relates to where you talk about similar shortening above.}
 There are two reasons for this phenomenon.
-First, as contact tracing is performed up to time $t$, any infection events that occur after time $t$ are not observed.
-Any individuals that are infected before time $t$ can only complete infection events that are shorter than time $t$ and the distribution of all infection events that occur before time $t$ will be concentrated on shorter intervals.
-In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrisnc generation-interval distribution:
+First, any infection events that occur after the contact tracing period canonot be observed due to right censoring and the observed generation-interval will be shorter.
+In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), 
+the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrisnc generation-interval distribution:
 \begin{equation}
-\tsub{c}{exp}(\tau) \propto g(\tau) \exp(-r\tau),
+\tsub{c}{exp}(\tau) \propto g(\tau) \exp(-r\tau).
 \end{equation}
-Second, number of susceptibles decrease over the course of an epidemic and any infector is less likely to infect susceptible individuals through long generation intervals than it would have in a fully susceptible population \citep{champredon2015intrinsic}.
+As backward generation-interval distributions are always shorter than the intrinsic generation-interval distribution during the exponential growth period, their weighted average (i.e., the censored intervals) will also be shorter.
+
+Second, number of susceptibles decrease over the course of an epidemic, forward generation-interval distribution at any given period is always shorter than the intrinsic distribution \citep{champredon2015intrinsic}.
+Asymptotically, we can ignore the effect of right-censoring, and the censored generation intervals can be thought of as a weighted average of the forward generation intervals.
 As a result, even if contact tracing is performed through an entire epidemic, mean generation interval will be underestimated.
@@
 
-\subsection{Local depletion of susceptibles}
+\section{Generation-interval distributions across space}
 
-\swp{cite trapman somewhere}
-The intrinsic generation-interval distribution is often defined as the distribution of times at which secondary infections occur [CITE].
-This definition implicitly assumes that all infectious contacts made give rise to infection. 
-Given limited contact network, a susceptible individual can be contacted multiple times and the depletion of susceptibles occur locally.
-Infectious contacts give rise to an infection only when the susceptible individual is contacted for the first time.
-Therefore, the intrinsic generation-interval distribution is actually the distribution of times at which infectious contacts are made.
+Given limited contact structure, a susceptible individual can be contacted multiple times;
+infectious contacts give rise to infection only when the susceptible individual is contacted for the first time.
+Then, the conditional kernel (conditional on the infection occuring) of an individual $a$ is given by
+\begin{equation}
+\hat{k}_a(\tau) = k_a(\tau) \exp \left(- \delta_a \int_0^\tau k_a(s) ds\right),
+\end{equation}
+where $\delta_a$ is the dilution.
+The individual kernel, $k_a(s)$, measures the intrinsic infectiousness of an individual $a$,
+whereas $\delta_a k_a(s)$ measures the infectiousness of an individual $a$ for a particular susceptible individual.
+Then, the population-level conditional kernel is:
+\begin{equation}
+\hat{k}(\tau) = \int \hat{k}_a(\tau) dA,
+\end{equation}
+and the conditional generation-interval distribution, defined to be the distribution of times at which secondary infections occur from a single primary infector, is:
+\begin{equation}
+\hat{g}(\tau) = \frac{\hat{k}(\tau)}{\int \hat{k} \tau}.
+\end{equation}
+\swp{Need to say something about trapman!!!}
 
-The \emph{effective} generation-interval distribution -- the distribution of times at which secondary infections occur from a single primary infector -- depends on the probability that a susceptible individual has not been contacted yet and the intrinsic generation-interval distribution.
-Given a per pair contact rate $\beta(t)$ that does not vary across pairs for a single primary infector, the effective generation-interval distribution is given by \swp{Need to double check this and/or make it more general.}
-\begin{equation}
-\tsub{g}{eff}(\tau) \propto g(\tau) \exp\left(-\int_0^{\tau} \beta(t) dt\right).
-\end{equation}
-Typically, when homogeneous mixing is assumed,
-the per pair contact rate is sufficiently small ($\beta \approx \RR/N$ where $N$ is the population size) that the effect of multiple contacts is negligible.
-
-The difference between effective and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a tree network with artificially high $\RR$ (Figure ???
+The difference between conditional and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a small tree network, in which a single infected individual is connected to multiple susceptible individuals, with artificially high $\RR$ (Figure ???
 to be made).
-We observe that the distribution of generation times matches with the effective generation-interval distribution, whereas the distribution of contact times matches with the intrinsic generation-interval distribution.
+We observe that the distribution of realized generation times matches with the conditional generation-interval distribution, whereas the distribution of contact times matches with the intrinsic generation-interval distribution.
 On the other hand, if a stochastic infection processes are simulated on a small homogeneous network (e.g., a household) with all else being equal, the distribution of generation times is even shorter than the effective generation intervals because a susceptible individual may be contacted by multiple infectors (and only the first contact gives rise to infection).

### 2018-09-21 doc/network.tex
update

@@
 
+\subsection*{Conditional kernel}
+
 Given limited contact structure, a susceptible individual can be contacted multiple times;
@@
 \end{equation}
-\swp{Need to say something about trapman!!!}
+\swp{Need to say something about trapman!!! Maybe say this is a generalization of Trapman's work...? Is that too much?}
 
-The difference between conditional and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a small tree network, in which a single infected individual is connected to multiple susceptible individuals, with artificially high $\RR$ (Figure ???
+The difference between conditional and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a small tree network -- a single infected individual is connected to multiple susceptible individuals but susceptibles are not connected with each other -- with artificially high $\RR$ (Figure ???
 to be made).
 We observe that the distribution of realized generation times matches with the conditional generation-interval distribution, whereas the distribution of contact times matches with the intrinsic generation-interval distribution.
-On the other hand, if a stochastic infection processes are simulated on a small homogeneous network (e.g., a household) with all else being equal, the distribution of generation times is even shorter than the effective generation intervals because a susceptible individual may be contacted by multiple infectors (and only the first contact gives rise to infection).
+On the other hand, if a stochastic infection processes are simulated on a small homogeneous network with all else being equal, the distribution of generation times is even shorter than the effective generation intervals because a susceptible individual may be contacted by multiple infectors (and only the first contact gives rise to infection).
 \jd{Let's think about how to explain this. In particular, the first one is where we have our correction that works \ldots}.
 Hence, we expect there to be varying degrees of local depletion of susceptibles in a heterogeneous contact structure and the accurate measurement of generation interval will be more difficult.
-
-\subsection{Spatial spread of an epidemic}
 

### 2018-09-21 R/generation.R

@@ -1,0 +1,68 @@
+intrinsic.generation <- function(x, plot=TRUE, ...) {
+    ig <- unlist(x$intrinsic_generation)
+    if(plot) hist(ig, freq=FALSE, ...)
+    invisible(ig)
+}
+
+network.generation <- function(x, plot=TRUE,
+                               type=c("backward", "forward"),
+                               interval,
+                               interval.type=c("time", "cases"),
+                               ...) {
+    type <- match.arg(type)
+    interval.type <- match.arg(interval.type)
+    
+    if (missing(interval)) {
+        interval <- c(0, max(x$data$time))
+    } else if (interval.type=="cases") {
+        interval <- x$data$time[match(interval, x$data$infected)]
+    }
+    
+    cohort <- which(x$t_infected >= interval[1] & x$t_infected <= interval[2])
+    
+    if (type == "backward") {
+        infector <- x$infected_by[cohort]
+        ng <- x$t_infected[cohort] -  x$t_infected[infector]
+        ng <- ng[!is.na(ng)]
+    } else {
+        ng <- unlist(lapply(cohort, function(z) {
+            infectee <- which(z==x$infected_by)
+            x$t_infected[infectee] - x$t_infected[z]
+        }))
+    }
+    
+    if (plot) hist(ng, freq=FALSE, ...)
+    
+    invisible(ng)
+}
+
+generation.data <- function(x,
+                            tmax,
+                            method=c("backward", "forward")
+                            ) {
+    method <- match.arg(method)
+    
+    if (method=="backward") {
+        index <- which(x$t_infected < tmax)
+        
+        index <- index[order(x$t_infected[index])]
+        
+        data.frame(
+            index=index,
+            t_infected=x$t_infected[index],
+            infected_by=x$infected_by[index],
+            generation=x$t_infected[index]-x$t_infected[x$infected_by[index]]
+        )
+    } else {
+        index <- which(x$t_infected < tmax)
+        
+        index <- index[order(x$t_infected[index])]
+        
+        data.frame(
+            index=x$infected_by[index],
+            t_infected=x$t_infected[x$infected_by[index]],
+            infected=index,
+            generation=x$t_infected[index]-x$t_infected[x$infected_by[index]]
+        )
+    }
+}

### 2018-09-21 R/seir.R

@@ -1,0 +1,257 @@
+##' @param x a vector of nodes
+##' @param size number of nodes to pick at random
+sample2 <- function(x, size) {
+    if(length(x)==1) {
+        rep(x, size)
+    } else{
+        sample(x, size, replace=TRUE)
+    }
+}
+
+##' Simulate an epidemic until done or imax reached
+##' @param g igraph object
+##' @param beta contact rate
+##' @param sigma 1/(latent period)
+##' @param gamma 1/(infectious period)
+##' @param imax maximum number of infected individuals
+seir <- function(g,
+                 beta,
+                 sigma,
+                 gamma,
+                 I0,
+                 initial_infected,
+                 seed=NULL,
+                 imax,
+                 keep.intrinsic=FALSE){
+    if(class(g) != "igraph"){
+        stop("g must be an igraph object")
+    }
+    
+    if (!is.null(seed)) set.seed(seed)
+    
+    V <- as.vector(V(g))
+    
+    if (missing(initial_infected)) {
+        if (missing(I0)) stop("specify the initial conditions")
+        
+        initial_infected <- sample(V, I0)    
+    }
+
+    I0 <- length(initial_infected)
+    
+    if (missing(imax)) imax <- length(V)
+    
+    t <- 0
+    
+    queue_v <- queue_t <- queue_infector <- rep(NA, I0)
+    
+    queue_v[1:I0] <- initial_infected 
+    queue_t[1:I0] <- 0
+    
+    t_infected <- t_infectious <- t_recovered <- rep(NA, length(V))
+    t_infected[initial_infected] <- 0
+    
+    t_gillespie <- NULL
+    c_infected <- 0
+    
+    if (keep.intrinsic) {
+        intrinsic_generation <- vector('list', length(V))
+    } else {
+        intrinsic_generation <- NULL
+    }
+    
+    done <- rep(FALSE, length(V))
+    infected_by <- rep(NA, length(V))
+    
+    stop <- FALSE
+    
+    while (!stop) {
+        j.index <- which.min(queue_t)
+        j <- queue_v[j.index]
+        
+        infected_by[j] <- queue_infector[j.index]
+        t_infected[j] <- queue_t[j.index]
+        
+        t <- queue_t[j.index]; t_gillespie <- c(t_gillespie, t)
+        
+        latent <- rexp(1, sigma)
+        t_infectious[j] <- t + latent
+        
+        c_infected <- c_infected +1
+        
+        n <- as.vector(neighbors(g, j))
[…]

### 2018-09-21 R/seminr.R

@@ -1,0 +1,243 @@
+##' @param x a vector of nodes
+##' @param size number of nodes to pick at random
+sample2 <- function(x, size) {
+    if(length(x)==1) {
+        rep(x, size)
+    } else{
+        sample(x, size, replace=TRUE)
+    }
+}
+
+##' Simulate an epidemic until done or imax reached
+##' @param g igraph object
+##' @param beta contact rate
+##' @param sigma 1/(latent period)
+##' @param m.sigma number of boxes for latent period
+##' @param gamma 1/(infectious period)
+##' @param n.gamma number of boxes for infectious period
+##' @param imax maximum number of infected individuals
+seminr <- function(g,
+                   beta,
+                   sigma, m.sigma,
+                   gamma, n.gamma,
+                   I0,
+                   initial_infected,
+                   seed=NULL,
+                   imax,
+                   keep.intrinsic=FALSE){
+    if(class(g) != "igraph"){
+        stop("g must be an igraph object")
+    }
+    
+    if (!is.null(seed)) set.seed(seed)
+    
+    V <- as.vector(V(g))
+    
+    if (missing(initial_infected)) {
+        if (missing(I0)) stop("specify the initial conditions")
+        
+        initial_infected <- sample(V, I0)    
+    }
+
+    I0 <- length(initial_infected)
+    
+    if (missing(imax)) imax <- length(V)
+    
+    t <- 0
+    
+    queue_v <- queue_t <- queue_infector <- rep(NA, I0)
+    
+    queue_v[1:I0] <- initial_infected 
+    queue_t[1:I0] <- 0
+    
+    t_infected <- t_infectious <- t_recovered <- rep(NA, length(V))
+    t_infected[initial_infected] <- 0
+    
+    t_gillespie <- NULL
+    c_infected <- 0
+    
+    if (keep.intrinsic) {
+        intrinsic_generation <- vector('list', length(V))
+    } else {
+        intrinsic_generation <- NULL
+    }
+    
+    done <- rep(FALSE, length(V))
+    infected_by <- rep(NA, length(V))
+    
+    stop <- FALSE
+    
+    while (!stop) {
+        j.index <- which.min(queue_t)
+        j <- queue_v[j.index]
+        
+        infected_by[j] <- queue_infector[j.index]
+        t_infected[j] <- queue_t[j.index]
+        
+        t <- queue_t[j.index]; t_gillespie <- c(t_gillespie, t)
+        
[…]

### 2018-09-21 analysis/ebola_param.R

@@ -1,0 +1,16 @@
+beta <- 2/5
+sigma <- 1/11.4
+gamma <- 1/5
+N <- 60000
+
+true.mean <- 1/sigma + 1/gamma
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.cv <- sqrt(true.var)/true.mean
+
+true.R <- beta/gamma

### 2018-09-21 analysis/ebola_param_seminr.R

@@ -8,7 +8,7 @@ N <- 60000
 
 true.mean <- 1/sigma + 1/gamma
 
-nsamp <- 1e6
+nsamp <- 1e5
 
 lat <- rgamma(nsamp, m.sigma, sigma*m.sigma)
 inf <- rgamma(nsamp, n.gamma, gamma*n.gamma)

### 2018-09-21 analysis/local_effect.R

@@ -1,31 +1,29 @@
 library(igraph)
-source("ebola_param_seminr.R")
-source("../R/seminr.R")
+source("ebola_param.R")
+source("../R/seir.R")
 source("../R/generation.R")
 
 beta <- 1
 
 N <- 5
 
-nsim <- 1000
+nsim <- 100
 
 set.seed(101)
 reslist_full <- vector('list', nsim)
 i <- 1
 while (i <= nsim) {
-	rr <- seminr.full(N, beta/(N-1), sigma, m.sigma, gamma, n.gamma, I0 = 1, imax=10,
+	rr <- seir.full(N, beta/(N-1), sigma, gamma, I0 = 1, imax=10,
 					  keep.intrinsic = TRUE)
 	reslist_full[[i]] <- rr
 	i <- i+1
 }
 
-g <- graph.tree(5, children=4)
-
 set.seed(101)
 reslist_local <- vector('list', nsim)
 i <- 1
 while (i <= nsim) {
-	rr <- seminr(g, beta/(N-1), sigma, m.sigma, gamma, n.gamma, initial_infected=1, imax=10,
+	rr <- seir.local(g, beta/(N-1), sigma, m.sigma, gamma, n.gamma, initial_infected=1, imax=10,
 				 keep.intrinsic = TRUE)
 	
 	if (nrow(rr$data) > 1) {

### 2018-09-21 analysis/temporal_effect.R

@@ -3,17 +3,16 @@ library(dplyr)
 library(ggplot2); theme_set(theme_bw())
 library(gridExtra)
 
-source("ebola_param_seminr.R")
+source("ebola_param.R")
 
 ## ode
 
 seir.ode <- function(t, y, parms) {
     with(as.list(c(y, parms)), {
         dS <- -beta/N*S*I
-        dE1 <- beta/N*S*I - m.sigma*sigma*E1
-        dE2 <- m.sigma*sigma*E1 - m.sigma*sigma*E2
-        dI <- m.sigma*sigma*E2 - gamma*I
-        list(c(dS, dE1, dE2, dI))
+        dE <- beta/N*S*I - sigma*E
+        dI <- sigma*E - gamma*I
+        list(c(dS, dE, dI))
     })
 }
 
@@ -21,10 +20,9 @@ seir.ode <- function(t, y, parms) {
 seirs.ode <- function(t, y, parms) {
 	with(as.list(c(y, parms)), {
 		dS <- gamma * I -beta/N*S*I
-		dE1 <- beta/N*S*I - m.sigma*sigma*E1
-		dE2 <- m.sigma*sigma*E1 - m.sigma*sigma*E2
-		dI <- m.sigma*sigma*E2 - gamma*I
-		list(c(dS, dE1, dE2, dI))
+		dE <- beta/N*S*I - sigma*E
+		dI <- sigma*E - gamma*I
+		list(c(dS, dE, dI))
 	})
 }
 
@@ -37,7 +35,7 @@ sumfun <- function(out,
 	
 	out$incidence <- incidence
 	
-	genden <- hist(gen, breaks=t, plot=FALSE)$density
+	genden <- intrinsic_fun(t)
 	
 	censor <- function(tau, tmax) {
 		m <- which(t==tmax)
@@ -68,8 +66,8 @@ sumfun <- function(out,
 	)
 }
 
-parms <- c(beta=beta, sigma=sigma, m.sigma=m.sigma, gamma=gamma, N=N)
-y <- c(S = N-10, E1=0, E2=0, I=10)
+parms <- c(beta=beta, sigma=sigma, gamma=gamma, N=N)
+y <- c(S = N-10, E=0, I=10)
 
 dt <- 0.1
 tmax <- 305

### 2018-09-21 doc/network.tex
new figure

@@
 
-The difference between conditional and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a small tree network -- a single infected individual is connected to multiple susceptible individuals but susceptibles are not connected with each other -- with artificially high $\RR$ (Figure ???
-to be made).
+The difference between conditional and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a small tree network -- a single infected individual is connected to multiple susceptible individuals but susceptibles are not connected with each other -- with artificially high $\RR$ (Figure~\ref{fig:local}).
 We observe that the distribution of realized generation times matches with the conditional generation-interval distribution, whereas the distribution of contact times matches with the intrinsic generation-interval distribution.
@@
 Hence, we expect there to be varying degrees of local depletion of susceptibles in a heterogeneous contact structure and the accurate measurement of generation interval will be more difficult.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{../analysis/local_effect.pdf}
+\caption{Fill this out.}
+\label{fig:local}
+\end{figure}
 

### 2018-09-21 R/seir.R

@@ -252,6 +252,8 @@ seir.local <- function(n,
     c_time <- latent + cumsum(time_between)
     
     v <- sample2(1:n, ncontact)
-    
-    c(c_time[which(!duplicated(v))])
+    list(
+    	conditional=c(head(c_time, -1)[which(!duplicated(v))]),
+    	intrinsic=head(c_time,-1)
+    )
 }

### 2018-09-21 analysis/ebola_param.R

@@ -2,11 +2,16 @@ beta <- 2/5
 sigma <- 1/11.4
 gamma <- 1/5
 N <- 60000
+lambda <- beta/(N-1)
 
 true.mean <- 1/sigma + 1/gamma
 
 intrinsic_fun <- function(tau) {
 	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+conditional_fun <- function(tau) {
+	sigma*(gamma+lambda)/(gamma+lambda-sigma) * (exp(-sigma*tau)-exp(-(gamma+lambda)*tau))
 }
 
 true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]

### 2018-09-21 analysis/local_effect.R

@@ -1,20 +1,21 @@
-library(igraph)
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
 source("ebola_param.R")
 source("../R/seir.R")
 source("../R/generation.R")
 
 beta <- 1
-
 N <- 5
+lambda <- beta/(N-1)
 
-nsim <- 100
+nsim <- 5000
 
 set.seed(101)
 reslist_full <- vector('list', nsim)
 i <- 1
 while (i <= nsim) {
-	rr <- seir.full(N, beta/(N-1), sigma, gamma, I0 = 1, imax=10,
-					  keep.intrinsic = TRUE)
+	rr <- seir.full(N, beta/(N-1), sigma, gamma, I0 = 1, imax=10)
 	reslist_full[[i]] <- rr
 	i <- i+1
 }
@@ -23,31 +24,84 @@ set.seed(101)
 reslist_local <- vector('list', nsim)
 i <- 1
 while (i <= nsim) {
-	rr <- seir.local(g, beta/(N-1), sigma, m.sigma, gamma, n.gamma, initial_infected=1, imax=10,
-				 keep.intrinsic = TRUE)
+	rr <- seir.local(4, beta/(N-1), sigma, gamma)
 	
-	if (nrow(rr$data) > 1) {
-		reslist_local[[i]] <- rr
-		i <- i+1
-	}
+	reslist_local[[i]] <- rr
+	i <- i+1
 }
 
-gendata_full <- lapply(reslist_full, network.generation, type="backward", plot=FALSE)
-gendata_full <- unlist(gendata_full)
-gendata_local <- lapply(reslist_local, network.generation, type="backward", plot=FALSE)
-gendata_local <- unlist(gendata_local)
-mean(gendata_full)
-mean(gendata_local)
+gendata_full <- reslist_full %>%
+	lapply(network.generation, type="backward", plot=FALSE) %>%
+	unlist
 
-t <- seq(0, 110, 1)
-genden <- hist(gen, breaks=t, plot=FALSE)$density
+gendata_cond <- reslist_local %>%
+	lapply("[[", "conditional") %>%
+	unlist
 
+gendata_int <- reslist_local %>%
+	lapply("[[", "intrinsic") %>%
+	unlist
+	
+gendata_mean <- data.frame(
+	type=c("intrinsic", "conditional", "homogeneous"),
+	value=c(mean(gendata_int), mean(gendata_cond), mean(gendata_full))
+)
 
[…]

### 2018-09-21 doc/network.tex
add comment

@@
 \jd{Let's think about how to explain this. In particular, the first one is where we have our correction that works \ldots}.
+\swp{I think things are clearer when we introduce conditional kernel?}
 Hence, we expect there to be varying degrees of local depletion of susceptibles in a heterogeneous contact structure and the accurate measurement of generation interval will be more difficult.

### 2018-09-23 doc/network.tex
bunch of files

@@
 \begin{figure}
-\includegraphics[width=\textwidth]{../analysis/temporal_effect.pdf}
+\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
 \caption{Fill this out.}
@@
 \end{equation}
-\swp{Need to say something about trapman!!! Maybe say this is a generalization of Trapman's work...? Is that too much?}
+\swp{Need to say something about trapman. Maybe say this is a generalization of Trapman's work...? Is that too much? Do we need to give an example of what conditional kernels look like?}
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{../analysis/local_effect.pdf}
+\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
 \caption{Fill this out.}
@@
 
-\section{Statistical approach to estimating generation-interval distribution}
+\section{Statistical approach to inferring generation-interval distribution}
 
-The observed generation-interval distribution is a weighted intrinsic generation-interval distribution (equation~\ref{eq:obsg}),
-Then, the intrinsic generation interval can be recovered by taking the inverse weights:
+\swp{I kind of want to keep the derivations in the methods (which comes after the discussion) and keep the description of methods simple here so that we don't scare people with the derivations...?}
+
+It is clear that aggregating generation intervals through space and time yields a distribution that is different from the intrinsic distribution.
+When the epidemic is growing, the proportion of susceptible individuals remain approximately constant and the intrinsic (or conditional) generation-interval distributions can be inferred from the observed intervals.
+In this section, we investigate two statistical methods for recovering the intrinsic (or conditional) generation-interval distribution.
+Detailed derivations of the two methods can be found in section??.
+
+\swp{maybe call it something else besides nonparametric method; it's still possible to make assumptions about the distribution and make inference}
+We refer to the first method as the nonparametric method as it does not rely on any assumptions about the generation-interval distributions.
+As the observed generation-interval distribution is a weighted distribution of the intrinsic distribution, the intrinsic distribution can be recovered by taking the inverse weight.
+Hence, the nonparametric method relies on the observed generation intervals and the exponential growth rate:
 \begin{equation}
-g(\tau) \propto g_t(\tau) \frac{1}{\int_{0}^t i(s-\tau) S(s) ds}
+g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau).
 \end{equation}
-However, this method requires a knowledge of susceptible dynamics and may not be feasible in practice.
 
-During exponential growth period, we can write $i(\tau) \propto \exp(r t)$, where $r$ is the exponential growth rate.
-Assuming that $S(t) \approx 1$, the observed generation-interval distribution during growth period can be written as follows:
+While the non-parametric method is simple, it does not use all available information from contact tracing data: it does not take into account who infected whom.
+Assuming that infection occur as a poisson process in the beginning of an epidemic, we can obtain the following likelihood for the observed infection process:
 \begin{equation}
-\tsub{g}{exp}(\tau) \propto g(\tau) \exp(-r\tau),
+\mathcal{L}(\RR, \theta) = \prod_{j=1}^N \left(\RR^{n_j} \exp \left(- \RR \int_0^{\tsub{t}{censor} - t_j} g(s;\theta) ds \right) \prod_{i=1}^{n_j} g(\tau_{i, j}; \theta) \right),
 \end{equation}
-Taking the inverse weight, we obtain the following expression for the intrinsic generation-interval distribution:
-\begin{equation}
-g(\tau) \propto \tsub{g}{exp}(\tau) \exp(r\tau)
-\end{equation}
-and the reproductive number:
-\begin{equation}
-\RR = \int_0^\infty \tsub{g}{exp}(\tau) \exp(r\tau) d\tau.
-\end{equation}
-Therefore, applying the Lotka-Euler equation using the observed generation interval via contact tracing results in underestimation of reproductive number:
-\begin{equation}
-\int_0^\infty \tsub{g}{exp}(\tau) \exp(r\tau) d\tau > \left(\int_0^\infty \tsub{g}{exp}(\tau) \exp(-r\tau) d\tau\right)^{-1}
-\end{equation}
-This method provides a non-parametric approach for inferring the intrinsic generation-interval distribution and the reproductive number from contact tracing data.
+where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation interval distribution $g$.
+A similar method was proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports and does not make assumptions about the shape of the generation-interval distribution.
 
-\swp{Need to rewrite this section. It's merely a place holder for now:}
-While the non-parametric method is simple, it does not use all available information from contact tracing data.
-In particular, it does not take into account who infected whom.
-Assuming a poisson process, we can obtain a likelihood for observing infections:
-\begin{equation}
-\RR^{n_e} \cdot \prod g(\tau_e) \cdot \exp \left(- \RR \int_0^{c - t_{\tiny\textrm{inf}}} g(s) ds \right)
-\end{equation}
-This method requires us to make an assumption about the generation-interval distributions... See example...
+\swp{I might say something about independence transmission events in the results -- see forsberg 2008...}
+
+First, we compared the four methods by simulating a stochastmic epidemic model homogeneous network with 100000 individuals using Ebola-like parameters for 15 weeks (Figure~?).
 
@@
 
+The first method is a non-parametric method.
+Recall that the observed generation-interval distribution is a weighted intrinsic generation-interval distribution (equation~\ref{eq:obsg}). 
+Then, the intrinsic generation-interval distribution can be obtained by taking the inverse weight:
+\begin{equation}
+g(\tau) \propto c_t(\tau) \frac{1}{\int_{0}^t i(s-\tau) S(s) ds}
+\end{equation}
+However, this method requires a knowledge of susceptible dynamics and may difficult to use in practice.
+Instead, during the exponential period, the generation-interval distribution is
+\begin{equation}
+g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau)
+\end{equation}
+and the reproductive number is
+\begin{equation}
+\RR = \int \tsub{c}{exp}(\tau) \exp(r\tau) d\tau.
+\end{equation}
+Note that applying the Lotka-Euler equation directly to the observed generation intervals will always underestimate the reproductive number:
+\begin{equation}
+\int_0^\infty \tsub{g}{exp}(\tau) \exp(r\tau) d\tau > \left(\int_0^\infty \tsub{g}{exp}(\tau). \exp(-r\tau) d\tau\right)^{-1}
+\end{equation}
+
+Finally, we obtain an estimator for mean generation interval and the reproductive number:
+\begin{equation}
+\begin{aligned}
+\hat{G} &= \frac{\sum_{i} \exp(r c_i) c_i}{\sum_{i} \exp(r c_i)},\\
+\hat{\RR} &= \sum_{i} \exp(r c_i),
+\end{aligned}
+\end{equation}
+where $r$ is the exponential growth rate and $c_i$ is an individual generation interval sample.
+
 \bibliography{network}

### 2018-09-23 R/estimate.R

@@ -1,0 +1,37 @@
+estfun_observed <- function(gen, r) {
+	data.frame(
+		mean=mean(gen),
+		RR=1/mean(exp(-r*gen))
+	)
+}
+
+estfun_nonparametric <- function(gen, r) {
+	data.frame(
+		mean=weighted.mean(gen, exp(r*gen)),
+		RR=mean(exp(r*gen))
+	)
+}
+
+bootfun <- function(gen, 
+					logr, logr.sd,
+					estfun,
+					nsim=1000) {
+	bootlist <- vector('list', nsim)
+	
+	for (i in 1:nsim) {
+		r <- exp(rnorm(1, logr, logr.sd))
+		
+		bootgen <- sample(gen, length(gen), replace=TRUE)
+		
+		bootlist[[i]] <- estfun(bootgen, r/7)
+	}
+	
+	bootdata <- do.call("rbind", bootlist) 
+	
+	est <- rbind(estfun(gen, exp(logr)/7), 
+				 apply(bootdata, 2, quantile, c(0.025, 0.975), na.rm=TRUE))
+
+	rownames(est) <- c("estimate", "lwr", "upr")
+	
+	est
+}

### 2018-09-23 R/seminr.R

@@ -24,6 +24,7 @@ seminr <- function(g,
                    initial_infected,
                    seed=NULL,
                    imax,
+				   tmax,
                    keep.intrinsic=FALSE){
     if(class(g) != "igraph"){
         stop("g must be an igraph object")
@@ -114,7 +115,7 @@ seminr <- function(g,
         queue_infector <- queue_infector[filter2]
         queue_t <- queue_t[filter2]
         
-        stop <- (c_infected == length(V) || all(done[queue_v]) || c_infected == imax)
+        stop <- (c_infected == length(V) || all(done[queue_v]) || c_infected == imax || t > tmax)
     }
     
     return(
@@ -141,6 +142,7 @@ seminr.full <- function(size,
                         I0,
                         seed = NULL,
                         imax,
+						tmax,
                         keep.intrinsic=FALSE){
     
     if (!is.null(seed)) set.seed(seed)
@@ -224,7 +226,7 @@ seminr.full <- function(size,
         queue_infector <- queue_infector[filter2]
         queue_t <- queue_t[filter2]
         
-        stop <- (c_infected == length(V) || all(done[queue_v]) || c_infected == imax)
+        stop <- (c_infected == length(V) || all(done[queue_v]) || c_infected == imax || t > tmax)
     }
     
     return(

### 2018-09-23 analysis/ebola_param.R

@@ -1,21 +1,0 @@
-beta <- 2/5
-sigma <- 1/11.4
-gamma <- 1/5
-N <- 60000
-lambda <- beta/(N-1)
-
-true.mean <- 1/sigma + 1/gamma
-
-intrinsic_fun <- function(tau) {
-	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
-}
-
-conditional_fun <- function(tau) {
-	sigma*(gamma+lambda)/(gamma+lambda-sigma) * (exp(-sigma*tau)-exp(-(gamma+lambda)*tau))
-}
-
-true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
-
-true.cv <- sqrt(true.var)/true.mean
-
-true.R <- beta/gamma

### 2018-09-23 analysis/ebola_param_seminr.R

@@ -1,22 +1,0 @@
-set.seed(101)
-beta <- 2/5
-sigma <- 1/11.4
-m.sigma <- round(1.75)
-gamma <- 1/5
-n.gamma <- round((5/4.7)^2)
-N <- 60000
-
-true.mean <- 1/sigma + 1/gamma
-
-nsamp <- 1e5
-
-lat <- rgamma(nsamp, m.sigma, sigma*m.sigma)
-inf <- rgamma(nsamp, n.gamma, gamma*n.gamma)
-
-gen <- lat + runif(nsamp, 0, sample(inf, prob=inf, replace=TRUE)) 
-
-true.var <- var(gen)
-
-true.shape <- true.mean^2/true.var
-
-true.R <- beta/gamma

### 2018-09-23 analysis/local_effect.R

@@ -1,107 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2); theme_set(theme_bw())
-source("ebola_param.R")
-source("../R/seir.R")
-source("../R/generation.R")
-
-beta <- 1
-N <- 5
-lambda <- beta/(N-1)
-
-nsim <- 5000
-
-set.seed(101)
-reslist_full <- vector('list', nsim)
-i <- 1
-while (i <= nsim) {
-	rr <- seir.full(N, beta/(N-1), sigma, gamma, I0 = 1, imax=10)
-	reslist_full[[i]] <- rr
-	i <- i+1
-}
-
-set.seed(101)
-reslist_local <- vector('list', nsim)
-i <- 1
-while (i <= nsim) {
-	rr <- seir.local(4, beta/(N-1), sigma, gamma)
-	
-	reslist_local[[i]] <- rr
-	i <- i+1
-}
-
-gendata_full <- reslist_full %>%
-	lapply(network.generation, type="backward", plot=FALSE) %>%
-	unlist
-
-gendata_cond <- reslist_local %>%
-	lapply("[[", "conditional") %>%
-	unlist
-
-gendata_int <- reslist_local %>%
-	lapply("[[", "intrinsic") %>%
-	unlist
-	
-gendata_mean <- data.frame(
-	type=c("intrinsic", "conditional", "homogeneous"),
-	value=c(mean(gendata_int), mean(gendata_cond), mean(gendata_full))
-)
-
-true_mean <- data.frame(
-	key=c("intrinsic", "conditional"),
-	value=c(1/sigma+1/gamma, 1/sigma + 1/(gamma+lambda))
-)
-	
-gendata_total <- list(
-	intrinsic=data.frame(
-		value=gendata_int
-	),
-	conditional=data.frame(
-		value=gendata_cond
-	),
-	homogeneous=data.frame(
-		value=gendata_full
-	) 
-) %>%
-	bind_rows(.id="type") %>%
-	mutate(type=factor(type, levels=c("intrinsic", "conditional", "homogeneous")))
-
-tvec <- seq(0, 80, by=0.1)
-
-dendata <- list(
-	data.frame(
-		key="intrinsic",
-		x=tvec,
-		density=intrinsic_fun(tvec)
-	),
-	data.frame(
-		key="conditional",
-		x=tvec,
-		density=conditional_fun(tvec)
-	) 
-) %>%
-	bind_rows
-
-gglocal <- ggplot(gendata_total) +
[…]

### 2018-09-23 analysis/seminr_growth.R

@@ -1,0 +1,22 @@
+library(epigrowthfit)
+
+load("../sim/seminr_sim.rda")
+
+fitlist <- vector('list', length(reslist))
+
+for (i in 1:length(reslist)) {
+	print(i)
+	sim <- reslist[[i]]
+
+	week <- floor(sim$data$time/7) + 1
+	
+	fitdata <- data.frame(
+		week=1:15,
+		cases=as.vector(table(head(week, -1)))
+	)
+	
+	fitlist[[i]] <- epigrowthfit(fitdata$week, fitdata$cases,
+						model="logistic")
+}
+
+save(fitlist, file="seminr_growth.rda")

### 2018-09-23 analysis/seminr_nonparametric.R

@@ -1,0 +1,38 @@
+source("../sim/ebola_param_seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+
+load("../sim/seminr_sim.rda")
+
+load("seminr_growth.rda")
+
+estlist <- vector('list', length(reslist))
+
+for (i in 1:length(reslist)) {
+	print(i)
+	sim <- reslist[[i]]
+	fit <- fitlist[[i]]
+	
+	logr <- fit@mle2@coef[1]
+	logr.sd <- sqrt(diag(fit@mle2@vcov)[1])
+	
+	if (is.na(logr.sd)) {
+		logr.sd <- max((log(fit@growthRate[3]) - log(fit@growthRate[1]))/1.96,
+					   (log(fit@growthRate[1]) - log(fit@growthRate[2]))/1.96)	
+	} 
+	
+	gen <- network.generation(sim, interval=c(0, 7*15), plot=FALSE)
+	
+	est <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_nonparametric)))
+	
+	est$type <- c("mean", "RR")
+	
+	est[,4] <- c(est[1,2] < true.mean && true.mean < est[1,3], 
+				 est[2,2] < true.R && true.R < est[2,3])
+	
+	colnames(est)[4] <- "coverage"
+	
+	estlist[[i]] <- est
+}
+
+save("estlist", file="seminr_nonparametric.rda")

### 2018-09-23 analysis/seminr_observed.R

@@ -1,0 +1,38 @@
+source("../sim/ebola_param_seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+
+load("../sim/seminr_sim.rda")
+
+load("seminr_growth.rda")
+
+estlist <- vector('list', length(reslist))
+
+for (i in 1:length(reslist)) {
+	print(i)
+	sim <- reslist[[i]]
+	fit <- fitlist[[i]]
+	
+	logr <- fit@mle2@coef[1]
+	logr.sd <- sqrt(diag(fit@mle2@vcov)[1])
+	
+	if (is.na(logr.sd)) {
+		logr.sd <- max((log(fit@growthRate[3]) - log(fit@growthRate[1]))/1.96,
+					   (log(fit@growthRate[1]) - log(fit@growthRate[2]))/1.96)	
+	} 
+	
+	gen <- network.generation(sim, interval=c(0, 7*15), plot=FALSE)
+	
+	est <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_observed)))
+	
+	est$type <- c("mean", "RR")
+	
+	est[,4] <- c(est[1,2] < true.mean && true.mean < est[1,3], 
+				 est[2,2] < true.R && true.R < est[2,3])
+	
+	colnames(est)[4] <- "coverage"
+	
+	estlist[[i]] <- est
+}
+
+save("estlist", file="seminr_observed.rda")

### 2018-09-23 analysis/temporal_effect.R

@@ -1,126 +1,0 @@
-library(deSolve)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("ebola_param.R")
-
-## ode
-
-seir.ode <- function(t, y, parms) {
-    with(as.list(c(y, parms)), {
-        dS <- -beta/N*S*I
-        dE <- beta/N*S*I - sigma*E
-        dI <- sigma*E - gamma*I
-        list(c(dS, dE, dI))
-    })
-}
-
-## we want to simulate equilibrium state!
-seirs.ode <- function(t, y, parms) {
-	with(as.list(c(y, parms)), {
-		dS <- gamma * I -beta/N*S*I
-		dE <- beta/N*S*I - sigma*E
-		dI <- sigma*E - gamma*I
-		list(c(dS, dE, dI))
-	})
-}
-
-sumfun <- function(out,
-				   tmax,
-				   t,
-				   tauby=5) {
-	incidence <- beta*out$S*out$I/N
-	c.incidence <- cumsum(incidence)
-	
-	out$incidence <- incidence
-	
-	genden <- intrinsic_fun(t)
-	
-	censor <- function(tau, tmax) {
-		m <- which(t==tmax)
-		total <- c.incidence[m]
-		
-		i <- which(t==tau)
-		
-		beta*sum(incidence[1:(m-i+1)] * out$S[i:m]/N)/total*genden[i]
-	}
-	
-	mean.censor <- function(tmax) {
-		if (tmax<=0){
-			return(0)
-		} else{
-			t <- t[t <= tmax]
-			
-			dc <- sapply(t, censor, tmax=tmax)
-			return(sum(dc*t)/sum(dc))
-		}
-	}
-	
-	tau <- seq(0, tmax, tauby)
-	cm <- sapply(tau, mean.censor)
-	
-	censor.df <- data.frame(
-		t=tau
-		, mean=cm
-	)
-}
-
-parms <- c(beta=beta, sigma=sigma, gamma=gamma, N=N)
-y <- c(S = N-10, E=0, I=10)
-
-dt <- 0.1
-tmax <- 305
-t <- seq(0, tmax, dt)
-out1 <- as.data.frame(ode(y, t, seir.ode, parms))
-out2 <- as.data.frame(ode(y, t, seirs.ode, parms))
[…]

### 2018-09-23 fig/local_effect.R

@@ -1,0 +1,107 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+source("../sim/ebola_param.R")
+source("../R/seir.R")
+source("../R/generation.R")
+
+beta <- 1
+N <- 5
+lambda <- beta/(N-1)
+
+nsim <- 5000
+
+set.seed(101)
+reslist_full <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+	rr <- seir.full(N, beta/(N-1), sigma, gamma, I0 = 1, imax=10)
+	reslist_full[[i]] <- rr
+	i <- i+1
+}
+
+set.seed(101)
+reslist_local <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+	rr <- seir.local(4, beta/(N-1), sigma, gamma)
+	
+	reslist_local[[i]] <- rr
+	i <- i+1
+}
+
+gendata_full <- reslist_full %>%
+	lapply(network.generation, type="backward", plot=FALSE) %>%
+	unlist
+
+gendata_cond <- reslist_local %>%
+	lapply("[[", "conditional") %>%
+	unlist
+
+gendata_int <- reslist_local %>%
+	lapply("[[", "intrinsic") %>%
+	unlist
+	
+gendata_mean <- data.frame(
+	type=c("intrinsic", "conditional", "homogeneous"),
+	value=c(mean(gendata_int), mean(gendata_cond), mean(gendata_full))
+)
+
+true_mean <- data.frame(
+	key=c("intrinsic", "conditional"),
+	value=c(1/sigma+1/gamma, 1/sigma + 1/(gamma+lambda))
+)
+	
+gendata_total <- list(
+	intrinsic=data.frame(
+		value=gendata_int
+	),
+	conditional=data.frame(
+		value=gendata_cond
+	),
+	homogeneous=data.frame(
+		value=gendata_full
+	) 
+) %>%
+	bind_rows(.id="type") %>%
+	mutate(type=factor(type, levels=c("intrinsic", "conditional", "homogeneous")))
+
+tvec <- seq(0, 80, by=0.1)
+
+dendata <- list(
+	data.frame(
+		key="intrinsic",
+		x=tvec,
+		density=intrinsic_fun(tvec)
+	),
+	data.frame(
+		key="conditional",
+		x=tvec,
+		density=conditional_fun(tvec)
+	) 
+) %>%
+	bind_rows
+
+gglocal <- ggplot(gendata_total) +
[…]

### 2018-09-23 fig/temporal_effect.R

@@ -1,0 +1,126 @@
+library(deSolve)
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("../sim/ebola_param.R")
+
+## ode
+
+seir.ode <- function(t, y, parms) {
+    with(as.list(c(y, parms)), {
+        dS <- -beta/N*S*I
+        dE <- beta/N*S*I - sigma*E
+        dI <- sigma*E - gamma*I
+        list(c(dS, dE, dI))
+    })
+}
+
+## we want to simulate equilibrium state!
+seirs.ode <- function(t, y, parms) {
+	with(as.list(c(y, parms)), {
+		dS <- gamma * I -beta/N*S*I
+		dE <- beta/N*S*I - sigma*E
+		dI <- sigma*E - gamma*I
+		list(c(dS, dE, dI))
+	})
+}
+
+sumfun <- function(out,
+				   tmax,
+				   t,
+				   tauby=5) {
+	incidence <- beta*out$S*out$I/N
+	c.incidence <- cumsum(incidence)
+	
+	out$incidence <- incidence
+	
+	genden <- intrinsic_fun(t)
+	
+	censor <- function(tau, tmax) {
+		m <- which(t==tmax)
+		total <- c.incidence[m]
+		
+		i <- which(t==tau)
+		
+		beta*sum(incidence[1:(m-i+1)] * out$S[i:m]/N)/total*genden[i]
+	}
+	
+	mean.censor <- function(tmax) {
+		if (tmax<=0){
+			return(0)
+		} else{
+			t <- t[t <= tmax]
+			
+			dc <- sapply(t, censor, tmax=tmax)
+			return(sum(dc*t)/sum(dc))
+		}
+	}
+	
+	tau <- seq(0, tmax, tauby)
+	cm <- sapply(tau, mean.censor)
+	
+	censor.df <- data.frame(
+		t=tau
+		, mean=cm
+	)
+}
+
+parms <- c(beta=beta, sigma=sigma, gamma=gamma, N=N)
+y <- c(S = N-10, E=0, I=10)
+
+dt <- 0.1
+tmax <- 305
+t <- seq(0, tmax, dt)
+out1 <- as.data.frame(ode(y, t, seir.ode, parms))
+out2 <- as.data.frame(ode(y, t, seirs.ode, parms))
+out3 <- as.data.frame(ode(unlist(tail(out2, 1))[-1] , t, seirs.ode, parms))
[…]

### 2018-09-23 sim/ebola_param.R

@@ -1,0 +1,21 @@
+beta <- 2/5
+sigma <- 1/11.4
+gamma <- 1/5
+N <- 60000
+lambda <- beta/(N-1)
+
+true.mean <- 1/sigma + 1/gamma
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+conditional_fun <- function(tau) {
+	sigma*(gamma+lambda)/(gamma+lambda-sigma) * (exp(-sigma*tau)-exp(-(gamma+lambda)*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.cv <- sqrt(true.var)/true.mean
+
+true.R <- beta/gamma

### 2018-09-23 sim/ebola_param_seminr.R

@@ -1,0 +1,22 @@
+set.seed(101)
+beta <- 2/5
+sigma <- 1/11.4
+m.sigma <- round(1.75)
+gamma <- 1/5
+n.gamma <- round((5/4.7)^2)
+N <- 100000
+
+true.mean <- 1/sigma + 1/gamma
+
+nsamp <- 1e5
+
+lat <- rgamma(nsamp, m.sigma, sigma*m.sigma)
+inf <- rgamma(nsamp, n.gamma, gamma*n.gamma)
+
+gen <- lat + runif(nsamp, 0, sample(inf, prob=inf, replace=TRUE)) 
+
+true.var <- var(gen)
+
+true.shape <- true.mean^2/true.var
+
+true.R <- beta/gamma

### 2018-09-23 sim/seminr_sim.R

@@ -1,0 +1,19 @@
+source("../sim/ebola_param_seminr.R")
+source("../R/seminr.R")
+source("../R/generation.R")
+
+nsim <- 100
+
+set.seed(101)
+reslist <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+    print(i)
+    rr <- seminr.full(N, beta/(N-1), sigma, m.sigma, gamma, n.gamma, I0 = 10, tmax=7*15)
+    if(nrow(rr$data) > 100) {
+        reslist[[i]] <- rr
+        i <- i+1
+    }
+}
+
+save("reslist", file="seminr_sim.rda")

### 2018-09-24 R/estimate.R
files

@@ -35,3 +35,49 @@ bootfun <- function(gen,
 	
 	est
 }
+
+minuslogl.full <- function(log.R, log.mean, log.shape, data, tmax) {
+	R <- exp(log.R)
+	mean <- exp(log.mean)
+	shape <- exp(log.shape)
+	scale <- mean/shape
+	
+	t_censor <- tmax-data$t_infected
+	
+	nll <- -(sum(
+		log.R +
+			dgamma(data$generation, shape=shape, scale=scale, log=TRUE),
+		na.rm=TRUE
+	) + sum(
+		-R * pgamma(t_censor, shape=shape, scale=scale),
+		na.rm=TRUE
+	))
+	
+	if (nll==0) return(Inf) else return(nll)
+}
+
+parametricfun <- function(R, mean, shape,
+						  data,
+						  tmax) {
+	m <- mle2(minuslogl.full, 
+		 start=list(log.R=log(R), log.mean=log(mean), log.shape=log(shape)),
+		 method="Nelder-Mead",
+		 optimizer="optim",
+		 data=list(data=data, tmax=tmax),
+		 control=list(maxit=10000)
+	)
+	
+	pp <- profile(m, 1:2)
+	
+	ci <- exp(confint(pp))
+	
+	est <- rbind(
+		exp(coef(m)[1:2]),
+		t(ci)
+	)
+	
+	rownames(est) <- c("estimate", "lwr", "upr")
+	
+	cbind(est[,2], est[,1])
+}
+

### 2018-09-24 analysis/seminr_growth.R

@@ -2,21 +2,32 @@ library(epigrowthfit)
 
 load("../sim/seminr_sim.rda")
 
-fitlist <- vector('list', length(reslist))
+weeks <- 3:5*3
 
-for (i in 1:length(reslist)) {
-	print(i)
-	sim <- reslist[[i]]
+fitlist <- vector('list', length(weeks))
 
-	week <- floor(sim$data$time/7) + 1
+for (j in 1:length(weeks)) {
 	
-	fitdata <- data.frame(
-		week=1:15,
-		cases=as.vector(table(head(week, -1)))
-	)
+	sublist <- vector('list', length(reslist))
 	
-	fitlist[[i]] <- epigrowthfit(fitdata$week, fitdata$cases,
-						model="logistic")
+	for (i in 1:length(reslist)) {
+		print(i)
+		sim <- reslist[[i]]
+		
+		week <- floor(sim$data$time/7) + 1
+		
+		fitdata <- data.frame(
+			week=1:15,
+			cases=as.vector(table(head(week, -1)))
+		)
+		
+		fitdata <- fitdata[1:weeks[j],]
+		
+		sublist[[i]] <- epigrowthfit(fitdata$week, fitdata$cases,
+									 model="logistic")
+	}
+	
+	fitlist[[j]] <- sublist
 }
 
 save(fitlist, file="seminr_growth.rda")

### 2018-09-24 analysis/seminr_nonparametric.R

@@ -6,33 +6,50 @@ load("../sim/seminr_sim.rda")
 
 load("seminr_growth.rda")
 
-estlist <- vector('list', length(reslist))
+weeks <- 3:5*3
 
-for (i in 1:length(reslist)) {
-	print(i)
-	sim <- reslist[[i]]
-	fit <- fitlist[[i]]
+estlist <- vector('list', length(weeks))
+
+set.seed(101)
+for (j in 1:length(weeks)) {
+	sublist <- vector('list', length(reslist))
 	
-	logr <- fit@mle2@coef[1]
-	logr.sd <- sqrt(diag(fit@mle2@vcov)[1])
+	for (i in 1:length(reslist)) {
+		print(i)
+		sim <- reslist[[i]]
+		fit <- fitlist[[j]][[i]]
+		
+		logr <- fit@mle2@coef[1]
+		logr.sd <- sqrt(diag(fit@mle2@vcov)[1])
+		
+		if (is.na(logr.sd) || logr.sd <= 0) {
+			logr.sd <- max((log(fit@growthRate[3]) - log(fit@growthRate[1]))/1.96,
+						   (log(fit@growthRate[1]) - log(fit@growthRate[2]))/1.96)	
+		} 
+		
+		gen <- network.generation(sim, interval=c(0, 7*weeks[j]), plot=FALSE)
+		
+		est <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_nonparametric)))
+		
+		est[,4] <- c(est[1,2] < true.mean && true.mean < est[1,3], 
+					 est[2,2] < true.R && true.R < est[2,3])
+		
+		colnames(est)[4] <- "coverage"
+		
+		est$type <- c("mean", "RR")
+		
+		est$sim <- i
+		
+		sublist[[i]] <- est
+	}
 	
-	if (is.na(logr.sd)) {
-		logr.sd <- max((log(fit@growthRate[3]) - log(fit@growthRate[1]))/1.96,
-					   (log(fit@growthRate[1]) - log(fit@growthRate[2]))/1.96)	
-	} 
+	subdf <- do.call("rbind", sublist)
 	
-	gen <- network.generation(sim, interval=c(0, 7*15), plot=FALSE)
+	rownames(subdf) <- NULL
 	
-	est <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_nonparametric)))
+	subdf$weeks <- weeks[j]
 	
[…]

### 2018-09-24 analysis/seminr_parametric.R

@@ -1,35 +1,24 @@
+library(bbmle)
 source("../sim/ebola_param_seminr.R")
 source("../R/generation.R")
 source("../R/estimate.R")
 
 load("../sim/seminr_sim.rda")
 
-load("seminr_growth.rda")
-
 weeks <- 3:5*3
 
 estlist <- vector('list', length(weeks))
 
-set.seed(101)
 for (j in 1:length(weeks)) {
 	sublist <- vector('list', length(reslist))
 	
 	for (i in 1:length(reslist)) {
 		print(i)
 		sim <- reslist[[i]]
-		fit <- fitlist[[j]][[i]]
 		
-		logr <- fit@mle2@coef[1]
-		logr.sd <- sqrt(diag(fit@mle2@vcov)[1])
+		gendata <- generation.data(sim, tmax=7*weeks[j])
 		
-		if (is.na(logr.sd) || logr.sd <= 0) {
-			logr.sd <- max((log(fit@growthRate[3]) - log(fit@growthRate[1]))/1.96,
-						   (log(fit@growthRate[1]) - log(fit@growthRate[2]))/1.96)	
-		} 
-		
-		gen <- network.generation(sim, interval=c(0, 7*weeks[j]), plot=FALSE)
-		
-		est <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_observed)))
+		est <- as.data.frame(t(parametricfun(true.R, true.mean, true.shape, gendata, 7*weeks[j])))
 		
 		est[,4] <- c(est[1,2] < true.mean && true.mean < est[1,3], 
 					 est[2,2] < true.R && true.R < est[2,3])
@@ -52,4 +41,4 @@ for (j in 1:length(weeks)) {
 	estlist[[j]] <- subdf
 }
 
-save("estlist", file="seminr_observed.rda")
+save("estlist", file="seminr_parametric.rda")

### 2018-09-24 fig/compare_methods.R

@@ -1,0 +1,74 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("../sim/ebola_param_seminr.R")
+
+dir <- c("../analysis/")
+files <- c("seminr_observed.rda", "seminr_nonparametric.rda", "seminr_parametric.rda")
+
+comblist <- vector('list', 3)
+
+for (i in 1:length(files)) {
+	load(paste0(dir, files[i]))
+	
+	comblist[[i]] <- estlist
+}
+
+names(comblist) <- c("observed", "nonparametric", "parametric")
+
+combdata <- comblist %>%
+	lapply(bind_rows) %>%
+	bind_rows(.id="method") %>%
+	mutate(method=factor(method, levels=c("observed", "nonparametric", "parametric")),
+		   weeks=factor(weeks, levels=3:5*3),
+		   type=factor(type, labels=c("mean generation interval", "reproductive number")))
+
+coverdata <- combdata %>%
+	group_by(method, type, weeks) %>%
+	summarize(coverage=mean(coverage, na.rm=TRUE))
+
+estdata <- combdata %>%
+	group_by(method, type, weeks) %>%
+	summarize(
+		median=median(estimate),
+		lwr=quantile(estimate, 0.025),
+		upr=quantile(estimate, 0.975)
+	) %>%
+	filter(!(method=="parametric" & weeks==3))
+
+truedata <- data.frame(
+	type=c("mean generation interval", "reproductive number"),
+	value=c(true.mean, true.R)
+)
+
+g1 <- ggplot(estdata) +
+	geom_point(aes(weeks, median, col=method, shape=method), position=position_dodge(width=0.2)) +
+	geom_errorbar(aes(weeks, ymin=lwr, ymax=upr, col=method), width=0.2, position=position_dodge(width=0.2)) + 
+	geom_hline(data=truedata, aes(yintercept=value), lty=2) +
+	scale_y_continuous("estimate") +
+	facet_wrap(~type, scale="free_y") +
+	theme(
+		axis.title.x = element_blank(),
+		strip.background = element_blank(),
+		legend.position = "top",
+		legend.title = element_blank()
+	)
+
+g2 <- ggplot(coverdata) +
+	geom_point(aes(weeks, coverage, col=method, shape=method)) +
+	geom_path(aes(as.numeric(weeks), coverage, col=method)) +
[…]

### 2018-09-24 doc/network.tex
update

@@
 
-First, we compared the four methods by simulating a stochastmic epidemic model homogeneous network with 100000 individuals using Ebola-like parameters for 15 weeks (Figure~?).
+First, we compared how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
+As expected, we find that using the observed generation-interval distributions always underestimates both the mean generation interval and the reproductive number;
+due to initial overestimation of the growth rate,
+
+\begin{figure}
+\includegraphics[width=\textwidth]{../fig/example.pdf}
+\caption{Fill this out.}
+\label{fig:example}
+\end{figure}
 

### 2018-09-24 fig/example.R

@@ -1,0 +1,152 @@
+library(epigrowthfit)
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+source("../sim/ebola_param_seminr.R")
+source("../R/seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+
+weekmax <- 28
+weekmin <- 9
+
+set.seed(101)
+rr <- seminr.full(N, beta/(N-1), sigma, m.sigma, gamma, n.gamma, I0 = 10, tmax=7*weekmax)
+
+weeks <- weekmin:weekmax
+
+reslist <- vector('list', length(weeks))
+
+for (i in 1:length(weeks)) {
+	print(i)
+	week <- weeks[i]
+	
+	obsdata <- rr$data %>%
+		filter(time < 7*week)
+ 	
+	fitdata <- data.frame(
+		week=1:week,
+		cases=as.vector(table(floor(obsdata$time/7)))
+	)
+	
+	epifit <- epigrowthfit(fitdata$week, fitdata$cases,
+						   model="logistic")
+	
+	logr <- epifit@mle2@coef[1]
+	logr.sd <- sqrt(diag(epifit@mle2@vcov)[1])
+	
+	if (is.na(logr.sd) || logr.sd <= 0) {
+		logr.sd <- max((log(epifit@growthRate[3]) - log(epifit@growthRate[1]))/1.96,
+					   (log(epifit@growthRate[1]) - log(epifit@growthRate[2]))/1.96)	
+	} 
+	
+	gen <- network.generation(rr, plot=FALSE, interval=c(0, 7*week))
+	
+	gendata <- generation.data(rr, tmax=7*week)
+	
+	nonparaest <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_nonparametric)))
+	
+	nonparaest$type <- c("mean", "RR")
+	nonparaest$method <- "nonparametric"
+	
+	obsest <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_observed)))
+	
+	obsest$type <- c("mean", "RR")
+	obsest$method <- "observed"
+	
+	paraest <- as.data.frame(t(parametricfun(true.R, true.mean, true.shape, gendata, 7*week)))
+	
+	paraest$type <- c("mean", "RR")
+	paraest$method <- "parametric"
+	
+	growthest <- data.frame(
+		estimate=epifit@growthRate[1],
[…]

### 2018-09-26 doc/network.tex
tweak

--- 2018-09-23 doc/network.tex
@@
 
-\swp{I might say something about independence transmission events in the results -- see forsberg 2008...}
+\swp{I might say something about independence transmission events in the results if I decide to compare forsberg2008 as well -- see forsberg 2008...}
 
-First, we compared the four methods by simulating a stochastmic epidemic model homogeneous network with 100000 individuals using Ebola-like parameters for 15 weeks (Figure~?).
+We compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
+During the initial exponential growth period (approximately between 9 and 20 weeks), both nonparametric and parametric methods yield similar estimates of mean generation interval and reproductive number.
+Howeve, we observe opposite trends after the growth period (approximately after 20 weeks).
+As the nonparametric method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
+On the other hand, as the parametric method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible results in underestimation of both mean generation interval and reproductive number.
+we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but the observed estimates always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulations.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{../fig/example.pdf}
+\caption{Fill this out.}
+\label{fig:example}
+\end{figure}
+
+Then, we compare accuracy and reliability of the two methods during the initial growth period using 100 stochastic simulations on a homogeneous network (Figure~\ref{fig:test}).
+We find that the parametric method provides the most consistent (least variable) estimates.
+The nonparametric method overestimates reproductive number initially but this is likely driven by the initial overestimation of the growth rate (Figure~\ref{fig:example}).
+
+\swp{May need to explain coverage...}
+Surprisingly, both methods fail to reach nominal coverage.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{../fig/compare_methods.pdf}
+\caption{Fill this out.}
+\label{fig:test}
+\end{figure}
 

### 2018-09-26 doc/network.tex
tweak

@@
 where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation interval distribution $g$.
-A similar method was proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports and does not make assumptions about the shape of the generation-interval distribution.
+A similar method was proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports rather than the observed generation intervals.
 
 \swp{I might say something about independence transmission events in the results if I decide to compare forsberg2008 as well -- see forsberg 2008...}
+\swp{I got stuck on forsberg 2008.
+I get weird estimates with wide confidence intervals but I feel like that's what their method is supposed to do (based on figure 3 in their paper).
+It doesn't use generation interval either so I think we can just ignore it?}
 
@@
 
-\swp{May need to explain coverage...}
+\swp{May need to explain what coverage is...}
 Surprisingly, both methods fail to reach nominal coverage.

### 2018-09-26 doc/network.tex
update

@@
 
-We compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
+First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
 During the initial exponential growth period (approximately between 9 and 20 weeks), both nonparametric and parametric methods yield similar estimates of mean generation interval and reproductive number.
@@
 On the other hand, as the parametric method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible results in underestimation of both mean generation interval and reproductive number.
-we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but the observed estimates always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulations.
+we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but the observed estimates always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation.
 
@@
 \swp{May need to explain what coverage is...}
-Surprisingly, both methods fail to reach nominal coverage.
+Surprisingly, both methods fail to attain nominal coverage;
+throughout the estimation period, both methods attain approximately 70\% coverage.
+Even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution, making a wrong distributional assumption leads to underconfidence.
+To confirm that the undercoverage is caused by distribuional assumption, we simulated an epidemic using gamma kernel and found that using the true shape yields nominal coverage (see Appendix).
+These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].
+\swp{Why does nonparametric method undercover?}
 

### 2018-09-26 doc/network.tex
Autocommit (doc)

@@
-\documentclass{article}
+\documentclass[12pt]{article}
 \title{Inferring generation-interval distributions from contact tracing data}
@@
 Aggregating generation intervals across heterogeneous environment also affects the observed generation intervals.
-Infection is more likely to spread quicker than on average among closely related inidividuals, and the local decrease in susceptibles result in shorter realized generation intervals.
+Infection is more likely to spread quicker than on average among closely related individuals, and the local decrease in susceptibles result in shorter realized generation intervals.
+\jd{Not following, and not sure we want to go here for now.}
 \cite{trapman2016inferring} gave a similar argument to explain the difference in the the reproductive numbers between a homogeneously mixing population and a heterogeneously (network structured) mixing population.
@@
 
-\subsection*{Right-censored interval}
+\subsection*{Right-censored intervals}
 

### 2018-09-26 doc/network.tex
Meeting with Daniel

@@
 \documentclass[12pt]{article}
-\title{Inferring generation-interval distributions from contact tracing data}
+\title{Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}}
 \author{Sang Woo Park, David Champredon and Jonathan Dushoff}
@@
 In practice, generation intervals are measured by contact tracing throughout the course of an epidemic, usually aggregated to increase the amount of information available; we refer to these as \emph{censored} (or right-censored) intervals.
-\jd{I kind of feel like, if you go on forever, aggregated GIs approach intrinsic GIs.
-This obviously isn't true for a single epidemic (observed GIs are shorter overall).
-On the other hand, it obviously \emph{is} true if you add births and get to a stable equilibrium.
-So I guess the remaining question is, do we get convergence if we have stable cycles?
-What we think about this will affect how we write the rest of this P.} \swp{Going to leave this comment here so that we can come back later.}
-If observation goes on throughout the disease cycle, then (something about something).
-On the other hand, if calculations are being done based on early-outbreak data, the available censored data is best thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions.
+When calculations are based on early-outbreak data, the available (censored) observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions.
 
+\jd{Think about this P. It's really more about networks than about heterogeneity.}
 Aggregating generation intervals across heterogeneous environment also affects the observed generation intervals.
-Infection is more likely to spread quicker than on average among closely related individuals, and the local decrease in susceptibles result in shorter realized generation intervals.
-\jd{Not following, and not sure we want to go here for now.}
+In the presence of heterogeneity, susceptibility will be lower in the neighborhood of infected individuals, resulting in shorter realized generation intervals.
 \cite{trapman2016inferring} gave a similar argument to explain the difference in the the reproductive numbers between a homogeneously mixing population and a heterogeneously (network structured) mixing population.
 Their result can be interpreted as a consequence of changes in generation-interval distributions: shorter generation intervals lead to smaller reproductive number [CITE]. 
+\jd{JD needs to write clearly what he thinks about the Trapman insight. Trapman is an anagram for rampant.}
 
@@
 In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), 
-the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrisnc generation-interval distribution:
+the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrinsic generation-interval distribution:
 \begin{equation}
@@
 
-Second, number of susceptibles decrease over the course of an epidemic, forward generation-interval distribution at any given period is always shorter than the intrinsic distribution \citep{champredon2015intrinsic}.
+Second, since the number of susceptibles decreases over the course of an epidemic, the forward generation-interval distribution for a given epidemic period is always shorter than the intrinsic distribution \citep{champredon2015intrinsic}.
 Asymptotically, we can ignore the effect of right-censoring, and the censored generation intervals can be thought of as a weighted average of the forward generation intervals.
 As a result, even if contact tracing is performed through an entire epidemic, mean generation interval will be underestimated.
+\jd{Simplify main argument? Add caveat about times when susceptibles are increasing? Leave out and stay focused on outbreaks?}
 
@@
 
-When an epidemic is in an equilibrium state (i.e., incidence $i(t)$ and the number of susceptibles $S(t)$ is constant over time), the observed generation-interval distribution is equivalent to the intrinsic generation-interval distribution.
-On the other hand, when an epidemic is in a cycle, we get ???  \swp{TODO: Continue this exciting research!}
+When a disease is at (or near) endemic equilibrium, the number of susceptibles remains (approximately) constant over time, and the observed generation-interval distribution should be similar to the intrinsic generation-interval distribution.
 
@@
 \hat{g}(\tau) = \frac{\hat{k}(\tau)}{\int \hat{k} \tau}.
+\label{eq:conditional}
 \end{equation}
@@
 
-The difference between conditional and intrinsic generation-interval distribution can be best demonstrated by simulating stochastic infection processes on a small tree network -- a single infected individual is connected to multiple susceptible individuals but susceptibles are not connected with each other -- with artificially high $\RR$ (Figure~\ref{fig:local}).
-We observe that the distribution of realized generation times matches with the conditional generation-interval distribution, whereas the distribution of contact times matches with the intrinsic generation-interval distribution.
-On the other hand, if a stochastic infection processes are simulated on a small homogeneous network with all else being equal, the distribution of generation times is even shorter than the effective generation intervals because a susceptible individual may be contacted by multiple infectors (and only the first contact gives rise to infection).
-\jd{Let's think about how to explain this. In particular, the first one is where we have our correction that works \ldots}.
-\swp{I think things are clearer when we introduce conditional kernel?}
-Hence, we expect there to be varying degrees of local depletion of susceptibles in a heterogeneous contact structure and the accurate measurement of generation interval will be more difficult.
+The difference between conditional and intrinsic generation-interval distribution can be demonstrated by simulating stochastic infection processes on a small tree network (i.e, a single infected individual is connected to multiple susceptible individuals who are not connected with each other) with artificially high $\RR$.
+Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of generation times matches with the conditional generation-interval distribution (Figure~\ref{fig:local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
+
+The conditional generation interval \eqref{conditional} only explains some of the reduction in generation times that occurs on most networks, however.
+Generation intervals are also shortened by indirect connections: if a susceptible individual is infected through another route while the focal individual is infected, this will shorten the generation interval \jd{This s.~is too repetitive.}
+Simulations on a small homogeneous network confirm this additional effect (\fref{local}, right panel). 
+In general, estimating the effects of most networks on observed generation intervals is expected to be a hard problem.
 
@@
 \swp{I kind of want to keep the derivations in the methods (which comes after the discussion) and keep the description of methods simple here so that we don't scare people with the derivations...?}
+\jd{Yes.}
 
@@
 
+\clearpage
+
+\section{Discussion}
+
 \section{Methods}

### 2018-10-02 doc/network.tex
minor edits

@@
 \usepackage{adjustbox}
+
+\newcommand{\eref}[1]{(\ref{eq:#1})}
+\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
+\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
+\newcommand{\sref}[1]{Sec.~\ref{#1}}
+\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
+\newcommand{\tref}[1]{Table~\ref{tab:#1}}
+\newcommand{\tlab}[1]{\label{tab:#1}}
 
@@
 \jd{Think about this P. It's really more about networks than about heterogeneity.}
-Aggregating generation intervals across heterogeneous environment also affects the observed generation intervals.
-In the presence of heterogeneity, susceptibility will be lower in the neighborhood of infected individuals, resulting in shorter realized generation intervals.
-\cite{trapman2016inferring} gave a similar argument to explain the difference in the the reproductive numbers between a homogeneously mixing population and a heterogeneously (network structured) mixing population.
-Their result can be interpreted as a consequence of changes in generation-interval distributions: shorter generation intervals lead to smaller reproductive number [CITE]. 
+\swp{How about this?}
+Observed generation intervals also vary across space.
+On a network-structured population, repeated contacts mask long generation intervals and susceptibility will decrease quicker in a closely connected neighborhood of infected individuals.
+Therefore, realized generation intervals will be shorter and sampling generation intervals over a heterogeneous environment can misrepresent the intrinsic generation-interval distribution.
+\cite{trapman2016inferring} derived relationships between the growth rate, $r$, and the reproductive number, $\RR$, on various network structures and showed that $\RR$ will always be greater in a homogeneous network given a fixed $r$ value;
+as generation-interval distributions define relationship between $r$ and $\RR$, 
+difference between the intrinsic and realized generation intervals can be used to explain the results obtained by \cite{trapman2016inferring}.
 \jd{JD needs to write clearly what he thinks about the Trapman insight. Trapman is an anagram for rampant.}
@@
 \end{equation}
-Finally, the renewal equation model can be used to model incidence over time:
+Then, the renewal equation model can be used to model incidence over time:
 \begin{equation}
-i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR S(\tau) \int g(s) i(\tau-s) ds
+i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds
 \end{equation}
 where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
+Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation [CITE]:
+\begin{equation}
+\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
+\end{equation}
 
@@
 There are two reasons for this phenomenon.
-First, any infection events that occur after the contact tracing period canonot be observed due to right censoring and the observed generation-interval will be shorter.
+First, any infection events that occur after the contact tracing period canonot be observed due to right censoring and short generation intervals are more likely to be observed.
 In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), 
@@
 \end{equation}
-As backward generation-interval distributions are always shorter than the intrinsic generation-interval distribution during the exponential growth period, their weighted average (i.e., the censored intervals) will also be shorter.
-
-Second, since the number of susceptibles decreases over the course of an epidemic, the forward generation-interval distribution for a given epidemic period is always shorter than the intrinsic distribution \citep{champredon2015intrinsic}.
-Asymptotically, we can ignore the effect of right-censoring, and the censored generation intervals can be thought of as a weighted average of the forward generation intervals.
+% As backward generation-interval distributions are always shorter than the intrinsic generation-interval distribution during the exponential growth period, their weighted average (i.e., the censored intervals) will also be shorter.
+Second, decreasing number of susceptibles over the course of an epidemic make long infections less likely to occur, and the realized intervals will always be shorter \citep{champredon2015intrinsic}.
 As a result, even if contact tracing is performed through an entire epidemic, mean generation interval will be underestimated.
 \jd{Simplify main argument? Add caveat about times when susceptibles are increasing? Leave out and stay focused on outbreaks?}
+\swp{Simplified. Let's focus on outbreaks for now.}
 
@@
 \section{Statistical approach to inferring generation-interval distribution}
-
-\swp{I kind of want to keep the derivations in the methods (which comes after the discussion) and keep the description of methods simple here so that we don't scare people with the derivations...?}
-\jd{Yes.}
 

### 2018-10-02 R/estimate.R

@@ -56,6 +56,24 @@ minuslogl.full <- function(log.R, log.mean, log.shape, data, tmax) {
 	if (nll==0) return(Inf) else return(nll)
 }
 
+minuslogl.white <- function(log.R, log.mean, log.shape, data, k=50) {
+	R <- exp(log.R)
+	mean <- exp(log.mean)
+	shape <- exp(log.shape)
+	scale <- mean/shape
+	
+	t <- nrow(data)
+	
+	p_j <- diff(pgamma(0:k, shape=shape, scale=scale))
+	p_j <- p_j/sum(p_j)
+	
+	mu_t <- sapply(2:t, function(x) {
+		R * sum(data$cases[(x-1):max(x-k, 1)] * p_j[1:min(x-1, k)])
+	})
+	
+	-sum(dpois(tail(data$cases, -1), mu_t, log=TRUE))
+}
+
 parametricfun <- function(R, mean, shape,
 						  data,
 						  tmax) {
@@ -81,3 +99,27 @@ parametricfun <- function(R, mean, shape,
 	cbind(est[,2], est[,1])
 }
 
+whitefun <- function(R, mean, shape,
+					 data,
+					 k=50) {
+	m <- mle2(minuslogl.white, 
+			  start=list(log.R=log(R), log.mean=log(mean), log.shape=log(shape)),
+			  method="Nelder-Mead",
+			  optimizer="optim",
+			  data=list(data=data, k=k),
+			  control=list(maxit=10000)
+	)
+	
+	pp <- profile(m, 1:2)
+	
+	ci <- exp(confint(pp))
+	
+	est <- rbind(
+		exp(coef(m)[1:2]),
+		t(ci)
+	)
+	
+	rownames(est) <- c("estimate", "lwr", "upr")
+	
+	cbind(est[,2], est[,1])
+}

### 2018-10-02 analysis/measles.R

@@ -82,7 +82,7 @@ y <- c(S=0.05, E=0.0001, I=0.0001, R=1-0.05-0.0001-0.0001) * 5e7
 dt <- 1
 dt2 <- 0.2
 tmax <- 43 * 365
-tmax2 <- 4 * 365
+tmax2 <- 12 * 365
 t <- seq(0, tmax, dt)
 t2 <- seq(0, tmax2, dt2)
 out1 <- as.data.frame(rk(y, t, measles.ode, parms))
@@ -94,7 +94,7 @@ out2$incidence <- with(as.list(parms), {
 	b0 * (1 + b1 * cos(2 * pi * t2/period)) * out2$S * out2$I/N
 })
 
-gensum <- sumfun(out2, tmax2, t2, parms, tauby=10)
+gensum <- sumfun(out2, tmax2, t2, parms, tauby=100)
 
 g1 <- ggplot(out2, aes(time, S)) +
 	geom_line(lwd=1) +
@@ -140,7 +140,6 @@ g3 <- ggplot(gensum, aes(t, mean)) +
 		strip.background = element_blank(),
 		panel.spacing = unit(0, "cm")
 	)
-	
 
 gg_temporal <- arrangeGrob(g1, g2, g3, nrow=3)
 

### 2018-10-03 doc/network.tex
Working at redeemer

@@
 
-\jd{Think about this P. It's really more about networks than about heterogeneity.}
-\swp{How about this?}
 Observed generation intervals also vary across space.
-On a network-structured population, repeated contacts mask long generation intervals and susceptibility will decrease quicker in a closely connected neighborhood of infected individuals.
-Therefore, realized generation intervals will be shorter and sampling generation intervals over a heterogeneous environment can misrepresent the intrinsic generation-interval distribution.
-\cite{trapman2016inferring} derived relationships between the growth rate, $r$, and the reproductive number, $\RR$, on various network structures and showed that $\RR$ will always be greater in a homogeneous network given a fixed $r$ value;
-as generation-interval distributions define relationship between $r$ and $\RR$, 
-difference between the intrinsic and realized generation intervals can be used to explain the results obtained by \cite{trapman2016inferring}.
-\jd{JD needs to write clearly what he thinks about the Trapman insight. Trapman is an anagram for rampant.}
+In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in  the neighborhood of infected individuals than in the general population.
+This means that contacts made late in infection are more likely to be ``masked'' due to contacts already being infected.
+Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
+
+This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
+\jd{Maybe moving on a good direction, but needs more work. The problem is that we need to talk clearly about the spatial correction without misleading people about the temporal correction.}
 
@@
 
+\jd{Have to set this up to make sure we're not sounding too mathy.}
 Assume that contact tracing is performed from the beginning of an epidemic to time $t$.
@@
 % As backward generation-interval distributions are always shorter than the intrinsic generation-interval distribution during the exponential growth period, their weighted average (i.e., the censored intervals) will also be shorter.
-Second, decreasing number of susceptibles over the course of an epidemic make long infections less likely to occur, and the realized intervals will always be shorter \citep{champredon2015intrinsic}.
+Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized intervals will always be shorter \citep{champredon2015intrinsic}.
 As a result, even if contact tracing is performed through an entire epidemic, mean generation interval will be underestimated.
-\jd{Simplify main argument? Add caveat about times when susceptibles are increasing? Leave out and stay focused on outbreaks?}
-\swp{Simplified. Let's focus on outbreaks for now.}
 

### 2018-10-03 doc/network.tex
push

--- 2018-10-02 doc/network.tex
@@
 \tsub{c}{exp}(\tau) \propto g(\tau) \exp(-r\tau).
+\label{eq:exp}
 \end{equation}
@@
 \includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
-\caption{Fill this out.}
+\caption{\textbf{Temporal varitaion in the mean observed generation interval.}
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the observed mean generation interval was calculated the course of an epidemic (see methods).
+}
 \label{fig:censor}
@@
 When a disease is at (or near) endemic equilibrium, the number of susceptibles remains (approximately) constant over time, and the observed generation-interval distribution should be similar to the intrinsic generation-interval distribution.
+\swp{Probably don't need a figure for this}
 
@@
 In this section, we investigate two statistical methods for recovering the intrinsic (or conditional) generation-interval distribution.
-Detailed derivations of the two methods can be found in section??.
+Detailed derivations of the two methods can be found in section.
 
 \swp{maybe call it something else besides nonparametric method; it's still possible to make assumptions about the distribution and make inference}
-We refer to the first method as the nonparametric method as it does not rely on any assumptions about the generation-interval distributions.
-As the observed generation-interval distribution is a weighted distribution of the intrinsic distribution, the intrinsic distribution can be recovered by taking the inverse weight.
-Hence, the nonparametric method relies on the observed generation intervals and the exponential growth rate:
+We refer to the first method as the population-level method as it relies on the observed distribution without taking into account who infected whom.
+As the observed generation-interval distribution is a weighted distribution of the intrinsic distribution, the intrinsic distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
+Hence, the population-level method relies on the observed generation intervals and the exponential growth rate:
 \begin{equation}

### 2018-10-03 fig/temporal_effect.R

@@ -90,11 +90,13 @@ outdata <- bind_rows(out1, out3, .id="type") %>%
 		type=factor(type, label=c("epidemic", "equilibrium"))
 	)
 
-g1 <- ggplot(outdata, aes(t, incidence)) +
+## We probably don't need to show the endemic case.. seems trivial...
+
+g1 <- ggplot(outdata %>% filter(type=="epidemic"), aes(t, incidence)) +
     geom_line(lwd=1.2) +
     scale_x_continuous(expand=c(0,0), limits=c(0, 300), breaks=seq(0, 250, by=50)) +
-    scale_y_continuous("daily incidence", expand=c(0,0), limits=c(-5, 2000)) +
-	facet_wrap(~type) +
+    scale_y_continuous("daily incidence", expand=c(0,0), limits=c(-5, 700)) +
+	# facet_wrap(~type) +
     theme(
         panel.grid=element_blank(),
         axis.title.x=element_blank(),
@@ -106,21 +108,21 @@ g1 <- ggplot(outdata, aes(t, incidence)) +
 gendata <- gensum %>%
 	bind_rows(.id="type")
 
-g2 <- ggplot(gendata, aes(t, mean)) +
+g2 <- ggplot(gendata %>% filter(type==1), aes(t, mean)) +
     geom_line(lwd=1.2) +
     geom_hline(yintercept = 1/sigma + 1/gamma, lty=2) +
     scale_x_continuous("time (days)", expand=c(0,0), limits=c(0, 300), breaks=seq(0, 250, by=50)) +
     scale_y_continuous("mean observed GI", expand=c(0,0), limits=c(0, 1/sigma+1/gamma+4)) +
-	facet_wrap(~type) +
+	# facet_wrap(~type) +
 	theme(
         panel.grid=element_blank(),
-        axis.title.y = element_text(margin = margin(t = 0, r = 13.2, b = 0, l = 0)),
+        axis.title.y = element_text(margin = margin(t = 0, r = 8.1, b = 0, l = 0)),
         plot.margin=unit(c(5.5, 10, 5.5, 5.5), "points"),
         strip.background = element_blank(),
         strip.text = element_blank(),
         panel.spacing = unit(0, "cm")
     )
 
-gg_temporal <- arrangeGrob(g1, g2)
+gg_temporal <- arrangeGrob(g1, g2, nrow=2)
 
 ggsave("temporal_effect.pdf", gg_temporal, width=6, height=4)

### 2018-10-03 doc/network.tex
Autocommit (doc)

@@
 \includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
-\caption{\textbf{Temporal varitaion in the mean observed generation interval.}
+\caption{\textbf{Temporal variation in the mean observed generation interval.}
 A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the observed mean generation interval was calculated the course of an epidemic (see methods).
@@
 During the initial exponential growth period (approximately between 9 and 20 weeks), both nonparametric and parametric methods yield similar estimates of mean generation interval and reproductive number.
-Howeve, we observe opposite trends after the growth period (approximately after 20 weeks).
+However, we observe opposite trends after the growth period (approximately after 20 weeks).
 As the nonparametric method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.

### 2018-10-09 doc/network.tex
trying to make things a little clearer

@@
 
+\subsection{Intrinsic kernel}
+
 We begin by introducing an analytical framework.
-Let $K_a(\tau)$ be an individual level kernel, i.e., the rate at which secondary infections are caused by an individual $a$ \citep{svensson2015influence}.
-\swp{Need to say something more about Sven; come back later}
-Then, the unconditional, population-level kernel is given by integrating over individual variations:
+Let $K_a(\tau)$ be an individual-level intrinsic kernel, i.e., the rate at which secondary infections are expected to be caused by an individual $a$ \citep{svensson2015influence}; \swp{Need to say something more about Sven; come back later}
+an individual-level intrinsic kernel describes an intrinsic infectiousness of an infected individual.
+The population-level intrinsic kernel is given by integrating over individual variations:
 \begin{equation}
@@
 \end{equation}
-Then, the intrinsic generation-interval is simply the population-level kernel normalized by the basic reproductive number:
+The intrinsic generation-interval is simply the population-level kernel normalized by the basic reproductive number:
 \begin{equation}
@@
 
-\subsection*{Right-censored intervals}
+\subsection{Right-censored intervals}
 
@@
 
-\subsection*{Conditional kernel}
+\subsection{Egocentric kernel}
 
-Given limited contact structure, a susceptible individual can be contacted multiple times;
+Within a limited contact structure, a susceptible individual can be contacted multiple times;
 infectious contacts give rise to infection only when the susceptible individual is contacted for the first time.
-Then, the conditional kernel (conditional on the infection occuring) of an individual $a$ is given by
+In particular, we can look at it from an egocentric point of view -- i.e., from a perspective of a single infector.
+Then, we can define the egocentric kernel, i.e., the rate at which secondary infections are realized by a single primary case $a$:
 \begin{equation}
-\hat{k}_a(\tau) = k_a(\tau) \exp \left(- \delta_a \int_0^\tau k_a(s) ds\right),
+\hat{K}_a(\tau) = K_a(\tau) \exp \left(- \delta_a \int_0^\tau K_a(s) ds\right),
 \end{equation}
-where $\delta_a$ is the dilution.
-The individual kernel, $k_a(s)$, measures the intrinsic infectiousness of an individual $a$,
-whereas $\delta_a k_a(s)$ measures the infectiousness of an individual $a$ for a particular susceptible individual.
-Then, the population-level conditional kernel is:
+where $K_a(\tau)$ is the individual-level intrinsic kernel and $e^{- \delta_a \int_0^\tau K_a(s) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by individual $a$.
+The dilution term, $\delta_a$, models how contacts are distributed among susceptible acquaintances.
+For example, when infectious contacts are distributed equally in a homogeneous population, we have $\delta_a = 1/(N-1)$, where $N$ is the population size.
+
+The population-level egocentric kernel is given by integrating over individual variations:
 \begin{equation}
-\hat{k}(\tau) = \int \hat{k}_a(\tau) dA,
+\hat{K}(\tau) = \int \hat{K}_a(\tau) dA,
 \end{equation}
-and the conditional generation-interval distribution, defined to be the distribution of times at which secondary infections occur from a single primary infector, is:
+and the population-level egocentric generation-interval distribution is:
 \begin{equation}
-\hat{g}(\tau) = \frac{\hat{k}(\tau)}{\int \hat{k} \tau}.
+\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K} \tau}.
 \label{eq:conditional}
 \end{equation}
-\swp{Need to say something about trapman. Maybe say this is a generalization of Trapman's work...? Is that too much? Do we need to give an example of what conditional kernels look like?}
+The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an average primary case.
+For convenience, egocentric distribution refers to the population-level distribution, unless noted otherwise.
 
-The difference between conditional and intrinsic generation-interval distribution can be demonstrated by simulating stochastic infection processes on a small tree network (i.e, a single infected individual is connected to multiple susceptible individuals who are not connected with each other) with artificially high $\RR$.
-Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of generation times matches with the conditional generation-interval distribution (Figure~\ref{fig:local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
+For example, consider a susceptible-exposed-infected-recovered (SEIR) model.
+This model assumes that latent and infectious periods are exponentially distributed and can describe wide range of diseases.
+It has been shown that the intrinsic generation-interval distribution that corresponds to this model is given by [CITE]:
+\begin{equation}
+g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma t} - e^{-\sigma t}\right),
+\end{equation}
+where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
+Assuming that per-pair contact rate is $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution is given by:
+\begin{equation}
+\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)t} - e^{-\sigma t}\right)
+\end{equation}
+In this scenario, per-pair contact rate is effectively decreasing mean infectious period: an infected individual is can no longer infect anyone (hence no longer infectious) when all its susceptible acquaintances have ben infected.
 
-The conditional generation interval \eqref{conditional} only explains some of the reduction in generation times that occurs on most networks, however.
-Generation intervals are also shortened by indirect connections: if a susceptible individual is infected through another route while the focal individual is infected, this will shorten the generation interval \jd{This s.~is too repetitive.}
-Simulations on a small homogeneous network confirm this additional effect (\fref{local}, right panel). 
-In general, estimating the effects of most networks on observed generation intervals is expected to be a hard problem.
+The difference between egocentric and intrinsic generation-interval distribution can be demonstrated by simulating stochastic infection processes on a small tree network (i.e, a single infected individual is connected to multiple susceptible individuals who are not connected with each other) with artificially high per-pair contact rate.
+Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of realized generation times matches with the egocentric generation-interval distribution (\fref{fig:local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
 
@@
 \end{figure}
+
+The egocentric generation interval \eqref{conditional} only explains some of the reduction in generation times that occurs on most networks, however.
+Generation intervals are also shortened by indirect connections: a susceptible individual can be infected through another route before the focal individual makes infectious contacts.
+Simulations on a small homogeneous network confirm this additional effect (\fref{local}, right panel). 
+In general, estimating the effects of most networks on observed generation intervals is expected to be a hard problem.
+
+\subsection{Linking growth rate and reproductive number}
 

### 2018-10-10 doc/network.tex
a little more

@@
 Simulations on a small homogeneous network confirm this additional effect (\fref{local}, right panel). 
+The indirect spatial effect can occur locally (e.g., in a household) or globally (e.g., in a well-mixed population); global spatial effect provides an equivalent explanation for the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}).
 In general, estimating the effects of most networks on observed generation intervals is expected to be a hard problem.
@@
 \subsection{Linking growth rate and reproductive number}
+
+The intrinsic generation-interval distribution provides a link between the speed (exponential growth rate, $r$) and the strength (reproductive number, $\RR$) of an epidemic via Euler-Lotka equation [CITE] but assumes a homogeneously mixing population.
+Instead, we can derive a relationship between $r$ and $\RR$ that accounts for the egocentric effect using the egocentric generation-interval distribution:
+\begin{equation}
+\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
+\end{equation}
+This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure. \swp{Need to confirm; Maybe prove it in appendix or something...how?}
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\hat{\RR}$ will always be smaller than $\RR$.
 

### 2018-10-10 doc/network.tex
forgot to push

@@
 \jd{Maybe moving on a good direction, but needs more work. The problem is that we need to talk clearly about the spatial correction without misleading people about the temporal correction.}
+\swp{Should we just talk about this in the paper rather than in intro? It feels like we're going into detail too much}
 
-\swp{Will rewrite this paragraph after writing the rest of it:}
+\swp{Will rewrite this paragraph after writing rest of the paper:}
 In this study, we explore the temporal and spatial variation in the observed generation-interval distributions obtained from contact tracing.
@@
 
-\section{Statistical approach to inferring generation-interval distribution}
+\section{Inferring generation-interval distributions from contact tracing data}
+
+When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) global spatial effect.
+Right-censoring effect and egocentric effect are well-understood and can be corrected for.
+While local and global spatial effects are difficult to estimate, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
+Both local and global effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differ:
+local spatial effect will be important early in an epidemic whereas the global effect will be small until after the exponential growth period has ended.
 

### 2018-11-05 doc/network.tex
update

@@
 However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
-Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can easily be estimated from incidence data [CITE].
+Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can easily be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
 These two quantities are linked by generation-interval distributions \citep{wallinga2007generation}.
@@
 
-\section{Inferring generation-interval distributions from contact tracing data}
+\section{Inferring generation-interval distributions from a contact tracing data}
 
 When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) global spatial effect.
-Right-censoring effect and egocentric effect are well-understood and can be corrected for.
-While local and global spatial effects are difficult to estimate, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
-Both local and global effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differ:
-local spatial effect will be important early in an epidemic whereas the global effect will be small until after the exponential growth period has ended.
+Right-censoring effect and egocentric effect are well-understood.
+While local and global spatial effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
+Both local and global effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
+local spatial effect will be important early in an epidemic whereas the global effect will be small until sufficient amount of susceptible individuals become infected.
+As a result, both local and egocentric effects affect the initial spread of a disease and therefore the realized reproductive number.
 
-It is clear that aggregating generation intervals through space and time yields a distribution that is different from the intrinsic distribution.
-When the epidemic is growing, the proportion of susceptible individuals remain approximately constant and the intrinsic (or conditional) generation-interval distributions can be inferred from the observed intervals.
-In this section, we investigate two statistical methods for recovering the intrinsic (or conditional) generation-interval distribution.
-Detailed derivations of the two methods can be found in section.
+Since the right-censoring effect is a sampling artifact, it must be corrected for.
+On the other hand, local and egocentric effects will be embedded within realized generation intervals and do not need to be corrected for as they reflect the characteristic of a realized epidemic.
+We claim that accounting for the right-censoring effect is sufficient and any necessary spatial component will be accounted for implicitly.
+Hereafter, we will refer to temporally corrected distribution as the effective generation-interval distribution to contrast from the intrinsic generation-interval distribution; only in a large homogeneously mixing population, the corrected generation-interval distribution will be equivalent to the intrinsic generation-interval distribution.
+In this section, we investigate two statistical methods for correcting for temporal bias in a contact tracing data.
+Detailed derivations of the two methods can be found in section ???.
 
-\swp{maybe call it something else besides nonparametric method; it's still possible to make assumptions about the distribution and make inference}
-We refer to the first method as the population-level method as it relies on the observed distribution without taking into account who infected whom.
-As the observed generation-interval distribution is a weighted distribution of the intrinsic distribution, the intrinsic distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
+We refer to the first method as the population-level method as it relies on the  observed distribution aggregated across population.
+As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
 Hence, the population-level method relies on the observed generation intervals and the exponential growth rate:
@@
 
-While the non-parametric method is simple, it does not use all available information from contact tracing data: it does not take into account who infected whom.
-Assuming that infection occur as a poisson process in the beginning of an epidemic, we can obtain the following likelihood for the observed infection process:
+While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.
+Instead, we can model the initial spread of an epidemic using a poisson process on an individual-level and obtain the following likelihod::
 \begin{equation}
@@
 \end{equation}
-where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation interval distribution $g$.
+where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation-interval distribution $g$.
 A similar method was proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports rather than the observed generation intervals.
@@
 First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
-During the initial exponential growth period (approximately between 9 and 20 weeks), both nonparametric and parametric methods yield similar estimates of mean generation interval and reproductive number.
+During the initial exponential growth period (approximately between 9 and 20 weeks), both population-level and individual-level methods yield similar estimates of mean generation interval and reproductive number.
 However, we observe opposite trends after the growth period (approximately after 20 weeks).
-As the nonparametric method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
-On the other hand, as the parametric method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible results in underestimation of both mean generation interval and reproductive number.
-we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but the observed estimates always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation.
+As the population-level method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
+On the other hand, as the individual-level method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible is translated to underestimation of both mean generation interval and reproductive number.
+we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
 
@@
 
-Then, we compare accuracy and reliability of the two methods during the initial growth period using 100 stochastic simulations on a homogeneous network (Figure~\ref{fig:test}).
-We find that the parametric method provides the most consistent (least variable) estimates.
-The nonparametric method overestimates reproductive number initially but this is likely driven by the initial overestimation of the growth rate (Figure~\ref{fig:example}).
+We compare accuracy and reliability of the two methods during the initial growth period using 100 stochastic simulations on a homogeneous network (Figure~\ref{fig:test}).
+We find that the individual-level method provides the most consistent (least variable) estimates.
+The population method overestimates reproductive number initially but this is likely driven by the initial overestimation of the growth rate (Figure~\ref{fig:example}).
 
 \swp{May need to explain what coverage is...}
-Surprisingly, both methods fail to attain nominal coverage;
-throughout the estimation period, both methods attain approximately 70\% coverage.
-Even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution, making a wrong distributional assumption leads to underconfidence.
-To confirm that the undercoverage is caused by distribuional assumption, we simulated an epidemic using gamma kernel and found that using the true shape yields nominal coverage (see Appendix).
+We also compare coverage probabilities of these methods. 
+Coverage probabilities are defined as the proportion of onfidence intervals that contain the true value in a repeated simulations; for example, a 95\% confidence interval should attain 95\% coverage probability by its definition.
+Surprisingly, both methods only attain approximately 70\% coverage.
+Even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution, making a wrong distributional assumption leads to narrow confidence intervals.
+To confirm that the undercoverage is caused by distribuional assumption, we simulated an epidemic using gamma intrinsic generation-interval and found that fitting the true shape yields nominal coverage (see Appendix).
 These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].

### 2018-11-07 doc/network.tex
planning to do this
Working after meeting

@@
 \newcommand{\RR}{\ensuremath{{\mathcal R}}}
+\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
 \newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}
@@
 However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
-Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can easily be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
+Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
 These two quantities are linked by generation-interval distributions \citep{wallinga2007generation}.
@@
 
-Early in the epidemic when depletion of susceptible is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
+Early in the epidemic when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
 As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
@@
 A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution.
-When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long.
-\swp{Where can I cite champredon? First sentence? Last sentence? Or don't cite DC because I already did in the previous paragraph?}
+When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{champredon2015intrinsic}.
 
@@
 Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
+This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
 
-This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
-\jd{Maybe moving on a good direction, but needs more work. The problem is that we need to talk clearly about the spatial correction without misleading people about the temporal correction.}
-\swp{Should we just talk about this in the paper rather than in intro? It feels like we're going into detail too much}
-
-\swp{Will rewrite this paragraph after writing rest of the paper:}
 In this study, we explore the temporal and spatial variation in the observed generation-interval distributions obtained from contact tracing.
-We show that using the observed generation-interval distribution directly will always underestimate the reproductive number \swp{need to confirm this statement when we're done with the ms}. 
-We provide a statistical framework of recovering the intrinsic generation-interval distribution from the observed generation-interval distribution.
+We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
+We discuss statistical approaches to estimating the intrinsic distributions and \RR\ from observed generation-interval distributions.
 
@@
 an individual-level intrinsic kernel describes an intrinsic infectiousness of an infected individual.
+\jd{Sven has earlier stuff I think that influenced me (and that probably has this framework.}
 The population-level intrinsic kernel is given by integrating over individual variations:
@@
 \end{equation}
+\jd{Do we need to worry about assumptions for the rest of this \P?
+Once we have an aspect space we need to assume that individual properties are independent of risk of infection.
+Also, since that's wrong on our networks, do we have to figure out why we're approximately saved?
+I guess because of scaling?}
 The basic reproductive number -- average number of secondary cases caused by an average primary case in a fully susceptible population -- is defined as: 
@@
 \end{equation}
-The intrinsic generation-interval is simply the population-level kernel normalized by the basic reproductive number:
+The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
 \begin{equation}
@@
 
-\jd{Have to set this up to make sure we're not sounding too mathy.}
+\jd{Make sure we're not sounding too mathy.}
 Assume that contact tracing is performed from the beginning of an epidemic to time $t$.
@@
 \end{equation}
-We note that the expression in the denominater is equivalent to cumulative incidence at time $t$.
-The intuition behind this is that we are normalizing acrosss all incidence before time $t$.
+The expression in the denominator is equivalent to cumulative incidence at time $t$; the intuition here is that we are normalizing acrosss all incidence before time $t$.
 Then, we have
@@
 In general, estimating the effects of most networks on observed generation intervals is expected to be a hard problem. 
+\jd{The global spatial effect is sort of the non-spatial effect: the overall response to decreasing susceptibles, like in Champredon.}
 
@@
 This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure. \swp{Need to confirm; Maybe prove it in appendix or something...how?}
-As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\hat{\RR}$ will always be smaller than $\RR$.
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$.
+\jd{\Rhat\ will always be smaller than the incorrect version of \RR\ estimated from the intrinsic distribution. Also, larger than the true version of \RR\ based on a true spatial $g$.}
 
@@
 
+Finally, we simulate 100 epidemics on an empirical network [CITE] and compare the estimates of different methods.
+
 \clearpage

### 2018-11-12 doc/network.tex
lots of files

@@
 
+\section{Appendix}
+
+These are notes for now ... Need to move them elsewhere and change notation... but this is part of our figure for empirical network ...
+
+$$
+1 = \kappa \phi_L(\alpha) \frac{\lambda^{(net)}}{\alpha + \lambda^{(net)}} (1 - \phi_I(\alpha+\lambda^{(net)}))
+$$
+where in our specific example,
+$$
+\phi_L(\alpha) = \frac{4 \delta^2}{(2 \delta + \alpha)^2}
+$$
+and so we have
+$$
+1 = \frac{4 \delta^2}{(2 \delta + \alpha)^2} \frac{\kappa \lambda^{(net)}}{\lambda^{(net)} + \alpha + \gamma},
+$$
+which suggests that
+$$
+\lambda^{(net)} = \frac{(2\delta + \alpha)^2 (\alpha + \gamma)}{(\kappa - 1) 4 \delta^2 - 4 \delta \alpha - \alpha^2}
+$$
+We can substitute into the following equation to obtain the $r$ and $\RR$ relationship:
+$$
+\RR = \kappa \frac{\lambda^{(net)}}{\lambda^{(net)} + \gamma}
+$$
+
 \bibliography{network}

### 2018-11-12 R/estimate.R

@@ -36,6 +36,37 @@ bootfun <- function(gen,
 	est
 }
 
+minuslogl.gamma <- function(log.mean, log.shape, gen) {
+	mean <- exp(log.mean)
+	shape <- exp(log.shape)
+	rate <- shape/mean
+	
+	-sum(dgamma(gen, shape=shape, rate=rate, log=TRUE))
+}
+
+populationfun <- function(mean, shape, gen, r) {
+	m <- mle2(minuslogl.gamma, 
+			  start=list(log.mean=log(mean), log.shape=log(shape)),
+			  method="Nelder-Mead",
+			  optimizer="optim",
+			  data=list(gen=gen),
+			  control=list(maxit=10000)
+	)
+	
+	cc <- exp(coef(m))
+	
+	mean <- cc[1]
+	shape <- cc[2]
+	rate <- shape/mean
+	adj.rate <- rate - r
+	adj.mean <- shape/adj.rate
+	
+	data.frame(
+		estimate=unname(c((1+r*adj.mean/shape)^(shape), adj.mean, shape)),
+		type=c("RR", "mean", "shape")
+	) 
+}
+
 minuslogl.full <- function(log.R, log.mean, log.shape, data, tmax) {
 	R <- exp(log.R)
 	mean <- exp(log.mean)
@@ -56,24 +87,6 @@ minuslogl.full <- function(log.R, log.mean, log.shape, data, tmax) {
 	if (nll==0) return(Inf) else return(nll)
 }
 
-minuslogl.white <- function(log.R, log.mean, log.shape, data, k=50) {
-	R <- exp(log.R)
-	mean <- exp(log.mean)
-	shape <- exp(log.shape)
-	scale <- mean/shape
-	
-	t <- nrow(data)
-	
-	p_j <- diff(pgamma(0:k, shape=shape, scale=scale))
-	p_j <- p_j/sum(p_j)
-	
-	mu_t <- sapply(2:t, function(x) {
-		R * sum(data$cases[(x-1):max(x-k, 1)] * p_j[1:min(x-1, k)])
-	})
-	
-	-sum(dpois(tail(data$cases, -1), mu_t, log=TRUE))
-}
-
 parametricfun <- function(R, mean, shape,
 						  data,
 						  tmax) {
@@ -83,31 +96,6 @@ parametricfun <- function(R, mean, shape,
 		 optimizer="optim",
 		 data=list(data=data, tmax=tmax),
[…]

### 2018-11-12 analysis/cmp_compare.R

@@ -1,0 +1,116 @@
+library(igraph)
+library(dplyr)
+library(bbmle)
+library(epigrowthfit)
+source("../sim/ebola_param_seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+source("../R/empirical.R")
+
+load("../sim/cmp_seminr_sim.rda")
+
+cmpGraph <- read.table("../data/CA-CondMat.txt")
+cmpGraph <- graph.data.frame(cmpGraph)
+cmpGraph <- as.undirected(cmpGraph)
+cmpGraph <- delete_vertex_attr(cmpGraph, "name")
+
+d <- degree(cmpGraph)
+
+kappa <- var(d)/mean(d) + mean(d) -1
+
+beta <- gamma/7
+
+## This is SEIR
+## R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
+
+## This is SE2IR
+R0fun.network <- function(r) {
+	lambda <- (2 * sigma+r)^2 * (r + gamma)/((kappa-1) * 4 * sigma^2 - 4 * sigma * r - r^2)
+	
+	kappa * lambda/(lambda + gamma)
+}
+
+
+
+censor.gi <- lapply(
+	reslist
+	, network.generation
+	, plot=FALSE
+	, interval=c(10, 400)
+	, interval.type="cases"
+)
+
+growth <- lapply(reslist, function(x){
+	day <- floor(x$data$time) + 1
+	
+	fitdata <- data.frame(
+		day=unique(day),
+		cases=as.vector(table(day))
+	)
+	
+	fitdata <- fitdata[cumsum(fitdata$cases) <= 1000,]
+	
+	ee <- epigrowthfit(fitdata$day, fitdata$cases,
+				 model="logistic")
+	
+	dd <- as.data.frame(t(ee@growthRate))
+	
+	dd
+}) %>%
+	bind_rows(.id="sim")
+
+individual.est <- lapply(reslist, function(x){
+	tmax <- x$data[1000-9,1]
+	gendata <- generation.data(x, tmax=tmax)
+	
+	est <- as.data.frame(t(parametricfun(true.R, true.mean, true.shape, gendata, tmax)))
+	
+	est$type <- c("mean", "RR")
+	
+	est
+}) %>%
+	bind_rows(.id="sim")
+
+population.est <- lapply(1:50, function(x){
+	cg <- censor.gi[[x]]
[…]

### 2018-11-12 sim/cmp_seminr_sim.R

@@ -1,0 +1,28 @@
+library(igraph)
+source("ebola_param_seminr.R")
+source("../R/seminr.R")
+
+cmpGraph <- read.table("../data/CA-CondMat.txt")
+cmpGraph <- graph.data.frame(cmpGraph)
+cmpGraph <- as.undirected(cmpGraph)
+cmpGraph <- delete_vertex_attr(cmpGraph, "name")
+
+nsim <- 50
+
+beta <- gamma/7
+
+set.seed(101)
+reslist <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+	print(i)
+	rr <- seminr(cmpGraph, beta, sigma, m.sigma, gamma, n.gamma, I0 = 10, tmax=2000)
+	if(nrow(rr$data) > 1000) {
+		reslist[[i]] <- rr
+		i <- i+1
+	}
+	
+	save("reslist", file="cmp_seminr_sim.rda")
+}
+
+save("reslist", file="cmp_seminr_sim.rda")

### 2018-11-12 R/empirical.R
fig

@@ -19,7 +19,7 @@ infection.generation <- function(x) {
 ##' @param x simulation result
 ##' @param reference reference generation (first generation when there are n infected people)
 trapman.R0 <- function(x,
-                       reference=100,
+                       reference=75,
                        include.initial=FALSE) {
     gen <- infection.generation(x)
     if (!include.initial) gen <- gen[-1]
@@ -29,7 +29,7 @@ trapman.R0 <- function(x,
 }
 
 empirical.R0 <- function(x,
-                         n=100) {
+                         n=75) {
     o <- order(x$t_infected)
     tmax <- tail(x$data$time, 1)
     

### 2018-11-12 analysis/cmp_compare.R

@@ -30,8 +30,6 @@ R0fun.network <- function(r) {
 	kappa * lambda/(lambda + gamma)
 }
 
-
-
 censor.gi <- lapply(
 	reslist
 	, network.generation
@@ -105,12 +103,22 @@ intrinsic.est <- lapply(growth$value, function(x) 1/mean(exp(-x*gen)))
 
 network.est <- lapply(growth$value, R0fun.network)
 
+trapman.est <- lapply(reslist, trapman.R0)
 
-#plot(density(unlist(intrinsic.est)), ylim=c(0, 0.8))
-#lines(density(unlist(network.est)), col=4)
-#lines(density(sapply(individual.est, function(x) x[2,1])), col=2)
-#lines(density((observed.est %>% filter(type=="RR"))$estimate), col=3)
+empirical.est <- lapply(reslist, empirical.R0)
 
+RRdata <- data.frame(
+	observed=observed.est$estimate[observed.est$type=="RR"],
+	individual=individual.est$estimate[individual.est$type=="RR"],
+	intrinsic=unlist(intrinsic.est),
+	network=unlist(network.est),
+	trapman=unlist(trapman.est),
+	empirical=unlist(empirical.est)
+)
 
+gendata <- data.frame(
+	observed=observed.est$estimate[observed.est$type=="mean"],
+	individual=individual.est$estimate[individual.est$type=="mean"]
+)
 
-
+save("RRdata", "gendata", file="cmp_compare.rda")

### 2018-11-12 fig/cmp.R

@@ -1,0 +1,39 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("../sim/ebola_param_seminr.R")
+
+load("../analysis/cmp_compare.rda")
+
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+RRgroup <- data.frame(
+	key=c("contact\ntracing", "individual\ncorrection", "tree\nbased", "empirical", "local\ncorrection", "intrinsic"),
+	group=factor(c("tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
+				 levels=c("tracing based", "empirical", "individual based"))
+)
+
+RRdata2 <- RRdata %>%
+	gather(key, value) %>%
+	mutate(key=factor(key, 
+					  levels=c("observed", "individual", "trapman", "empirical", "network", "intrinsic"),
+					  labels=c("contact\ntracing", "individual\ncorrection", 
+					  		 "tree\nbased", "empirical", "local\ncorrection", "intrinsic"))) %>%
+	merge(RRgroup)
+
+ggR <- ggplot(RRdata2) +
+	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
+	ylab("Reproductive number") +
+	facet_wrap(~group, scale="free_x") +
+	theme(
+		panel.spacing = unit(0, "lines"),
+		strip.background = element_blank(),
+		legend.position = "none",
+		axis.title.x=element_blank()
+	)
+
+ggsave("cmp_reproductive.pdf", ggR, width=6, height=4)
+

### 2018-11-12 doc/network.tex
A few extra comments

@@
 \caption{\textbf{Temporal variation in the mean observed generation interval.}
-A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the observed mean generation interval was calculated the course of an epidemic (see methods).
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the observed mean generation interval was calculated over the course of an epidemic (see methods).
+\jd{Meaning, left-censored?}
 }
@@
 While local and global spatial effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
+\jd{Any objection to conceptualizing global as mean susceptible depletion? Then it's easy to measure. It's the non-egocentric non-global that's the problem.}
 Both local and global effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
@@
 \end{equation}
+While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.
 
-While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.
-Instead, we can model the initial spread of an epidemic using a poisson process on an individual-level and obtain the following likelihod::
+The individual-level method uses the poisson process to estimate a likelihood for the observations from each individual:
 \begin{equation}
@@
 It doesn't use generation interval either so I think we can just ignore it?}
+\jd{If it really is similar in some important way just stick with what you've already said. Otherwise, ignore it.}
 

### 2018-11-12 analysis/cmp_compare.R
push

@@ -18,7 +18,7 @@ d <- degree(cmpGraph)
 
 kappa <- var(d)/mean(d) + mean(d) -1
 
-beta <- gamma/7
+beta <- 0.4/5
 
 ## This is SEIR
 ## R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
@@ -48,10 +48,9 @@ growth <- lapply(reslist, function(x){
 	
 	fitdata <- fitdata[cumsum(fitdata$cases) <= 1000,]
 	
-	ee <- epigrowthfit(fitdata$day, fitdata$cases,
-				 model="logistic")
+	ee <- MASS::glm.nb(cases~day, data=fitdata)
 	
-	dd <- as.data.frame(t(ee@growthRate))
+	dd <- data.frame(r=coef(ee)[2])
 	
 	dd
 }) %>%
@@ -61,7 +60,7 @@ individual.est <- lapply(reslist, function(x){
 	tmax <- x$data[1000-9,1]
 	gendata <- generation.data(x, tmax=tmax)
 	
-	est <- as.data.frame(t(parametricfun(true.R, true.mean, true.shape, gendata, tmax)))
+	est <- as.data.frame(t(parametricfun(5, true.mean, true.shape, gendata, tmax)))
 	
 	est$type <- c("mean", "RR")
 	
@@ -99,9 +98,9 @@ observed.est <- lapply(1:50, function(x){
 }) %>%
 	bind_rows(.id="sim")
 
-intrinsic.est <- lapply(growth$value, function(x) 1/mean(exp(-x*gen)))
+intrinsic.est <- lapply(growth$r, function(x) 1/mean(exp(-x*gen)))
 
-network.est <- lapply(growth$value, R0fun.network)
+network.est <- lapply(growth$r, R0fun.network)
 
 trapman.est <- lapply(reslist, trapman.R0)
 
@@ -110,6 +109,7 @@ empirical.est <- lapply(reslist, empirical.R0)
 RRdata <- data.frame(
 	observed=observed.est$estimate[observed.est$type=="RR"],
 	individual=individual.est$estimate[individual.est$type=="RR"],
+	population=population.est$estimate[population.est$type=="RR"],
 	intrinsic=unlist(intrinsic.est),
 	network=unlist(network.est),
 	trapman=unlist(trapman.est),
@@ -118,7 +118,8 @@ RRdata <- data.frame(
 
 gendata <- data.frame(
[…]

### 2018-11-12 fig/cmp.R

@@ -11,23 +11,24 @@ scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,p
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
 
 RRgroup <- data.frame(
-	key=c("contact\ntracing", "individual\ncorrection", "tree\nbased", "empirical", "local\ncorrection", "intrinsic"),
-	group=factor(c("tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
+	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "local\ncorrection", "intrinsic"),
+	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
 				 levels=c("tracing based", "empirical", "individual based"))
 )
 
 RRdata2 <- RRdata %>%
 	gather(key, value) %>%
 	mutate(key=factor(key, 
-					  levels=c("observed", "individual", "trapman", "empirical", "network", "intrinsic"),
-					  labels=c("contact\ntracing", "individual\ncorrection", 
+					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
+					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
 					  		 "tree\nbased", "empirical", "local\ncorrection", "intrinsic"))) %>%
-	merge(RRgroup)
+	merge(RRgroup) %>%
+	filter(key != "tree\nbased")
 
 ggR <- ggplot(RRdata2) +
 	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
 	ylab("Reproductive number") +
-	facet_wrap(~group, scale="free_x") +
+	facet_grid(~group, scale="free", space="free_x") +
 	theme(
 		panel.spacing = unit(0, "lines"),
 		strip.background = element_blank(),
@@ -36,4 +37,3 @@ ggR <- ggplot(RRdata2) +
 	)
 
 ggsave("cmp_reproductive.pdf", ggR, width=6, height=4)
-

### 2018-11-12 sim/cmp_seminr_sim.R

@@ -9,7 +9,7 @@ cmpGraph <- delete_vertex_attr(cmpGraph, "name")
 
 nsim <- 50
 
-beta <- gamma/7
+beta <- 0.4/5
 
 set.seed(101)
 reslist <- vector('list', nsim)

### 2018-11-12 R/estimate.R
new fig

@@ -98,16 +98,18 @@ parametricfun <- function(R, mean, shape,
 		 control=list(maxit=10000)
 	)
 	
-	pp <- profile(m, 1:2)
+	pp <- profile(m)
 	
 	ci <- exp(confint(pp))
 	
 	est <- rbind(
-		exp(coef(m)[1:2]),
+		exp(coef(m)),
 		t(ci)
 	)
 	
 	rownames(est) <- c("estimate", "lwr", "upr")
 	
-	cbind(est[,2], est[,1])
+	colnames(est) <- NULL
+	
+	est
 }

### 2018-11-12 analysis/cmp_compare.R

@@ -62,11 +62,20 @@ individual.est <- lapply(reslist, function(x){
 	
 	est <- as.data.frame(t(parametricfun(5, true.mean, true.shape, gendata, tmax)))
 	
-	est$type <- c("mean", "RR")
+	est$type <- c("RR", "mean", "shape")
 	
 	est
 }) %>%
 	bind_rows(.id="sim")
+
+individual.RR <- lapply(1:50, function(x){
+	gg <- growth[x,2]
+	
+	mean <- filter(individual.est, type=="mean")[x,2]
+	shape <- filter(individual.est, type=="shape")[x,2]
+	
+	(1 + gg * 1/shape * mean)^(shape)
+})
 
 population.est <- lapply(1:50, function(x){
 	cg <- censor.gi[[x]]
@@ -108,7 +117,7 @@ empirical.est <- lapply(reslist, empirical.R0)
 
 RRdata <- data.frame(
 	observed=observed.est$estimate[observed.est$type=="RR"],
-	individual=individual.est$estimate[individual.est$type=="RR"],
+	individual=unlist(individual.RR),
 	population=population.est$estimate[population.est$type=="RR"],
 	intrinsic=unlist(intrinsic.est),
 	network=unlist(network.est),

### 2019-01-15 doc/network.tex
trying to get back to working on this ms

@@
 An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
-Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic [CITE].
+Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \cite{anderson1991infectious, diekmann1990definition}.
 However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
 Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
-These two quantities are linked by generation-interval distributions \citep{wallinga2007generation}.
+These two quantities are linked by generation-interval distributions \citep{wallinga2007generation, park2019practical}.
 
 Generation interval is defined as the time between when a person becomes infected and when that person infects another person.
-Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time}.
+Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.
 There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals are measured by evaluating the infectiousness of infected people,
@@
 
-\swp{Going to rename the sections... but I think I like the current order...}
+\swp{Might rename the sections later... but I think I like the current order...}
 \section{Generation-interval distributions across time}
@@
 Let $K_a(\tau)$ be an individual-level intrinsic kernel, i.e., the rate at which secondary infections are expected to be caused by an individual $a$ \citep{svensson2015influence}; \swp{Need to say something more about Sven; come back later}
+\swp{TODO: read \cite{svensson2007note} and \cite{tomba2010some}}
 an individual-level intrinsic kernel describes an intrinsic infectiousness of an infected individual.
@@
 I guess because of scaling?}
+\swp{What do you mean by individual properties?}
 The basic reproductive number -- average number of secondary cases caused by an average primary case in a fully susceptible population -- is defined as: 
@@
 where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
-Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation [CITE]:
+Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation \citep{lotka1907relation}:
 \begin{equation}
@@
 \jd{Make sure we're not sounding too mathy.}
+% For an emerging epidemic, it is reasonable to assume that contact tracing is performed from the beginning of an epidemic.
 Assume that contact tracing is performed from the beginning of an epidemic to time $t$.
@@
 \end{equation}
-Then, the censored interval at time $t$ is given by
+Then, the censored generation-interval at time $t$ is given by
 \begin{equation}
@@
 In particular, we can look at it from an egocentric point of view -- i.e., from a perspective of a single infector.
-Then, we can define the egocentric kernel, i.e., the rate at which secondary infections are realized by a single primary case $a$:
+We define the egocentric kernel, i.e., the rate at which secondary infections are realized by a single primary case $a$ in the absence of other infectors:
 \begin{equation}
@@
 \end{equation}
-In this scenario, per-pair contact rate is effectively decreasing mean infectious period: an infected individual is can no longer infect anyone (hence no longer infectious) when all its susceptible acquaintances have ben infected.
+In this scenario, per-pair contact rate effectively decreases mean infectious period: an infected individual can no longer infect anyone (hence no longer infectious) when all its susceptible acquaintances have ben infected.
 
 The difference between egocentric and intrinsic generation-interval distribution can be demonstrated by simulating stochastic infection processes on a small tree network (i.e, a single infected individual is connected to multiple susceptible individuals who are not connected with each other) with artificially high per-pair contact rate.
-Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of realized generation times matches with the egocentric generation-interval distribution (\fref{fig:local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
+Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of realized generation times matches with the egocentric generation-interval distribution (\fref{local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
 
@@
 
-The egocentric generation interval \eqref{conditional} only explains some of the reduction in generation times that occurs on most networks, however.
+The egocentric generation interval \eref{conditional} only explains some of the reduction in generation times that occurs on most networks, however.
 Generation intervals are also shortened by indirect connections: a susceptible individual can be infected through another route before the focal individual makes infectious contacts.
 Simulations on a small homogeneous network confirm this additional effect (\fref{local}, right panel).
-The indirect spatial effect can occur locally (e.g., in a household) or globally (e.g., in a well-mixed population); global spatial effect provides an equivalent explanation for the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}).
-In general, estimating the effects of most networks on observed generation intervals is expected to be a hard problem. 
+
+In general, there are two factors that cause further reduction in the mean generation interval: (1) the local spatial effect (e.g., in a household) and (2) the overall depletion of susceptible individuals in the population;
+the latter effect provides an equivalent explanation for the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}). 
+Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the effects of most networks on observed generation intervals is expected to be a hard problem.
 \jd{The global spatial effect is sort of the non-spatial effect: the overall response to decreasing susceptibles, like in Champredon.}
+\swp{Tried to fix this}
 
@@
 
-The intrinsic generation-interval distribution provides a link between the speed (exponential growth rate, $r$) and the strength (reproductive number, $\RR$) of an epidemic via Euler-Lotka equation [CITE] but assumes a homogeneously mixing population.
+The intrinsic generation-interval distribution provides a link between the speed (exponential growth rate, $r$) and the strength (reproductive number, $\RR$) of an epidemic via Euler-Lotka equation \citep{lotka1907relation} but assumes a homogeneously mixing population.
 Instead, we can derive a relationship between $r$ and $\RR$ that accounts for the egocentric effect using the egocentric generation-interval distribution:
@@
 This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure. \swp{Need to confirm; Maybe prove it in appendix or something...how?}
-As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$.
-\jd{\Rhat\ will always be smaller than the incorrect version of \RR\ estimated from the intrinsic distribution. Also, larger than the true version of \RR\ based on a true spatial $g$.}
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
+Similarly, \Rhat\ will always be greater than the \emph{true} reproduction number \RR\ as it does not account for other spatial effects that can decrease the mean generation interval.
 
@@
 
-When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) global spatial effect.
+When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) mean susceptible depletion effect.
 Right-censoring effect and egocentric effect are well-understood.
-While local and global spatial effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
-\jd{Any objection to conceptualizing global as mean susceptible depletion? Then it's easy to measure. It's the non-egocentric non-global that's the problem.}
-Both local and global effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
-local spatial effect will be important early in an epidemic whereas the global effect will be small until sufficient amount of susceptible individuals become infected.
-As a result, both local and egocentric effects affect the initial spread of a disease and therefore the realized reproductive number.
+While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
+Both local spatial and susceptible depletion effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
+local spatial effect will be important early in an epidemic whereas the susceptible depletion will be small until sufficient amount of susceptible individuals become infected.
+As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local spatial and egocentric effects.
 
 Since the right-censoring effect is a sampling artifact, it must be corrected for.
-On the other hand, local and egocentric effects will be embedded within realized generation intervals and do not need to be corrected for as they reflect the characteristic of a realized epidemic.
+On the other hand, local and egocentric effects will be embedded within realized generation intervals and do not need to be corrected for as they reflect the unobserved characteristic of a realized epidemic.
 We claim that accounting for the right-censoring effect is sufficient and any necessary spatial component will be accounted for implicitly.
@@
 
-The individual-level method uses the poisson process to estimate a likelihood for the observations from each individual:
+The individual-level method assumes that the initial exponential growth phase of an epidemic can be approximated by a Poisson process to derive a likelihood for the observations from each individual:
 \begin{equation}
@@
 where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation-interval distribution $g$.
-A similar method was proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports rather than the observed generation intervals.
+A similar method has beem proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports rather than the observed generation intervals.
 
@@
 we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
+Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproduction number; this is due to initial overestimation of the initial growth rate and we generally expect reproduction to be underestimated.
+\swp{Maybe another figure in the supp}
 
@@
 We find that the individual-level method provides the most consistent (least variable) estimates.
-The population method overestimates reproductive number initially but this is likely driven by the initial overestimation of the growth rate (Figure~\ref{fig:example}).
+The population method overestimates reproductive number initially but this is likely to be driven by the initial overestimation of the growth rate as discussed above (Figure~\ref{fig:example}).
 
-\swp{May need to explain what coverage is...}
 We also compare coverage probabilities of these methods. 
-Coverage probabilities are defined as the proportion of onfidence intervals that contain the true value in a repeated simulations; for example, a 95\% confidence interval should attain 95\% coverage probability by its definition.
+Coverage probabilities are defined as the proportion of confidence intervals that contain the true value in a repeated simulations; for example, a 95\% confidence interval should attain 95\% coverage probability by its definition.
 Surprisingly, both methods only attain approximately 70\% coverage.
@@
 These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].
-\swp{Why does nonparametric method undercover?}
 

### 2019-01-20 doc/network.tex
trying to do some work

@@
 An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
-Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \cite{anderson1991infectious, diekmann1990definition}.
+Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
 However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
-Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
+Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
 These two quantities are linked by generation-interval distributions \citep{wallinga2007generation, park2019practical}.
@@
 In this study, we explore the temporal and spatial variation in the observed generation-interval distributions obtained from contact tracing.
-We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
-We discuss statistical approaches to estimating the intrinsic distributions and \RR\ from observed generation-interval distributions.
+We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
+We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
 
-\swp{Might rename the sections later... but I think I like the current order...}
+\swp{I might rename the sections later but I think I like the current order.}
 \section{Generation-interval distributions across time}
 
-\subsection{Intrinsic kernel}
+\subsection{Intrinsic generation-interval distribution}
+
+Generation-interval distributions are often viewed from a population-level pespective, as a distribution describing \emph{expected} infection times of \emph{an average} infected individuals, but we can distinguish population-level distributions from individual-level distributions;
+making this distinction clear will be particularly useful when we discuss spatial components.
+An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.
 
 We begin by introducing an analytical framework.
-Let $K_a(\tau)$ be an individual-level intrinsic kernel, i.e., the rate at which secondary infections are expected to be caused by an individual $a$ \citep{svensson2015influence}; \swp{Need to say something more about Sven; come back later}
-\swp{TODO: read \cite{svensson2007note} and \cite{tomba2010some}}
+Let $K_a(\tau)$ be an individual-level intrinsic kernel, i.e., the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence};
 an individual-level intrinsic kernel describes an intrinsic infectiousness of an infected individual.
-\jd{Sven has earlier stuff I think that influenced me (and that probably has this framework.}
 The population-level intrinsic kernel is given by integrating over individual variations:
@@
 \swp{What do you mean by individual properties?}
+Similar to the inidividual-level kernel, the population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} individual.
+
 The basic reproductive number -- average number of secondary cases caused by an average primary case in a fully susceptible population -- is defined as: 
@@
 As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
-Similarly, \Rhat\ will always be greater than the \emph{true} reproduction number \RR\ as it does not account for other spatial effects that can decrease the mean generation interval.
+Similarly, \Rhat\ will always be greater than the \emph{true} reproductive number \RR\ as it does not account for other spatial effects that can decrease the mean generation interval.
 
@@
 As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
-Hence, the population-level method relies on the observed generation intervals and the exponential growth rate:
+Hence, the population-level method relies on the observed generation intervals and the exponential growth rate \citep{tomba2010some}:
 \begin{equation}
@@
 we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
-Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproduction number; this is due to initial overestimation of the initial growth rate and we generally expect reproduction to be underestimated.
+Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproductive number; this is due to initial overestimation of the initial growth rate and we generally expect reproductive to be underestimated.
 \swp{Maybe another figure in the supp}
@@
 
+\begin{figure}
+\includegraphics[width=\textwidth]{../fig/compare_methods.pdf}
+\caption{Fill this out.}
+\label{fig:test}
+\end{figure}
+
 We also compare coverage probabilities of these methods. 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{../fig/compare_methods.pdf}
+\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
 \caption{Fill this out.}
-\label{fig:test}
+\label{fig:cmp}
 \end{figure}
 
-Finally, we simulate 100 epidemics on an empirical network [CITE] and compare the estimates of different methods.
+%% TODO: put SEIR in main text and show SE2IR in appendix for consistency
+Finally, we simulate 100 epidemics with Ebola-liked parameters on an empirical network [CITE] and compare the estimates of reproductive number based on different methods with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
+As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
+estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
+Direct estimates based on contact tracing data severely underestimates the empirical estimates.
+Both population and individual corrections provide similar estimates to empirical reproductive numbers.
+For smaller values of \RR, we expect the differences to become smaller.
 

### 2019-01-20 fig/individual_and_population.R

@@ -1,0 +1,46 @@
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+source("../sim/ebola_param.R")
+
+lat <- 1/sigma
+inf <- 1/gamma
+
+t <- seq(0, 35, by=0.001)
+
+dd_individual <- data.frame(
+	time=t,
+	density=ifelse(t>lat & t < lat+inf, 2/inf, 0),
+	type="Individual-level kernel"
+)
+
+dd_population <- data.frame(
+	time=t,
+	density=2 * intrinsic_fun(t),
+	type="Population-level kernel"
+)
+
+dd_population_poly <- data.frame(
+	time=c(t, 35),
+	density=c(2 * intrinsic_fun(t), 0)
+)
+
+g1 <- ggplot(dd_individual) +
+	geom_line(aes(time, density), lwd=1.2) +
+	scale_x_continuous("Time (days)", expand=c(0, 0)) +
+	scale_y_continuous(expression(K[a](tau)), expand=c(0,0), limit=c(0, 0.45)) +
+	facet_wrap(~type) +
+	theme(
+		strip.background = element_blank(),
+		panel.grid = element_blank()
+	)
+
+g2 <- g1 %+% dd_population
+
+grid.arrange(
+	g1 + geom_polygon(aes(time, density), alpha=0.3),
+	g2 + geom_polygon(data=dd_population_poly, aes(time, density), alpha=0.3),
+	nrow=1
+)
+
+
+

### 2019-01-20 sim/cmp_seir_sim.R

@@ -16,13 +16,13 @@ reslist <- vector('list', nsim)
 i <- 1
 while (i <= nsim) {
 	print(i)
-	rr <- seminr(cmpGraph, beta, sigma, m.sigma, gamma, n.gamma, I0 = 10, tmax=2000)
+	rr <- seminr(cmpGraph, beta, sigma, 1, gamma, 1, I0 = 10, tmax=2000)
 	if(nrow(rr$data) > 1000) {
 		reslist[[i]] <- rr
 		i <- i+1
 	}
 	
-	save("reslist", file="cmp_seminr_sim.rda")
+	save("reslist", file="cmp_seir_sim.rda")
 }
 
-save("reslist", file="cmp_seminr_sim.rda")
+save("reslist", file="cmp_seir_sim.rda")

### 2019-01-20 doc/network.tex
bit more work

@@
 
-Generation-interval distributions are often viewed from a population-level pespective, as a distribution describing \emph{expected} infection times of \emph{an average} infected individuals, but we can distinguish population-level distributions from individual-level distributions;
+\begin{figure}
+\includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
+\caption{\textbf{Comparison of individual- and population-level kernels.}
+(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days. 
+(Right) a population-level kernal of infected individuals with exponentially distributed latent and infectious period of 11.4 and 5 days, respectively. 
+Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
+}
+\label{fig:indpop}
+
+\end{figure}
+
+Generation-interval distributions are often viewed from a population-level pespective, as a distribution describing \emph{expected} infection times of \emph{an average} infected individuals, but we can distinguish population-level distributions from individual-level distributions (\fref{indpop});
 making this distinction clear will be particularly useful when we discuss spatial components.
 An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.
-
-We begin by introducing an analytical framework.
-Let $K_a(\tau)$ be an individual-level intrinsic kernel, i.e., the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence};
-an individual-level intrinsic kernel describes an intrinsic infectiousness of an infected individual.
-The population-level intrinsic kernel is given by integrating over individual variations:
+On the other hand, the population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
 \begin{equation}
@@
 \end{equation}
+This distribution describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
 \jd{Do we need to worry about assumptions for the rest of this \P?
@@
 \swp{What do you mean by individual properties?}
-Similar to the inidividual-level kernel, the population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} individual.
 
-The basic reproductive number -- average number of secondary cases caused by an average primary case in a fully susceptible population -- is defined as: 
+The basic reproductive number -- average number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
 \begin{equation}
@@
 \end{equation}
+
 Then, the renewal equation model can be used to model incidence over time:

### 2019-01-20 fig/individual_and_population.R

@@ -25,7 +25,7 @@ dd_population_poly <- data.frame(
 )
 
 g1 <- ggplot(dd_individual) +
-	geom_line(aes(time, density), lwd=1.2) +
+	geom_line(aes(time, density), lwd=1.1) +
 	scale_x_continuous("Time (days)", expand=c(0, 0)) +
 	scale_y_continuous(expression(K[a](tau)), expand=c(0,0), limit=c(0, 0.45)) +
 	facet_wrap(~type) +
@@ -36,11 +36,14 @@ g1 <- ggplot(dd_individual) +
 
 g2 <- g1 %+% dd_population
 
-grid.arrange(
+gtot <- arrangeGrob(
 	g1 + geom_polygon(aes(time, density), alpha=0.3),
-	g2 + geom_polygon(data=dd_population_poly, aes(time, density), alpha=0.3),
+	g2 + 
+		geom_polygon(data=dd_population_poly, aes(time, density), alpha=0.3) +
+		scale_y_continuous(expression(K(tau)), expand=c(0,0), limit=c(0, 0.45)),
 	nrow=1
 )
 
+ggsave("individual_and_population.pdf", gtot, width=6, height=3)
 
 

### 2019-01-20 doc/network.tex
commit

@@
 
-\begin{figure}
+\begin{figure}[t]
 \includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
@@
 
-Generation-interval distributions are often viewed from a population-level pespective, as a distribution describing \emph{expected} infection times of \emph{an average} infected individuals, but we can distinguish population-level distributions from individual-level distributions (\fref{indpop});
+Generation-interval distributions are often viewed from a population-level pespective but we can distinguish population-level distributions from individual-level distributions (\fref{indpop}); 
 making this distinction clear will be particularly useful when we discuss spatial components.
 An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.
-On the other hand, the population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
+The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
 \begin{equation}
@@
 \end{equation}
-This distribution describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
+The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
 \jd{Do we need to worry about assumptions for the rest of this \P?
@@
 
+There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
 The basic reproductive number -- average number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
@@
 
-Then, the renewal equation model can be used to model incidence over time:
+Since the intrinsic generation-interval distribution describes time distribution of new infections, it can be used to model incidence over time:
 \begin{equation}
@@
 where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
-Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation \citep{lotka1907relation}:
+This model, also referred to as the renewal equation, represents wide range of epidemic models.
+Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation, which provides a direct link between speed and strength of an epidemic.
+ \citep{lotka1907relation}:
 \begin{equation}
@@
 
-\subsection{Right-censored intervals}
+\subsection{Right-censored generation-interval distribution}
 

### 2019-01-20 doc/network.tex
bib

@@
 where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
-This model, also referred to as the renewal equation, represents wide range of epidemic models.
-Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation, which provides a direct link between speed and strength of an epidemic.
- \citep{lotka1907relation}:
+This model, also referred to as the renewal equation, represents wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
+Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
 \begin{equation}

### 2019-01-21 doc/network.tex
come back later

@@
 \jd{Make sure we're not sounding too mathy.}
+Generation intervals can be sampled during an outbreak by tracing infected individuals backward.
+As a result of backward sampling, it is impossible to sample from infections that occur beyond the sampling time and the generation-interval samples suffer from right censoring.
+In this section, we discuss the effect of right-censoring on the observed generation-interval distribution.
+
+Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution, but the backward generation-interval distribution only consider infections that are single generation apart from a focal time point.
+Instead, contact tracing can be performed over multiple generations, all the way back to the primary case.
+
 % For an emerging epidemic, it is reasonable to assume that contact tracing is performed from the beginning of an epidemic.

### 2019-01-22 doc/network.tex
more progress

--- 2019-01-20 doc/network.tex
@@
 \jd{Make sure we're not sounding too mathy.}
-% For an emerging epidemic, it is reasonable to assume that contact tracing is performed from the beginning of an epidemic.
-Assume that contact tracing is performed from the beginning of an epidemic to time $t$.
-The number of infection occuring at time $s$ caused by infectors who were themselves infected at time $s-\tau$ is given by
+\swp{I tried.}
+Generation intervals can be sampled during an outbreak by tracing infected individuals backward.
+As a result of backward sampling, it is impossible to sample from infections that occur beyond the sampling time and the generation-interval samples suffer from right censoring.
+In this section, we discuss the effect of right-censoring on the observed generation-interval distribution.
+
+Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time period.
+Instead, contact tracing can be performed over multiple generations, tracing all the way back to the primary case.
+Hence, to obtain a theoretical sampling distribution for the right-censored generation intervals, we want to consider all infections that occur from the beginning of an epidemic to a time period when contact tracing was performed.
+
+\begin{figure}[t]
+\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
+\caption{\textbf{Temporal variation in the mean observed generation interval.}
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the observed (right-censored) mean generation interval was calculated over the course of an epidemic (see methods).
+}
+\label{fig:censor}
+\end{figure}
+
+The number of infection occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
 \begin{equation}
@@
 \end{equation}
-Then, total number of secondary infections that are $\tau$ time steps apart and occur before time $t$:
+Then, total number of secondary infections that are $\tau$ time steps apart and occur before time $t$ can be obtained by integrating the above expression:
 \begin{equation} 
@@
 \end{equation}
-Then, the censored generation-interval at time $t$ is given by
-\begin{equation}
-c_t(\tau)= \frac{\RR \int_\tau^t i(s-\tau) g(\tau) S(s) ds}{\RR \int_0^t \int_x^t i(s-x) g(x) S(s) ds dx}.
-\end{equation}
-The expression in the denominator is equivalent to cumulative incidence at time $t$; the intuition here is that we are normalizing acrosss all incidence before time $t$.
-Then, we have
+The distribution of the secondary infections essentially represent the realized generation-interval distribution.
+Normalizing by the total number of cases before time $t$, we obtain a theoretical sampling distribution for the right-censored (censored at time $t$) generation intervals:
 \begin{equation}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
-\caption{\textbf{Temporal variation in the mean observed generation interval.}
-A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the observed mean generation interval was calculated over the course of an epidemic (see methods).
-\jd{Meaning, left-censored?}
-}
-\label{fig:censor}
-\end{figure}
-
-When a disease is at (or near) endemic equilibrium, the number of susceptibles remains (approximately) constant over time, and the observed generation-interval distribution should be similar to the intrinsic generation-interval distribution.
-\swp{Probably don't need a figure for this}
+When a disease is at (or near) endemic equilibrium, the number of susceptibles remains (approximately) constant over time, and we expect the observed generation-interval distribution to be similar to the intrinsic generation-interval distribution.
 
@@
 
-Within a limited contact structure, a susceptible individual can be contacted multiple times;
-infectious contacts give rise to infection only when the susceptible individual is contacted for the first time.
-In particular, we can look at it from an egocentric point of view -- i.e., from a perspective of a single infector.
+An infected individual may contact the same susceptible individual multiple times but  infectious contacts give rise to infection only when the susceptible individual is contacted for the first time.
+Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average, in a limited network.
+To understand the effect of multiple contacts on realized generation intervals, we have to consider the infection process from an egocentric point of view, only taking into account infectious contacts made by a single infector.
 We define the egocentric kernel, i.e., the rate at which secondary infections are realized by a single primary case $a$ in the absence of other infectors:
@@
 \end{equation}
-The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an average primary case.
+The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case.
 For convenience, egocentric distribution refers to the population-level distribution, unless noted otherwise.
@@
 This model assumes that latent and infectious periods are exponentially distributed and can describe wide range of diseases.
-It has been shown that the intrinsic generation-interval distribution that corresponds to this model is given by [CITE]:
+The intrinsic generation-interval distribution that corresponds to this model can be written has [CITE]:
 \begin{equation}
@@
 
-The difference between egocentric and intrinsic generation-interval distribution can be demonstrated by simulating stochastic infection processes on a small tree network (i.e, a single infected individual is connected to multiple susceptible individuals who are not connected with each other) with artificially high per-pair contact rate.
+The difference between egocentric and intrinsic generation-interval distribution can be demonstrated by simulating stochastic infection processes on a small tree network (i.e, a single infected individual is connected to multiple susceptible individuals who are not connected with each other) with high per-pair contact rate.
 Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of realized generation times matches with the egocentric generation-interval distribution (\fref{local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
@@
 Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the effects of most networks on observed generation intervals is expected to be a hard problem.
+\cite{tomba2010some} showed that an analytical expression for the mean generation interval, which takes into account the local spatial effect, can be derived for a household with one primary case and two susceptible individuals, but solving for higher order structures seems both infeasible and impractical.
 \jd{The global spatial effect is sort of the non-spatial effect: the overall response to decreasing susceptibles, like in Champredon.}
@@
 As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
-Similarly, \Rhat\ will always be greater than the \emph{true} reproductive number \RR\ as it does not account for other spatial effects that can decrease the mean generation interval.
+Similarly, \Rhat\ will always be greater than the \emph{true} reproductive number \RR\ as it does not account for other effects (e.g., local spatial effects and mean susceptible-depletion effects) that can decrease the mean generation interval.
 
@@
 
-When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) mean susceptible depletion effect.
+When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) mean susceptible-depletion effect.
 Right-censoring effect and egocentric effect are well-understood.
@@
 Since the right-censoring effect is a sampling artifact, it must be corrected for.
-On the other hand, local and egocentric effects will be embedded within realized generation intervals and do not need to be corrected for as they reflect the unobserved characteristic of a realized epidemic.
+Local and egocentric effects are expected to be embedded within realized generation intervals and do not need to be corrected for as they reflect the unobserved characteristic of a realized epidemic.
 We claim that accounting for the right-censoring effect is sufficient and any necessary spatial component will be accounted for implicitly.
@@
 As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
-Hence, the population-level method relies on the observed generation intervals and the exponential growth rate \citep{tomba2010some}:
+Hence, the population-level method relies on the observed generation intervals and the exponential growth rate \citep{tomba2010some, nishiura2010time}:
 \begin{equation}
@@
 \end{equation}
+When the susceptible dynamics is known, this equation can be generalized by using \eref{obsg}.
 While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.
@@
 where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation-interval distribution $g$.
-A similar method has beem proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports rather than the observed generation intervals.
+A similar method has beem proposed by \cite{forsberg2008likelihood} but thier method uses on discretized incidence reports rather than the observed generation intervals.
 
-\swp{I might say something about independence transmission events in the results if I decide to compare forsberg2008 as well -- see forsberg 2008...}
-\swp{I got stuck on forsberg 2008.
-I get weird estimates with wide confidence intervals but I feel like that's what their method is supposed to do (based on figure 3 in their paper).
-It doesn't use generation interval either so I think we can just ignore it?}
-\jd{If it really is similar in some important way just stick with what you've already said. Otherwise, ignore it.}
-
-First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
+First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (\fref{example}).
 During the initial exponential growth period (approximately between 9 and 20 weeks), both population-level and individual-level methods yield similar estimates of mean generation interval and reproductive number.

### 2019-01-22 doc/network.tex
caption

@@
 \includegraphics[width=\textwidth]{../fig/local_effect.pdf}
-\caption{Fill this out.}
+\caption{
+\textbf{Comparison of spatial effects on the realized generation intervals.}
+(Left) The intrinsic generation-interval distribution corresponds to the distribution of all infectious contacts that were made, regardless of the outcome (i.e., whether it resulted in infection or not); 5000 stochastic simulations were performed on a tree network with 5 nodes (1 infector and 4 susceptibles).
+(Middle) The egocentric generation-interval distribution corresponds to the distribution of all infectious contacts made by a single primary infector that gave rise to infection; 5000 stochastic simulations were performed on a tree network with 5 nodes (1 infector and 4 susceptibles).
+(Right) The realized generation-interval distribution corresponds to the distribution of all infection times throughout an epidemic; 5000 stochastic simulations were performed on a homogeneous network with 5 nodes (1 infector and 4 susceptibles).
+\swp{Parameters? High per-pair contact? Need to describe and make the figure prettier}
+}
 \label{fig:local}
@@
 \includegraphics[width=\textwidth]{../fig/example.pdf}
-\caption{Fill this out.}
+\caption{\swp{Need to re-do this figure; initial overestimation of little $r$ is weird}}
 \label{fig:example}
@@
 \includegraphics[width=\textwidth]{../fig/compare_methods.pdf}
-\caption{Fill this out.}
+\caption{\swp{Need to re-do this figure; initial overestimation of little $r$ is weird}.}
 \label{fig:test}
@@
 \includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
-\caption{Fill this out.}
+\caption{\textbf{Comparison of estimates of reproductive number based on various methods}
+Using the observed generation-interval distributions without correcting for right-censoring severely underestimates the reproductive number.
+Similarly, using intrinsic and egocentric generation-interval distributions without accounting for local spatial effects overestimates the reproductive number.
+Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
+100 stochastic simulations were performed on an empirical network [CITE] using Ebola-like parameters (latent period = 11.4 days and infectious period = 5 days).
+}
 \label{fig:cmp}

### 2019-01-22 analysis/measles.R
smallpox

@@ -1,146 +1,0 @@
-library(deSolve)
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-intrinsic_fun <- function(tau, parm) {
-	with(as.list(parm), {
-		sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
-	})
-}
-
-## ode
-
-measles.ode <- function(t, y, parms) {
-	with(as.list(c(y, parms)), {
-		
-		beta <- b0 * (1 + b1 * cos(2 * pi * t/period))
-		
-		incidence <- beta * S * I/N
-		
-		dS <- mu * (N-S) - incidence
-		dE <- incidence - (sigma+mu)*E
-		dI <- sigma*E - (gamma+mu)*I
-		dR <- (gamma+mu)*I - mu*R
-		list(c(dS, dE, dI, dR))
-	})
-}
-
-sumfun <- function(out,
-				   tmax,
-				   t,
-				   parm,
-				   tauby=5) {
-	incidence <- out$incidence
-	c.incidence <- cumsum(incidence)
-	
-	genden <- intrinsic_fun(t, parm)
-	
-	censor <- function(tau, tmax) {
-		m <- which(t==tmax)
-		total <- c.incidence[m]
-		
-		i <- which(t==tau)
-		
-		sum(incidence[1:(m-i+1)] * out$S[i:m]/parm[["N"]])*genden[i]
-	}
-	
-	mean.censor <- function(tmax) {
-		if (tmax<=0){
-			return(0)
-		} else{
-			t <- t[t <= tmax]
-			
-			dc <- sapply(t, censor, tmax=tmax)
-			return(sum(dc*t)/sum(dc))
-		}
-	}
-	
-	tau <- seq(0, tmax, tauby)
-	cm <- sapply(tau, mean.censor)
-	
-	censor.df <- data.frame(
-		t=tau
-		, mean=cm
-	)
-	
-	censor.df
-}
-
-parms <- c(
-	b0=1250/365,
-	b1=0.15,
-	sigma=1/8,
-	gamma=1/5,
-	mu=1/50/365,
-	period=365,
-	N=5e7
-)
-
-y <- c(S=0.05, E=0.0001, I=0.0001, R=1-0.05-0.0001-0.0001) * 5e7
-
-dt <- 1
-dt2 <- 0.2
[…]

### 2019-01-22 analysis/measles_param.R

@@ -1,23 +1,0 @@
-set.seed(101)
-beta <- 2/5
-mu <- 1/50
-sigma <- 1/11.4
-m.sigma <- round(1.75)
-gamma <- 1/5
-n.gamma <- round((5/4.7)^2)
-N <- 5e7
-
-true.mean <- 1/sigma + 1/gamma
-
-nsamp <- 1e6
-
-lat <- rgamma(nsamp, m.sigma, sigma*m.sigma)
-inf <- rgamma(nsamp, n.gamma, gamma*n.gamma)
-
-gen <- lat + runif(nsamp, 0, sample(inf, prob=inf, replace=TRUE)) 
-
-true.var <- var(gen)
-
-true.shape <- true.mean^2/true.var
-
-true.R <- beta/gamma

### 2019-01-22 fig/smallpox.R

@@ -1,0 +1,75 @@
+library(dplyr)
+
+sp <- read.csv("../data/Smallpox_1951_Tilburg_NL.csv")
+
+sp2 <- sp %>%
+	mutate(exposure=as.numeric(as.Date(Exposure_date)),
+		   symptom=as.numeric(as.Date(DOD))) %>%
+	mutate(exposure=exposure-min(exposure, na.rm=TRUE),
+		   symptom=symptom-min(symptom, na.rm=TRUE))
+
+dd <- data.frame(
+	time=sp2$exposure,
+	gen=sp2$exposure - sp2$exposure[match(sp2$IDSource, sp2$ID)],
+	serial=sp2$symptom - sp2$symptom[match(sp2$IDSource, sp2$ID)],
+	source=sp2$IDSource,
+	recipient=sp2$ID,
+	type=sp2$Type
+) %>%
+	filter(!is.na(gen), gen > 0) %>% ## there is one case where gen < 0 so I get rid of it
+	arrange(time)
+
+dd_contact <- dd %>%
+	group_by(time) %>%
+	summarize(
+		sum=sum(gen),
+		weight=length(gen)
+	) %>%
+	mutate(
+		mean=cumsum(sum)/cumsum(weight)
+	)
+
+dd_case <- dd %>%
+	filter(type=="Case") %>%
+	group_by(time) %>%
+	summarize(
+		sum=sum(gen),
+		weight=length(gen)
+	) %>%
+	mutate(
+		mean=cumsum(sum)/cumsum(weight)
+	)
+
+dd_serial <- dd %>%
+	filter(type=="Case", !is.na(serial)) %>%
+	group_by(time) %>%
+	summarize(
+		sum=sum(serial),
+		weight=length(serial)
+	) %>%
+	mutate(
+		mean=cumsum(sum)/cumsum(weight)
+	)
+
+pdf("smallpox.pdf", width=10, height=3)
+par(mfrow=c(1, 3))
+plot(dd_contact$time, dd_contact$mean, type="l",
+	 xlab="Time (days since the first case)",
+	 ylab="Mean generation interval",
+	 ylim=c(13, 18.5))
+title("All contacts")
+abline(h=17.7, lty=2)
+
+plot(dd_case$time, dd_case$mean, type="l",
+	 xlab="Time (days since the first case)",
+	 ylab="Mean generation interval",
+	 ylim=c(13, 18.5))
+title("Symptomatic cases")
+abline(h=17.7, lty=2)
+
[…]

### 2019-01-29 doc/network.tex
smallpox
Got started (at least!)

@@
 
+%% SWP:
+%% Got data from
+%% https://www.ncbi.nlm.nih.gov/pubmed/29087987
+%% Try and analyze
+%% See what's different and understand what they did
+
 An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
@@
 while \emph{observed} generation intervals refer to the time between actual infection events.
-Observed generation intervals in turn can be \emph{aggregated} across time, measured \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others, or measured \emph{backward in time} by look at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
+Observed generation intervals in turn can be \emph{aggregated} across time, measured \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others, or measured \emph{backward in time} by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
 % When aggregated generation intervals are observed until a given time in an ongoing epidemic (often via contact tracing), we refer to these as \emph{censored} (or right-censored) intervals.
 
-Early in the epidemic when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
+\jd{Consider rewriting to use early-late as the first-level structure. Early is the part that people focus on most. We could even throw in censorship.}
+Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
 As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
@@
 
-In practice, generation intervals are measured by contact tracing throughout the course of an epidemic, usually aggregated to increase the amount of information available; we refer to these as \emph{censored} (or right-censored) intervals.
-When calculations are based on early-outbreak data, the available (censored) observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions.
+In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, these measurements are \emph{censored} (or right-censored) at the observation time.
+These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, they observed censored intervals will be shorter than intrinsic distributions early in the epidemic.
 
@@
 (Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days. 
-(Right) a population-level kernal of infected individuals with exponentially distributed latent and infectious period of 11.4 and 5 days, respectively. 
+(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively. 
 Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
@@
 
-Since the intrinsic generation-interval distribution describes time distribution of new infections, it can be used to model incidence over time:
+We assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
 \begin{equation}
-i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds
+i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds.
 \end{equation}
-where $i(\tau)$ is incidence at time $\tau$ and $S(\tau)$ is the proportion of the population susceptible.
-This model, also referred to as the renewal equation, represents wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
-Finally, assuming exponentially growing incidence ($i(t) = i(0) \exp(r t)$), we obtain the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
+This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
+Over a period of time where $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
 \begin{equation}

### 2019-01-31 doc/network.tex
Autocommit (doc)

@@
 
-\jd{Make sure we're not sounding too mathy.}
-\swp{I tried.}
-Generation intervals can be sampled during an outbreak by tracing infected individuals backward.
-As a result of backward sampling, it is impossible to sample from infections that occur beyond the sampling time and the generation-interval samples suffer from right censoring.
-In this section, we discuss the effect of right-censoring on the observed generation-interval distribution.
+\jd{Mathematically, early censored and early backward intervals are the same, but conceptually they're just different. In particular, censoring occurs even with forward tracing.}
+\jd{During exponential growth, the theoretical distributions are the same; maybe we should do a tiny bit of that math here so that we can talk more clearly about it.}
+\jd{We need to think about exposition. Maybe simple formulae for all of our interval types, both in general and under the assumption of exponential growth.}
+\jd{We should be saying that backward $\to$ censored in the limit at the same time we talk about the association below.}
 
+Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
 Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time period.
@@
 Hence, to obtain a theoretical sampling distribution for the right-censored generation intervals, we want to consider all infections that occur from the beginning of an epidemic to a time period when contact tracing was performed.
+\jd{I hope that none of these papers actually make the mistake that you imply, it's all about the limit.}
 
@@
 
-An infected individual may contact the same susceptible individual multiple times but  infectious contacts give rise to infection only when the susceptible individual is contacted for the first time.
+An infected individual may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
 Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average, in a limited network.
-To understand the effect of multiple contacts on realized generation intervals, we have to consider the infection process from an egocentric point of view, only taking into account infectious contacts made by a single infector.
-We define the egocentric kernel, i.e., the rate at which secondary infections are realized by a single primary case $a$ in the absence of other infectors:
+To explore the effect of multiple contacts on realized generation intervals, we first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
+We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case $a$ in the absence of other infectors:
 \begin{equation}
@@
 \end{equation}
-The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case.
-For convenience, egocentric distribution refers to the population-level distribution, unless noted otherwise.
+The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for brevity we will often omit ``population-level''.
 
 For example, consider a susceptible-exposed-infected-recovered (SEIR) model.
-This model assumes that latent and infectious periods are exponentially distributed and can describe wide range of diseases.
+This widely used model assumes that latent and infectious periods are exponentially distributed.
 The intrinsic generation-interval distribution that corresponds to this model can be written has [CITE]:
@@
 
-The difference between egocentric and intrinsic generation-interval distribution can be demonstrated by simulating stochastic infection processes on a small tree network (i.e, a single infected individual is connected to multiple susceptible individuals who are not connected with each other) with high per-pair contact rate.
-Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of realized generation times matches with the egocentric generation-interval distribution (\fref{local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
+This calculation can be validated by simulating stochastic infection processes on a ``star'' network (i.e, a single infected individual at the center connected to multiple susceptible individuals who are not connected with each other).
+Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of realized generation times matches the egocentric generation-interval distribution (\fref{local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
 
@@
 \caption{
-\textbf{Comparison of spatial effects on the realized generation intervals.}
-(Left) The intrinsic generation-interval distribution corresponds to the distribution of all infectious contacts that were made, regardless of the outcome (i.e., whether it resulted in infection or not); 5000 stochastic simulations were performed on a tree network with 5 nodes (1 infector and 4 susceptibles).
-(Middle) The egocentric generation-interval distribution corresponds to the distribution of all infectious contacts made by a single primary infector that gave rise to infection; 5000 stochastic simulations were performed on a tree network with 5 nodes (1 infector and 4 susceptibles).
-(Right) The realized generation-interval distribution corresponds to the distribution of all infection times throughout an epidemic; 5000 stochastic simulations were performed on a homogeneous network with 5 nodes (1 infector and 4 susceptibles).
-\swp{Parameters? High per-pair contact? Need to describe and make the figure prettier}
+\textbf{Spatial effects on realized generation intervals.}
+Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
+(Left) The intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible.
+(Middle) The egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
+(Right) General spatial distributions are shorter than egocentric distributions, because contacts can be wasted when susceptibles become infected through other routesor (simulated on a homogeneous network).
+All graphs were generated using 5000 stochastic simulations were on a network with 5 nodes (1 infector and 4 susceptibles).
 }
@@
 
-In general, there are two factors that cause further reduction in the mean generation interval: (1) the local spatial effect (e.g., in a household) and (2) the overall depletion of susceptible individuals in the population;
+In general, there are two factors that cause spatial reduction in the mean generation interval: local spatial effects (e.g., in a household or neighborhood) and (2) population-level effects; 
 the latter effect provides an equivalent explanation for the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}).
-Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the effects of most networks on observed generation intervals is expected to be a hard problem.
+\jd{I'll have to think about this, or you'll have to explain.}
+Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks.
 \cite{tomba2010some} showed that an analytical expression for the mean generation interval, which takes into account the local spatial effect, can be derived for a household with one primary case and two susceptible individuals, but solving for higher order structures seems both infeasible and impractical.

### 2019-01-31 doc/network.tex
Just a few more thoughts

@@
 As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
-Similarly, \Rhat\ will always be greater than the \emph{true} reproductive number \RR\ as it does not account for other effects (e.g., local spatial effects and mean susceptible-depletion effects) that can decrease the mean generation interval.
+Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
 
@@
 When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) mean susceptible-depletion effect.
-Right-censoring effect and egocentric effect are well-understood.
+We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 

### 2019-02-01 doc/network.tex
A few more comments; sorry it's so hard to find time.

@@
 When the susceptible dynamics is known, this equation can be generalized by using \eref{obsg}.
-While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.
+The population-level method is simple and intuitive, but does not take into account who infected whom; we might therefore expect the individual-level method to be more powerful.
 
-The individual-level method assumes that the initial exponential growth phase of an epidemic can be approximated by a Poisson process to derive a likelihood for the observations from each individual:
+The individual-level method \jd{I don't see how we're assuming this:} assumes that the initial exponential growth phase of an epidemic can be approximated by a Poisson process to derive a likelihood for the observations from each individual:
 \begin{equation}
@@
 where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation-interval distribution $g$.
-A similar method has beem proposed by \cite{forsberg2008likelihood} but thier method uses on discretized incidence reports rather than the observed generation intervals.
+\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
 
@@
 However, we observe opposite trends after the growth period (approximately after 20 weeks).
+\jd{I kind of feel we're gettting off track here.
+The main focus should be on early fits, and we should be describing how cool \fref{cmp} is before we wander off into these woods.}
 As the population-level method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
 On the other hand, as the individual-level method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible is translated to underestimation of both mean generation interval and reproductive number.
-we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
+we observe gradual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
 Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproductive number; this is due to initial overestimation of the initial growth rate and we generally expect reproductive to be underestimated.

### 2019-02-05 doc/network.tex
A few more comments

@@
 
-In general, there are two factors that cause spatial reduction in the mean generation interval: local spatial effects (e.g., in a household or neighborhood) and (2) population-level effects; 
+\jd{I think there's room to be clearer here.
+All three depletion effects are similar.
+Should we call them all depletion?: egocentric, local, global?} In general, there are two factors that cause spatial reduction in the mean generation interval: local spatial effects (e.g., in a household or neighborhood) and (2) population-level effects;
 the latter effect provides an equivalent explanation for the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}).
 \jd{I'll have to think about this, or you'll have to explain.}
-Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks.
-\cite{tomba2010some} showed that an analytical expression for the mean generation interval, which takes into account the local spatial effect, can be derived for a household with one primary case and two susceptible individuals, but solving for higher order structures seems both infeasible and impractical.
+Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks, although there has been interesting work on spatial cases, particularly on hierarchical networks of households nested within communities (e.g.,  \cite{tomba2010some}).
+\jd{Don't think we need more than that about Tomba; I wonder if any of Frank Ball's household work could be relevant.}
 \jd{The global spatial effect is sort of the non-spatial effect: the overall response to decreasing susceptibles, like in Champredon.}
 \swp{Tried to fix this}
+\jd{K. I've added at least one new comment in this direction Feb 05 (Tue).}
 
@@
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
+\jd{How about: Both local and global depletion\ldots?}
 Both local spatial and susceptible depletion effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:

### 2019-02-26 doc/network.tex
writing

@@
 Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
-However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
+However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic \swp{Can't figure out what I should cite here; maybe nothing}.
 Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
@@
 In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, these measurements are \emph{censored} (or right-censored) at the observation time.
-These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, they observed censored intervals will be shorter than intrinsic distributions early in the epidemic.
+These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.
 
 Observed generation intervals also vary across space.
-In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population.
-This means that contacts made late in infection are more likely to be ``masked'' due to contacts already being infected.
+In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
+This means that contacts made late in infection are more likely to be ``masked'' due to infecious contacts that were made earlier.
 Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
@@
 
-In this study, we explore the temporal and spatial variation in the observed generation-interval distributions obtained from contact tracing.
+In this study, we explore the spatiotemporal variation in the observed generation-interval distributions obtained from contact tracing.
 We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
 We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
+\swp{Come back and fix this}
 
@@
 I guess because of scaling?}
-\swp{What do you mean by individual properties?}
+\swp{What do you mean by individual properties or even scaling?}
 
 There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
-The basic reproductive number -- average number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
+The basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
 \begin{equation}
@@
 \jd{We should be saying that backward $\to$ censored in the limit at the same time we talk about the association below.}
+\swp{I tried}
 
 Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
-Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time period.
+Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time.
 Instead, contact tracing can be performed over multiple generations, tracing all the way back to the primary case.
-Hence, to obtain a theoretical sampling distribution for the right-censored generation intervals, we want to consider all infections that occur from the beginning of an epidemic to a time period when contact tracing was performed.
+The backward generation-interval distribution is mathematically equivalent to the censored generation-interval distribution only during the exponential growth phase, when depletion of the susceptible population is negligible.
+However, the exponential growth assumption breaks down rapidly, and it is important to be able to conceptually distinguish the censored generation-interval from the backward generation-interval.
 \jd{I hope that none of these papers actually make the mistake that you imply, it's all about the limit.}
+\swp{Sure, in the limit they're all same. But we're often not in the limit and contact tracing can be performed beyond the limit. We can make this point.}
+
+The number of infection occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
+\begin{equation}
+i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
+\end{equation}
+The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
+\begin{equation}
+b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-x) g(x) d\tau}
+\end{equation}
+On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
+Then, the censored generation-interval distribution is given by
+\begin{equation}
+c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(\tau) \int_0^t i(s-\tau) S(s) ds d\tau}
+\end{equation}
+Alternatively, the censored generatio-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
+\begin{equation}
+c_t(\tau) \propto \int_0^t b_s(\tau) S(s) ds.
+\end{equation}
+Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$.
 
@@
 
-The number of infection occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
-\begin{equation}
-i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
-\end{equation}
-Then, total number of secondary infections that are $\tau$ time steps apart and occur before time $t$ can be obtained by integrating the above expression:
-\begin{equation} 
-\RR \int_\tau^t i(s-\tau) g(\tau) S(s) ds.
-\end{equation}
-The distribution of the secondary infections essentially represent the realized generation-interval distribution.
-Normalizing by the total number of cases before time $t$, we obtain a theoretical sampling distribution for the right-censored (censored at time $t$) generation intervals:
-\begin{equation}
-c_t(\tau) = \frac{\RR \int_\tau^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t i(s) ds}.
-\end{equation}
-For convenience, we ignore normalizing constants and write
-\begin{equation}\label{eq:obsg}
-c_t(\tau) \propto g(\tau) \int_{0}^t i(s-\tau) S(s) ds.
-\end{equation}
-
 For a single epidemic, the observed mean generation interval through contact tracing will always be shorter than intrinsic mean generation interval (Figure~\ref{fig:censor}).
@@
 First, any infection events that occur after the contact tracing period canonot be observed due to right censoring and short generation intervals are more likely to be observed.
-In particular, when an epidemic is growing exponentially ($i(t) \propto \exp(rt)$), 
-the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrinsic generation-interval distribution:
+In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
+the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrinsic generation-interval distribution \citep{britton2019estimation}:
 \begin{equation}

### 2019-02-26 doc/network.tex
more

@@
 \jd{Consider rewriting to use early-late as the first-level structure. Early is the part that people focus on most. We could even throw in censorship.}
+\swp{Not sure what you mean... I tried to focus on early-late throughout the manuscript}
 Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
@@
 An infected individual may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
-Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average, in a limited network.
+Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average, in a limited contact network.
 To explore the effect of multiple contacts on realized generation intervals, we first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
@@
 This widely used model assumes that latent and infectious periods are exponentially distributed.
-The intrinsic generation-interval distribution that corresponds to this model can be written has [CITE]:
+The intrinsic generation-interval distribution that corresponds to this model can be written has \citep{svensson2015influence}:
 \begin{equation}
@@
 
-\jd{I think there's room to be clearer here.
-All three depletion effects are similar.
-Should we call them all depletion?: egocentric, local, global?} In general, there are two factors that cause spatial reduction in the mean generation interval: local spatial effects (e.g., in a household or neighborhood) and (2) population-level effects;
-the latter effect provides an equivalent explanation for the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}).
-\jd{I'll have to think about this, or you'll have to explain.}
+\jd{I think there's room to be clearer here. All three depletion effects are similar. Should we call them all depletion?: egocentric, local, global?} 
+\swp{I like it}
+In general, spatial reduction in the mean generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
+Egocentric depletion, as discussed previously, is caused by an infected individual making multiple contacts to the same individual.
+Local depletion refers to a depletion of susceptible individuals in a household or neighborhood;
+this effect can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
+Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}).
+\jd{I'll have to think about this, or you'll have to explain.} \swp{we changed our local and global explanations so this should be consistent what we said above.}
 Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks, although there has been interesting work on spatial cases, particularly on hierarchical networks of households nested within communities (e.g.,  \cite{tomba2010some}).
@@
 \jd{K. I've added at least one new comment in this direction Feb 05 (Tue).}
+\swp{Tried to fix this}
 
@@
 
+\swp{I might want to move this paragraph to after equation 13. It doesn't feel like it fits here.}
 The intrinsic generation-interval distribution provides a link between the speed (exponential growth rate, $r$) and the strength (reproductive number, $\RR$) of an epidemic via Euler-Lotka equation \citep{lotka1907relation} but assumes a homogeneously mixing population.
@@
 \end{equation}
-This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure. \swp{Need to confirm; Maybe prove it in appendix or something...how?}
+This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure.
+\swp{Need to confirm; Maybe prove it in appendix or something...
+will get back to this later}
 As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.

### 2019-02-26 doc/network.tex
upload

@@
 \documentclass[12pt]{article}
+\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
+
 \title{Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}}
@@
 Then, the censored generation-interval distribution is given by
-\begin{equation}
+\begin{equation}\label{eq:obsg}
 c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(\tau) \int_0^t i(s-\tau) S(s) ds d\tau}
@@
 Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
-(Left) The intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible.
-(Middle) The egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
-(Right) General spatial distributions are shorter than egocentric distributions, because contacts can be wasted when susceptibles become infected through other routesor (simulated on a homogeneous network).
+(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible (simulated on a star network).
+(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
+(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
 All graphs were generated using 5000 stochastic simulations were on a network with 5 nodes (1 infector and 4 susceptibles).
@@
 
-When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) mean susceptible-depletion effect.
+When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
-\jd{How about: Both local and global depletion\ldots?}
-Both local spatial and susceptible depletion effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
-local spatial effect will be important early in an epidemic whereas the susceptible depletion will be small until sufficient amount of susceptible individuals become infected.
-As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local spatial and egocentric effects.
+\jd{How about: Both local and global depletion\ldots?} \swp{Done.}
+Both local and global depletion effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
+local depletion will be important early in an epidemic whereas global depletion will be small until sufficient amount of susceptible individuals become infected.
+As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local structure of the contact network near the source of an outbreak.
 
-Since the right-censoring effect is a sampling artifact, it must be corrected for.
-Local and egocentric effects are expected to be embedded within realized generation intervals and do not need to be corrected for as they reflect the unobserved characteristic of a realized epidemic.
+Since the right-censoring effect is a sampling bias, it must be corrected for.
+Any spatial effects are expected to be embedded within realized generation intervals and do not need to be corrected for as they reflect the unobserved characteristic of a realized epidemic.
 We claim that accounting for the right-censoring effect is sufficient and any necessary spatial component will be accounted for implicitly.
@@
 In this section, we investigate two statistical methods for correcting for temporal bias in a contact tracing data.
-Detailed derivations of the two methods can be found in section ???.
+Detailed derivations of the two methods can be found in Methods.
+\swp{TODO: Write methods}
 
-We refer to the first method as the population-level method as it relies on the observed distribution aggregated across population.
+\begin{figure}[ht]
+\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
+\caption{\textbf{Comparison of estimates of reproductive number based on various methods}
+Using the observed generation-interval distributions without correcting for right-censoring severely underestimates the reproductive number.
+Similarly, using intrinsic and egocentric generation-interval distributions without accounting for local spatial effects overestimates the reproductive number.
+Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
+Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters (see Methods).
+}
+\label{fig:cmp}
+\end{figure}
+
+We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
 As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
@@
 \end{equation}
-When the susceptible dynamics is known, this equation can be generalized by using \eref{obsg}.
-The population-level method is simple and intuitive, but does not take into account who infected whom; we might therefore expect the individual-level method to be more powerful.
+When the susceptible dynamics is known, this equation can be generalized by using \eref{obsg}; we do not pursue this direction in this study.
+The population-level method is simple and intuitive, but does not take into account who infected whom.
+%% we might therefore expect the individual-level method to be more powerful.
+%% SWP: We haven't introduced the individual-level method yet...
 
-The individual-level method \jd{I don't see how we're assuming this:} assumes that the initial exponential growth phase of an epidemic can be approximated by a Poisson process to derive a likelihood for the observations from each individual:
+The individual-level method assumes that the initial exponential growth phase of an epidemic can be approximated by a Poisson process to derive a likelihood for the observations from each individual:
 \begin{equation}
-\mathcal{L}(\RR, \theta) = \prod_{j=1}^N \left(\RR^{n_j} \exp \left(- \RR \int_0^{\tsub{t}{censor} - t_j} g(s;\theta) ds \right) \prod_{i=1}^{n_j} g(\tau_{i, j}; \theta) \right),
+\begin{aligned}
+&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
+&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
+\end{aligned}
 \end{equation}
-where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation-interval distribution $g$.
+where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by infector $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
 \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
+
+To test these methods, we simulate 100 epidemics with Ebola-liked parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
+\cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
+however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
+Therefore, we expect our empirical estimate to better reflect the infection process during the exponential growth phase.
+\swp{TODO: Compare Trapman vs our estimate of empirical R0 in Appendix; Trapman estimate is much lower.}
+
+As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
+estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
+Direct estimates based on contact tracing data severely underestimates the empirical estimates.
+While both population- and individual-level corrections provide similar estimates to the empirical reproductive number,
+population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
+For smaller values of \RR, we expect the differences to become smaller. 
+\swp{TODO: In Appendix, we present the same figure using gamma-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}
+
+\section{Discussion}
+
+\pagebreak
+
+\section{Junk}
 
@@
 These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].
-
-\begin{figure}
-\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
-\caption{\textbf{Comparison of estimates of reproductive number based on various methods}
-Using the observed generation-interval distributions without correcting for right-censoring severely underestimates the reproductive number.
-Similarly, using intrinsic and egocentric generation-interval distributions without accounting for local spatial effects overestimates the reproductive number.
-Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
-100 stochastic simulations were performed on an empirical network [CITE] using Ebola-like parameters (latent period = 11.4 days and infectious period = 5 days).
-}
-\label{fig:cmp}
-\end{figure}
-
-%% TODO: put SEIR in main text and show SE2IR in appendix for consistency
-Finally, we simulate 100 epidemics with Ebola-liked parameters on an empirical network [CITE] and compare the estimates of reproductive number based on different methods with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
-As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
-estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
-Direct estimates based on contact tracing data severely underestimates the empirical estimates.
-Both population and individual corrections provide similar estimates to empirical reproductive numbers.
-For smaller values of \RR, we expect the differences to become smaller.
 

### 2019-02-26 analysis/cmp_compare.R

@@ -7,7 +7,19 @@ source("../R/generation.R")
 source("../R/estimate.R")
 source("../R/empirical.R")
 
-load("../sim/cmp_seminr_sim.rda")
+load("../sim/cmp_seir_sim.rda")
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+conditional_fun <- function(tau) {
+	sigma*(gamma+lambda)/(gamma+lambda-sigma) * (exp(-sigma*tau)-exp(-(gamma+lambda)*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.shape <- true.mean^2/true.var
 
 cmpGraph <- read.table("../data/CA-CondMat.txt")
 cmpGraph <- graph.data.frame(cmpGraph)
@@ -21,14 +33,14 @@ kappa <- var(d)/mean(d) + mean(d) -1
 beta <- 0.4/5
 
 ## This is SEIR
-## R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
+R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
 
 ## This is SE2IR
-R0fun.network <- function(r) {
-	lambda <- (2 * sigma+r)^2 * (r + gamma)/((kappa-1) * 4 * sigma^2 - 4 * sigma * r - r^2)
-	
-	kappa * lambda/(lambda + gamma)
-}
+## R0fun.network <- function(r) {
+##	lambda <- (2 * sigma+r)^2 * (r + gamma)/((kappa-1) * 4 * sigma^2 - 4 * sigma * r - r^2)
+##	
+##	kappa * lambda/(lambda + gamma)
+## }
 
 censor.gi <- lapply(
 	reslist
@@ -64,11 +76,13 @@ individual.est <- lapply(reslist, function(x){
 	
 	est$type <- c("RR", "mean", "shape")
 	
+	print(est)
+	
 	est
 }) %>%
 	bind_rows(.id="sim")
 
-individual.RR <- lapply(1:50, function(x){
+individual.RR <- lapply(1:100, function(x){
 	gg <- growth[x,2]
 	
 	mean <- filter(individual.est, type=="mean")[x,2]
[…]

### 2019-02-26 fig/cmp.R

@@ -36,4 +36,4 @@ ggR <- ggplot(RRdata2) +
 		axis.title.x=element_blank()
 	)
 
-ggsave("cmp_reproductive.pdf", ggR, width=6, height=4)
+ggsave("cmp_reproductive.pdf", ggR, width=6, height=3)

### 2019-02-26 fig/local_effect.R

@@ -5,6 +5,9 @@ source("../sim/ebola_param.R")
 source("../R/seir.R")
 source("../R/generation.R")
 
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
 beta <- 1
 N <- 5
 lambda <- beta/(N-1)
@@ -43,12 +46,12 @@ gendata_int <- reslist_local %>%
 	unlist
 	
 gendata_mean <- data.frame(
-	type=c("intrinsic", "conditional", "homogeneous"),
+	type=c("Intrinsic", "Egocentric", "Homogeneous"),
 	value=c(mean(gendata_int), mean(gendata_cond), mean(gendata_full))
 )
 
 true_mean <- data.frame(
-	key=c("intrinsic", "conditional"),
+	key=c("Intrinsic", "Egocentric"),
 	value=c(1/sigma+1/gamma, 1/sigma + 1/(gamma+lambda))
 )
 	
@@ -64,18 +67,20 @@ gendata_total <- list(
 	) 
 ) %>%
 	bind_rows(.id="type") %>%
-	mutate(type=factor(type, levels=c("intrinsic", "conditional", "homogeneous")))
+	mutate(type=factor(type, 
+					   levels=c("intrinsic", "conditional", "homogeneous"),
+					   labels=c("Intrinsic", "Egocentric", "Homogeneous") ))
 
 tvec <- seq(0, 80, by=0.1)
 
 dendata <- list(
 	data.frame(
-		key="intrinsic",
+		key="Intrinsic",
 		x=tvec,
 		density=intrinsic_fun(tvec)
 	),
 	data.frame(
-		key="conditional",
+		key="Egocentric",
 		x=tvec,
 		density=conditional_fun(tvec)
 	) 
@@ -84,11 +89,11 @@ dendata <- list(
 
 gglocal <- ggplot(gendata_total) +
 	geom_histogram(aes(value, y=..density..), bins=30, col="black", fill=NA) +
-	geom_line(data=dendata, aes(x, density, col=key), lwd=1.2) +
-	geom_vline(data=true_mean, aes(xintercept=value, col=key), lwd=1.2) +
-	geom_vline(data=gendata_mean, aes(xintercept=value), lty=2, lwd=1.2) +
[…]

### 2019-02-26 fig/smallpox.R

@@ -73,3 +73,8 @@ plot(dd_serial$time, dd_serial$mean, type="l",
 title("Symptomatic cases")
 abline(h=17.7, lty=2)
 dev.off()
+
+
+## ppl who got smallpox are going to be examined more carefully
+## read Wallinga
+## 

### 2019-03-08 doc/network.tex
updates

@@
 
+\section{Example: Middle East Respiratory Syndrome Coronavirus (MERS-CoV)}
+
+\swp{This entire section is missing citations}
+
+We revisit Middle East Respiratory Syndrome Coronavirus (MERS-Cov) outbreak from South Korea in 2015 to explore ideas presented in this study \citep{cowling2015preliminary}. 
+The epidemic is characterized by intra- and inter-hospital transmission events and a few super spreading events. 
+We expect this kind of scenario to provide a realistic example of an epidemic on a network where we can observe spatial effects on multiple scales (e.g., stronger local depletion effect in intra-hospital transmission than in inter-hosptial transmission).
+The data was obtained from the R package \texttt{outbreaks} \citep{outbreakspackage}.
+\swp{"This dataset is meant for teaching purposes; it represents neither the final outbreak investigation results nor a consolidated and complete description of the transmission chain." This is what the description says.
+Is this OK?}
+
 \section{Discussion}
+
+Estimating reproductive number based on generation-interval distributions
 

### 2019-03-08 fig/mers.R

@@ -1,0 +1,42 @@
+library(dplyr)
+library(outbreaks)
+library(ggplot2); theme_set(theme_bw())
+
+ll <- mers_korea_2015$linelist
+cc <- mers_korea_2015$contacts
+
+toDate <- as.numeric(as.Date(ll$dt_onset[match(cc$to, ll$id)]))
+toDate <- toDate - 16569
+fromDate <- as.numeric(as.Date(ll$dt_onset[match(cc$from, ll$id)]))
+fromDate <- fromDate - 16569
+
+serial_data <- data.frame(
+	time=toDate,
+	gen=toDate - fromDate,
+	within=ll$loc_hosp[match(cc$to, ll$id)] == ll$loc_hosp[match(cc$from, ll$id)],
+	hospital=ll$loc_hosp[match(cc$to, ll$id)]
+)
+
+boxplot(gen~within, data=serial_data)
+t.test(gen~within, data=serial_data)
+
+ggplot(serial_data) +
+	geom_boxplot(aes(within, gen))
+
+dd <- data.frame(
+	time=ll$dt_onset[match(cc$to, ll$id)],
+	gen=toDate - fromDate
+) %>%
+	arrange(time) %>%
+	group_by(time) %>%
+	summarize(
+		sum=sum(gen),
+		weight=length(gen)
+	) %>%
+	mutate(mean=cumsum(sum)/cumsum(weight))
+
+plot(dd$time, dd$mean)
+
+plot(table(ll$dt_onset))
+
+

### 2019-03-11 doc/network.tex
remove a section
a little more

--- 2019-02-26 doc/network.tex
@@
 Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
-However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic \swp{Can't figure out what I should cite here; maybe nothing}.
+However, directly measuring the reproductive number often requires detailed knowledge of the natural history of a disease and may not be feasible early in an epidemic \swp{Can't figure out what I should cite here; maybe nothing}.
 Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
@@
 For smaller values of \RR, we expect the differences to become smaller. 
-\swp{TODO: In Appendix, we present the same figure using gamma-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}
+\swp{TODO: In Appendix, we present the same figure using Erlang-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}
 
 \section{Discussion}
+
+The intrinsic generation-interval distribution provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\mathcal R$) of an epidemic \citep{wallinga2007generation, park2019practical}.
+This link has been widely adopted in outbreak analyses as it allows us to conveniently estimate the reproductive number from quantities that are easier to measure.
+However, estimating the \emph{correct} generation-interval distribution is challenging: the observed generation-interval distribution varies depending on how and when it is measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+Here, we decompose variation in the observed generation intervals into spatial and temporal components under a realistic setting where contact tracing data are aggregated over time.
+We use simple models to distinguish among different sources of variations in the observed generation intervals.
+We provide two methods for correcting temporal bias and test their accuracy under a realistic setting using simulated epidemics.
+
+Generation intervals can be measured either forward, by tracing infectees, or backward, by tracing infectors.
+
+\section{Methods}
 

### 2019-03-11 analysis/cmp_compare.R

@@ -30,7 +30,7 @@ d <- degree(cmpGraph)
 
 kappa <- var(d)/mean(d) + mean(d) -1
 
-beta <- 0.4/5
+beta <- 0.13/5
 
 ## This is SEIR
 R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
@@ -125,9 +125,9 @@ intrinsic.est <- lapply(growth$r, function(x) 1/mean(exp(-x*gen)))
 
 network.est <- lapply(growth$r, R0fun.network)
 
-trapman.est <- lapply(reslist, trapman.R0)
+trapman.est <- lapply(reslist, trapman.R0, 100)
 
-empirical.est <- lapply(reslist, empirical.R0)
+empirical.est <- lapply(reslist, empirical.R0, 100)
 
 RRdata <- data.frame(
 	observed=observed.est$estimate[observed.est$type=="RR"],

### 2019-03-11 fig/cmp.R

@@ -5,10 +5,21 @@ library(gridExtra)
 
 source("../sim/ebola_param_seminr.R")
 
-load("../analysis/cmp_compare.rda")
-
 scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+genlist <- list()
+RRlist <- list()
+
+load("../analysis/cmp_compare.rda")
+
+genlist$big <- gendata
+RRlist$big <- RRdata
+
+load("../analysis/cmp_compare_2.rda")
+
+genlist$small <- gendata
+RRlist$small <- RRdata
 
 RRgroup <- data.frame(
 	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "local\ncorrection", "intrinsic"),
@@ -16,8 +27,20 @@ RRgroup <- data.frame(
 				 levels=c("tracing based", "empirical", "individual based"))
 )
 
-RRdata2 <- RRdata %>%
-	gather(key, value) %>%
+RRdata2 <- RRlist %>%
+	bind_rows(.id="type") %>%
+	gather(key, value, -type) %>%
+	mutate(key=factor(key, 
+					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
+					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
+					  		 "tree\nbased", "empirical", "local\ncorrection", "intrinsic"))) %>%
+	merge(RRgroup) %>%
+	filter(key != "tree\nbased") %>%
+	mutate(type=factor(type, levels=c("small", "big"), labels=c("Smaller epidemic", "Larger epidemic")))
+
+gendata2 <- genlist %>%
+	bind_rows(.id="type") %>%
+	gather(key, value, -type) %>%
 	mutate(key=factor(key, 
 					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
 					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
@@ -25,12 +48,12 @@ RRdata2 <- RRdata %>%
 	merge(RRgroup) %>%
 	filter(key != "tree\nbased")
 
-ggR <- ggplot(RRdata2) +
+ggR <- ggplot(RRdata2 %>% filter(type=="Larger epidemic")) +
[…]

### 2019-03-14 doc/network.tex
trying to make progress

@@
 
-The intrinsic generation-interval distribution provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\mathcal R$) of an epidemic \citep{wallinga2007generation, park2019practical}.
-This link has been widely adopted in outbreak analyses as it allows us to conveniently estimate the reproductive number from quantities that are easier to measure.
-However, estimating the \emph{correct} generation-interval distribution is challenging: the observed generation-interval distribution varies depending on how and when it is measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\mathcal R$) of an epidemic \citep{wallinga2007generation, park2019practical}.
+This link has been widely adopted in outbreak analyses as it allows us to conveniently estimate the reproductive number from measurable quantities.
+However, determining the \emph{correct} generation-interval distribution is challenging: observed generation-interval distributions can vary depending on how and when it is measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
 Here, we decompose variation in the observed generation intervals into spatial and temporal components under a realistic setting where contact tracing data are aggregated over time.
 We use simple models to distinguish among different sources of variations in the observed generation intervals.
-We provide two methods for correcting temporal bias and test their accuracy under a realistic setting using simulated epidemics.
+We provide two methods for correcting temporal bias and test their accuracy using simulations.
 
-Generation intervals can be measured either forward, by tracing infectees, or backward, by tracing infectors.
+Generation intervals can be measured either forward (by tracing the infectees of an infected individual) or backward (by tracing the infector of an infected individual).
+Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart.
+Instead, we consider a slightly different scenario in which generation intervals are measured over multiple generations.
+This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
+In this case, the observed generation intervals are derived all infections that occured before the last date on which contact tracing was performed and are subject to right-censoring -- it is not possible to trace infections that did not happen yet.
+
+When the epidemic is growing exponentially, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution.
 
 \section{Methods}
+
+\subsection{Deterministic SEIR model}
+
+The Susceptible-Exposed-Infected-Recovered (SEIR)
+
+\subsection{Stochastic SEIR model on a network}
+
+\subsection{Derivation of the individual-based method}
 

### 2019-03-19 doc/network.tex
Worked through time; space next

@@
 
-\jd{Consider rewriting to use early-late as the first-level structure. Early is the part that people focus on most. We could even throw in censorship.}
-\swp{Not sure what you mean... I tried to focus on early-late throughout the manuscript}
 Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
-As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
+As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
 Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
@@
 
-In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, these measurements are \emph{censored} (or right-censored) at the observation time.
+In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, measurements are ``censored'': we don't know what happens after the observation time.
 These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.
@@
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
-This means that contacts made late in infection are more likely to be ``masked'' due to infecious contacts that were made earlier.
+This means that contacts made late in infection are more likely to be ``masked'' due to infectious contacts that were made earlier.
 Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
@@
 
-In this study, we explore the spatiotemporal variation in the observed generation-interval distributions obtained from contact tracing.
+In this study, we explore spatiotemporal variation in observed generation-interval distributions.
 We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
@@
 
-Generation-interval distributions are often viewed from a population-level pespective but we can distinguish population-level distributions from individual-level distributions (\fref{indpop}); 
-making this distinction clear will be particularly useful when we discuss spatial components.
-An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.
+Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence};
+making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
+An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
 The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
@@
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
-\jd{Do we need to worry about assumptions for the rest of this \P?
-Once we have an aspect space we need to assume that individual properties are independent of risk of infection.
-Also, since that's wrong on our networks, do we have to figure out why we're approximately saved?
-I guess because of scaling?}
-\swp{What do you mean by individual properties or even scaling?}
+\jd{This is assuming that we know how to average across infected individuals.
+Do we need to assume that individual kernel properties are independent of risk of infection?
+If we have high-degree individuals, does it already break that assumption?}
 
@@
 \end{equation}
+\jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
 The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
@@
 We assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
+\jd{Do we need to be clear when we do and don't assume this?}
 \begin{equation}
@@
 
-\jd{Mathematically, early censored and early backward intervals are the same, but conceptually they're just different. In particular, censoring occurs even with forward tracing.}
-\jd{During exponential growth, the theoretical distributions are the same; maybe we should do a tiny bit of that math here so that we can talk more clearly about it.}
-\jd{We need to think about exposition. Maybe simple formulae for all of our interval types, both in general and under the assumption of exponential growth.}
-\jd{We should be saying that backward $\to$ censored in the limit at the same time we talk about the association below.}
-\swp{I tried}
+Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
+Several studies have associated the backward generation-interval distribution with the early-epidemic censored generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}.
+The censored generation-interval distribution in fact averages over backward generation-interval distributions up until the observation time.
+While we mathematically expect these underlying distributions to be the same (and thus the same distribution as their average) during the exponential growth phase, these assumptions can break down (due to susceptible depletion or other temporal change).
+This, it is useful to conceptually distinguish the censored distribution the backward distribution.
 
-Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
-Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time.
-Instead, contact tracing can be performed over multiple generations, tracing all the way back to the primary case.
-The backward generation-interval distribution is mathematically equivalent to the censored generation-interval distribution only during the exponential growth phase, when depletion of the susceptible population is negligible.
-However, the exponential growth assumption breaks down rapidly, and it is important to be able to conceptually distinguish the censored generation-interval from the backward generation-interval.
-\jd{I hope that none of these papers actually make the mistake that you imply, it's all about the limit.}
-\swp{Sure, in the limit they're all same. But we're often not in the limit and contact tracing can be performed beyond the limit. We can make this point.}
-
-The number of infection occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
+The density of infections occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
 \begin{equation}
@@
 The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
+\jd{Is it nicer to keep $s$ in the numerator? Otherwise, why do we bother with $s$ at all?}
 \begin{equation}
@@
 \end{equation}
+
+\jd{Can we just skip this \P, or put in an appendix or something?}
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
@@
 \end{equation}
-Alternatively, the censored generatio-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
+Thus, the censored generation-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
 \begin{equation}

### 2019-03-19 doc/network.tex
meeting

@@
 
-We assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
-\jd{Do we need to be clear when we do and don't assume this?}
+\jd{Rewrite these equations a bit more consistently. What are we using for: index infection, current infection? $\tau$ is difference between them. Is it simpler just to use ' for dummies?}
+In the homogeneous model, we assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
 \begin{equation}

### 2019-03-20 doc/network.tex
Trying to simplify

@@
 We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case $a$ in the absence of other infectors:
+\jd{We can't use dilution directly in the hazard term: it needs to be multiplied by something like big R.
+I would put $R\delta$ there.
+The extreme dilution examples would be mean degree (if highly connected people infect that much more) or individual degree (if they don't infect more at all.)}
 \begin{equation}
@@
 \end{equation}
-In this scenario, per-pair contact rate effectively decreases mean infectious period: an infected individual can no longer infect anyone (hence no longer infectious) when all its susceptible acquaintances have ben infected.
+In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate: infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again).
 
 This calculation can be validated by simulating stochastic infection processes on a ``star'' network (i.e, a single infected individual at the center connected to multiple susceptible individuals who are not connected with each other).
-Simulations confirm that in this case the distribution of contact times matches the intrinsic generation-interval distribution, while the distribution of realized generation times matches the egocentric generation-interval distribution (\fref{local}) -- generation times are shorter on average because contacts with already-infected susceptibles do not lead to new infections.
+Simulations confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution, while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (\fref{local}).
 

### 2019-03-20 doc/network.tex
Through space. Handing back to Daniel

@@
 We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
-We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
+We discuss statistical approaches to estimating effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
 \swp{Come back and fix this}
@@
 
+\jd{This is confusing to me.
+Wouldn't the natural next step be a random network?
+Homogeneous won't always be shorter than egocentric, I guess.
+Or are you simulating for a long time?
+Would it be better to simulate for a shorter time, and add another layer?
+I was thinking homogeneous $\to$ random $\to$ structured?
+Doing homogeneous during the exponential period should give us the same as egocentric, right?
+Maybe we need two figures?
+Or maybe we don't need a star network at all: first panel of current figure can be done on any network, and second panel can be done on a homogeneous network with a shortish time window?}
 \begin{figure}
@@
 
-\jd{I think there's room to be clearer here. All three depletion effects are similar. Should we call them all depletion?: egocentric, local, global?} 
-\swp{I like it}
 In general, spatial reduction in the mean generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
@@
 this effect can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
-Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the underestimation of the mean generation interval, sampled across an entire epidemic (\fref{censor}).
-\jd{I'll have to think about this, or you'll have to explain.} \swp{we changed our local and global explanations so this should be consistent what we said above.}
-Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks, although there has been interesting work on spatial cases, particularly on hierarchical networks of households nested within communities (e.g.,  \cite{tomba2010some}).
-\jd{Don't think we need more than that about Tomba; I wonder if any of Frank Ball's household work could be relevant.}
-\jd{The global spatial effect is sort of the non-spatial effect: the overall response to decreasing susceptibles, like in Champredon.}
-\swp{Tried to fix this}
-\jd{K. I've added at least one new comment in this direction Feb 05 (Tue).}
-\swp{Tried to fix this}
+Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
+
+Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks -- although there has been interesting work on spatial cases, particularly on hierarchical networks of households nested within communities (e.g.,  \cite{tomba2010some}).
+\jd{I wonder if any of Frank Ball's household work could be relevant.}
 
@@
 
-\section{Inferring generation-interval distributions from a contact tracing data}
+\section{Effective distributions}
 
@@
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
-While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
-\jd{How about: Both local and global depletion\ldots?} \swp{Done.}
+While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and estimated reproductive numbers. 
 Both local and global depletion effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
@@
 Since the right-censoring effect is a sampling bias, it must be corrected for.
-Any spatial effects are expected to be embedded within realized generation intervals and do not need to be corrected for as they reflect the unobserved characteristic of a realized epidemic.
-We claim that accounting for the right-censoring effect is sufficient and any necessary spatial component will be accounted for implicitly.
-Hereafter, we will refer to temporally corrected distribution as the effective generation-interval distribution to contrast from the intrinsic generation-interval distribution; only in a large homogeneously mixing population, the corrected generation-interval distribution will be equivalent to the intrinsic generation-interval distribution.
+In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
+We therefore expect that if we correct observed generation intervals for the right-censoring effect, spatial components will be accounted for implicitly, and we can use these temporally corrected intervals to correctly link $r$ and $\RR$.
+and we can obtain principled esi.
+Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution.
+The effective distribution incorporates spatial effects.
+In a large homogeneously mixing population, and during the early phase an an epidemic, the effective distribution will be equivalent to the intrinsic distribution.
+\jd{In a network with limited degree, they differ due to egocentric effects; in a non-random network they differ due to local effects; beyond the early period, they differ due to global effects.}
+
+\section{Inferring generation-interval distributions from a contact tracing data}
+
 In this section, we investigate two statistical methods for correcting for temporal bias in a contact tracing data.
@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods}
-Using the observed generation-interval distributions without correcting for right-censoring severely underestimates the reproductive number.
+Using the observed generation-interval distributions without correcting for right-censoring severely \jd{Use a log scale on the y axis to make this clearer} underestimates the reproductive number.
 Similarly, using intrinsic and egocentric generation-interval distributions without accounting for local spatial effects overestimates the reproductive number.
@@
 
-To test these methods, we simulate 100 epidemics with Ebola-liked parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
+To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
 \cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
@@
 
-The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\mathcal R$) of an epidemic \citep{wallinga2007generation, park2019practical}.
+The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\RR$) of an epidemic \citep{wallinga2007generation, park2019practical}.
 This link has been widely adopted in outbreak analyses as it allows us to conveniently estimate the reproductive number from measurable quantities.

### 2019-03-26 doc/network.tex
should have commited earlier

--- 2019-03-19 doc/network.tex
@@
 
+\jd{Consider rewriting to use early-late as the first-level structure. Early is the part that people focus on most. We could even throw in censorship.}
+\swp{Not sure what you mean... I tried to focus on early-late throughout the manuscript}
 Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
-As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
+As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
 Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
@@
 
-In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, measurements are ``censored'': we don't know what happens after the observation time.
+In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, these measurements are \emph{censored} (or right-censored) at the observation time.
 These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.
@@
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
-This means that contacts made late in infection are more likely to be ``masked'' due to infectious contacts that were made earlier.
+This means that contacts made late in infection are more likely to be ``masked'' due to infecious contacts that were made earlier.
 Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
@@
 
-In this study, we explore spatiotemporal variation in observed generation-interval distributions.
+In this study, we explore the spatiotemporal variation in the observed generation-interval distributions obtained from contact tracing.
 We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
@@
 
-Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence};
-making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
-An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
+Generation-interval distributions are often viewed from a population-level pespective but we can distinguish population-level distributions from individual-level distributions (\fref{indpop}); 
+making this distinction clear will be particularly useful when we discuss spatial components.
+An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.
 The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
@@
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
-\jd{This is assuming that we know how to average across infected individuals.
-Do we need to assume that individual kernel properties are independent of risk of infection?
-If we have high-degree individuals, does it already break that assumption?}
+\jd{Do we need to worry about assumptions for the rest of this \P?
+Once we have an aspect space we need to assume that individual properties are independent of risk of infection.
+Also, since that's wrong on our networks, do we have to figure out why we're approximately saved?
+I guess because of scaling?}
+\swp{What do you mean by individual properties or even scaling?}
 
@@
 \end{equation}
-\jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
 The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
@@
 
-\jd{Rewrite these equations a bit more consistently. What are we using for: index infection, current infection? $\tau$ is difference between them. Is it simpler just to use ' for dummies?}
-In the homogeneous model, we assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
+We assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
 \begin{equation}
@@
 
+\jd{Mathematically, early censored and early backward intervals are the same, but conceptually they're just different. In particular, censoring occurs even with forward tracing.}
+\jd{During exponential growth, the theoretical distributions are the same; maybe we should do a tiny bit of that math here so that we can talk more clearly about it.}
+\jd{We need to think about exposition. Maybe simple formulae for all of our interval types, both in general and under the assumption of exponential growth.}
+\jd{We should be saying that backward $\to$ censored in the limit at the same time we talk about the association below.}
+\swp{I tried}
+
 Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
-Several studies have associated the backward generation-interval distribution with the early-epidemic censored generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}.
-The censored generation-interval distribution in fact averages over backward generation-interval distributions up until the observation time.
-While we mathematically expect these underlying distributions to be the same (and thus the same distribution as their average) during the exponential growth phase, these assumptions can break down (due to susceptible depletion or other temporal change).
-This, it is useful to conceptually distinguish the censored distribution the backward distribution.
+Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time.
+Instead, contact tracing can be performed over multiple generations, tracing all the way back to the primary case.
+The backward generation-interval distribution is mathematically equivalent to the censored generation-interval distribution only during the exponential growth phase, when depletion of the susceptible population is negligible.
+However, the exponential growth assumption breaks down rapidly, and it is important to be able to conceptually distinguish the censored generation-interval from the backward generation-interval.
+\jd{I hope that none of these papers actually make the mistake that you imply, it's all about the limit.}
+\swp{Sure, in the limit they're all same. But we're often not in the limit and contact tracing can be performed beyond the limit. We can make this point.}
 
-The density of infections occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
+The number of infection occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
 \begin{equation}
@@
 The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
-\jd{Is it nicer to keep $s$ in the numerator? Otherwise, why do we bother with $s$ at all?}
 \begin{equation}
@@
 \end{equation}
-
-\jd{Can we just skip this \P, or put in an appendix or something?}
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
@@
 \end{equation}
-Thus, the censored generation-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
+Alternatively, the censored generatio-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
 \begin{equation}
@@
 This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
-In this case, the observed generation intervals are derived all infections that occured before the last date on which contact tracing was performed and are subject to right-censoring -- it is not possible to trace infections that did not happen yet.
+In this case, the observed generation intervals are derived all infections that occured before the last date on which contact tracing was performed and are subject to right-censoring -- it is not possible to trace individuals that have not been incted yet.
 
-When the epidemic is growing exponentially, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution.
+When the epidemic is growing exponentially, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-censoring.
+Near the end of an epidemic, the effect of right-censoring becomes negligible but the observed generation intervals are still shorter on average than the intrinsic generation intervals.
+
+We further classify susceptible depletion into three levels: egocentric, local, and global.
+Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
+Accounting for the egocentric effect allows us to reinterpret the results by Trapman et al using arguments based on generation intervals.
+Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
+Global susceptible depletion refers to the overall decrease in the susceptible pool in the population.
+
+Susecptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.
+
+Limitations
+\begin{itemize}
+	\item Exponential growth; polynomial
+	\item Extending the framework (accounting for control)
+	\item Low coverage of gamma
+	\item Serial interval
+\end{itemize}
 
@@
 
-\subsection{Derivation of the individual-based method}
-
-\pagebreak
-
-\section{Junk}
-
-First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (\fref{example}).
-During the initial exponential growth period (approximately between 9 and 20 weeks), both population-level and individual-level methods yield similar estimates of mean generation interval and reproductive number.
-However, we observe opposite trends after the growth period (approximately after 20 weeks).
-\jd{I kind of feel we're gettting off track here.
-The main focus should be on early fits, and we should be describing how cool \fref{cmp} is before we wander off into these woods.}
-As the population-level method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
-On the other hand, as the individual-level method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible is translated to underestimation of both mean generation interval and reproductive number.
-we observe gradual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).
-Around the 14th week, reproductive number estimate based on the observed generation interval approximately matches the true reproductive number; this is due to initial overestimation of the initial growth rate and we generally expect reproductive to be underestimated.
-\swp{Maybe another figure in the supp}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{../fig/example.pdf}
-\caption{\swp{Need to re-do this figure; initial overestimation of little $r$ is weird}}
-\label{fig:example}
-\end{figure}
-
-We compare accuracy and reliability of the two methods during the initial growth period using 100 stochastic simulations on a homogeneous network (Figure~\ref{fig:test}).
-We find that the individual-level method provides the most consistent (least variable) estimates.
-The population method overestimates reproductive number initially but this is likely to be driven by the initial overestimation of the growth rate as discussed above (Figure~\ref{fig:example}).
-
-\begin{figure}
-\includegraphics[width=\textwidth]{../fig/compare_methods.pdf}
-\caption{\swp{Need to re-do this figure; initial overestimation of little $r$ is weird}.}
-\label{fig:test}
-\end{figure}
-
-We also compare coverage probabilities of these methods. 
-Coverage probabilities are defined as the proportion of confidence intervals that contain the true value in a repeated simulations; for example, a 95\% confidence interval should attain 95\% coverage probability by its definition.
-Surprisingly, both methods only attain approximately 70\% coverage.
-Even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution, making a wrong distributional assumption leads to narrow confidence intervals.
-To confirm that the undercoverage is caused by distribuional assumption, we simulated an epidemic using gamma intrinsic generation-interval and found that fitting the true shape yields nominal coverage (see Appendix).
-These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].
-
-\clearpage
-
-\section{Discussion}
-
-\section{Methods}
-
-The first method is a non-parametric method.
-Recall that the observed generation-interval distribution is a weighted intrinsic generation-interval distribution (equation~\ref{eq:obsg}). 
-Then, the intrinsic generation-interval distribution can be obtained by taking the inverse weight:
-\begin{equation}
-g(\tau) \propto c_t(\tau) \frac{1}{\int_{0}^t i(s-\tau) S(s) ds}
-\end{equation}
-However, this method requires a knowledge of susceptible dynamics and may difficult to use in practice.
-Instead, during the exponential period, the generation-interval distribution is
-\begin{equation}
-g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau)
-\end{equation}
-and the reproductive number is
-\begin{equation}
-\RR = \int \tsub{c}{exp}(\tau) \exp(r\tau) d\tau.
-\end{equation}
-Note that applying the Lotka-Euler equation directly to the observed generation intervals will always underestimate the reproductive number:
-\begin{equation}
-\int_0^\infty \tsub{g}{exp}(\tau) \exp(r\tau) d\tau > \left(\int_0^\infty \tsub{g}{exp}(\tau). \exp(-r\tau) d\tau\right)^{-1}
-\end{equation}
-
-Finally, we obtain an estimator for mean generation interval and the reproductive number:
-\begin{equation}
-\begin{aligned}
-\hat{G} &= \frac{\sum_{i} \exp(r c_i) c_i}{\sum_{i} \exp(r c_i)},\\
-\hat{\RR} &= \sum_{i} \exp(r c_i),
-\end{aligned}
-\end{equation}
-where $r$ is the exponential growth rate and $c_i$ is an individual generation interval sample.
-
-\section{Appendix}
-
-These are notes for now ... Need to move them elsewhere and change notation... but this is part of our figure for empirical network ...
-
-$$
-1 = \kappa \phi_L(\alpha) \frac{\lambda^{(net)}}{\alpha + \lambda^{(net)}} (1 - \phi_I(\alpha+\lambda^{(net)}))
-$$
-where in our specific example,
-$$
-\phi_L(\alpha) = \frac{4 \delta^2}{(2 \delta + \alpha)^2}
-$$
-and so we have
-$$
-1 = \frac{4 \delta^2}{(2 \delta + \alpha)^2} \frac{\kappa \lambda^{(net)}}{\lambda^{(net)} + \alpha + \gamma},
-$$
-which suggests that
-$$
-\lambda^{(net)} = \frac{(2\delta + \alpha)^2 (\alpha + \gamma)}{(\kappa - 1) 4 \delta^2 - 4 \delta \alpha - \alpha^2}
-$$
-We can substitute into the following equation to obtain the $r$ and $\RR$ relationship:
-$$
-\RR = \kappa \frac{\lambda^{(net)}}{\lambda^{(net)} + \gamma}
-$$
-
 \bibliography{network}

### 2019-03-26 doc/network.tex
things

@@
 Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
-However, directly measuring the reproductive number often requires detailed knowledge of the natural history of a disease and may not be feasible early in an epidemic \swp{Can't figure out what I should cite here; maybe nothing}.
-Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
-These two quantities are linked by generation-interval distributions \citep{wallinga2007generation, park2019practical}.
+However, directly measuring the reproductive number often requires detailed knowledge of the natural history of a disease and may not be feasible early in an epidemic \citep{dietz1993estimation}.
+Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
+These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, wallinga2007generation, roberts2007model,  park2019practical}.
 
-Generation interval is defined as the time between when a person becomes infected and when that person infects another person.
+Generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
 Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.
@@
 An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
-The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
+\swp{Individuals may vary in terms of their prognosis (e.g., duration of an infectious period) and their infectiousness, which can depend both on their intrinsic infectivity as well as their neighborhood.}
+Then, the population-level kernel is given by integrating over these individual variations:
 \begin{equation}
@@
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
+\swp{
+In practice, integrating over the aspect space $A$ is expected to be a difficult problem, especially when an epidemic is driven by a few highly infectious individuals.
+For now, we assume that individual kernel properties are independent of risk of infection;
+we re-visit this assumption later in the discussion.
+}
 \jd{This is assuming that we know how to average across infected individuals.
@@
 \begin{equation}
-\RR_0 = \int K(t).
+\RR_0 = \int K(\tau) d\tau.
 \end{equation}
 \jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
+\swp{Not sure what to do here.}
 The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
 \begin{equation}
-g(\tau) = \frac{K(t)}{\RR_0}.
+g(\tau) = \frac{K(\tau)}{\RR_0}.
 \end{equation}
@@
 \jd{Rewrite these equations a bit more consistently. What are we using for: index infection, current infection? $\tau$ is difference between them. Is it simpler just to use ' for dummies?}
-In the homogeneous model, we assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
+\swp{I tried. $t$ is index, $\tau$ is difference. $s$ is used sometimes for dummies.}
+In a homogeneously mixing population, we can assume that the disease incidence $i(t)$ is the product of the current infectiousness of previously infected individuals and the proportion of the population susceptible, $S(t)$.
 \begin{equation}
-i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds.
+i(t) = S(t) \int K(s) i(t-s) ds = \RR \int g(s) i(t-s) ds.
 \end{equation}
 This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
-Over a period of time where $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
+Over a period of time where the proportion of susceptible $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
 \begin{equation}
@@
 
-The density of infections occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
+The density of infections occuring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
 \begin{equation}
-i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
+i_{t-\tau}(t) = \RR i(t-\tau) g(\tau) S(t)
 \end{equation}
 The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
-\jd{Is it nicer to keep $s$ in the numerator? Otherwise, why do we bother with $s$ at all?}
 \begin{equation}
-b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-x) g(x) d\tau}
+b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-s) g(s) ds}
 \end{equation}
@@
 \jd{Can we just skip this \P, or put in an appendix or something?}
-On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
+\swp{We decided to ask David for opinion}
+On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$.
 Then, the censored generation-interval distribution is given by
@@
 \end{equation}
-Thus, the censored generation-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
+Alternatively, the censored generation-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
 \begin{equation}
@@
 \end{equation}
-Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$.
+Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., when the epidemic is growing exponentially.
 
@@
 
-Since the right-censoring effect is a sampling bias, it must be corrected for.
+Since the right-censoring effect is a sampling bias, it must be taken into account.
 In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
@@
 \includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
-\caption{\textbf{Comparison of estimates of reproductive number based on various methods}
-Using the observed generation-interval distributions without correcting for right-censoring severely \jd{Use a log scale on the y axis to make this clearer} underestimates the reproductive number.
+\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
+Using the observed generation-interval distributions without correcting for right-censoring severely underestimates the reproductive number.
 Similarly, using intrinsic and egocentric generation-interval distributions without accounting for local spatial effects overestimates the reproductive number.

### 2019-03-26 fig/cmp.R

@@ -50,7 +50,7 @@ gendata2 <- genlist %>%
 
 ggR <- ggplot(RRdata2 %>% filter(type=="Larger epidemic")) +
 	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
-	ylab("Reproductive number") +
+	scale_y_log10("Reproductive number", breaks=c(2.5, 5, 10, 20)) +
 	facet_grid(~group, scale="free", space="free_x") +
 	theme(
 		panel.spacing.x = unit(0, "lines"),

### 2019-04-28 doc/network.tex
minor edit
edits

@@
 Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
-As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
+As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution \citep{kenah2008generation, tomba2010some, champredon2015intrinsic}.
 Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
-A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution.
-When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{champredon2015intrinsic}.
+A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{champredon2015intrinsic, britton2019estimation}.
+When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{tomba2010some, champredon2015intrinsic}.
 
-In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, measurements are ``censored'': we don't know what happens after the observation time.
-These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.
+In practice, generation intervals are often measured by aggregating available information from contact tracing. 
+Whereas forward and backward generation intervals represent infections that are one generation away from the focal cohort, these aggregated generation intervals represent infections across multiple cohorts.
+In this case, the time of infection of a cohort can range anywhere from the beginning of an epidemic to the most recent infection.
+As a result, while an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
+These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.
 
-Observed generation intervals also vary across space.
+Realized generation intervals also vary across space.
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
@@
 
-In this study, we explore spatiotemporal variation in observed generation-interval distributions.
+In this study, we explore spatiotemporal variation in generation intervals measured through contact tracing.
 We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
@@
 An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
-\swp{Individuals may vary in terms of their prognosis (e.g., duration of an infectious period) and their infectiousness, which can depend both on their intrinsic infectivity as well as their neighborhood.}
-Then, the population-level kernel is given by integrating over these individual variations:
+Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on their intrinsic infectivity as well as their neighborhood.
+
+The population-level kernel is given by integrating over these individual variations:
 \begin{equation}
@@
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
-\swp{
-In practice, integrating over the aspect space $A$ is expected to be a difficult problem, especially when an epidemic is driven by a few highly infectious individuals.
-For now, we assume that individual kernel properties are independent of risk of infection;
-we re-visit this assumption later in the discussion.
-}
+In order to integrate over the aspect space $A$, we have to assume that (1) theoretical distribution of individual kernel properties are known exactly and (2) the individual properties are independent of risk of infection;
+therefore, it only makes sense to define this integral for simple epidemic models for homogeneously mixing population, such as the Susceptible-Infected-Recovered (SIR) model (or its extensions).
+In practice, it is unlikely that averaging across an empirical distribution will give us the correct population-level kernel, especially when an epidemic is driven by a few highly infectious individuals.
+For brevity, we use this notation to distinguish the difference between the individual- and the population-level kernels.
 \jd{This is assuming that we know how to average across infected individuals.
@@
 If we have high-degree individuals, does it already break that assumption?}
+\swp{Can we say this?}
 
@@
 \jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
-\swp{Not sure what to do here.}
-The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
+Finally, the intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
 \begin{equation}
@@
 \end{equation}
+These definitions of the basic reproductive number and the intrinsic generation-interval distribution assume homogeneous mixing, as discussed above.
+\swp{Can we just say this? I'm not sure what else we can say.}
 
 \jd{Rewrite these equations a bit more consistently. What are we using for: index infection, current infection? $\tau$ is difference between them. Is it simpler just to use ' for dummies?}
-\swp{I tried. $t$ is index, $\tau$ is difference. $s$ is used sometimes for dummies.}
+\swp{I tried. $t$ is current time, $\tau$ is time difference. $s$ and $x$ are used for integrals.}
 In a homogeneously mixing population, we can assume that the disease incidence $i(t)$ is the product of the current infectiousness of previously infected individuals and the proportion of the population susceptible, $S(t)$.
 \begin{equation}
-i(t) = S(t) \int K(s) i(t-s) ds = \RR \int g(s) i(t-s) ds.
+i(t) = S(t) \int K(s) i(t-s) ds = \RR \int g(s) i(t-s) ds,
 \end{equation}
+where $\RR = S(t) \RR_0$ is the effective reproductive number.
 This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
-Over a period of time where the proportion of susceptible $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
+Over a period of time where the proportion of susceptible $S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
 \begin{equation}
-\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
+\frac{1}{\RR} = \int g(s) \exp(-r s) ds.
 \end{equation}
@@
 \begin{equation}\label{eq:obsg}
-c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(\tau) \int_0^t i(s-\tau) S(s) ds d\tau}
+c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(x) \int_0^t i(s-x) S(s) ds dx}
 \end{equation}
@@
 
-\subsection{Egocentric kernel}
-
 An infected individual may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
@@
 The extreme dilution examples would be mean degree (if highly connected people infect that much more) or individual degree (if they don't infect more at all.)}
+\swp{That seems redundant. $K_a$ already has big (individual) R. Why do we multiply by another R? I added more. Is this better?}
 \begin{equation}
@@
 The dilution term, $\delta_a$, models how contacts are distributed among susceptible acquaintances.
-For example, when infectious contacts are distributed equally in a homogeneous population, we have $\delta_a = 1/(N-1)$, where $N$ is the population size.
+Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \citep{trapman2016inferring}.
+Then, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$ and the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
 
 The population-level egocentric kernel is given by integrating over individual variations:
-\begin{equation}
+\begin{equation}\label{eq:ego}
 \hat{K}(\tau) = \int \hat{K}_a(\tau) dA,
@@
 \begin{equation}
-\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K} \tau}.
+\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
 \label{eq:conditional}
 \end{equation}
-The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for brevity we will often omit ``population-level''.
+The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; 
+for brevity we will often omit ``population-level''.
+Finally, the egocentric reproductive number can be estimated by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
+\begin{equation}
+\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
+\end{equation}
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
+Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
 
-For example, consider a susceptible-exposed-infected-recovered (SEIR) model.
-This widely used model assumes that latent and infectious periods are exponentially distributed.
+For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
 The intrinsic generation-interval distribution that corresponds to this model can be written has \citep{svensson2015influence}:
@@
 where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
-Assuming that per-pair contact rate is $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution is given by:
+Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
 \begin{equation}
@@
 
-Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks -- although there has been interesting work on spatial cases, particularly on hierarchical networks of households nested within communities (e.g.,  \cite{tomba2010some}).
+Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks -- although there has been interesting work on special cases, particularly on hierarchical networks of households nested within communities (e.g.,  \cite{tomba2010some}).
 \jd{I wonder if any of Frank Ball's household work could be relevant.}
+\swp{Come back.}
 
-\subsection{Linking growth rate and reproductive number}
-
-\swp{I might want to move this paragraph to after equation 13. It doesn't feel like it fits here.}
-The intrinsic generation-interval distribution provides a link between the speed (exponential growth rate, $r$) and the strength (reproductive number, $\RR$) of an epidemic via Euler-Lotka equation \citep{lotka1907relation} but assumes a homogeneously mixing population.
-Instead, we can derive a relationship between $r$ and $\RR$ that accounts for the egocentric effect using the egocentric generation-interval distribution:
-\begin{equation}
-\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
-\end{equation}
-This is equivalent to the relationship between $r$ and $\RR$ that \cite{trapman2016inferring} derived under a network structure.
-\swp{Need to confirm; Maybe prove it in appendix or something...
-will get back to this later}
-As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
-Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
-
-\section{Effective distributions}
+\section{Inferring generation-interval distributions from a contact tracing data}
 
@@
 
-Since the right-censoring effect is a sampling bias, it must be taken into account.
+Since the right-censoring effect is a sampling bias, it must be taken into account explicitly.
 In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
-We therefore expect that if we correct observed generation intervals for the right-censoring effect, spatial components will be accounted for implicitly, and we can use these temporally corrected intervals to correctly link $r$ and $\RR$.
-and we can obtain principled esi.
+We therefore expect that if we correct observed generation intervals for the right-censoring effect, spatial components will be taken into account implicitly. 
+We can use these temporally corrected intervals to correctly link $r$ and $\RR$.
 Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution. 
@@
 \jd{In a network with limited degree, they differ due to egocentric effects; in a non-random network they differ due to local effects; beyond the early period, they differ due to global effects.}
-
-\section{Inferring generation-interval distributions from a contact tracing data}
-
-In this section, we investigate two statistical methods for correcting for temporal bias in a contact tracing data.
-Detailed derivations of the two methods can be found in Methods.
-\swp{TODO: Write methods}
 
@@
 
+Here, we investigate two methods for correcting for temporal bias in a contact tracing data.
 We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
 As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
-Hence, the population-level method relies on the observed generation intervals and the exponential growth rate \citep{tomba2010some, nishiura2010time}:
+Hence, the population-level method relies on the observed generation intervals and the exponential growth rate \citep{tomba2010some, nishiura2010time, britton2019estimation}:
 \begin{equation}
@@
 \end{equation}
-When the susceptible dynamics is known, this equation can be generalized by using \eref{obsg}; we do not pursue this direction in this study.
+When the susceptible dynamics is known, this equation can be generalized by using \eref{obsg}; we do not pursue this direction here.
 The population-level method is simple and intuitive, but does not take into account who infected whom.
-%% we might therefore expect the individual-level method to be more powerful.
-%% SWP: We haven't introduced the individual-level method yet...
 
-The individual-level method assumes that the initial exponential growth phase of an epidemic can be approximated by a Poisson process to derive a likelihood for the observations from each individual:
+The individual-level method assumes that the initial exponential growth phase of an epidemic can be approximated by a nonhomogeneous Poisson process.
+
+to derive a likelihood for the observations from each individual:
 \begin{equation}

### 2019-04-28 doc/network.tex
going to work on it again tomorrow...

@@
 
-The individual-level method assumes that the initial exponential growth phase of an epidemic can be approximated by a nonhomogeneous Poisson process.
+\swp{Cite something about poisson process}
+We refer to the second method as the individual-level method because it relies on individual contact information.
+We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution.
+Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is proportional to
+\begin{equation}
+\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta),
+\end{equation}
+where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
+We use gamma distribution to model the effective generation-interval distribution;
+parameters $\theta$ correspond to the mean and shape parameters of a gamma distribution.
 
-to derive a likelihood for the observations from each individual:
+The entire data likelihood of the individual-level method can be written as:
 \begin{equation}
@@
 \end{equation}
-where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by infector $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
+where $N$ is the total number of infected individuals.
+Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
 \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
 
+\swp{Probabaly move simulations to appendix}
 To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).

### 2019-04-30 doc/network.tex
making changes here and there

@@
 
-%% SWP:
-%% Got data from
+%% SWP: see or cite
 %% https://www.ncbi.nlm.nih.gov/pubmed/29087987
-%% Try and analyze
-%% See what's different and understand what they did
 
@@
 
-Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
-As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution \citep{kenah2008generation, tomba2010some, champredon2015intrinsic}.
+Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
+As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
+and the distribution of forward generation intervals will be shorter than the intrinsic distribution \citep{kenah2008generation, tomba2010some, champredon2015intrinsic}.
 Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
-A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{champredon2015intrinsic, britton2019estimation}.
+A susceptible individual is thus relatively more likely to be infected by a newly infected individual, 
+and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{champredon2015intrinsic, britton2019estimation}.
 When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{tomba2010some, champredon2015intrinsic}.
@@
 In practice, generation intervals are often measured by aggregating available information from contact tracing. 
-Whereas forward and backward generation intervals represent infections that are one generation away from the focal cohort, these aggregated generation intervals represent infections across multiple cohorts.
-In this case, the time of infection of a cohort can range anywhere from the beginning of an epidemic to the most recent infection.
+Whereas forward and backward generation intervals represent infections that are one generation apart from a cohort, these aggregated generation intervals represent infections across multiple cohorts; 
+the time of infection of a cohort can range anywhere from the beginning of an epidemic to the most recent infection.
 As a result, while an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
@@
 
-Realized generation intervals also vary across space.
+Realized generation intervals can also vary across space.
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
@@
 
-\swp{I might rename the sections later but I think I like the current order.}
-\section{Generation-interval distributions across time}
-
-\subsection{Intrinsic generation-interval distribution}
+\section{Intrinsic generation-interval distributions}
 
@@
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
-In order to integrate over the aspect space $A$, we have to assume that (1) theoretical distribution of individual kernel properties are known exactly and (2) the individual properties are independent of risk of infection;
-therefore, it only makes sense to define this integral for simple epidemic models for homogeneously mixing population, such as the Susceptible-Infected-Recovered (SIR) model (or its extensions).
-In practice, it is unlikely that averaging across an empirical distribution will give us the correct population-level kernel, especially when an epidemic is driven by a few highly infectious individuals.
-For brevity, we use this notation to distinguish the difference between the individual- and the population-level kernels.
+Here, we assume that the individual properties are independent of risk of infection.
+In practice, it is unlikely that naively averaging across an empirical distribution will give us the correct population-level kernel, especially when an epidemic is driven by a few highly infectious individuals.
 \jd{This is assuming that we know how to average across infected individuals.
@@
 There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
-The basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
+Assuming that a population mixes homogeneously, the basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- can be defined as:
 \begin{equation}
@@
 \jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
+\swp{Can we just say this? I'm not sure what else we can say.}
 Finally, the intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
@@
 \end{equation}
-These definitions of the basic reproductive number and the intrinsic generation-interval distribution assume homogeneous mixing, as discussed above.
-\swp{Can we just say this? I'm not sure what else we can say.}
 
@@
 
-\subsection{Right-censored generation-interval distribution}
+\section{Generation-interval distributions across time}
 
-Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
-Several studies have associated the backward generation-interval distribution with the early-epidemic censored generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}.
-The censored generation-interval distribution in fact averages over backward generation-interval distributions up until the observation time.
-While we mathematically expect these underlying distributions to be the same (and thus the same distribution as their average) during the exponential growth phase, these assumptions can break down (due to susceptible depletion or other temporal change).
-This, it is useful to conceptually distinguish the censored distribution the backward distribution.
+When generation intervals are measured through contact tracing during an outbreak, 
+the observed intervals are ``right-censored'', because intervals that end after the observation time are not observed. 
+Several studies have associated the backward generation-interval distribution with the early-epidemic observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation};
+in fact, the censored generation-interval distribution averages over backward generation-interval distributions up until the observation time.
+While we mathematically expect these underlying distributions to be the same (and thus the same distribution as their average) during the exponential growth phase, these assumptions can break down as an epidemic progresses.
+Here, we want to conceptually distinguish the censored distribution the backward distribution.
 
-The density of infections occuring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
+The density of new infections occuring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
 \begin{equation}
@@
 \swp{We decided to ask David for opinion}
-On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$.
+On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$.
 Then, the censored generation-interval distribution is given by
@@
 
-The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\RR$) of an epidemic \citep{wallinga2007generation, park2019practical}.
-This link has been widely adopted in outbreak analyses as it allows us to conveniently estimate the reproductive number from measurable quantities.
-However, determining the \emph{correct} generation-interval distribution is challenging: observed generation-interval distributions can vary depending on how and when it is measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
-Here, we decompose variation in the observed generation intervals into spatial and temporal components under a realistic setting where contact tracing data are aggregated over time.
-We use simple models to distinguish among different sources of variations in the observed generation intervals.
-We provide two methods for correcting temporal bias and test their accuracy using simulations.
+The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\RR$) of an epidemic \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
+However, observed generation-interval distributions can vary depending on how and when it is measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
+determining which distribution correctly links $r$ and $\RR$ can be challenging.
+Here, we explore how observed generation intervals, measured through contact tracing, vary across time and space.
+Whereas temporal variations represent observation bias, spatial variations reflect the contact structure of a realized outbreak.
+Therefore, correcting for temporal variation, but not for spatial variation, in an observed generation-interval distribution provides the correct link between $r$ and $\RR$, allowing us to estimate the reproductive number $\RR$ that is consistent with the realized infection process.
 
 Generation intervals can be measured either forward (by tracing the infectees of an infected individual) or backward (by tracing the infector of an infected individual).
-Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart.
+Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart from a single cohort.
 Instead, we consider a slightly different scenario in which generation intervals are measured over multiple generations.
 This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
-In this case, the observed generation intervals are derived all infections that occured before the last date on which contact tracing was performed and are subject to right-censoring -- it is not possible to trace individuals that have not been incted yet.
+In this case, the observed generation intervals are derived from all infections that occured before the last date on which contact tracing was performed.
 
-When the epidemic is growing exponentially, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-censoring.
-Near the end of an epidemic, the effect of right-censoring becomes negligible but the observed generation intervals are still shorter on average than the intrinsic generation intervals.
+The observed generation intervals are subject to right-censoring -- it is not possible to trace individuals that have not been incted yet.
+During an ongoing outbreak, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-censoring.
+Near the end of an outbreak, the effect of right-censoring becomes negligible but the observed generation intervals are still shorter on average than the intrinsic generation intervals. 
+The remaining bias can be explained by depletion of the susceptible population.
 
-We further classify susceptible depletion into three levels: egocentric, local, and global.
+We classify susceptible depletion into three levels: egocentric, local, and global.
 Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
-Accounting for the egocentric effect allows us to reinterpret the results by Trapman et al using arguments based on generation intervals.
+Accounting for the egocentric effect allows us to reinterpret the results by \cite{trapman2016inferring} using arguments based on generation intervals.
 Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
 Global susceptible depletion refers to the overall decrease in the susceptible pool in the population.
-
-Susecptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.
+Egocentric and local depletion effects are important
 

### 2019-04-30 doc/network.tex
trying to make things better

@@
 \swp{We decided to ask David for opinion}
+\swp{I tried to rewrite the section to make things clearer; it seems to make this paragraph less boring and more relevant?}
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$.
@@
 \end{equation}
-Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., when the epidemic is growing exponentially.
+Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., when the epidemic is growing exponentially;
+however, this assumption quickly breaks down as the epidemic progresses (Figure~\ref{fig:censor}).
 
@@
 \caption{\textbf{Temporal variation in the mean observed generation interval.}
-A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the observed (right-censored) mean generation interval was calculated over the course of an epidemic (see methods).
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the backward and the censored mean generation interval were calculated over the course of an epidemic.
 }
@@
 Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized intervals will always be shorter \citep{champredon2015intrinsic}.
-As a result, even if contact tracing is performed through an entire epidemic, mean generation interval will be underestimated.
+As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
 
-When a disease is at (or near) endemic equilibrium, the number of susceptibles remains (approximately) constant over time, and we expect the observed generation-interval distribution to be similar to the intrinsic generation-interval distribution.
+% When a disease is at (or near) endemic equilibrium, the number of susceptibles remains (approximately) constant over time, and we expect the observed generation-interval distribution to be similar to the intrinsic generation-interval distribution.
 

### 2019-04-30 fig/temporal_effect.R

@@ -1,4 +1,5 @@
 library(deSolve)
+library(tidyr)
 library(dplyr)
 library(ggplot2); theme_set(theme_bw())
 library(gridExtra)
@@ -14,16 +15,6 @@ seir.ode <- function(t, y, parms) {
         dI <- sigma*E - gamma*I
         list(c(dS, dE, dI))
     })
-}
-
-## we want to simulate equilibrium state!
-seirs.ode <- function(t, y, parms) {
-	with(as.list(c(y, parms)), {
-		dS <- gamma * I -beta/N*S*I
-		dE <- beta/N*S*I - sigma*E
-		dI <- sigma*E - gamma*I
-		list(c(dS, dE, dI))
-	})
 }
 
 sumfun <- function(out,
@@ -46,6 +37,13 @@ sumfun <- function(out,
 		beta*sum(incidence[1:(m-i+1)] * out$S[i:m]/N)/total*genden[i]
 	}
 	
+	backward <- function(tau, tmax) {
+		m <- which(zapsmall((tmax-tau)-t)==0)
+		i <- which(t==tau)
+		
+		incidence[m] * genden[i]
+	}
+	
 	mean.censor <- function(tmax) {
 		if (tmax<=0){
 			return(0)
@@ -57,12 +55,25 @@ sumfun <- function(out,
 		}
 	}
 	
+	mean.backward <- function(tmax) {
+		if (tmax<=0){
+			return(0)
+		} else{
+			t <- t[t <= tmax]
+			
+			dc <- sapply(t, backward, tmax=tmax)
+			return(sum(dc*t)/sum(dc))
+		}
+	}
+	
 	tau <- seq(0, tmax, tauby)
 	cm <- sapply(tau, mean.censor)
+	bm <- sapply(tau, mean.backward)
 	
 	censor.df <- data.frame(
 		t=tau
-		, mean=cm
+		, censored=cm
+		, backward=bm
 	)
 }
 
@@ -73,26 +84,20 @@ dt <- 0.1
 tmax <- 305
 t <- seq(0, tmax, dt)
 out1 <- as.data.frame(ode(y, t, seir.ode, parms))
-out2 <- as.data.frame(ode(y, t, seirs.ode, parms))
-out3 <- as.data.frame(ode(unlist(tail(out2, 1))[-1] , t, seirs.ode, parms))
[…]

### 2019-04-30 doc/network.tex
stuff

@@
 
-An infected individual may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
+Infected individuals may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
 Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average, in a limited contact network.
+
 To explore the effect of multiple contacts on realized generation intervals, we first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
@@
 Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \citep{trapman2016inferring}.
-Then, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$ and the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
+Then, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
 
@@
 As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
-Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
+% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
 
@@
 
+\swp{Shouldn't we put this earlier?}
 In general, spatial reduction in the mean generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
@@
 \jd{I wonder if any of Frank Ball's household work could be relevant.}
-\swp{Come back.}
+\swp{Will come back.}
 

### 2019-04-30 doc/network.tex
stuff

@@
 Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
-Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters (see Methods).
+Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters.
 }
@@
 Global susceptible depletion refers to the overall decrease in the susceptible pool in the population.
-Egocentric and local depletion effects are important
-
-Limitations
-\begin{itemize}
-	\item Exponential growth; polynomial
-	\item Extending the framework (accounting for control)
-	\item Low coverage of gamma
-	\item Serial interval
-\end{itemize}
-
-\section{Methods}
-
-\subsection{Deterministic SEIR model}
-
-The Susceptible-Exposed-Infected-Recovered (SEIR)
-
-\subsection{Stochastic SEIR model on a network}
 

### 2019-04-30 doc/network.tex
small changes

@@
 (Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
-All graphs were generated using 5000 stochastic simulations were on a network with 5 nodes (1 infector and 4 susceptibles).
+All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles).
 }
@@
 
-\swp{Shouldn't we put this earlier?}
 In general, spatial reduction in the mean generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
@@
 Whereas temporal variations represent observation bias, spatial variations reflect the contact structure of a realized outbreak.
-Therefore, correcting for temporal variation, but not for spatial variation, in an observed generation-interval distribution provides the correct link between $r$ and $\RR$, allowing us to estimate the reproductive number $\RR$ that is consistent with the realized infection process.
+Therefore, correcting for temporal variation, but not for spatial variation, in an observed generation-interval distribution provides a link between $r$ and $\RR$ which allows us to estimate the reproductive number $\RR$ that is consistent with the realized infection process.
 

### 2019-04-30 doc/network.tex
new paragraph

@@
 
-\begin{figure}[t]
+\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
@@
 A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the backward and the censored mean generation interval were calculated over the course of an epidemic.
+The dotted horizontal line represents the mean intrinsic generation interval.
 }
@@
 determining which distribution correctly links $r$ and $\RR$ can be challenging.
-Here, we explore how observed generation intervals, measured through contact tracing, vary across time and space.
-Whereas temporal variations represent observation bias, spatial variations reflect the contact structure of a realized outbreak.
-Therefore, correcting for temporal variation, but not for spatial variation, in an observed generation-interval distribution provides a link between $r$ and $\RR$ which allows us to estimate the reproductive number $\RR$ that is consistent with the realized infection process.
+Here, we analyze how observed generation intervals, measured through contact tracing, vary across time and space.
+Temporal variations represent observation bias, whereas spatial variations reflect the contact structure of a realized outbreak.
+Therefore, correcting for temporal variation, but not for spatial variation, in an observed generation-interval distribution provides the correct link between $r$ and $\RR$.
 
@@
 This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
-In this case, the observed generation intervals are derived from all infections that occured before the last date on which contact tracing was performed.
+In this case, the generation intervals are measured from all observed cases that occured before the last date on which contact tracing was performed.
 
@@
 
+Susecptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.
+Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
+Therefore, we expect the observed generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the reproductive number by simply accounting for the right-censoring.
+We refer to the temporally-corrected distribution as the effective generation-interval distribution.
+
+We compare two methods for estimating the effective generation-interval distribution and assume that the effective generation-interval distribution follows a gamma distribution.
+The gamma approximation of the generation interval distribution has widely used due to its simplicity;
+we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$.
+However, further investigation of our methods suggests that making a wrong distributional assumption leads to narrow confidence intervals, even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model), making a wrong distributional assumption leads to narrow confidence intervals \swp{Appendix}.
+These results are particularly alarming because it is impossible to know the true shape of the distribution [CITE].
+
 \bibliography{network}

### 2019-04-30 doc/network.tex
Preparing.

@@
 Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.
-There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals are measured by evaluating the infectiousness of infected people,
+There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals are based on the infectiousness of infected people,
 %% without regard to actual infection events,
@@
 the time of infection of a cohort can range anywhere from the beginning of an epidemic to the most recent infection.
-As a result, while an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
+While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
 These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.
@@
 Realized generation intervals can also vary across space.
+\jd{Realized generation intervals are also affected by spatial and network structure.}
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
@@
 \swp{Can we say this?}
+\jd{Let's discuss the whole thing. Maybe we can refer to a theoretical average?}
 
@@
 in fact, the censored generation-interval distribution averages over backward generation-interval distributions up until the observation time.
-While we mathematically expect these underlying distributions to be the same (and thus the same distribution as their average) during the exponential growth phase, these assumptions can break down as an epidemic progresses.
-Here, we want to conceptually distinguish the censored distribution the backward distribution.
+During the exponential growth phase, we expect the backward distributions to be the same (and thus the censored distribution should match them).
+This connection will break down as the epidemic progresses, however, so we will to distinguish the censored distribution from the backward distribution.
 

### 2019-04-30 doc/network.tex
may cause conflicts

--- 2019-04-30 doc/network.tex
@@
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
-While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and estimated reproductive numbers. 
-Both local and global depletion effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
-local depletion will be important early in an epidemic whereas global depletion will be small until sufficient amount of susceptible individuals become infected.
-As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local structure of the contact network near the source of an outbreak.
+While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and estimated reproductive numbers: 
+both local and global depletion effects reduce number of infections that occur and shorten generation intervals.
+% SWP: I would rather put these in Discussion?
+% but the time scale on which they matter differs:
+% local depletion will be important early in an epidemic whereas global depletion will be small until sufficient amount of susceptible individuals become infected.
+% As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local structure of the contact network near the source of an outbreak.
 
@@
 
-\swp{Cite something about poisson process}
 We refer to the second method as the individual-level method because it relies on individual contact information.
-We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution.
-Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is proportional to
+We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution \citep{daley2007introduction}.
+Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
 \begin{equation}
-\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta),
+\frac{\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta)}{n_j!},
 \end{equation}
@@
 Generation intervals can be measured either forward (by tracing the infectees of an infected individual) or backward (by tracing the infector of an infected individual).
-Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart from a single cohort.
-Instead, we consider a slightly different scenario in which generation intervals are measured over multiple generations.
+Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart from a single cohort \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+Instead, we considered a slightly different scenario in which generation intervals are measured over multiple generations.
 This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
@@
 
-We classify susceptible depletion into three levels: egocentric, local, and global.
+We classified susceptible depletion into three levels: egocentric, local, and global.
 Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
-Accounting for the egocentric effect allows us to reinterpret the results by \cite{trapman2016inferring} using arguments based on generation intervals.
+Accounting for the egocentric effect allowed us to reinterpret the results by \cite{trapman2016inferring} using arguments based on generation intervals.
 Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
@@
 Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
-Therefore, we expect the observed generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the reproductive number by simply accounting for the right-censoring.
-We refer to the temporally-corrected distribution as the effective generation-interval distribution.
+Therefore, we predicted the observed generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the \emph{effective} generation-interval distribution by simply accounting for the right-censoring.
+Simulation studies confirmed our prediction: using the effective generation-interval distribution provides a proper link between $r$ and $\RR$.
 
@@
 
+\swp{Conclusion?}
+
+\section{Methods}
+
 \bibliography{network}

### 2019-04-30 doc/network.tex

@@
 \swp{I tried. $t$ is current time, $\tau$ is time difference. $s$ and $x$ are used for integrals.}
+\jd{How about: $s$ as a dummy for $t$, $\sigma$ as a dummy for $\tau$?}
+
 In a homogeneously mixing population, we can assume that the disease incidence $i(t)$ is the product of the current infectiousness of previously infected individuals and the proportion of the population susceptible, $S(t)$.
@@
 \swp{I tried to rewrite the section to make things clearer; it seems to make this paragraph less boring and more relevant?}
+
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$.

### 2019-04-30 doc/network.tex
Merging with Daniel

@@
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
-While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and estimated reproductive numbers: 
-both local and global depletion effects reduce number of infections that occur and shorten generation intervals.
-% SWP: I would rather put these in Discussion?
-% but the time scale on which they matter differs:
-% local depletion will be important early in an epidemic whereas global depletion will be small until sufficient amount of susceptible individuals become infected.
-% As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local structure of the contact network near the source of an outbreak.
+While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and estimated reproductive numbers. 
+Both local and global depletion effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
+local depletion will be important early in an epidemic whereas global depletion will be small until sufficient amount of susceptible individuals become infected.
+As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local structure of the contact network near the source of an outbreak.
 
@@
 
+\swp{Cite something about poisson process}
 We refer to the second method as the individual-level method because it relies on individual contact information.
-We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution \citep{daley2007introduction}.
-Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
+We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution.
+Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is proportional to
 \begin{equation}
-\frac{\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta)}{n_j!},
+\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta),
 \end{equation}
@@
 Generation intervals can be measured either forward (by tracing the infectees of an infected individual) or backward (by tracing the infector of an infected individual).
-Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart from a single cohort \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
-Instead, we considered a slightly different scenario in which generation intervals are measured over multiple generations.
+Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart from a single cohort.
+Instead, we consider a slightly different scenario in which generation intervals are measured over multiple generations.
 This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
@@
 
-We classified susceptible depletion into three levels: egocentric, local, and global.
+We classify susceptible depletion into three levels: egocentric, local, and global.
 Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
-Accounting for the egocentric effect allowed us to reinterpret the results by \cite{trapman2016inferring} using arguments based on generation intervals.
+Accounting for the egocentric effect allows us to reinterpret the results by \cite{trapman2016inferring} using arguments based on generation intervals.
 Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
@@
 Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
-Therefore, we predicted the observed generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the \emph{effective} generation-interval distribution by simply accounting for the right-censoring.
-Simulation studies confirmed our prediction: using the effective generation-interval distribution provides a proper link between $r$ and $\RR$.
+Therefore, we expect the observed generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the reproductive number by simply accounting for the right-censoring.
+We refer to the temporally-corrected distribution as the effective generation-interval distribution.
 
@@
 
-\swp{Conclusion?}
-
-\section{Methods}
-
 \bibliography{network}

### 2019-04-30 doc/network.tex
Meeting with Daniel

--- 2019-04-30 doc/network.tex
@@
 
-The observed generation intervals are subject to right-censoring -- it is not possible to trace individuals that have not been incted yet.
+The observed generation intervals are subject to right-censoring -- it is not possible to trace individuals that have not been infected yet.
 During an ongoing outbreak, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-censoring.
@@
 
-Susecptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.
+Susceptible depletion happening at all three levels shortens realized generation intervals but acts on different time scales.
 Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
 Therefore, we predicted the observed generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the \emph{effective} generation-interval distribution by simply accounting for the right-censoring.
-Simulation studies confirmed our prediction: using the effective generation-interval distribution provides a proper link between $r$ and $\RR$.
+Simulation studies confirmed our prediction: using the effective generation-interval distribution provides the correct link between $r$ and $\RR$.
 
@@
 we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$.
-However, further investigation of our methods suggests that making a wrong distributional assumption leads to narrow confidence intervals, even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model), making a wrong distributional assumption leads to narrow confidence intervals \swp{Appendix}.
+However, further investigation of our methods suggests that making a wrong distributional assumption leads to overly narrow confidence intervals, even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model) \swp{Appendix}.
 These results are particularly alarming because it is impossible to know the true shape of the distribution [CITE].
@@
 
+\jd{Let me try.}
+
 \section{Methods}
+
+\jd{Maybe a bit more detail here.}
 

### 2019-05-06 doc/network.tex
commit

--- 2019-04-30 doc/network.tex
@@
 \swp{I tried. $t$ is current time, $\tau$ is time difference. $s$ and $x$ are used for integrals.}
-\jd{How about: $s$ as a dummy for $t$, $\sigma$ as a dummy for $\tau$?}
-
 In a homogeneously mixing population, we can assume that the disease incidence $i(t)$ is the product of the current infectiousness of previously infected individuals and the proportion of the population susceptible, $S(t)$.
@@
 \end{equation}
-The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
+The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i(t-\tau) g(s)$:
 \begin{equation}
@@
 \swp{I tried to rewrite the section to make things clearer; it seems to make this paragraph less boring and more relevant?}
-
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$.

### 2019-05-06 doc/network.tex
resolving comments

--- 2019-04-30 doc/network.tex
@@
 
-Realized generation intervals can also vary across space.
-\jd{Realized generation intervals are also affected by spatial and network structure.}
+Realized generation intervals are also affected by spatial and network structure.
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
@@
 In practice, it is unlikely that naively averaging across an empirical distribution will give us the correct population-level kernel, especially when an epidemic is driven by a few highly infectious individuals.
-\jd{This is assuming that we know how to average across infected individuals.
-Do we need to assume that individual kernel properties are independent of risk of infection?
-If we have high-degree individuals, does it already break that assumption?}
-\swp{Can we say this?}
-\jd{Let's discuss the whole thing. Maybe we can refer to a theoretical average?}
 
@@
 \end{equation}
-\jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
-\swp{Can we just say this? I'm not sure what else we can say.}
 Finally, the intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
@@
 \end{equation}
-
-\jd{Rewrite these equations a bit more consistently. What are we using for: index infection, current infection? $\tau$ is difference between them. Is it simpler just to use ' for dummies?}
-\swp{I tried. $t$ is current time, $\tau$ is time difference. $s$ and $x$ are used for integrals.}
-\jd{How about: $s$ as a dummy for $t$, $\sigma$ as a dummy for $\tau$?}
 
@@
 \end{equation}
-The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
+The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i(t-\tau) g(\tau)$:
 \begin{equation}
-b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-s) g(s) ds}
+b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-\sigma) g(\sigma) d\sigma}
 \end{equation}
@@
 \begin{equation}\label{eq:obsg}
-c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(x) \int_0^t i(s-x) S(s) ds dx}
+c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(x) \int_0^t i(s-\sigma) S(s) ds d\sigma}
 \end{equation}
@@
 We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case $a$ in the absence of other infectors:
-\jd{We can't use dilution directly in the hazard term: it needs to be multiplied by something like big R.
-I would put $R\delta$ there.
-The extreme dilution examples would be mean degree (if highly connected people infect that much more) or individual degree (if they don't infect more at all.)}
-\swp{That seems redundant. $K_a$ already has big (individual) R. Why do we multiply by another R? I added more. Is this better?}
 \begin{equation}
@@
 \swp{Conclusion?}
-
 \jd{Let me try.}
@@
 
-\jd{Maybe a bit more detail here.}
-
 \bibliography{network}

### 2019-05-06 doc/network.tex
trying to write methods

@@
 We refer to the second method as the individual-level method because it relies on individual contact information.
-We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate is equal to the effective generation-interval distribution \citep{daley2007introduction}.
+We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR g(t - t_j)$, where $g(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
+We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
 Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
@@
 where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
-We use gamma distribution to model the effective generation-interval distribution;
-parameters $\theta$ correspond to the mean and shape parameters of a gamma distribution.
 
@@
 Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
+Since the estimate of $\RR$ is sensitive to under-reporting, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate.
 \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
 
-\swp{Probabaly move simulations to appendix}
-To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
+To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
 \cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
@@
 
+The generation-interval-based approach to estimating the reproductive number assumes that an epidemic grows exponentially \citep{wallinga2007generation, park2019practical}.
+In practice, heterogeneity in population structure can lead to subexponential growth \citep{chowell2016characterizing, viboud2016generalized}.
+
 \swp{Conclusion?}
@@
 
+\subsection{Deterministic SEIR model}
+
+\swp{CITE}
+To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
+The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
+While we use the standard SEIR model, which assumes exponentially distributed latent and infectious period, throughout this paper, we present the extended version of the model, which allows for Erlang-distributed latent and infectious periods (we present some results based on the extended model in Appendix):
+\begin{equation}
+\begin{aligned}
+\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
+\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
+\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
+\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
+\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} I_n) && \text{for } n = 2, 3, \dots, n_I\\
+\end{aligned}
+\end{equation}
+where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
+Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, and $1/\gamma$ is the mean infectious period.
+
+\subsection{Stochastic SEIR model}
+
+\swp{What do I cite here? Gillespie?}
+We use individual-based simulations of the stochastic SEIR model on a network to study variation in realized generation intervals.
+Every node of a network is numbered from 1 to $N$, where $N$ is the total number of nodes. 
+Each node represents an individual.
+Likewise, an edge of a network represents acquaintanceship; infected individuals can only contact their acquaintances.
+
+First, we begin by randomly selecting initially infected individuals.
+For each infected individual $i$, we randomly draw a random number $E_i$ from an Erlang distribution with mean $1/\sigma$ and shape $n_E$.
+Given a total contact rate of $k_i \lambda$, where $k_i$ is the number of acquaintances and $\lambda$ is the per-pair contact rate, and a transition rate of $n_I \gamma$ between each infectious stages,
+the total number of contacts that an indiviudal $i$ in infectious stage $j$ makes before moving to infectious stage $j+1$ follows a geometric distribution with probability $n_I\gamma/(k_i \lambda + n_I\gamma)$:
+\begin{equation}
+C_{i, j} \sim \mathrm{Geometric}\left(\frac{n_I\gamma}{k_i \lambda + n_I\gamma}\right).
+\end{equation}
+For each contact, a contactee is uniformly sampled from the set of acquaintances of the individual $i$.
+
+\subsection{Empirical network}
+
 \bibliography{network}

### 2019-05-06 doc/network.tex
update

@@
 
-\swp{What do I cite here? Gillespie?}
-We use individual-based simulations of the stochastic SEIR model on a network to study variation in realized generation intervals.
-Every node of a network is numbered from 1 to $N$, where $N$ is the total number of nodes. 
-Each node represents an individual.
-Likewise, an edge of a network represents acquaintanceship; infected individuals can only contact their acquaintances.
+\swp{What do I cite here?}
+We develop an algorithm to simulate an individual-based SEIR model on a network to keep track of generation intervals.
+Our algorithm is based on the Gillespie algorithm.
+We model individuals with nodes on a network and their acquaintanceships with edges; infected individuals can only contact their acquaintances.
 
-First, we begin by randomly selecting initially infected individuals.
-For each infected individual $i$, we randomly draw a random number $E_i$ from an Erlang distribution with mean $1/\sigma$ and shape $n_E$.
-Given a total contact rate of $k_i \lambda$, where $k_i$ is the number of acquaintances and $\lambda$ is the per-pair contact rate, and a transition rate of $n_I \gamma$ between each infectious stages,
-the total number of contacts that an indiviudal $i$ in infectious stage $j$ makes before moving to infectious stage $j+1$ follows a geometric distribution with probability $n_I\gamma/(k_i \lambda + n_I\gamma)$:
+First, we begin by randomly selecting initially infected individuals; these individuals are assumed to be infected at $t = 0$.
+For each infected individual $i$, we randomly draw a random number $E_i$ from an Erlang distribution with mean $1/\sigma$ and shape $n_E$ to represent the length of their latent period.
+Given a total contact rate of $S_i \lambda$, where $S_i$ is the number of susceptible acquaintances and $\lambda$ is the per-pair contact rate, and a transition rate of $n_I \gamma$ between each infectious stages,
+the total number of contacts that an indiviudal $i$ in infectious stage $j$ makes before moving to infectious stage $j+1$ follows a geometric distribution with probability $n_I\gamma/(S_i \lambda + n_I\gamma)$:
 \begin{equation}
-C_{i, j} \sim \mathrm{Geometric}\left(\frac{n_I\gamma}{k_i \lambda + n_I\gamma}\right).
+C_{i, j} \sim \mathrm{Geometric}\left(\frac{n_I\gamma}{S_i \lambda + n_I\gamma}\right).
 \end{equation}
 For each contact, a contactee is uniformly sampled from the set of acquaintances of the individual $i$.
+Then, the time between two events follows an exponential distribution with rate $S_i \lambda + n_I\gamma$:
+\begin{equation}
+T_{i, j, k} \sim \mathrm{Exponential}\left(S_i \lambda + n_I\gamma\right),
+\end{equation}
+where $k = 1, 2, \dots, C_{i, j}, C_{i, j} + 1$. Here, the $(C_{i, j} + 1)$-th event is the transition from the infectious stage from $j$ to $j+1$.
+This process is repeated until the individual fully recovers (i.e., goes through all infectious stages).
+Then, the time at which individual $i$ makes $k$-th contact is given by the sum of current time $t$, the length of latent period $E_i$, and the time between all events (either a contact or transistion between states) caused by this individual prior to the contact:
+\begin{equation}
+\begin{cases}
+t + E_i + \sum_{n=1}^k T_{i, j, n}, & \text{if } j = 1,\\
+t + E_i + \sum_{m=1}^{j-1} \sum_{n=1}^{C_{i, m} + 1} T_{i, m, n} + \sum_{n=1}^k T_{i, j, n}, & \text{if } j > 1,
+\end{cases}
+\end{equation}
 

### 2019-05-06 doc/network.tex
working on methods

@@
 This process is repeated until the individual fully recovers (i.e., goes through all infectious stages).
-Then, the time at which individual $i$ makes $k$-th contact is given by the sum of current time $t$, the length of latent period $E_i$, and the time between all events (either a contact or transistion between states) caused by this individual prior to the contact:
+Then, the (yet to be realized) generation time between individual $i$ and the $k$-th contactee (during infectious stage $j$) is equal the sum of the length of latent period $E_i$ and the time between all events (either a contact or transistion between states) caused by this individual prior to the contact:
 \begin{equation}
-\begin{cases}
-t + E_i + \sum_{n=1}^k T_{i, j, n}, & \text{if } j = 1,\\
-t + E_i + \sum_{m=1}^{j-1} \sum_{n=1}^{C_{i, m} + 1} T_{i, m, n} + \sum_{n=1}^k T_{i, j, n}, & \text{if } j > 1,
+G_{i,j,k} = \begin{cases}
+E_i + \sum_{n=1}^k T_{i, j, n}, & \text{if } j = 1,\\
+E_i + \sum_{m=1}^{j-1} \sum_{n=1}^{C_{i, m} + 1} T_{i, m, n} + \sum_{n=1}^k T_{i, j, n}, & \text{if } j > 1.
 \end{cases}
 \end{equation}
+Finally, the time at which this individual is contacted by individual $i$ is equal to the sum of current time $t$ and the generation time $G_{i,j,k}$.
+
+After repeating the contact process for all initially infected individuals, all contacts are put into a sorted queue.
+The first person in the queue becomes infected, and the current time is updated to infection time of this individual. 
+Any subsequent contacts made to this individual are removed from the queue because they will no longer be effective.
+We repeat the contact process for this newly infected individual.
+Then, new contacts are added to the sorted queue.
+Simulation continues until there are no more contacts are left in the queue.
 

### 2019-05-07 doc/network.tex
tweak
Meeting with Daniel

@@
 \swp{TODO: In Appendix, we present the same figure using Erlang-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}
+\jd{TODO: Somehting about Erlang/Gmma}
 
@@
 First, we begin by randomly selecting initially infected individuals; these individuals are assumed to be infected at $t = 0$.
-For each infected individual $i$, we randomly draw a random number $E_i$ from an Erlang distribution with mean $1/\sigma$ and shape $n_E$ to represent the length of their latent period.
-Given a total contact rate of $S_i \lambda$, where $S_i$ is the number of susceptible acquaintances and $\lambda$ is the per-pair contact rate, and a transition rate of $n_I \gamma$ between each infectious stages,
-the total number of contacts that an indiviudal $i$ in infectious stage $j$ makes before moving to infectious stage $j+1$ follows a geometric distribution with probability $n_I\gamma/(S_i \lambda + n_I\gamma)$:
-\begin{equation}
-C_{i, j} \sim \mathrm{Geometric}\left(\frac{n_I\gamma}{S_i \lambda + n_I\gamma}\right).
-\end{equation}
-For each contact, a contactee is uniformly sampled from the set of acquaintances of the individual $i$.
-Then, the time between two events follows an exponential distribution with rate $S_i \lambda + n_I\gamma$:
-\begin{equation}
-T_{i, j, k} \sim \mathrm{Exponential}\left(S_i \lambda + n_I\gamma\right),
-\end{equation}
-where $k = 1, 2, \dots, C_{i, j}, C_{i, j} + 1$. Here, the $(C_{i, j} + 1)$-th event is the transition from the infectious stage from $j$ to $j+1$.
-This process is repeated until the individual fully recovers (i.e., goes through all infectious stages).
-Then, the (yet to be realized) generation time between individual $i$ and the $k$-th contactee (during infectious stage $j$) is equal the sum of the length of latent period $E_i$ and the time between all events (either a contact or transistion between states) caused by this individual prior to the contact:
-\begin{equation}
-G_{i,j,k} = \begin{cases}
-E_i + \sum_{n=1}^k T_{i, j, n}, & \text{if } j = 1,\\
-E_i + \sum_{m=1}^{j-1} \sum_{n=1}^{C_{i, m} + 1} T_{i, m, n} + \sum_{n=1}^k T_{i, j, n}, & \text{if } j > 1.
-\end{cases}
-\end{equation}
-Finally, the time at which this individual is contacted by individual $i$ is equal to the sum of current time $t$ and the generation time $G_{i,j,k}$.
+For each infected individual $i$, we randomly draw the latent period $E_i$ from an Erlang distribution with mean $1/\sigma$ and shape $n_E$.
+We then construct the random infectious period and infectious contact times simultaneously as follows.
+For each of the $n_I$ stages of the Erlang infectious period, we draw the number of effective contacts from a geometric distribution with mean $S_i \lambda / n_I\gamma$, where $S_i$ is the number
+of susceptible acquaintances and $\lambda$ is the per-pair contact rate.
+We then choose the time between consecutive events (the chosen number of contacts, followed by exit from the given stage of infection) from an exponential distribution with rate $S_i \lambda + n_I\gamma$.
+For each contact, a contactee is uniformly sampled from the set of susceptible acquaintances of the individual $i$.
+The infectious period $I_i$ is the sum of all of these waiting times.
+
+[[Add flowchart?]]
 
@@
 Then, new contacts are added to the sorted queue.
-Simulation continues until there are no more contacts are left in the queue.
+The simulation continues until there are no more contacts are left in the queue.
 

### 2019-05-08 doc/network.tex
more work

@@
 Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
-These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, wallinga2007generation, roberts2007model,  park2019practical}.
+These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.
 
@@
 Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.
-There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals are based on the infectiousness of infected people,
-%% without regard to actual infection events,
-while \emph{observed} generation intervals refer to the time between actual infection events.
-Observed generation intervals in turn can be \emph{aggregated} across time, measured \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others, or measured \emph{backward in time} by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
+There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals measure the infectiousness of an infected individual,
+while \emph{realized} generation intervals refer to the time between actual infection events.
+Realized generation intervals in turn can be measured either \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others or \emph{backward} in time by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
 % When aggregated generation intervals are observed until a given time in an ongoing epidemic (often via contact tracing), we refer to these as \emph{censored} (or right-censored) intervals.
@@
 While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
-These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.
+These censored observations can be thought of as a weighted average of backward generation-interval distributions;
+like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.
 
-Realized generation intervals are also affected by spatial and network structure.
+Realized generation intervals are also affected by spatial structure.
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
 This means that contacts made late in infection are more likely to be ``masked'' due to infectious contacts that were made earlier.
-Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
+As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
+These observations allow us to make the following prediction: removing the right-censoring bias from the observed generation intervals will give us a distribution that contains information about the population structure; using this will allow us to correctly infer $\RR$ from $r$.
 
 In this study, we explore spatiotemporal variation in generation intervals measured through contact tracing.
-We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
-We discuss statistical approaches to estimating effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
-\swp{Come back and fix this}
+We extend previous frameworks to study how the censored generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
+We classify sources of spatial variation in generation intervals into three levels (egocentric, local, and global) and discuss how they affect the observed generation-interval distributions.
+Finally, we propose two methods for accounting for temporal bias and test our prediction using simulations.
 
@@
 \swp{We decided to ask David for opinion}
-\swp{I tried to rewrite the section to make things clearer; it seems to make this paragraph less boring and more relevant?}
-
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$.
@@
 The dilution term, $\delta_a$, models how contacts are distributed among susceptible acquaintances.
+
 Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \citep{trapman2016inferring}.
-Then, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
+In this case, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
+This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
+for brevity, we do not pursue these directions here.
 
@@
 \begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{K}_a(\tau) dA,
+\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
 \end{equation}
-and the population-level egocentric generation-interval distribution is:
+\citep{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
+The population-level egocentric generation-interval distribution is:
 \begin{equation}
@@
 \end{equation}
-The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; 
-for brevity we will often omit ``population-level''.
+The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
 Finally, the egocentric reproductive number can be estimated by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
@@
 \end{equation}
-As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution.
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
+this generation-interval-based argument provides a clear biological interpretation for the same result presented by \cite{trapman2016inferring}.
 % Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
@@
 Or maybe we don't need a star network at all: first panel of current figure can be done on any network, and second panel can be done on a homogeneous network with a shortish time window?}
+\swp{I think we were confused about this comment when we talked about it. I think random network is unnecessary?}
 \begin{figure}
@@
 
-Although the effect of the overall depletion of susceptible individuals is relatively well-understood \citep{champredon2015intrinsic}, estimating the local effects on observed generation intervals is expected to be a hard problem on realistic networks -- although there has been interesting work on special cases, particularly on hierarchical networks of households nested within communities (e.g.,  \cite{tomba2010some}).
-\jd{I wonder if any of Frank Ball's household work could be relevant.}
-\swp{Will come back.}
-
 \section{Inferring generation-interval distributions from a contact tracing data}
@@
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
-While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and estimated reproductive numbers: 
+While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
 both local and global depletion effects reduce number of infections that occur and shorten generation intervals.
@@
 
-Since the right-censoring effect is a sampling bias, it must be taken into account explicitly.
+Since the right-censoring effect is a sampling bias, we want to take it into account explicitly.
 In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
@@
 In a large homogeneously mixing population, and during the early phase an an epidemic, the effective distribution will be equivalent to the intrinsic distribution.
-\jd{In a network with limited degree, they differ due to egocentric effects; in a non-random network they differ due to local effects; beyond the early period, they differ due to global effects.}
 

### 2019-05-08 doc/network.tex
a bit more

@@
 
+To simulate epidemics on a realistic network, we use the `condensed matter physics' network from the Stanford Large Network Dataset Collection \citep{leskovec2016snap}.
+This graph describes co-authorship among anyone who submitted a paper to Condensed Matter category in the arXiv between January 1993 and April 2003 \citep{leskovec2007graph}.
+It consists of 23133 nodes and 93497 edges.
+\cite{trapman2016inferring} used the same network to simulate epidemics.
+
+\subsection{Measuring the exponential growth rate}
+
+\section{Appendix}
+
 \bibliography{network}

### 2019-05-09 doc/network.tex
tweak

@@
 There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals measure the infectiousness of an infected individual,
-while \emph{realized} generation intervals refer to the time between actual infection events.
-Realized generation intervals in turn can be measured either \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others or \emph{backward} in time by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
+while \emph{observed} generation intervals refer to the time between actual infection events.
+Observed generation intervals in turn can be measured either \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others or \emph{backward} in time by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
 % When aggregated generation intervals are observed until a given time in an ongoing epidemic (often via contact tracing), we refer to these as \emph{censored} (or right-censored) intervals.

### 2019-05-09 doc/network.tex
Trying to work…

@@
 \newcommand{\tlab}[1]{\label{tab:#1}}
+\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}
 
@@
 
-\begin{figure}[t]
+\begin{figure}[!pbth]
 \includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!pbth]
 \includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
@@
 
-\jd{This is confusing to me.
-Wouldn't the natural next step be a random network?
-Homogeneous won't always be shorter than egocentric, I guess.
-Or are you simulating for a long time?
-Would it be better to simulate for a shorter time, and add another layer?
-I was thinking homogeneous $\to$ random $\to$ structured?
-Doing homogeneous during the exponential period should give us the same as egocentric, right?
-Maybe we need two figures?
-Or maybe we don't need a star network at all: first panel of current figure can be done on any network, and second panel can be done on a homogeneous network with a shortish time window?}
-\swp{I think we were confused about this comment when we talked about it. I think random network is unnecessary?}
-\begin{figure}
+\jd{I want to keep thinking about network choice. Would a random network be useful? Are we careful enough about time windows (exponential phase vs (roughly) full single epidemic?}
+\begin{figure}[!pbth]
 \includegraphics[width=\textwidth]{../fig/local_effect.pdf}
@@
 
-\begin{figure}[ht]
+\begin{figure}[!pbth]
 \includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
@@
 \swp{TODO: In Appendix, we present the same figure using Erlang-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}
-\jd{TODO: Somehting about Erlang/Gmma}
 
@@
 The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\RR$) of an epidemic \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
-However, observed generation-interval distributions can vary depending on how and when it is measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
+However, observed generation-interval distributions can vary depending on how and when they are measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
 determining which distribution correctly links $r$ and $\RR$ can be challenging.
-Here, we analyze how observed generation intervals, measured through contact tracing, vary across time and space.
-Temporal variations represent observation bias, whereas spatial variations reflect the contact structure of a realized outbreak.
-Therefore, correcting for temporal variation, but not for spatial variation, in an observed generation-interval distribution provides the correct link between $r$ and $\RR$.
+Here, we analyze how observed generation intervals, measured through contact tracing, differ from intrinsic generation intervals.
+Changes due to temporal censoring reflect observation bias, whereas changes due to spatial or network structure reflect the dynamics of the outbreak.
+Thus correcting the observed distribution for temporal, but not spatial, effects provides the correct link between $r$ and $\RR$.
 
-Generation intervals can be measured either forward (by tracing the infectees of an infected individual) or backward (by tracing the infector of an infected individual).
-Previous studies have characterized changes in these distributions over time by looking at infections that are one generation apart from a single cohort \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
-Instead, we considered a slightly different scenario in which generation intervals are measured over multiple generations.
-This assumption better reflects real intervention strategies because contact tracing is often performed throughout an ongoing epidemic to prevent further infections. 
-In this case, the generation intervals are measured from all observed cases that occured before the last date on which contact tracing was performed.
-
-The observed generation intervals are subject to right-censoring -- it is not possible to trace individuals that have not been infected yet.
+Observed generation intervals are subject to right-censoring -- it is not possible to trace individuals that have not been infected yet.
+These right-censored distributions can be thought of as averages of ``backward'' generation intervals (measured by tracing the infectors of a cohort of infected individuals) \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
 During an ongoing outbreak, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-censoring.
-Near the end of an outbreak, the effect of right-censoring becomes negligible but the observed generation intervals are still shorter on average than the intrinsic generation intervals.
-The remaining bias can be explained by depletion of the susceptible population.
+Early in the outbreak, censored intervals are expected to match the early-outbreak backward intervals.
+Near the end of an outbreak, the effect of right-censoring becomes negligible but the observed generation intervals are still shorter on average than intrinsic generation intervals, because of depletion of the susceptible population.
 
@@
 The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
-While we use the standard SEIR model, which assumes exponentially distributed latent and infectious period, throughout this paper, we present the extended version of the model, which allows for Erlang-distributed latent and infectious periods (we present some results based on the extended model in Appendix):
+We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.
+This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model:
+\jd{Mention that our main-text models use exponential later instead.}
 \begin{equation}
@@
 We then construct the random infectious period and infectious contact times simultaneously as follows.
-For each of the $n_I$ stages of the Erlang infectious period, we draw the number of effective contacts from a geometric distribution with mean $S_i \lambda / n_I\gamma$, where $S_i$ is the number
+For each of the $n_I$ stages of the infectious period, we draw the number of effective contacts from a geometric distribution with mean $S_i \lambda / n_I\gamma$, where $S_i$ is the number
 of susceptible acquaintances and $\lambda$ is the per-pair contact rate.

### 2019-05-19 doc/network.tex
minor tweaks to wording

--- 2019-05-09 doc/network.tex
@@
 As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
-and the distribution of forward generation intervals will be shorter than the intrinsic distribution \citep{kenah2008generation, tomba2010some, champredon2015intrinsic}.
+and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution \citep{kenah2008generation, tomba2010some, champredon2015intrinsic}.
 Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
@@
 and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{champredon2015intrinsic, britton2019estimation}.
-When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{tomba2010some, champredon2015intrinsic}.
+When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{tomba2010some, champredon2015intrinsic}.
 
@@
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
-These observations allow us to make the following prediction: removing the right-censoring bias from the observed generation intervals will give us a distribution that contains information about the population structure; using this will allow us to correctly infer $\RR$ from $r$.
+These observations allow us to make the following prediction: removing the censoring bias from the observed generation intervals will give us a distribution that contains information about the population structure; using this distribution should allow us to correctly infer $\RR$ from $r$.
 

### 2019-05-19 doc/network.tex
little more tweak

@@
 An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
-Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on their intrinsic infectivity as well as their neighborhood.
+Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the infectiousness of a disesase as well as the contact structure around them.
 
@@
 Here, we assume that the individual properties are independent of risk of infection.
-In practice, it is unlikely that naively averaging across an empirical distribution will give us the correct population-level kernel, especially when an epidemic is driven by a few highly infectious individuals.
+In practice, it is unlikely that naively averaging across an empirical distribution will correspond to the correct population-level kernel, especially when an epidemic is driven by a few highly infectious individuals.
 
-There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
+There are two components to an infection kernel: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
 Assuming that a population mixes homogeneously, the basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- can be defined as:
@@
 \swp{We decided to ask David for opinion}
-On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$.
+On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections ($i_{s-\tau}(s)$) that are $\tau$ time units apart from any cohort infection time $s$ before $t$.
 Then, the censored generation-interval distribution is given by
 \begin{equation}\label{eq:obsg}
-c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(x) \int_0^t i(s-\sigma) S(s) ds d\sigma}
+\begin{aligned}
+c_t(\tau) 
+&= \frac{\int_0^t i_{s-\tau}(s) ds}{\int_0^t \int_0^t i_{s-\sigma}(s) ds d\sigma}= \frac{\int_0^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t \int_0^t i(s-\sigma) g(x) S(s) ds d\sigma}\\
+\end{aligned}
 \end{equation}
-Alternatively, the censored generation-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
+Substituting $b_s(\tau) = i(s-\tau) g(\tau)$ allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
 \begin{equation}
@@
 \end{equation}
-Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., when the epidemic is growing exponentially;
-however, this assumption quickly breaks down as the epidemic progresses (Figure~\ref{fig:censor}).
+This relationship shows that the backward generation-interval distribution is equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., when the epidemic is growing exponentially.
 
@@
 \caption{\textbf{Temporal variation in the mean observed generation interval.}
-A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters and the backward and the censored mean generation interval were calculated over the course of an epidemic.
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters.
+The backward and the censored mean generation interval were calculated over the course of an epidemic.
 The dotted horizontal line represents the mean intrinsic generation interval.
@@
 There are two reasons for this phenomenon.
-First, any infection events that occur after the contact tracing period canonot be observed due to right censoring and short generation intervals are more likely to be observed.
+First, any infection events that occur after the contact tracing period cannot be observed due to right censoring and short generation intervals are more likely to be observed.
 In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
@@
 \end{equation}
-% As backward generation-interval distributions are always shorter than the intrinsic generation-interval distribution during the exponential growth period, their weighted average (i.e., the censored intervals) will also be shorter.
+During this period, the censored generation-interval distribution has the same mean as the backward generation-interval distribution.
 Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized intervals will always be shorter \citep{champredon2015intrinsic}.
 As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
+Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
 

### 2019-05-20 doc/network.tex
tweak

@@
 The effective distribution incorporates spatial effects. 
-In a large homogeneously mixing population, and during the early phase an an epidemic, the effective distribution will be equivalent to the intrinsic distribution.
+In a large homogeneously mixing population, the effective distribution will be equivalent to the intrinsic distribution.
 
@@
 To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
-\cite{trapman2016inferring} estimated the true reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
+\cite{trapman2016inferring} defined the empirical reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
 however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.

### 2019-05-20 doc/network.tex
a little more

@@
 The generation-interval-based approach to estimating the reproductive number assumes that an epidemic grows exponentially \citep{wallinga2007generation, park2019practical}.
-In practice, heterogeneity in population structure can lead to subexponential growth \citep{chowell2016characterizing, viboud2016generalized}.
+In practice, heterogeneity in population structure can lead to subexponential growth \citep{szendroi2004polynomial, chowell2015western, chowell2016growing, chowell2016characterizing, kiskowski2016modeling, viboud2016generalized};
+we therefore expect our simulations on an empirical network to be better characterized by subexponential growth models \citep{viboud2016generalized}.
+However, our simulations suggest that the initial exponential growth assumption still provides a viable approach for estimating the reproductive number.
+This conclusion is supported by \cite{chowell2015western} who found that epidemics can exhibit exponential growth patterns globally even if local epidemics may not.
 

### 2019-05-21 R/estimate.R
tweaks to the code

@@ -44,9 +44,9 @@ minuslogl.gamma <- function(log.mean, log.shape, gen) {
 	-sum(dgamma(gen, shape=shape, rate=rate, log=TRUE))
 }
 
-populationfun <- function(mean, shape, gen, r) {
+populationfun <- function(gen, r) {
 	m <- mle2(minuslogl.gamma, 
-			  start=list(log.mean=log(mean), log.shape=log(shape)),
+			  start=list(log.mean=log(mean(gen)), log.shape=log(mean(gen)^2/var(gen))),
 			  method="Nelder-Mead",
 			  optimizer="optim",
 			  data=list(gen=gen),

### 2019-05-21 doc/network.tex
DC comments on manuscript

@@
 Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
-However, directly measuring the reproductive number often requires detailed knowledge of the natural history of a disease and may not be feasible early in an epidemic \citep{dietz1993estimation}.
+However, directly measuring the reproductive number often requires detailed knowledge of the natural history of a disease 
+\dc{and also knowing the transmission chains; but that means doing some sort of contact tracing...}
+ and may not be feasible early in an epidemic \citep{dietz1993estimation}.
 Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
@@
 
+\dc{you make several assertions that are difficult to understand that early in the text.
+It really doesn't flow well, it was hard to follow, evn fo me who has worked on this kind of things.
+I suggest you define first c_t.}
+
 When generation intervals are measured through contact tracing during an outbreak, 
@@
 in fact, the censored generation-interval distribution averages over backward generation-interval distributions up until the observation time.
+\dc{You define $c_t$ after saying this, this is confusing.}
 During the exponential growth phase, we expect the backward distributions to be the same (and thus the censored distribution should match them).
+\dc{again, this is explained after...
+reading doesn't flow well}
 This connection will break down as the epidemic progresses, however, so we will to distinguish the censored distribution from the backward distribution.
@@
 \swp{We decided to ask David for opinion}
+\dc{I'm not sure which paragraph you are mentionning. If it's the one preceding, well, see my comment above. If it's the one following, yes, keep it. 
+More generally, I would keep as much definitions as it is necessary for the average reader to understand.
+Even if many thing said up to now are in other papers (that you cite), it really helps to have everything re-defined and interpreted in this context.
+This is relatively deep stuff, so anything that helps is welcome, at the risk of being slightly repetitive with previous work.
+}
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections ($i_{s-\tau}(s)$) that are $\tau$ time units apart from any cohort infection time $s$ before $t$.
@@
 the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrinsic generation-interval distribution \citep{britton2019estimation}:
+\dc{notation c_{exp} is not really consistent with Equation (9) c_t}
 \begin{equation}
@@
 \end{equation}
+
+\dc{I'm surely missing something obvious, but how do you justify this equation from the one above? }
+
 In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate: infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again).
@@
 Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
+\dc{Would it be possible to define local depletion using a more rigorous network topology, rather than households/neighborhood?
+(Would it help?).
+The same way the egocentric is defined with a star-shaped network, can we define the local depletion with a (well-known?) shape?
+Global depletion may be fine as it is, as this is something we are used to in general (in other modelling frameworks).
+}
 
@@
 We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR g(t - t_j)$, where $g(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
+\dc{is effective GI distribution, $g$, the same as the intrinsic? The notation implies it is... but I'm confused now.}
 We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
@@
 
+\dc{that reminds me of Cauchemez's 2006 AJE paper: ``Estimating in Real Time the Efficacy of Measures to Control Emerging Communicable''. Not sure if it's really relevant, but just in case...  }
+
 To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
@@
 estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
+\dc{Figure?
+I guess this is work in progress...}
 Direct estimates based on contact tracing data severely underestimates the empirical estimates.

### 2019-05-21 doc/network.tex
need to commit before I merge

--- 2019-05-20 doc/network.tex
@@
 
+We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}: 
+\begin{equation}
+c(t) = \frac{K}{1 + \[(K/c_0) - 1\] e^{-rt}},
+\end{equation}
+where $K$ is the final size of an epidemic and $c_0$ is the initial number of cases.
+Fitting this curve directly to cumulative incidence can lead to overly confident result \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution.
+This assumption requires us to estimate the dispersion parameter $\theta$ of a negative binomial distribution.
+Parameters $r$, $K$, $c_0$, and $\theta$ are estimated simultaneously by maximizing the likelihood.
+
 \section{Appendix}

### 2019-05-21 doc/network.tex
Push before meeting

@@
 Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
-However, directly measuring the reproductive number often requires detailed knowledge of the natural history of a disease 
-\dc{and also knowing the transmission chains; but that means doing some sort of contact tracing...}
- and may not be feasible early in an epidemic \citep{dietz1993estimation}.
+\jd{Took out a comment about tracing and tried to clarify.}
+However, estimating the reproductive number directly from disease natural history requires detailed knowledge which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
 Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
@@
 
+\jd{Still too much forward stuff for me; we don't use it for anything.}
 Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
@@
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
-These observations allow us to make the following prediction: removing the censoring bias from the observed generation intervals will give us a distribution that contains information about the population structure; using this distribution should allow us to correctly infer $\RR$ from $r$.
+These observations allow us to make the following prediction: removing the censoring bias from observed generation intervals will give us a distribution that contains information about the population structure; using this distribution should allow us to correctly infer $\RR$ from $r$.
 
@@
 We extend previous frameworks to study how the censored generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
-We classify sources of spatial variation in generation intervals into three levels (egocentric, local, and global) and discuss how they affect the observed generation-interval distributions.
+We classify sources of spatial variation in generation intervals into three levels (egocentric, local, and global) and discuss how they affect observed generation-interval distributions.
 Finally, we propose two methods for accounting for temporal bias and test our prediction using simulations.
@@
 Here, we assume that the individual properties are independent of risk of infection.
-In practice, it is unlikely that naively averaging across an empirical distribution will correspond to the correct population-level kernel, especially when an epidemic is driven by a few highly infectious individuals.
+In practice, it is unlikely that naively averaging across an empirical distribution will correspond to the correct population-level kernel at a particular time, especially when an epidemic is driven by a few highly infectious individuals.
+\jd{I'd rather drop that clause here; we can get back to it in Discussion if necessary.}
 
 There are two components to an infection kernel: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
-Assuming that a population mixes homogeneously, the basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- can be defined as:
+Assuming that a population mixes homogeneously, we can write: 
 \begin{equation}
-\RR_0 = \int K(\tau) d\tau.
+K(\tau) = \RR_0 g(\tau),
 \end{equation}
-Finally, the intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
-\begin{equation}
-g(\tau) = \frac{K(\tau)}{\RR_0}.
-\end{equation}
+where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population), 
+and $g(\tau)$ is the intrinsic generation-interval distribution.
 
-In a homogeneously mixing population, we can assume that the disease incidence $i(t)$ is the product of the current infectiousness of previously infected individuals and the proportion of the population susceptible, $S(t)$.
+In a homogeneously mixing population, disease incidence $i(t)$ is the product of the current infectiousness of infected individuals and the proportion of the population susceptible, $S(t)$.
 \begin{equation}
@@
 
-\dc{you make several assertions that are difficult to understand that early in the text.
+%% \dc{you make several assertions that are difficult to understand that early in the text.
 It really doesn't flow well, it was hard to follow, evn fo me who has worked on this kind of things.
 I suggest you define first c_t.}
+%% \jd{I think that verbal explanations first are better for a larger fraction of people.
+Maybe we can just make the verbal explanations better.
+I agree that we should keep the horrible math if DC likes it.}
 
-When generation intervals are measured through contact tracing during an outbreak, 
-the observed intervals are ``right-censored'', because intervals that end after the observation time are not observed.
-Several studies have associated the backward generation-interval distribution with the early-epidemic observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation};
-in fact, the censored generation-interval distribution averages over backward generation-interval distributions up until the observation time.
+When generation intervals are estimated through contact tracing during an outbreak, infection events that have not happened yet are not observed. 
+This effect is called ``right-censoring''.
+We can understand the effect of right-censoring using backward generation-interval distributions \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}:
+the observed (censored) generation-interval is a weighted distribution averages over backward generation-interval distributions up until the observation time.
 \dc{You define $c_t$ after saying this, this is confusing.}
@@
 the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrinsic generation-interval distribution \citep{britton2019estimation}:
-\dc{notation c_{exp} is not really consistent with Equation (9) c_t}
+\dc{notation $c_{exp}$ is not really consistent with Equation (9) $c_t$}
+\jd{Use math formatting in comments or it can mess up the whole doc.}
 \begin{equation}
@@
 
-The population-level egocentric kernel is given by integrating over individual variations:
+The population-level egocentric kernel is found by integrating over individual variations:
 \begin{equation}\label{eq:ego}
@@
 \begin{equation}
-c(t) = \frac{K}{1 + \[(K/c_0) - 1\] e^{-rt}},
+c(t) = \frac{K}{1 + \left[(K/c_0) - 1\right] e^{-rt}},
 \end{equation}

### 2019-05-21 doc/network.tex
Very much in the middle …

@@
 We can understand the effect of right-censoring using backward generation-interval distributions \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}:
-the observed (censored) generation-interval is a weighted distribution averages over backward generation-interval distributions up until the observation time.
-\dc{You define $c_t$ after saying this, this is confusing.}
-During the exponential growth phase, we expect the backward distributions to be the same (and thus the censored distribution should match them).
+the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions up until the observation time.
+
+%% \dc{You define $c_t$ after saying this, this is confusing.}
+%% During the exponential growth phase, we expect the backward distributions to be the same (and thus the censored distribution should match them).
 \dc{again, this is explained after...
 reading doesn't flow well}
-This connection will break down as the epidemic progresses, however, so we will to distinguish the censored distribution from the backward distribution.
+%% This connection will break down as the epidemic progresses, however, so we will to distinguish the censored distribution from the backward distribution.
 
@@
 \end{equation}
-
-\jd{Can we just skip this \P, or put in an appendix or something?}
-\swp{We decided to ask David for opinion}
-\dc{I'm not sure which paragraph you are mentionning. If it's the one preceding, well, see my comment above. If it's the one following, yes, keep it. 
-More generally, I would keep as much definitions as it is necessary for the average reader to understand.
-Even if many thing said up to now are in other papers (that you cite), it really helps to have everything re-defined and interpreted in this context.
-This is relatively deep stuff, so anything that helps is welcome, at the risk of being slightly repetitive with previous work.
-}
 On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections ($i_{s-\tau}(s)$) that are $\tau$ time units apart from any cohort infection time $s$ before $t$.

### 2019-05-21 doc/network.tex
going over comments

--- 2019-05-21 doc/network.tex
@@
 \jd{Still too much forward stuff for me; we don't use it for anything.}
-Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
-As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
-and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution \citep{kenah2008generation, tomba2010some, champredon2015intrinsic}.
-Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on.
+\swp{Got rid of forward stuff with some justification (?)}
+% Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
+% As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
+% and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.
+Unlike the intrinsic generation-interval distribution, the observed generation-interval distributions depend on when and how they are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+Here, we describe how backward generation-interval distributions change over time (see \cite{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic} for discussion on forward generation intervals); backward measurements better reflect real outbreak scenarios where contact tracing is performed by identifying infectors of infected individuals.
+When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
 A susceptible individual is thus relatively more likely to be infected by a newly infected individual, 
-and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{champredon2015intrinsic, britton2019estimation}.
-When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{tomba2010some, champredon2015intrinsic}.
+and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
+When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and backward generation intervals will be long \citep{nishiura2010time, tomba2010some, champredon2015intrinsic}.
 
 In practice, generation intervals are often measured by aggregating available information from contact tracing. 
-Whereas forward and backward generation intervals represent infections that are one generation apart from a cohort, these aggregated generation intervals represent infections across multiple cohorts; 
+Whereas backward (and forward) generation intervals represent infections that are one generation apart from a cohort, these aggregated generation intervals represent infections across multiple cohorts;
 the time of infection of a cohort can range anywhere from the beginning of an epidemic to the most recent infection.
@@
 We classify sources of spatial variation in generation intervals into three levels (egocentric, local, and global) and discuss how they affect observed generation-interval distributions.
-Finally, we propose two methods for accounting for temporal bias and test our prediction using simulations.
+Finally, we comapre two methods for accounting for temporal bias and test our prediction using simulations.
 

### 2019-05-22 doc/network.tex
Belated (from meeting)
trying to merge

@@
 
-The population-level kernel is given by integrating over these individual variations:
+Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
 \begin{equation}
@@
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
-Here, we assume that the individual properties are independent of risk of infection.
-In practice, it is unlikely that naively averaging across an empirical distribution will correspond to the correct population-level kernel at a particular time, especially when an epidemic is driven by a few highly infectious individuals.
+% In practice, it is unlikely that naively averaging across an empirical distribution will correspond to the correct population-level kernel at a particular time, especially when an epidemic is driven by a few highly infectious individuals.
 \jd{I'd rather drop that clause here; we can get back to it in Discussion if necessary.}
+\swp{Dropped.}
 
@@
 where $\RR = S(t) \RR_0$ is the effective reproductive number.
-This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
+This model, also referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
 Over a period of time where the proportion of susceptible $S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:

### 2019-05-28 doc/network.tex
edits

@@
 However, estimating the reproductive number directly from disease natural history requires detailed knowledge which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
-Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can often be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
+Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
 These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.
@@
 \jd{Still too much forward stuff for me; we don't use it for anything.}
-\swp{Got rid of forward stuff with some justification (?)}
+\swp{Got rid of forward stuff and tried to clarify}
 % Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
@@
 % and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.
-Unlike the intrinsic generation-interval distribution, the observed generation-interval distributions depend on when and how they are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
-Here, we describe how backward generation-interval distributions change over time (see \cite{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic} for discussion on forward generation intervals); backward measurements better reflect real outbreak scenarios where contact tracing is performed by identifying infectors of infected individuals.
+Unlike the intrinsic generation-interval distribution, the shape of the observed generation-interval distributions depends on when and how they are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+Here, we briefly describe how backward generation-interval distributions change over time because backward measurements better reflect real outbreak scenarios where contact tracing is performed by identifying infectors of infected individuals.
+(see \cite{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic} for discussions on forward generation intervals).
 When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
@@
 In practice, generation intervals are often measured by aggregating available information from contact tracing. 
-Whereas backward (and forward) generation intervals represent infections that are one generation apart from a cohort, these aggregated generation intervals represent infections across multiple cohorts;
+Whereas backward generation intervals represent infections that are one generation apart from a cohort, these aggregated generation intervals represent infections across multiple cohorts; 
 the time of infection of a cohort can range anywhere from the beginning of an epidemic to the most recent infection.
@@
 An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
-Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the infectiousness of a disesase as well as the contact structure around them.
+Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the intrinsic infectiousness of a disesase as well as the contact structure around an infected individual.
 
@@
 
-There are two components to an infection kernel: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
+There are two components to an infection kernel: intrinsic infectiousness of an average infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
 Assuming that a population mixes homogeneously, we can write: 
@@
 \end{equation}
-On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections ($i_{s-\tau}(s)$) that are $\tau$ time units apart from any cohort infection time $s$ before $t$.
+On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$:
 Then, the censored generation-interval distribution is given by
@@
 
-For a single epidemic, the observed mean generation interval through contact tracing will always be shorter than intrinsic mean generation interval (Figure~\ref{fig:censor}).
+For a single outbreak, the observed mean generation interval through contact tracing will always be shorter than intrinsic mean generation interval (Figure~\ref{fig:censor}).
 There are two reasons for this phenomenon.
@@
 In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
-the censored generation-interval distribution is equivalent to the inverse exponentially weighted intrinsic generation-interval distribution \citep{britton2019estimation}:
+the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the inverse exponentially weighted intrinsic generation-interval distribution \citep{britton2019estimation}:
 \dc{notation $c_{exp}$ is not really consistent with Equation (9) $c_t$}
 \jd{Use math formatting in comments or it can mess up the whole doc.}
+\swp{Should I just write it as $c_t$? I can't come up with a better notation.}
 \begin{equation}
-\tsub{c}{exp}(\tau) \propto g(\tau) \exp(-r\tau).
+c_t(\tau) = b_t(\tau) \propto g(\tau) \exp(-r\tau).
 \label{eq:exp}
 \end{equation}
-During this period, the censored generation-interval distribution has the same mean as the backward generation-interval distribution.
+A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
 Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized intervals will always be shorter \citep{champredon2015intrinsic}.
@@
 Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
+Overall, we expect naively using the observed generation-interval distribution to underestimate the reproductive number.
 
@@
 \end{equation}
-\citep{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
+\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
 The population-level egocentric generation-interval distribution is:
@@
 where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
-Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
+Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution (see Methods for derivation):
 \begin{equation}
@@
 \end{equation}
-
+In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
 \dc{I'm surely missing something obvious, but how do you justify this equation from the one above? }
-
-In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate: infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again).
+\swp{Is this clearer?}
 
@@
 Egocentric depletion, as discussed previously, is caused by an infected individual making multiple contacts to the same individual.
-Local depletion refers to a depletion of susceptible individuals in a household or neighborhood;
-this effect can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
+Local depletion refers to a depletion of susceptible individuals in a household or neighborhood.
+Both the egocentric depletion and the local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
 Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
@@
 }
+\swp{I'm not sure if that would help.
+We just want to get the idea across that there's more than egocentric depletion during the early in the epidemic.
+We don't actually want to do theory based on local stuff.
+We want to get the idea across that its information is hidden in the GI distributions and we don't have to worry about the technical details.
+I tried to clarify the text a little more.}
 
 \section{Inferring generation-interval distributions from a contact tracing data}
+
+\swp{Should we move some of the explanation in this section (pop v.~individual ) to Methods?}
 
@@
 We therefore expect that if we correct observed generation intervals for the right-censoring effect, spatial components will be taken into account implicitly. 
-We can use these temporally corrected intervals to correctly link $r$ and $\RR$.
-Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution. 
-The effective distribution incorporates spatial effects. 
+When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$.
+Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution $\tsub{g}{eff}(t)$.
 In a large homogeneously mixing population, the effective distribution will be equivalent to the intrinsic distribution.

### 2019-05-29 doc/network.tex
trying to move things around...

@@
 
-Here, we investigate two methods for correcting for temporal bias in a contact tracing data.
+Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
 We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
-As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
-Hence, the population-level method relies on the observed generation intervals and the exponential growth rate \citep{tomba2010some, nishiura2010time, britton2019estimation}:
+When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval inverse exponentially weighted \eref{exp}; hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
 \begin{equation}
-g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau).
+\tsup{g}{exp}(\tau) \propto c_t(\tau) \exp(r\tau),
 \end{equation}
-When the susceptible dynamics is known, this equation can be generalized by using \eref{obsg}; we do not pursue this direction here.
-The population-level method is simple and intuitive, but does not take into account who infected whom.
-
+where $r$ is the exponential growth rate.
 We refer to the second method as the individual-level method because it relies on individual contact information.
-We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR g(t - t_j)$, where $g(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
-\dc{is effective GI distribution, $g$, the same as the intrinsic? The notation implies it is... but I'm confused now.}
-We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
-Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
-\begin{equation}
-\frac{\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta)}{n_j!},
-\end{equation}
-where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
-
-The entire data likelihood of the individual-level method can be written as:
-\begin{equation}
-\begin{aligned}
-&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
-&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
-\end{aligned}
-\end{equation}
-where $N$ is the total number of infected individuals.
-Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
-Since the estimate of $\RR$ is sensitive to under-reporting, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate.
-\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
-
-\dc{that reminds me of Cauchemez's 2006 AJE paper: ``Estimating in Real Time the Efficacy of Measures to Control Emerging Communicable''. Not sure if it's really relevant, but just in case...  }
+We model each infection as a non-homogeneous Poisson process arising from the infector; 
+incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
 
 To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
-\cite{trapman2016inferring} defined the empirical reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals;
+\cite{trapman2016inferring} defined the empirical reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals.
 however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
@@
 
-\subsection{Empirical network}
+\swp{TODO:}
+\subsection{Egocentric ...}
 
-To simulate epidemics on a realistic network, we use the `condensed matter physics' network from the Stanford Large Network Dataset Collection \citep{leskovec2016snap}.
-This graph describes co-authorship among anyone who submitted a paper to Condensed Matter category in the arXiv between January 1993 and April 2003 \citep{leskovec2007graph}.
-It consists of 23133 nodes and 93497 edges.
-\cite{trapman2016inferring} used the same network to simulate epidemics.
+\subsection{Population-level method}
+
+\subsection{Individual-level method}
+
+We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR g(t - t_j)$, where $g(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
+\dc{is effective GI distribution, $g$, the same as the intrinsic? The notation implies it is... but I'm confused now.}
+We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
+Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
+\begin{equation}
+\frac{\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta)}{n_j!},
+\end{equation}
+where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
+
+The entire data likelihood of the individual-level method can be written as:
+\begin{equation}
+\begin{aligned}
+&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
+&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
+\end{aligned}
+\end{equation}
+where $N$ is the total number of infected individuals.
+Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
+Since the estimate of $\RR$ is sensitive to under-reporting, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate.
+\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
+
+\dc{that reminds me of Cauchemez's 2006 AJE paper: ``Estimating in Real Time the Efficacy of Measures to Control Emerging Communicable''. Not sure if it's really relevant, but just in case...  }
 
@@
 
+\subsection{Empirical network}
+
+To simulate epidemics on a realistic network, we use the `condensed matter physics' network from the Stanford Large Network Dataset Collection \citep{leskovec2016snap}.
+This graph describes co-authorship among anyone who submitted a paper to Condensed Matter category in the arXiv between January 1993 and April 2003 \citep{leskovec2007graph}.
+It consists of 23133 nodes and 93497 edges.
+\cite{trapman2016inferring} used the same network to simulate epidemics.
+
 \section{Appendix}

### 2019-05-29 doc/network.tex
sentence
simplify text

@@
 incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
+For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$.
 
 To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
-\cite{trapman2016inferring} defined the empirical reproductive number by dividing the total number of infected individuals in generations $2$ to $k+1$ by the total number of infected individuals in generations $1$ to $k$, where the reference generation $k$ is defined to be the first generation that contains at least 75 infected individuals.
-however, individuals in later generations are potentially subject to stronger local depletion and even global depletion effects.
-Therefore, we expect our empirical estimate to better reflect the infection process during the exponential growth phase.
-\swp{TODO: Compare Trapman vs our estimate of empirical R0 in Appendix; Trapman estimate is much lower.}
-
 As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
@@
 I guess this is work in progress...}
-Direct estimates based on contact tracing data severely underestimates the empirical estimates.
-While both population- and individual-level corrections provide similar estimates to the empirical reproductive number,
+Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
+While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
 population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.

### 2019-05-30 doc/network.tex
need to think about conclusion more...

@@
 estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
-\dc{Figure?
-I guess this is work in progress...}
 Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
@@
 We compare two methods for estimating the effective generation-interval distribution and assume that the effective generation-interval distribution follows a gamma distribution.
-The gamma approximation of the generation interval distribution has widely used due to its simplicity;
-we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$.
-However, further investigation of our methods suggests that making a wrong distributional assumption leads to overly narrow confidence intervals, even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model) \swp{Appendix}.
-These results are particularly alarming because it is impossible to know the true shape of the distribution [CITE].
+The gamma approximation of the generation interval distribution has widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
+we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$ \citep{park2019practical}.
+However, further investigation of our methods suggests that making a wrong distributional assumption leads to slightly narrow confidence intervals, even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model) \swp{Appendix}.
+These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
 
-The generation-interval-based approach to estimating the reproductive number assumes that an epidemic grows exponentially \citep{wallinga2007generation, park2019practical}.
+The generation-interval-based approaches to estimating the reproductive number often assumes that an epidemic grows exponentially \citep{wearing2005appropriate, wallinga2007generation, roberts2007model, park2019practical}.
 In practice, heterogeneity in population structure can lead to subexponential growth \citep{szendroi2004polynomial, chowell2015western, chowell2016growing, chowell2016characterizing, kiskowski2016modeling, viboud2016generalized};
@@
 \jd{Let me try.}
+\swp{How about this?}
+Contact tracing provides an effective way of controlling the disease and gathering epidemiological information early in an outbreak.
+Out study demonstrates that generation intervals measured through contact tracing capture key characteristics (e.g., contact structure) an outbreak; 
+these implicit features allow us to infer the ``correct'' generation-interval distribution, which accurately represents how disease spreads in a population. 
+Future studies should carefully consider how underlying epidemiological patterns can be inferred from contact tracing data.
+Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.
 

### 2019-05-30 doc/network.tex
minor tweaks

@@
 
-\swp{CITE}
 To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
@@
 We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.
-This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model:
+This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
 \jd{Mention that our main-text models use exponential later instead.}
@@
 where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
-Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, and $1/\gamma$ is the mean infectious period.
+Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
 
@@
 
-\swp{What do I cite here?}
 We develop an algorithm to simulate an individual-based SEIR model on a network to keep track of generation intervals.
-Our algorithm is based on the Gillespie algorithm.
+Our algorithm is based on the Gillespie algorithm \citep{gillespie1977exact}.
 We model individuals with nodes on a network and their acquaintanceships with edges; infected individuals can only contact their acquaintances.
@@
 
-[[Add flowchart?]]
+\jd{Add flowchart?}
 

### 2019-05-30 doc/network.tex
i think im almost done

@@
 The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
-Finally, the egocentric reproductive number can be estimated by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
+Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
 \begin{equation}
 \frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
+\label{eq:egorR}
 \end{equation}
@@
 where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
-Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution (see Methods for derivation):
+Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
 \begin{equation}
-\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)t} - e^{-\sigma t}\right)
+\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right)
 \end{equation}
 In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
+In practice, directly using this distribution to link $r$ and $\RR$ is unrealistic because it requires that we know the per-pair contact rate. 
+Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary meterial section 1.4.2);
+we briefly describe this relationship in Methods.
 \dc{I'm surely missing something obvious, but how do you justify this equation from the one above? }
@@
 \jd{I want to keep thinking about network choice. Would a random network be useful? Are we careful enough about time windows (exponential phase vs (roughly) full single epidemic?}
+\swp{I might have an idea... let me think about this a bit more...}
 \begin{figure}[!pbth]
@@
 Egocentric depletion, as discussed previously, is caused by an infected individual making multiple contacts to the same individual.
-Local depletion refers to a depletion of susceptible individuals in a household or neighborhood.
+Local depletion refers to a depletion of susceptible individuals in a household or neighborhood;
+we can think of these structures as small homogeneous networks embedded in a larger population structre (and therefore we can expect similar effects as we see in \fref{local}, right panel).
 Both the egocentric depletion and the local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
@@
 When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$.
-Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution $\tsub{g}{eff}(t)$.
+Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution $\tsub{g}{eff}(\tau)$.
 In a large homogeneously mixing population, the effective distribution will be equivalent to the intrinsic distribution.
@@
 \begin{equation}
-\tsup{g}{exp}(\tau) \propto c_t(\tau) \exp(r\tau),
+\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau),
 \end{equation}
@@
 
-\swp{TODO:}
-\subsection{Egocentric ...}
+\subsection{Egocentric relationship between $r$ and $\RR$ (SEIR model)}
 
-\subsection{Population-level method}
+Here, we rederive the egocentric relationship between $r$ and $\RR$ as presented by \cite{trapman2016inferring} (see original source for detailed derivations). 
+Assume that latent and infectious periods are exponentially distributed with mean $1/\sigma$ and $1/\gamma$, respectively.
+Assuming a constant per-pair contact rate of $\lambda$ for any pair, the egocentric generation-interval distribution can be written as follows:
+\begin{equation}
+\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right).
+\end{equation}
+Substituting this equation into \eref{egorR}, we get
+\begin{equation}
+\hat\RR = \left(1 + \frac{r}{\sigma}\right) \left(1 + \frac{r}{\gamma + \lambda}\right),
+\end{equation}
+where $r$ is the exponential growth rate.
+Alternatively, the reproductive number can be expressed based on the degree distribution (mean $\mu$ and variance $\sigma^2$) of a network:
+\begin{equation}
+\hat\RR = \underbrace{\kappa \times \lambda}_{\text{average contact rate}} \times \underbrace{\frac{1}{\gamma + \lambda}}_{\text{mean effective infectious period}},
+\end{equation}
+where $\kappa = \sigma^2/\mu + \mu - 1$, referred to as the mean degree excess \citep{newman2003structure}, describes expected number of susceptible individuals that an average infected individual will encounter early in an outbreak.
+Combining the two equations, we get
+\begin{equation}
+\lambda = \frac{(\gamma + r) (\sigma + r)}{(\kappa - 1) \sigma - r},
+\end{equation}
+which completes the egocentric relationship between $r$ and $\RR$:
+\begin{equation}
+\hat\RR = \frac{\gamma + r}{\gamma \sigma/(\sigma + r) + r/\kappa}.
+\end{equation}
 
-\subsection{Individual-level method}
+\subsection{Population-level method for estimating generation-interval distributions from contact tracing data}
 
-We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR g(t - t_j)$, where $g(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
-\dc{is effective GI distribution, $g$, the same as the intrinsic? The notation implies it is... but I'm confused now.}
+Population-level method tries to estimate the effective generation-interval distribution by reverting the inverse exponential weighting in the observed generation-interval distribution:
+\begin{equation}
+\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau).
+\end{equation}
+In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we esimate the mean and the shape parameter of a gamma distribution by maximum likelihood.
+\swp{Come back}
+
+\subsection{Individual-level method for estimating generation-interval distributions from contact tracing data}
+
+We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
 We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
@@
 \begin{equation}
-\frac{\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta)}{n_j!},
+\frac{\RR^{n_j} \exp \big(- \RR \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
 \end{equation}
-where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $G$).
+where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $\tsub{G}{eff}$).
 
@@
 &\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
-&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR G(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} g(s_{i, j}; \theta) \right),
+&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
 \end{aligned}
@@
 This assumption requires us to estimate the dispersion parameter $\theta$ of a negative binomial distribution.
-Parameters $r$, $K$, $c_0$, and $\theta$ are estimated simultaneously by maximizing the likelihood.
+Parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
 

### 2019-05-31 doc/network.tex
tweak

@@
 \end{equation}
-In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we esimate the mean and the shape parameter of a gamma distribution by maximum likelihood.
-\swp{Come back}
+In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we esimatemean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
+Then, the effective generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
 
@@
 
-The entire data likelihood of the individual-level method can be written as:
+The entire data likelihood of the nonhomogeneous poisson process can be written as:
 \begin{equation}
@@
 \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
-
-\dc{that reminds me of Cauchemez's 2006 AJE paper: ``Estimating in Real Time the Efficacy of Measures to Control Emerging Communicable''. Not sure if it's really relevant, but just in case...  }
 

### 2019-05-31 doc/network.tex
stuff

@@
 Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
+Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.
 }
@@
 \caption{\textbf{Temporal variation in the mean observed generation interval.}
-A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters.
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
 The backward and the censored mean generation interval were calculated over the course of an epidemic.
@@
 (Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
-All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles).
+All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
+mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
+Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
 }
@@
 both local and global depletion effects reduce number of infections that occur and shorten generation intervals.
-% SWP: I would rather put these in Discussion?
-% but the time scale on which they matter differs:
-% local depletion will be important early in an epidemic whereas global depletion will be small until sufficient amount of susceptible individuals become infected.
-% As a result, we expect the initial spread of a disease and the realized reproductive number to be heavily dependent on local structure of the contact network near the source of an outbreak.
 
@@
 Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
-Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters.
+Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$.
+Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear. \swp{Appendix}
 }

### 2019-06-04 doc/network.tex
near complete?

@@
 \jd{I want to keep thinking about network choice. Would a random network be useful? Are we careful enough about time windows (exponential phase vs (roughly) full single epidemic?}
-\swp{I might have an idea... let me think about this a bit more...}
+\swp{Clarified time windows.}
 \begin{figure}[!pbth]
@@
 Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
+Each simulation is run until all individuals are either susceptible or has recovered.
 }
@@
 
-\swp{Should we move some of the explanation in this section (pop v.~individual ) to Methods?}
-
 When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-Using the observed generation-interval distributions without correcting for right-censoring severely underestimates the reproductive number.
-Similarly, using intrinsic and egocentric generation-interval distributions without accounting for local spatial effects overestimates the reproductive number.
+Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
+Similarly, using intrinsic and egocentric generation-interval distributions overestimates the reproductive number because they fail to account for local spatial effects.
 Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
 Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$.
-Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear. \swp{Appendix}
+Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
 }
@@
 For smaller values of \RR, we expect the differences to become smaller.
-\swp{TODO: In Appendix, we present the same figure using Erlang-distributed latent periods, which better corresponds to Ebola. Overall, our qualitative conclusions do not change.}
+In Appendix, we present the same figure using smaller \RR (Appendix A1) and using Erlang-distributed latent periods (Appendix A2), which better corresponds to Ebola.
+Overall, our qualitative conclusions do not change.
 
@@
 Contact tracing provides an effective way of controlling the disease and gathering epidemiological information early in an outbreak.
-Out study demonstrates that generation intervals measured through contact tracing capture key characteristics (e.g., contact structure) an outbreak; 
-these implicit features allow us to infer the ``correct'' generation-interval distribution, which accurately represents how disease spreads in a population. 
+Out study demonstrates that generation intervals measured through contact tracing can capture key characteristics (e.g., contact structure) an outbreak; 
+these implicit features allow us to infer the ``correct'' generation-interval distribution, which accurately describes how disease spreads in a population. 
 Future studies should carefully consider how underlying epidemiological patterns can be inferred from contact tracing data.
@@
 \jd{Add flowchart?}
+\swp{Not sure what kind of flowchart you're thinking about.. let me think more...}
 
@@
 Parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
+Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.
+\swp{This is using epigrowthfit. How do I cite it? It's not out yet...}
 
@@
 
+\pagebreak
+\appendix
 \section{Appendix}
+
+\subsection{Comparison of estimates of reproductive number -- smaller $\RR$}
+\pagebreak
+
+\begin{figure}[!pbth]
+\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
+\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
+This figure matches \fref{cmp} in the main text.
+A smaller per-pair contact rate ($\lambda = 0.0026 \textrm{ days}^{-1}$) is used to simulate epidemics.
+All other parameters are the same as in \fref{cmp}.
+}
+\label{fig:cmpsmall}
+\end{figure}
+
+\subsection{Comparison of estimates of reproductive number -- Erlang distributed latent periods}
+
+\begin{figure}[!pbth]
+\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
+\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
+This figure matches \fref{cmp} in the main text.
+Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
+All other parameters are the same as in \fref{cmp}.
+}
+\label{fig:cmpseminir}
+\end{figure}
+
+\pagebreak
 

### 2019-06-04 analysis/cmp_compare.R

@@ -97,17 +97,6 @@ population.est <- lapply(1:100, function(x){
 	cg <- censor.gi[[x]]
 	gg <- growth[x,2]
 	
-	data.frame(
-		estimate=c(weighted.mean(cg, exp(gg*cg)), mean(exp(gg*cg))),
-		type=c("mean", "RR")
-	)
-}) %>%
-	bind_rows(.id="sim")
-
-population.est2 <- lapply(1:100, function(x){
-	cg <- censor.gi[[x]]
-	gg <- growth[x,2]
-	
 	populationfun(cg, gg)
 }) %>%
 	bind_rows(.id="sim")
@@ -134,7 +123,7 @@ empirical.est <- lapply(reslist, empirical.R0, 100)
 RRdata <- data.frame(
 	observed=observed.est$estimate[observed.est$type=="RR"],
 	individual=unlist(individual.RR),
-	population=population.est2$estimate[population.est2$type=="RR"],
+	population=population.est$estimate[population.est$type=="RR"],
 	intrinsic=unlist(intrinsic.est),
 	network=unlist(network.est),
 	trapman=unlist(trapman.est),

### 2019-06-04 analysis/cmp_compare_seminr.R

@@ -1,0 +1,138 @@
+library(igraph)
+library(dplyr)
+library(bbmle)
+library(epigrowthfit)
+source("../sim/ebola_param_seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+source("../R/empirical.R")
+
+load("../sim/cmp_seminr_sim.rda")
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+conditional_fun <- function(tau) {
+	sigma*(gamma+lambda)/(gamma+lambda-sigma) * (exp(-sigma*tau)-exp(-(gamma+lambda)*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.shape <- true.mean^2/true.var
+
+cmpGraph <- read.table("../data/CA-CondMat.txt")
+cmpGraph <- graph.data.frame(cmpGraph)
+cmpGraph <- as.undirected(cmpGraph)
+cmpGraph <- delete_vertex_attr(cmpGraph, "name")
+
+d <- degree(cmpGraph)
+
+kappa <- var(d)/mean(d) + mean(d) -1
+
+## This is SEIR
+# R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
+
+## This is SE2IR
+R0fun.network <- function(r) {
+	lambda <- (2 * sigma+r)^2 * (r + gamma)/((kappa-1) * 4 * sigma^2 - 4 * sigma * r - r^2)
+	kappa * lambda/(lambda + gamma)
+}
+
+censor.gi <- lapply(
+	reslist
+	, network.generation
+	, plot=FALSE
+	, interval=c(10, 1010)
+	, interval.type="cases"
+)
+
+count <- 0
+
+growth <- lapply(reslist, function(x){
+	count <<- count + 1
+	print(count)
+	day <- floor(x$data$time) + 1
+	
+	fitdata <- data.frame(
+		day=unique(day),
+		cases=as.vector(table(day))
+	)
+	
+	ee <- epigrowthfit(time=fitdata$day,
+				 deaths=fitdata$cases,
+				 model="logistic")
+	
+	dd <- data.frame(r=coef(ee)[1])
+	
+	dd
+}) %>%
+	bind_rows(.id="sim")
+
[…]

### 2019-06-04 analysis/cmp_compare_small.R

@@ -1,0 +1,141 @@
+library(igraph)
+library(dplyr)
+library(bbmle)
+library(epigrowthfit)
+source("../sim/ebola_param_seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+source("../R/empirical.R")
+
+load("../sim/cmp_seir_sim_2.rda")
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+conditional_fun <- function(tau) {
+	sigma*(gamma+lambda)/(gamma+lambda-sigma) * (exp(-sigma*tau)-exp(-(gamma+lambda)*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.shape <- true.mean^2/true.var
+
+cmpGraph <- read.table("../data/CA-CondMat.txt")
+cmpGraph <- graph.data.frame(cmpGraph)
+cmpGraph <- as.undirected(cmpGraph)
+cmpGraph <- delete_vertex_attr(cmpGraph, "name")
+
+d <- degree(cmpGraph)
+
+kappa <- var(d)/mean(d) + mean(d) -1
+
+beta <- 0.13/5
+
+## This is SEIR
+R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
+
+## This is SE2IR
+## R0fun.network <- function(r) {
+##	lambda <- (2 * sigma+r)^2 * (r + gamma)/((kappa-1) * 4 * sigma^2 - 4 * sigma * r - r^2)
+##	
+##	kappa * lambda/(lambda + gamma)
+## }
+
+censor.gi <- lapply(
+	reslist
+	, network.generation
+	, plot=FALSE
+	, interval=c(10, 1010)
+	, interval.type="cases"
+)
+
+count <- 0
+
+growth <- lapply(reslist, function(x){
+	count <<- count + 1
+	print(count)
+	day <- floor(x$data$time) + 1
+	
+	fitdata <- data.frame(
+		day=unique(day),
+		cases=as.vector(table(day))
+	)
+	
+	ee <- epigrowthfit(time=fitdata$day,
+					   deaths=fitdata$cases,
+					   model="logistic")
+	
+	dd <- data.frame(r=coef(ee)[1])
+	
+	dd
+}) %>%
+	bind_rows(.id="sim")
+
[…]

### 2019-06-04 analysis/seminr_growth.R

@@ -1,33 +1,0 @@
-library(epigrowthfit)
-
-load("../sim/seminr_sim.rda")
-
-weeks <- 3:5*3
-
-fitlist <- vector('list', length(weeks))
-
-for (j in 1:length(weeks)) {
-	
-	sublist <- vector('list', length(reslist))
-	
-	for (i in 1:length(reslist)) {
-		print(i)
-		sim <- reslist[[i]]
-		
-		week <- floor(sim$data$time/7) + 1
-		
-		fitdata <- data.frame(
-			week=1:15,
-			cases=as.vector(table(head(week, -1)))
-		)
-		
-		fitdata <- fitdata[1:weeks[j],]
-		
-		sublist[[i]] <- epigrowthfit(fitdata$week, fitdata$cases,
-									 model="logistic")
-	}
-	
-	fitlist[[j]] <- sublist
-}
-
-save(fitlist, file="seminr_growth.rda")

### 2019-06-04 analysis/seminr_nonparametric.R

@@ -1,55 +1,0 @@
-source("../sim/ebola_param_seminr.R")
-source("../R/generation.R")
-source("../R/estimate.R")
-
-load("../sim/seminr_sim.rda")
-
-load("seminr_growth.rda")
-
-weeks <- 3:5*3
-
-estlist <- vector('list', length(weeks))
-
-set.seed(101)
-for (j in 1:length(weeks)) {
-	sublist <- vector('list', length(reslist))
-	
-	for (i in 1:length(reslist)) {
-		print(i)
-		sim <- reslist[[i]]
-		fit <- fitlist[[j]][[i]]
-		
-		logr <- fit@mle2@coef[1]
-		logr.sd <- sqrt(diag(fit@mle2@vcov)[1])
-		
-		if (is.na(logr.sd) || logr.sd <= 0) {
-			logr.sd <- max((log(fit@growthRate[3]) - log(fit@growthRate[1]))/1.96,
-						   (log(fit@growthRate[1]) - log(fit@growthRate[2]))/1.96)	
-		} 
-		
-		gen <- network.generation(sim, interval=c(0, 7*weeks[j]), plot=FALSE)
-		
-		est <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_nonparametric)))
-		
-		est[,4] <- c(est[1,2] < true.mean && true.mean < est[1,3], 
-					 est[2,2] < true.R && true.R < est[2,3])
-		
-		colnames(est)[4] <- "coverage"
-		
-		est$type <- c("mean", "RR")
-		
-		est$sim <- i
-		
-		sublist[[i]] <- est
-	}
-	
-	subdf <- do.call("rbind", sublist)
-	
-	rownames(subdf) <- NULL
-	
-	subdf$weeks <- weeks[j]
-	
-	estlist[[j]] <- subdf
-}
-
-save("estlist", file="seminr_nonparametric.rda")

### 2019-06-04 analysis/seminr_observed.R

@@ -1,55 +1,0 @@
-source("../sim/ebola_param_seminr.R")
-source("../R/generation.R")
-source("../R/estimate.R")
-
-load("../sim/seminr_sim.rda")
-
-load("seminr_growth.rda")
-
-weeks <- 3:5*3
-
-estlist <- vector('list', length(weeks))
-
-set.seed(101)
-for (j in 1:length(weeks)) {
-	sublist <- vector('list', length(reslist))
-	
-	for (i in 1:length(reslist)) {
-		print(i)
-		sim <- reslist[[i]]
-		fit <- fitlist[[j]][[i]]
-		
-		logr <- fit@mle2@coef[1]
-		logr.sd <- sqrt(diag(fit@mle2@vcov)[1])
-		
-		if (is.na(logr.sd) || logr.sd <= 0) {
-			logr.sd <- max((log(fit@growthRate[3]) - log(fit@growthRate[1]))/1.96,
-						   (log(fit@growthRate[1]) - log(fit@growthRate[2]))/1.96)	
-		} 
-		
-		gen <- network.generation(sim, interval=c(0, 7*weeks[j]), plot=FALSE)
-		
-		est <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_observed)))
-		
-		est[,4] <- c(est[1,2] < true.mean && true.mean < est[1,3], 
-					 est[2,2] < true.R && true.R < est[2,3])
-		
-		colnames(est)[4] <- "coverage"
-		
-		est$type <- c("mean", "RR")
-		
-		est$sim <- i
-		
-		sublist[[i]] <- est
-	}
-	
-	subdf <- do.call("rbind", sublist)
-	
-	rownames(subdf) <- NULL
-	
-	subdf$weeks <- weeks[j]
-	
-	estlist[[j]] <- subdf
-}
-
-save("estlist", file="seminr_observed.rda")

### 2019-06-04 analysis/seminr_parametric.R

@@ -1,44 +1,0 @@
-library(bbmle)
-source("../sim/ebola_param_seminr.R")
-source("../R/generation.R")
-source("../R/estimate.R")
-
-load("../sim/seminr_sim.rda")
-
-weeks <- 3:5*3
-
-estlist <- vector('list', length(weeks))
-
-for (j in 1:length(weeks)) {
-	sublist <- vector('list', length(reslist))
-	
-	for (i in 1:length(reslist)) {
-		print(i)
-		sim <- reslist[[i]]
-		
-		gendata <- generation.data(sim, tmax=7*weeks[j])
-		
-		est <- as.data.frame(t(parametricfun(true.R, true.mean, true.shape, gendata, 7*weeks[j])))
-		
-		est[,4] <- c(est[1,2] < true.mean && true.mean < est[1,3], 
-					 est[2,2] < true.R && true.R < est[2,3])
-		
-		colnames(est)[4] <- "coverage"
-		
-		est$type <- c("mean", "RR")
-		
-		est$sim <- i
-		
-		sublist[[i]] <- est
-	}
-	
-	subdf <- do.call("rbind", sublist)
-	
-	rownames(subdf) <- NULL
-	
-	subdf$weeks <- weeks[j]
-	
-	estlist[[j]] <- subdf
-}
-
-save("estlist", file="seminr_parametric.rda")

### 2019-06-04 fig/cmp.R

@@ -8,49 +8,26 @@ source("../sim/ebola_param_seminr.R")
 scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
 
-genlist <- list()
-RRlist <- list()
-
 load("../analysis/cmp_compare.rda")
 
-genlist$big <- gendata
-RRlist$big <- RRdata
-
-load("../analysis/cmp_compare_2.rda")
-
-genlist$small <- gendata
-RRlist$small <- RRdata
-
 RRgroup <- data.frame(
-	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "local\ncorrection", "intrinsic"),
+	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "egocentric", "intrinsic"),
 	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
 				 levels=c("tracing based", "empirical", "individual based"))
 )
 
-RRdata2 <- RRlist %>%
-	bind_rows(.id="type") %>%
-	gather(key, value, -type) %>%
+RRdata2 <- RRdata %>%
+	gather(key, value) %>%
 	mutate(key=factor(key, 
 					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
 					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
-					  		 "tree\nbased", "empirical", "local\ncorrection", "intrinsic"))) %>%
-	merge(RRgroup) %>%
-	filter(key != "tree\nbased") %>%
-	mutate(type=factor(type, levels=c("small", "big"), labels=c("Smaller epidemic", "Larger epidemic")))
-
-gendata2 <- genlist %>%
-	bind_rows(.id="type") %>%
-	gather(key, value, -type) %>%
-	mutate(key=factor(key, 
-					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
-					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
[…]

### 2019-06-04 fig/cmp_seminr.R

@@ -1,0 +1,39 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("../sim/ebola_param_seminr.R")
+
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+load("../analysis/cmp_compare_seminr.rda")
+
+RRgroup <- data.frame(
+	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "egocentric", "intrinsic"),
+	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
+				 levels=c("tracing based", "empirical", "individual based"))
+)
+
+RRdata2 <- RRdata %>%
+	gather(key, value) %>%
+	mutate(key=factor(key, 
+					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
+					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
+					  		 "tree\nbased", "empirical", "egocentric", "intrinsic"))) %>%
+	merge(RRgroup) %>%
+	filter(key != "tree\nbased")
+
+ggR <- ggplot(RRdata2) +
+	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
+	scale_y_log10("Reproductive number", breaks=c(1, 2, 4, 8, 16)) +
+	facet_grid(~group, scale="free", space="free_x") +
+	theme(
+		panel.spacing.x = unit(0, "lines"),
+		strip.background = element_blank(),
+		legend.position = "none",
+		axis.title.x=element_blank()
+	)
+
+ggsave("cmp_reproductive_seminr.pdf", ggR, width=6, height=3)

### 2019-06-04 fig/cmp_small.R

@@ -1,0 +1,39 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("../sim/ebola_param_seminr.R")
+
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+load("../analysis/cmp_compare_small.rda")
+
+RRgroup <- data.frame(
+	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "egocentric", "intrinsic"),
+	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
+				 levels=c("tracing based", "empirical", "individual based"))
+)
+
+RRdata2 <- RRdata %>%
+	gather(key, value) %>%
+	mutate(key=factor(key, 
+					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
+					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
+					  		 "tree\nbased", "empirical", "egocentric", "intrinsic"))) %>%
+	merge(RRgroup) %>%
+	filter(key != "tree\nbased")
+
+ggR <- ggplot(RRdata2) +
+	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
+	scale_y_log10("Reproductive number", breaks=c(1, 2, 3, 4, 8, 16)) +
+	facet_grid(~group, scale="free", space="free_x") +
+	theme(
+		panel.spacing.x = unit(0, "lines"),
+		strip.background = element_blank(),
+		legend.position = "none",
+		axis.title.x=element_blank()
+	)
+
+ggsave("cmp_reproductive_small.pdf", ggR, width=6, height=3)

### 2019-06-04 fig/compare_methods.R

@@ -1,74 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-source("../sim/ebola_param_seminr.R")
-
-dir <- c("../analysis/")
-files <- c("seminr_observed.rda", "seminr_nonparametric.rda", "seminr_parametric.rda")
-
-comblist <- vector('list', 3)
-
-for (i in 1:length(files)) {
-	load(paste0(dir, files[i]))
-	
-	comblist[[i]] <- estlist
-}
-
-names(comblist) <- c("observed", "nonparametric", "parametric")
-
-combdata <- comblist %>%
-	lapply(bind_rows) %>%
-	bind_rows(.id="method") %>%
-	mutate(method=factor(method, levels=c("observed", "nonparametric", "parametric")),
-		   weeks=factor(weeks, levels=3:5*3),
-		   type=factor(type, labels=c("mean generation interval", "reproductive number")))
-
-coverdata <- combdata %>%
-	group_by(method, type, weeks) %>%
-	summarize(coverage=mean(coverage, na.rm=TRUE))
-
-estdata <- combdata %>%
-	group_by(method, type, weeks) %>%
-	summarize(
-		median=median(estimate),
-		lwr=quantile(estimate, 0.025),
-		upr=quantile(estimate, 0.975)
-	) %>%
-	filter(!(method=="parametric" & weeks==3))
-
-truedata <- data.frame(
-	type=c("mean generation interval", "reproductive number"),
-	value=c(true.mean, true.R)
-)
-
-g1 <- ggplot(estdata) +
-	geom_point(aes(weeks, median, col=method, shape=method), position=position_dodge(width=0.2)) +
-	geom_errorbar(aes(weeks, ymin=lwr, ymax=upr, col=method), width=0.2, position=position_dodge(width=0.2)) + 
-	geom_hline(data=truedata, aes(yintercept=value), lty=2) +
-	scale_y_continuous("estimate") +
-	facet_wrap(~type, scale="free_y") +
-	theme(
-		axis.title.x = element_blank(),
-		strip.background = element_blank(),
-		legend.position = "top",
-		legend.title = element_blank()
-	)
-
-g2 <- ggplot(coverdata) +
-	geom_point(aes(weeks, coverage, col=method, shape=method)) +
-	geom_path(aes(as.numeric(weeks), coverage, col=method)) +
[…]

### 2019-06-04 fig/example.R

@@ -1,152 +1,0 @@
-library(epigrowthfit)
-library(dplyr)
-library(tidyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-source("../sim/ebola_param_seminr.R")
-source("../R/seminr.R")
-source("../R/generation.R")
-source("../R/estimate.R")
-
-weekmax <- 28
-weekmin <- 9
-
-set.seed(101)
-rr <- seminr.full(N, beta/(N-1), sigma, m.sigma, gamma, n.gamma, I0 = 10, tmax=7*weekmax)
-
-weeks <- weekmin:weekmax
-
-reslist <- vector('list', length(weeks))
-
-for (i in 1:length(weeks)) {
-	print(i)
-	week <- weeks[i]
-	
-	obsdata <- rr$data %>%
-		filter(time < 7*week)
- 	
-	fitdata <- data.frame(
-		week=1:week,
-		cases=as.vector(table(floor(obsdata$time/7)))
-	)
-	
-	epifit <- epigrowthfit(fitdata$week, fitdata$cases,
-						   model="logistic")
-	
-	logr <- epifit@mle2@coef[1]
-	logr.sd <- sqrt(diag(epifit@mle2@vcov)[1])
-	
-	if (is.na(logr.sd) || logr.sd <= 0) {
-		logr.sd <- max((log(epifit@growthRate[3]) - log(epifit@growthRate[1]))/1.96,
-					   (log(epifit@growthRate[1]) - log(epifit@growthRate[2]))/1.96)	
-	} 
-	
-	gen <- network.generation(rr, plot=FALSE, interval=c(0, 7*week))
-	
-	gendata <- generation.data(rr, tmax=7*week)
-	
-	nonparaest <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_nonparametric)))
-	
-	nonparaest$type <- c("mean", "RR")
-	nonparaest$method <- "nonparametric"
-	
-	obsest <- as.data.frame(t(bootfun(gen, logr, logr.sd, estfun_observed)))
-	
-	obsest$type <- c("mean", "RR")
-	obsest$method <- "observed"
-	
-	paraest <- as.data.frame(t(parametricfun(true.R, true.mean, true.shape, gendata, 7*week)))
-	
-	paraest$type <- c("mean", "RR")
-	paraest$method <- "parametric"
-	
-	growthest <- data.frame(
-		estimate=epifit@growthRate[1],
-		lwr=epifit@growthRate[2],
[…]

### 2019-06-04 fig/looksmall.R

@@ -1,87 +1,0 @@
-library(dplyr)
-library(readr)
-
-## line list
-ll <- read_csv(input_files[[1]])
-
-## merge 
-ml <- (ll
-	%>% left_join(
-		ll
-		, by=c("ID"="IDSource")
-		, suffix=c("one", "the_other")
-	) %>% mutate(gen = Exposure_dateone-Exposure_datethe_other)
-)
-
-print(summary(ml))
-print((ml$gen))
-
-quit()
-
-## data
-dd <- data.frame(
-	time=sp2$exposure,
-	gen=sp2$exposure - sp2$exposure[match(sp2$IDSource, sp2$ID)],
-	serial=sp2$symptom - sp2$symptom[match(sp2$IDSource, sp2$ID)],
-	source=sp2$IDSource,
-	recipient=sp2$ID,
-	type=sp2$Type
-) %>%
-	filter(!is.na(gen), gen > 0) %>% ## there is one case where gen < 0 so I get rid of it
-	arrange(time)
-
-## All contacts (as opposed to case)
-dd_contact <- dd %>%
-	group_by(time) %>%
-	summarize(
-		sum=sum(gen),
-		weight=length(gen)
-	) %>%
-	mutate(
-		mean=cumsum(sum)/cumsum(weight)
-	)
-
-dd_case <- dd %>%
-	filter(type=="Case") %>%
-	group_by(time) %>%
-	summarize(
-		sum=sum(gen),
-		weight=length(gen)
-	) %>%
-	mutate(
-		mean=cumsum(sum)/cumsum(weight)
-	)
-
-dd_serial <- dd %>%
-	filter(type=="Case", !is.na(serial)) %>%
-	group_by(time) %>%
-	summarize(
-		sum=sum(serial),
-		weight=length(serial)
-	) %>%
-	mutate(
-		mean=cumsum(sum)/cumsum(weight)
-	)
-
-pdf("smallpox.pdf", width=10, height=3)
-par(mfrow=c(1, 3))
-plot(dd_contact$time, dd_contact$mean, type="l",
-	 xlab="Time (days since the first case)",
-	 ylab="Mean generation interval",
-	 ylim=c(13, 18.5))
-title("All contacts")
-abline(h=17.7, lty=2)
-
-plot(dd_case$time, dd_case$mean, type="l",
-	 xlab="Time (days since the first case)",
-	 ylab="Mean generation interval",
-	 ylim=c(13, 18.5))
[…]

### 2019-06-04 fig/mers.R

@@ -1,42 +1,0 @@
-library(dplyr)
-library(outbreaks)
-library(ggplot2); theme_set(theme_bw())
-
-ll <- mers_korea_2015$linelist
-cc <- mers_korea_2015$contacts
-
-toDate <- as.numeric(as.Date(ll$dt_onset[match(cc$to, ll$id)]))
-toDate <- toDate - 16569
-fromDate <- as.numeric(as.Date(ll$dt_onset[match(cc$from, ll$id)]))
-fromDate <- fromDate - 16569
-
-serial_data <- data.frame(
-	time=toDate,
-	gen=toDate - fromDate,
-	within=ll$loc_hosp[match(cc$to, ll$id)] == ll$loc_hosp[match(cc$from, ll$id)],
-	hospital=ll$loc_hosp[match(cc$to, ll$id)]
-)
-
-boxplot(gen~within, data=serial_data)
-t.test(gen~within, data=serial_data)
-
-ggplot(serial_data) +
-	geom_boxplot(aes(within, gen))
-
-dd <- data.frame(
-	time=ll$dt_onset[match(cc$to, ll$id)],
-	gen=toDate - fromDate
-) %>%
-	arrange(time) %>%
-	group_by(time) %>%
-	summarize(
-		sum=sum(gen),
-		weight=length(gen)
-	) %>%
-	mutate(mean=cumsum(sum)/cumsum(weight))
-
-plot(dd$time, dd$mean)
-
-plot(table(ll$dt_onset))
-
-

### 2019-06-04 fig/smallpox.R

@@ -1,80 +1,0 @@
-library(dplyr)
-
-sp <- read.csv("../data/Smallpox_1951_Tilburg_NL.csv")
-
-sp2 <- sp %>%
-	mutate(exposure=as.numeric(as.Date(Exposure_date)),
-		   symptom=as.numeric(as.Date(DOD))) %>%
-	mutate(exposure=exposure-min(exposure, na.rm=TRUE),
-		   symptom=symptom-min(symptom, na.rm=TRUE))
-
-dd <- data.frame(
-	time=sp2$exposure,
-	gen=sp2$exposure - sp2$exposure[match(sp2$IDSource, sp2$ID)],
-	serial=sp2$symptom - sp2$symptom[match(sp2$IDSource, sp2$ID)],
-	source=sp2$IDSource,
-	recipient=sp2$ID,
-	type=sp2$Type
-) %>%
-	filter(!is.na(gen), gen > 0) %>% ## there is one case where gen < 0 so I get rid of it
-	arrange(time)
-
-dd_contact <- dd %>%
-	group_by(time) %>%
-	summarize(
-		sum=sum(gen),
-		weight=length(gen)
-	) %>%
-	mutate(
-		mean=cumsum(sum)/cumsum(weight)
-	)
-
-dd_case <- dd %>%
-	filter(type=="Case") %>%
-	group_by(time) %>%
-	summarize(
-		sum=sum(gen),
-		weight=length(gen)
-	) %>%
-	mutate(
-		mean=cumsum(sum)/cumsum(weight)
-	)
-
-dd_serial <- dd %>%
-	filter(type=="Case", !is.na(serial)) %>%
-	group_by(time) %>%
-	summarize(
-		sum=sum(serial),
-		weight=length(serial)
-	) %>%
-	mutate(
-		mean=cumsum(sum)/cumsum(weight)
-	)
-
-pdf("smallpox.pdf", width=10, height=3)
-par(mfrow=c(1, 3))
-plot(dd_contact$time, dd_contact$mean, type="l",
-	 xlab="Time (days since the first case)",
-	 ylab="Mean generation interval",
-	 ylim=c(13, 18.5))
-title("All contacts")
-abline(h=17.7, lty=2)
-
-plot(dd_case$time, dd_case$mean, type="l",
-	 xlab="Time (days since the first case)",
-	 ylab="Mean generation interval",
-	 ylim=c(13, 18.5))
-title("Symptomatic cases")
-abline(h=17.7, lty=2)
-
[…]

### 2019-06-04 sim/cmp_seir_sim_2.R

@@ -1,0 +1,28 @@
+library(igraph)
+source("ebola_param_seminr.R")
+source("../R/seminr.R")
+
+cmpGraph <- read.table("../data/CA-CondMat.txt")
+cmpGraph <- graph.data.frame(cmpGraph)
+cmpGraph <- as.undirected(cmpGraph)
+cmpGraph <- delete_vertex_attr(cmpGraph, "name")
+
+nsim <- 100
+
+beta <- 0.13/5
+
+set.seed(101)
+reslist <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+	print(i)
+	rr <- seminr(cmpGraph, beta, sigma, 1, gamma, 1, I0 = 10, tmax=2000)
+	if(nrow(rr$data) > 1000) {
+		reslist[[i]] <- rr
+		i <- i+1
+	}
+	
+	save("reslist", file="cmp_seir_sim_2.rda")
+}
+
+save("reslist", file="cmp_seir_sim_2.rda")

### 2019-06-04 sim/seminr_sim.R

@@ -1,19 +1,0 @@
-source("../sim/ebola_param_seminr.R")
-source("../R/seminr.R")
-source("../R/generation.R")
-
-nsim <- 100
-
-set.seed(101)
-reslist <- vector('list', nsim)
-i <- 1
-while (i <= nsim) {
-    print(i)
-    rr <- seminr.full(N, beta/(N-1), sigma, m.sigma, gamma, n.gamma, I0 = 10, tmax=7*15)
-    if(nrow(rr$data) > 100) {
-        reslist[[i]] <- rr
-        i <- i+1
-    }
-}
-
-save("reslist", file="seminr_sim.rda")

### 2019-06-04 doc/network.tex
...

@@
 Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
-Since the estimate of $\RR$ is sensitive to under-reporting, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate.
+Since the estimate of $\RR$ is sensitive to under-reporting, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate (using the Euler-Lotka equation).
 \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

### 2019-06-05 doc/network.tex
one more line
tweak
stuff

@@
 For smaller values of \RR, we expect the differences to become smaller.
-In Appendix, we present the same figure using smaller \RR (Appendix A1) and using Erlang-distributed latent periods (Appendix A2), which better corresponds to Ebola.
+In Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
 Overall, our qualitative conclusions do not change.
@@
 We compare two methods for estimating the effective generation-interval distribution and assume that the effective generation-interval distribution follows a gamma distribution.
-The gamma approximation of the generation interval distribution has widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
+The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
 we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$ \citep{park2019practical}.
-However, further investigation of our methods suggests that making a wrong distributional assumption leads to slightly narrow confidence intervals, even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model) \swp{Appendix}.
+However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).
 
@@
 \jd{Mention that our main-text models use exponential later instead.}
+\swp{Done.}
 \begin{equation}
@@
 Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
+While we use the SEIR model for simulations in our main text, we show results based on more realistic distributions in Appendix.
 
@@
 \subsection{Comparison of estimates of reproductive number -- smaller $\RR$}
-\pagebreak
 
-\begin{figure}[!pbth]
-\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../fig/cmp_reproductive_small.pdf}
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
@@
 
+\pagebreak
+
 \subsection{Comparison of estimates of reproductive number -- Erlang distributed latent periods}
 
-\begin{figure}[!pbth]
-\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../fig/cmp_reproductive_seminr.pdf}
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
@@
 
+\subsection{Testing the individual-based method on simulations on a homogeneous network}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../fig/full_coverage_fig.pdf}
+\caption{\textbf{Wrong distributional assumption may result in biased estimates of the parameters of a generation-interval distribution.}
+We simulate a stochastic SEIR model on a homogenous network with $10^5$ individuals using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
+Then, we apply the individual-based method based on the first 1000 infections.
+(A-B) A boxplot of 100 estimates of the mean generation intervals and their coefficient of variations (CV). 
+Dashed horizontal lines represent the true value.
+(C) Coverage probability of the mean generation intervals and their CVs based on the 95\% confidence interval.
+Dashed horizontal lines represent the 95\% coverage probability.
+(D) Estimated generation-interval distributions based on the gamma approximation and a true generation-interval distribution of an SEIR model.
+(E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.
+Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
+}
+\label{fig:cover}
+\end{figure}
+
+\pagebreak
+
 \bibliography{network}

### 2019-06-05 analysis/full_coverage.R

@@ -1,0 +1,39 @@
+library(igraph)
+library(dplyr)
+library(bbmle)
+source("../sim/ebola_param_seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+source("../R/empirical.R")
+
+load("../sim/full_seir_sim.rda")
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.shape <- true.mean^2/true.var
+
+individual.est <- lapply(reslist, function(x){
+	tmax <- x$data[1001,1]
+	gendata <- generation.data(x, tmax=tmax)
+	
+	est <- as.data.frame(t(parametricfun(true.R, true.mean, true.shape, gendata, tmax)))
+	
+	est$type <- c("RR", "mean", "shape")
+	
+	est$coverage <- c(
+		est$lwr[1] < true.R && true.R < est$upr[1],
+		est$lwr[2] < true.mean && true.mean < est$upr[2],
+		est$lwr[3] < true.shape && true.shape < est$upr[3]
+	)
+	
+	print(est)
+	
+	est
+}) %>%
+	bind_rows(.id="sim")
+
+save("individual.est", file="full_coverage.rda")

### 2019-06-05 fig/full_coverage_fig.R

@@ -1,0 +1,128 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+source("../sim/ebola_param_seminr.R")
+
+load("../analysis/full_coverage.rda")
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.shape <- true.mean^2/true.var
+
+coverdata <- individual.est %>%
+	group_by(type) %>%
+	summarize(
+		coverage=mean(coverage)
+	) %>%
+	filter(
+		type!="RR"
+	) %>%
+	mutate(
+		type=factor(type, labels=c("Mean", "CV"))
+	)
+
+g1 <- ggplot(filter(individual.est, type=="mean")) +
+	geom_boxplot(aes(y=estimate)) +
+	geom_hline(yintercept=true.mean, lty=2) +
+	scale_y_continuous("Mean generation time (days)") +
+	ggtitle("A") +
+	theme(
+		axis.text.x = element_blank(),
+		axis.ticks.x = element_blank()
+	)
+
+g2 <- ggplot(filter(individual.est, type=="shape")) +
+	geom_boxplot(aes(y=sqrt(1/estimate))) +
+	geom_hline(yintercept=sqrt(1/true.shape), lty=2)  +
+	scale_y_continuous("Coefficient of variation (CV)")  +
+	ggtitle("B") +
+	theme(
+		axis.text.x = element_blank(),
+		axis.ticks.x = element_blank()
+	)
+
+g3 <- ggplot(coverdata) +
+	geom_point(aes(type, coverage), shape=1, size=3) +
+	geom_hline(yintercept=0.95, lty=2) +
+	scale_y_continuous("Coverage probability", limits=c(0, 1)) +
+	scale_x_discrete("Parameters")  +
+	ggtitle("C")
+
+xvec <- seq(0, 51, by=0.1)
+
+realdensity <- data.frame(
+	x=xvec,
+	density=intrinsic_fun(xvec)
+)
+
+estdensity <- lapply(1:100, function(x) {
+	ii <- filter(individual.est, sim==x)
+	
[…]

### 2019-06-05 sim/ebola_param_seminr.R

@@ -1,9 +1,9 @@
 set.seed(101)
 beta <- 2/5
 sigma <- 1/11.4
-m.sigma <- round(1.75)
+m.sigma <- 2
 gamma <- 1/5
-n.gamma <- round((5/4.7)^2)
+n.gamma <- 1
 N <- 100000
 
 true.mean <- 1/sigma + 1/gamma

### 2019-06-05 sim/full_seir_sim.R

@@ -1,0 +1,22 @@
+library(igraph)
+source("ebola_param_seminr.R")
+source("../R/seminr.R")
+
+nsim <- 100
+
+set.seed(101)
+reslist <- vector('list', nsim)
+i <- 1
+while (i <= nsim) {
+	print(i)
+	rr <- seminr.full(N, beta/N, sigma, 1, gamma, 1, I0 = 10, imax=1010, tmax=200,
+					  keep.intrinsic=TRUE)
+	if(nrow(rr$data) > 1000) {
+		reslist[[i]] <- rr
+		i <- i+1
+	}
+	
+	save("reslist", file="full_seir_sim.rda")
+}
+
+save("reslist", file="full_seir_sim.rda")

### 2019-06-11 doc/network.tex
DC's edits and comments

@@
 
-\usepackage{graphics}
-\usepackage{adjustbox}
+\usepackage{graphicx}
+%\usepackage{adjustbox}
 
@@
 
-An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
-Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
+An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength, characterized by its reproductive number, \RR.
+The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
 \jd{Took out a comment about tracing and tried to clarify.}
@@
 and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
-When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and backward generation intervals will be long \citep{nishiura2010time, tomba2010some, champredon2015intrinsic}.
+When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and backward generation intervals will be longer than the intrinsic one \citep{nishiura2010time, tomba2010some, champredon2015intrinsic}.
 
@@
 Whereas backward generation intervals represent infections that are one generation apart from a cohort, these aggregated generation intervals represent infections across multiple cohorts; 
-the time of infection of a cohort can range anywhere from the beginning of an epidemic to the most recent infection.
+the time of infection of the infectors can range anywhere from the beginning of an epidemic to the most recent infection.
+\dc{I found the previous paragraph difficult to read and edited it to try to improve.
+The `cohorts` were not referring to the same ones (I think) and that was confusing.
+The bckwrd GI is observed at one given time and those infectees each had infectors from potentially different cohorts.}
 While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
-These censored observations can be thought of as a weighted average of backward generation-interval distributions;
+These censored observations can be thought of as a weighted average of backward generation-interval distributions; \dc{weighted by what?}
 like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.
@@
 Realized generation intervals are also affected by spatial structure.
-In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic.
-This means that contacts made late in infection are more likely to be ``masked'' due to infectious contacts that were made earlier.
+In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population even in the early phase of an epidemic.
+This means that contacts made late in an individual's infection are more likely to be ``masked'' due to infectious contacts that were made earlier \dc{Although I get you try to explain in simple terms what's going on with ``masked'', I think it's not really clear: masked by what?
+is it that subsequent contacts in the neighbourhood do not count given they were infected earlier?}.
 As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
@@
 We classify sources of spatial variation in generation intervals into three levels (egocentric, local, and global) and discuss how they affect observed generation-interval distributions.
-Finally, we comapre two methods for accounting for temporal bias and test our prediction using simulations.
+Finally, we compare two methods for accounting for temporal bias and test our prediction using simulations.
 
@@
 Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
-Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.
-}
+Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.}
 \label{fig:indpop}
-
 \end{figure}
@@
 making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
-An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
-Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the intrinsic infectiousness of a disesase as well as the contact structure around an infected individual.
+An individual-level intrinsic infection kernel $K_a(\tau)$ \dc{is it clearer to specify the kernel is about infections?} describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
+Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the intrinsic infectiousness of a disease as well as the contact structure around an infected individual.
 
@@
 \begin{equation}
-K(\tau) = \int K_a (\tau) dA.
+K(\tau) = \int K_a (\tau) da.
 \end{equation}
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
-% In practice, it is unlikely that naively averaging across an empirical distribution will correspond to the correct population-level kernel at a particular time, especially when an epidemic is driven by a few highly infectious individuals.
-\jd{I'd rather drop that clause here; we can get back to it in Discussion if necessary.}
-\swp{Dropped.}
+\dc{wouldn't it be better to say: ``The population-level kernel describes the rate at which secondary infections are expected to be caused by an infected individual, on \emph{average}.'' I have troubles describing an ``average  infected individual''...
+Oh no, but you say it again below...
+well, I keep this comment, do whatever you want!
+}
 
-There are two components to an infection kernel: intrinsic infectiousness of an average infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
+There are two components to an infection kernel: intrinsic infectiousness of an average infected individual (the basic reproductive number)\dc{again, for me it's the intrinsic infectiousness averaged over the population, that's the $\int da$ thing.
+I stop from now on complaining about this ;)} and time distribution of new infections (generation-interval distribution).
 Assuming that a population mixes homogeneously, we can write: 
@@
 
-In a homogeneously mixing population, disease incidence $i(t)$ is the product of the current infectiousness of infected individuals and the proportion of the population susceptible, $S(t)$.
+In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
 \begin{equation}
@@
 the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions up until the observation time.
+\dc{again, weighted by what? the binary variable ``observed/not observed''? It's disturbing that we talk about weights but never say what they are...}
 
-%% \dc{You define $c_t$ after saying this, this is confusing.}
-%% During the exponential growth phase, we expect the backward distributions to be the same (and thus the censored distribution should match them).
-\dc{again, this is explained after...
-reading doesn't flow well}
-%% This connection will break down as the epidemic progresses, however, so we will to distinguish the censored distribution from the backward distribution.
-
-The density of new infections occuring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
+The density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
 \begin{equation}
-i_{t-\tau}(t) = \RR i(t-\tau) g(\tau) S(t)
+i_{t-\tau}(t) = \RR \, i(t-\tau) g(\tau) S(t)
 \end{equation}
-The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i(t-\tau) g(\tau)$:
+As shown in \cite{champredon2015intrinsic} \dc{I added the reference for the reader who wants to see how exactly this is derived}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
 \begin{equation}
@@
 \end{equation}
-On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ before $t$:
-Then, the censored generation-interval distribution is given by
+On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ \dc{you mean:``[...] any cohort infected at time $s$[...]''?}  before $t$: $\int_0^t i_{s-\tau}(s) ds$.
+Then, the censored generation-interval distribution is given by normalizing \dc{I felt we could explain a bit more, hence edited accordingly}
 \begin{equation}\label{eq:obsg}
@@
 \end{equation}
-This relationship shows that the backward generation-interval distribution is equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., when the epidemic is growing exponentially.
+This relationship shows that the backward generation-interval distribution is equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., during the initial stage of the epidemic, when it is growing exponentially.
 
@@
 
-For a single outbreak, the observed mean generation interval through contact tracing will always be shorter than intrinsic mean generation interval (Figure~\ref{fig:censor}).
+For a single outbreak, the observed mean \dc{shouldn't it be ``censored'' instead of ``observed mean''??} generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
 There are two reasons for this phenomenon.
@@
 \swp{Should I just write it as $c_t$? I can't come up with a better notation.}
+\dc{what about $c_t^{\exp}$ ?}
 \begin{equation}
@@
 A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
-Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized intervals will always be shorter \citep{champredon2015intrinsic}.
+\dc{I'm confused... Figure~\ref{fig:censor} does not show they are the same. On the other hand, \autoref{eq:exp} is an equality... which one shall I believe?}
+Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized \dc{censored?} intervals will always be shorter \citep{champredon2015intrinsic}.
 As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
 Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
-Overall, we expect naively using the observed generation-interval distribution to underestimate the reproductive number.
+Overall, we expect, naively using the observed generation-interval distribution, to underestimate the reproductive number.
 
@@
 \begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
+\hat{K}(\tau) = \int \hat{K}_a(\tau) da.
 \end{equation}
@@
 Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
+\dc{but that is only during the exponential growth phase, isn't it?}
 \begin{equation}
@@
 For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
-The intrinsic generation-interval distribution that corresponds to this model can be written has \citep{svensson2015influence}:
+The intrinsic generation-interval distribution that corresponds to this model can be written has \citep{Champredon2018equivalence} \dc{changed to a more appropriate reference (and self-promotion!)}:
 \begin{equation}
@@
 we briefly describe this relationship in Methods.
-\dc{I'm surely missing something obvious, but how do you justify this equation from the one above? }
-\swp{Is this clearer?}
 
@@
 Local depletion refers to a depletion of susceptible individuals in a household or neighborhood;
-we can think of these structures as small homogeneous networks embedded in a larger population structre (and therefore we can expect similar effects as we see in \fref{local}, right panel).
-Both the egocentric depletion and the local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
+we can think of these structures as small homogeneous networks embedded in a larger population structure (and therefore we can expect similar effects as we see in \fref{local}, right panel).
+Both the egocentric and local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
 Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
-\dc{Would it be possible to define local depletion using a more rigorous network topology, rather than households/neighborhood?
-(Would it help?).
-The same way the egocentric is defined with a star-shaped network, can we define the local depletion with a (well-known?) shape?
-Global depletion may be fine as it is, as this is something we are used to in general (in other modelling frameworks).
-}
-\swp{I'm not sure if that would help.
-We just want to get the idea across that there's more than egocentric depletion during the early in the epidemic.
-We don't actually want to do theory based on local stuff.
-We want to get the idea across that its information is hidden in the GI distributions and we don't have to worry about the technical details.
-I tried to clarify the text a little more.}
 
@@
 We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
-When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval inverse exponentially weighted \eref{exp}; hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
+When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval inverse exponentially weighted \eref{exp}; \dc{the previous sentence doesn't read well; forgot words?}
+hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
 \begin{equation}
 \tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau),
+\label{eq:geff}
 \end{equation}
@@
 We refer to the second method as the individual-level method because it relies on individual contact information.
+\dc{Wait, I'm lost! what is the first method? what it the second method?!}
 We model each infection as a non-homogeneous Poisson process arising from the infector; 
@@
 For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$.
+
+\dc{I'm confused...
+\autoref{eq:geff} shows the adjustment needed because of the censoring, that is clear.
+Is this what you call ``population-level''?
+But does the paragraph above provide also an adjustment??
+The individual-level??
+}
+
+\dc{Oh OK, I'm back here after reading the Methods section.
+Please, give an explicit reference to that section, and also to the equations.
+As you could see with my comment above, I was lost, but having such references would have helped me followed...
+}
 
@@
 Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.
+
+\dc{I suggest a short paragraph putting this work in perspective to other studies (like for example https://www.ncbi.nlm.nih.gov/pubmed/29087987).
+You also did it with Trapman, etc.
+throughout the main text, but it would be nice to summarize }
 
@@
 (E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.
-Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
-}
+Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}}
 \label{fig:cover}

### 2019-06-12 doc/network.tex
A few more edits from DC

@@
 \end{equation}
-In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we esimatemean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
+In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
 Then, the effective generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
@@
 \begin{equation}
-\frac{\RR^{n_j} \exp \big(- \RR \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
+\frac{\RR^{n_j} \exp \big(- \RR\, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
 \end{equation}
@@
 
-The entire data likelihood of the nonhomogeneous poisson process can be written as:
+The entire data likelihood of the non-homogeneous poisson process can be written as:
 \begin{equation}
@@
 &\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
-&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
+&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
 \end{aligned}
@@
 \swp{This is using epigrowthfit. How do I cite it? It's not out yet...}
+\dc{isn't this based on ``Estimating Initial Epidemic Growth Rates'' from Ma et al BMB 2013?}
 
@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-This figure matches \fref{cmp} in the main text.
-Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
+This figure matches \fref{cmp} in the main text which used a SEIR model.
+Here, an Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
 All other parameters are the same as in \fref{cmp}.

### 2019-06-18 doc/network.tex
edits based on David's comments

@@
 In practice, generation intervals are often measured by aggregating available information from contact tracing. 
-Whereas backward generation intervals represent infections that are one generation apart from a cohort, these aggregated generation intervals represent infections across multiple cohorts; 
-the time of infection of the infectors can range anywhere from the beginning of an epidemic to the most recent infection.
+Whereas backward generation intervals represent infections that are one generation apart from a cohort of infectees, these aggregated generation intervals represent infections across multiple cohorts (i.e., the time of infection of the infectees can range anywhere from the beginning of an epidemic to the most recent infection).
 \dc{I found the previous paragraph difficult to read and edited it to try to improve.
@@
 The bckwrd GI is observed at one given time and those infectees each had infectors from potentially different cohorts.}
+\swp{I meant to say that because the backward GI is observed GI is observed at one given time, all infectees share the same infection time. Is this clearer?}
 While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
-These censored observations can be thought of as a weighted average of backward generation-interval distributions; \dc{weighted by what?}
+These censored observations can be thought of as a weighted average of backward generation-interval distributions (weighted by incidence); \dc{weighted by what?} \swp{equation 8}
 like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.
@@
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population even in the early phase of an epidemic.
-This means that contacts made late in an individual's infection are more likely to be ``masked'' due to infectious contacts that were made earlier \dc{Although I get you try to explain in simple terms what's going on with ``masked'', I think it's not really clear: masked by what?
+This means that contacts made late in an individual's infection are more likely to be ineffective due to infectious contacts that were made earlier (and so the contactee will have been infected already)
+\dc{Although I get you try to explain in simple terms what's going on with ``masked'', I think it's not really clear: masked by what?
 is it that subsequent contacts in the neighbourhood do not count given they were infected earlier?}.
+\swp{Tried to clarify}
 As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
@@
 making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
-An individual-level intrinsic infection kernel $K_a(\tau)$ \dc{is it clearer to specify the kernel is about infections?} describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
+An individual-level intrinsic infection kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
+\dc{is it clearer to specify the kernel is about infections?}
+\swp{Yes,}
 Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the intrinsic infectiousness of a disease as well as the contact structure around an infected individual.
@@
 \begin{equation}
-K(\tau) = \int K_a (\tau) da.
+K(\tau) = \int K_a (\tau) dA.
 \end{equation}
-The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
+\swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
+The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
 \dc{wouldn't it be better to say: ``The population-level kernel describes the rate at which secondary infections are expected to be caused by an infected individual, on \emph{average}.'' I have troubles describing an ``average  infected individual''...
@@
 }
+\swp{I think that's clearer too.}
 
-There are two components to an infection kernel: intrinsic infectiousness of an average infected individual (the basic reproductive number)\dc{again, for me it's the intrinsic infectiousness averaged over the population, that's the $\int da$ thing.
-I stop from now on complaining about this ;)} and time distribution of new infections (generation-interval distribution).
+\dc{again, for me it's the intrinsic infectiousness averaged over the population, that's the $\int da$ thing. I stop from now on complaining about this ;)}
+\swp{I think figure 1 shows that we're averaging across infectioussness as well as the time distribution.
+We also say above that ``Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness''}
+There are two components to an infection kernel: intrinsic infectiousness of an average infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
 Assuming that a population mixes homogeneously, we can write: 
@@
 We can understand the effect of right-censoring using backward generation-interval distributions \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}:
-the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions up until the observation time.
+the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions (weighted by incidence) up until the observation time.
 \dc{again, weighted by what? the binary variable ``observed/not observed''? It's disturbing that we talk about weights but never say what they are...}
+\swp{They were (and are) in equation 8. I made it explicit.}
 
@@
 As shown in \cite{champredon2015intrinsic} \dc{I added the reference for the reader who wants to see how exactly this is derived}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
-\begin{equation}
+\begin{equation}\label{eq:backward}
 b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-\sigma) g(\sigma) d\sigma}
 \end{equation}
-On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infection time $s$ \dc{you mean:``[...] any cohort infected at time $s$[...]''?}  before $t$: $\int_0^t i_{s-\tau}(s) ds$.
-Then, the censored generation-interval distribution is given by normalizing \dc{I felt we could explain a bit more, hence edited accordingly}
+On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infected at time $s$ before $t$: $\int_0^t i_{s-\tau}(s) ds$.
+Then, the censored generation-interval distribution is given by normalizing
 \begin{equation}\label{eq:obsg}
@@
 \end{equation}
-Substituting $b_s(\tau) = i(s-\tau) g(\tau)$ allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
+Substituting \eref{backward} allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
 \begin{equation}
-c_t(\tau) \propto \int_0^t b_s(\tau) S(s) ds.
+c_t(\tau) \propto \int_0^t b_s(\tau) i(s) ds.
 \end{equation}
-This relationship shows that the backward generation-interval distribution is equivalent to the censored generation-interval distribution when $S \approx 1$, i.e., during the initial stage of the epidemic, when it is growing exponentially.
 
@@
 
-For a single outbreak, the observed mean \dc{shouldn't it be ``censored'' instead of ``observed mean''??} generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
+\dc{shouldn't it be ``censored'' instead of ``observed mean''??}
+\swp{I think observed in OK. Observation through contact tracing is what causes censoring.}
+For a single outbreak, the observed mean  generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
 There are two reasons for this phenomenon.
@@
 \dc{what about $c_t^{\exp}$ ?}
+\swp{TODO: ask JD}
 \begin{equation}
@@
 \dc{I'm confused... Figure~\ref{fig:censor} does not show they are the same. On the other hand, \autoref{eq:exp} is an equality... which one shall I believe?}
-Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized \dc{censored?} intervals will always be shorter \citep{champredon2015intrinsic}.
+\swp{This is a notation issue... equality only holds during the exponential period. TODO: talk to Jonathan about notation.}
+Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized \dc{censored?} \swp{no, realized; there is no observation process in the second phenomenon} intervals will always be shorter \citep{champredon2015intrinsic}.
 As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
@@
 \begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{K}_a(\tau) da.
+\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
 \end{equation}
@@
 \dc{but that is only during the exponential growth phase, isn't it?}
+\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.}
 \begin{equation}
@@
 For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
-The intrinsic generation-interval distribution that corresponds to this model can be written has \citep{Champredon2018equivalence} \dc{changed to a more appropriate reference (and self-promotion!)}:
+The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{Champredon2018equivalence}:
 \begin{equation}
@@
 Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
-We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
-When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval inverse exponentially weighted \eref{exp}; \dc{the previous sentence doesn't read well; forgot words?}
+We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
+When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval to be inverse-exponentially weighted \eref{exp}; \dc{the previous sentence doesn't read well; forgot words?} \swp{Yes, forgot words.
+Not sure if ``inverse-exponentially weighted'' is clear, however.}
 hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
@@
 where $r$ is the exponential growth rate.
-We refer to the second method as the individual-level method because it relies on individual contact information.
-\dc{Wait, I'm lost! what is the first method? what it the second method?!}
-We model each infection as a non-homogeneous Poisson process arising from the infector; 
+We refer to the second method (Section~\ref{indlevel}) as the individual-level method because it relies on individual contact information.
+We model each infection as a non-homogeneous Poisson process arising from the infector \eref{nonhomogeneous}; 
 incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
 For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$.
-
-\dc{I'm confused...
-\autoref{eq:geff} shows the adjustment needed because of the censoring, that is clear.
-Is this what you call ``population-level''?
-But does the paragraph above provide also an adjustment??
-The individual-level??
-}
-
-\dc{Oh OK, I'm back here after reading the Methods section.
-Please, give an explicit reference to that section, and also to the equations.
-As you could see with my comment above, I was lost, but having such references would have helped me followed...
-}
 
@@
 \subsection{Population-level method for estimating generation-interval distributions from contact tracing data}
+\label{poplevel}
 
@@
 \subsection{Individual-level method for estimating generation-interval distributions from contact tracing data}
+\label{indlevel}
 
@@
 The entire data likelihood of the non-homogeneous poisson process can be written as:
-\begin{equation}
+\begin{equation}\label{eq:nonhomogeneous}
 \begin{aligned}
@@
 \dc{isn't this based on ``Estimating Initial Epidemic Growth Rates'' from Ma et al BMB 2013?}
+\swp{Yes, but they also have a package called epigrowthfit which is based on Ma et al. Maybe in acknowledgements?}
 

### 2019-06-19 doc/network.tex
abstract?

@@
 
+\section*{Abstract}
+
+Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, provides a link between two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
+Often, generation intervals are measured through contact tracing by identifying who infected whom;
+we study how these observed intervals vary both spatially and temporally.
+Temporal variation in generation intervals is caused by the censoring in the observation process through contact tracing, whereas spatial variation in generation intervals reflects the patterns of spread of disease.
+Therefore, correcting for temporal variation, but not for spatial variation, provides an ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
+
 \section{Introduction}
@@
 
-An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength, characterized by its reproductive number, \RR.
+An epidemic can be characterized by the exponential growth rate, $r$, and the reproductive number, \RR.
 The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.

### 2019-06-19 doc/network.tex
section numbering

@@
 \appendix
+\renewcommand\thefigure{\thesection.\arabic{figure}}
+\setcounter{figure}{0}    
 \section{Appendix}
@@
 (E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.
-Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}}
+Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
+}
 \label{fig:cover}

### 2019-06-19 doc/network.tex
Editing and resolving comments

@@
 The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
-\jd{Took out a comment about tracing and tried to clarify.}
 However, estimating the reproductive number directly from disease natural history requires detailed knowledge which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
@@
 
-\jd{Still too much forward stuff for me; we don't use it for anything.}
-\swp{Got rid of forward stuff and tried to clarify}
 % Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
@@
 % and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.
-Unlike the intrinsic generation-interval distribution, the shape of the observed generation-interval distributions depends on when and how they are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+While the intrinsic generation-interval distribution is often assumed to be fixed, the shape of the observed generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
 Here, we briefly describe how backward generation-interval distributions change over time because backward measurements better reflect real outbreak scenarios where contact tracing is performed by identifying infectors of infected individuals.
@@
 In practice, generation intervals are often measured by aggregating available information from contact tracing. 
-Whereas backward generation intervals represent infections that are one generation apart from a cohort of infectees, these aggregated generation intervals represent infections across multiple cohorts (i.e., the time of infection of the infectees can range anywhere from the beginning of an epidemic to the most recent infection).
-\dc{I found the previous paragraph difficult to read and edited it to try to improve.
-The `cohorts` were not referring to the same ones (I think) and that was confusing.
-The bckwrd GI is observed at one given time and those infectees each had infectors from potentially different cohorts.}
-\swp{I meant to say that because the backward GI is observed GI is observed at one given time, all infectees share the same infection time. Is this clearer?}
+%% Whereas backward generation intervals represent infections that are one generation apart from a cohort of infectees, these aggregated generation intervals represent infections across multiple cohorts (i.e., the time of infection of the infectees can range anywhere from the beginning of an epidemic to the most recent infection).
 While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
-These censored observations can be thought of as a weighted average of backward generation-interval distributions (weighted by incidence); \dc{weighted by what?} \swp{equation 8}
-like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.
+Thus, there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
+In fact, censored observations can be thought of as an average of backward generation-interval distributions, over the cohorts which have been infected before the observation time (and weighted by cohort size).
+\jd{Simplified radically.}
 
@@
 In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population even in the early phase of an epidemic.
-This means that contacts made late in an individual's infection are more likely to be ineffective due to infectious contacts that were made earlier (and so the contactee will have been infected already)
-\dc{Although I get you try to explain in simple terms what's going on with ``masked'', I think it's not really clear: masked by what?
-is it that subsequent contacts in the neighbourhood do not count given they were infected earlier?}.
-\swp{Tried to clarify}
+This means that contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already) 
 As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
-These observations allow us to make the following prediction: removing the censoring bias from observed generation intervals will give us a distribution that contains information about the population structure; using this distribution should allow us to correctly infer $\RR$ from $r$.
+These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will give us a corrected distribution that contains information about the population structure; using this corrected distribution should allow us to correctly infer $\RR$ from $r$.
 
@@
 making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
-An individual-level intrinsic infection kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
-\dc{is it clearer to specify the kernel is about infections?}
-\swp{Yes,}
-Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the intrinsic infectiousness of a disease as well as the contact structure around an infected individual.
+An individual-level intrinsic infection kernel $K_a(\tau)$ describes the rate at which an infected individual $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
+Individuals may vary in their course of infection (e.g., duration of latent and infectious periods) and their level of infectiousness, which can depend both on biological infectiousness as well contact patterns.
 
-Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
+Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations across an ``aspect space'' $A$ which describes the population distributions:
 \begin{equation}
@@
 \swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
+\jd{dA is better, and we obviously need to clarify.}
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
-\dc{wouldn't it be better to say: ``The population-level kernel describes the rate at which secondary infections are expected to be caused by an infected individual, on \emph{average}.'' I have troubles describing an ``average  infected individual''...
-Oh no, but you say it again below...
-well, I keep this comment, do whatever you want!
-}
-\swp{I think that's clearer too.}
 
-\dc{again, for me it's the intrinsic infectiousness averaged over the population, that's the $\int da$ thing. I stop from now on complaining about this ;)}
-\swp{I think figure 1 shows that we're averaging across infectioussness as well as the time distribution.
-We also say above that ``Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness''}
-There are two components to an infection kernel: intrinsic infectiousness of an average infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
 Assuming that a population mixes homogeneously, we can write: 
@@
 \end{equation}
-where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population), 
-and $g(\tau)$ is the intrinsic generation-interval distribution.
+where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population), 
+and $g(\tau)$ is the expected time distribution of those cases (the intrinsic generation-interval distribution).
 
@@
 the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions (weighted by incidence) up until the observation time.
-\dc{again, weighted by what? the binary variable ``observed/not observed''? It's disturbing that we talk about weights but never say what they are...}
-\swp{They were (and are) in equation 8. I made it explicit.}
 
@@
 \end{equation}
-As shown in \cite{champredon2015intrinsic} \dc{I added the reference for the reader who wants to see how exactly this is derived}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
+As shown in \cite{champredon2015intrinsic},
+%% \dc{I added the reference for the reader who wants to see how exactly this is derived}
+the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
 \begin{equation}\label{eq:backward}
@@
 
-\dc{shouldn't it be ``censored'' instead of ``observed mean''??}
-\swp{I think observed in OK. Observation through contact tracing is what causes censoring.}
 For a single outbreak, the observed mean  generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
 There are two reasons for this phenomenon.
-First, any infection events that occur after the contact tracing period cannot be observed due to right censoring and short generation intervals are more likely to be observed.
+First, longer generation intervals are more likely to be missed due to right censoring (and short generation intervals are more likely to be observed).
 In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
-the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the inverse exponentially weighted intrinsic generation-interval distribution \citep{britton2019estimation}:
+the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
 \dc{notation $c_{exp}$ is not really consistent with Equation (9) $c_t$}
-\jd{Use math formatting in comments or it can mess up the whole doc.}
 \swp{Should I just write it as $c_t$? I can't come up with a better notation.}
@@
 \swp{TODO: ask JD}
+\jd{I like $b_0$ for the initial backward. We can then use $c_0$ for the initial censored, I guess.}
 \begin{equation}
@@
 \swp{This is a notation issue... equality only holds during the exponential period. TODO: talk to Jonathan about notation.}
-Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized \dc{censored?} \swp{no, realized; there is no observation process in the second phenomenon} intervals will always be shorter \citep{champredon2015intrinsic}.
+Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
 As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
@@
 As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
-this generation-interval-based argument provides a clear biological interpretation for the same result presented by \cite{trapman2016inferring}.
+this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
 % Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
@@
 In practice, directly using this distribution to link $r$ and $\RR$ is unrealistic because it requires that we know the per-pair contact rate. 
-Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary meterial section 1.4.2);
+Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
 we briefly describe this relationship in Methods.
@@
 
-\jd{I want to keep thinking about network choice. Would a random network be useful? Are we careful enough about time windows (exponential phase vs (roughly) full single epidemic?}
-\swp{Clarified time windows.}
 \begin{figure}[!pbth]
@@
 We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
-When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval to be inverse-exponentially weighted \eref{exp}; \dc{the previous sentence doesn't read well; forgot words?} \swp{Yes, forgot words.
-Not sure if ``inverse-exponentially weighted'' is clear, however.}
+When an epidemic is growing exponentially, right-censoring causes the observed generation interval to be discounted by the exponential growth rate \eref{exp};
+\swp{Forgot words. Not sure if ``inverse-exponentially weighted'' is clear, however.}
+\jd{I fiddled the words -- but the equation ref is the key here; I think it's fine.}
 hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
@@
 where $r$ is the exponential growth rate.
+
 We refer to the second method (Section~\ref{indlevel}) as the individual-level method because it relies on individual contact information.

### 2019-06-19 doc/network.tex
Drastic changes to Conclusion

@@
 
-To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap} and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
+To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
+\jd{What's wrong with this reference (see rendering)?}
+and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
 As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
@@
 we therefore expect our simulations on an empirical network to be better characterized by subexponential growth models \citep{viboud2016generalized}.
-However, our simulations suggest that the initial exponential growth assumption still provides a viable approach for estimating the reproductive number.
-This conclusion is supported by \cite{chowell2015western} who found that epidemics can exhibit exponential growth patterns globally even if local epidemics may not.
+However, our simulations suggest that the initial exponential growth assumption still provides a useful approach for estimating the reproductive number.
+%% This conclusion is supported by \cite{chowell2015western} who found that epidemics can exhibit exponential growth patterns globally even if local epidemics may not.
 
-\swp{Conclusion?}
-\jd{Let me try.}
-\swp{How about this?}
-Contact tracing provides an effective way of controlling the disease and gathering epidemiological information early in an outbreak.
-Out study demonstrates that generation intervals measured through contact tracing can capture key characteristics (e.g., contact structure) an outbreak; 
-these implicit features allow us to infer the ``correct'' generation-interval distribution, which accurately describes how disease spreads in a population. 
-Future studies should carefully consider how underlying epidemiological patterns can be inferred from contact tracing data.
-Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.
+Estimating generation-interval distributions is a key, and often under-appreciated, component of disease modeling and forecasting.
+Different definitions, and different measurement approaches, produce different estimates of these distributions.
+We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
+These predictable differences can arise from temporal effects, local spatial (or network) effects and population-level effects.
+We have shown that correcting for temporal effects allows us to estimate a spatially informed distribution, which accurately describes how disease spreads in a population. 
+Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence spread of disease. 
+%% Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.
 

### 2019-06-19 doc/network.tex
Minor stuff at meeting
JD fiddles abstract

@@
 
-Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, provides a link between two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
-Often, generation intervals are measured through contact tracing by identifying who infected whom;
-we study how these observed intervals vary both spatially and temporally.
-Temporal variation in generation intervals is caused by the censoring in the observation process through contact tracing, whereas spatial variation in generation intervals reflects the patterns of spread of disease.
-Therefore, correcting for temporal variation, but not for spatial variation, provides an ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
+Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
+Often, generation intervals are measured through contact tracing by identifying who infected whom.
+We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales.
+We show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
 
 \section{Introduction}
-
-%% SWP: see or cite
-%% https://www.ncbi.nlm.nih.gov/pubmed/29087987
 
@@
 
-%% \dc{you make several assertions that are difficult to understand that early in the text.
-It really doesn't flow well, it was hard to follow, evn fo me who has worked on this kind of things.
-I suggest you define first c_t.}
-%% \jd{I think that verbal explanations first are better for a larger fraction of people.
-Maybe we can just make the verbal explanations better.
-I agree that we should keep the horrible math if DC likes it.}
-
 When generation intervals are estimated through contact tracing during an outbreak, infection events that have not happened yet are not observed. 
@@
 \end{equation}
-As shown in \cite{champredon2015intrinsic},
-%% \dc{I added the reference for the reader who wants to see how exactly this is derived}
-the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
+As shown in \cite{champredon2015intrinsic}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
 \begin{equation}\label{eq:backward}
@@
 
+\jd{\P\ about contact tracing.}
+\dc{I suggest a short paragraph putting this work in perspective to other studies (like for example https://www.ncbi.nlm.nih.gov/pubmed/29087987).
+You also did it with Trapman, etc.
+throughout the main text, but it would be nice to summarize }
+
 Estimating generation-interval distributions is a key, and often under-appreciated, component of disease modeling and forecasting.
@@
 These predictable differences can arise from temporal effects, local spatial (or network) effects and population-level effects.
-We have shown that correcting for temporal effects allows us to estimate a spatially informed distribution, which accurately describes how disease spreads in a population. 
+We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
 Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence spread of disease. 
 %% Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.
-
-\dc{I suggest a short paragraph putting this work in perspective to other studies (like for example https://www.ncbi.nlm.nih.gov/pubmed/29087987).
-You also did it with Trapman, etc.
-throughout the main text, but it would be nice to summarize }
 
@@
 This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
-\jd{Mention that our main-text models use exponential later instead.}
-\swp{Done.}
 \begin{equation}
@@
 The infectious period $I_i$ is the sum of all of these waiting times.
-
-\jd{Add flowchart?}
-\swp{Not sure what kind of flowchart you're thinking about.. let me think more...}
 

### 2019-06-20 doc/network.tex
a new paragraph?

@@
 we therefore expect our simulations on an empirical network to be better characterized by subexponential growth models \citep{viboud2016generalized}.
-However, our simulations suggest that the initial exponential growth assumption still provides a useful approach for estimating the reproductive number.
-%% This conclusion is supported by \cite{chowell2015western} who found that epidemics can exhibit exponential growth patterns globally even if local epidemics may not.
+However, our simulations suggest that the initial exponential growth assumption still provides a viable approach for estimating the reproductive number.
 
-\jd{\P\ about contact tracing.}
-\dc{I suggest a short paragraph putting this work in perspective to other studies (like for example https://www.ncbi.nlm.nih.gov/pubmed/29087987).
-You also did it with Trapman, etc.
-throughout the main text, but it would be nice to summarize }
+\swp{New paragraph:}
+Contact tracing provides an effective way of collecting epidemiological data and controlling an outbreak \citep{clarke1998contact, eames2003contact, donnelly2003epidemiological}.
+In particular, using tracing information allows us to infer real-time estimates of the time-varying reproductive number \citep{cauchemez2006estimating, hens2012robust, jewell2012enhancing, soetens2018real};
+generation-interval distributions, which can be either assumed or estimated, often play a central role in describing these tracing data.
+Our study illustrates that generation intervals measured through contact tracing data contain information about the underlying contact structure, which can be implicitly reflected in the estimates of the reproductive number;
+this perspective can be particularly useful for characterizing an epidemic because detailed information about the contact structure is often unavailable.
 
-Estimating generation-interval distributions is a key, and often under-appreciated, component of disease modeling and forecasting.
+Generation-interval distribution is a key, and often under-appreciated, component of disease modeling and forecasting.
 Different definitions, and different measurement approaches, produce different estimates of these distributions.
 We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
-These predictable differences can arise from temporal effects, local spatial (or network) effects and population-level effects.
+These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
 We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
 Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence spread of disease.
-%% Understanding the role of contact tracing and developing protocols that are informed by the theory will help researchers better prepare for future outbreaks.
 

### 2019-06-20 doc/network.tex
kind of done...?

@@
 \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
-
-\title{Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}}
-\author{Sang Woo Park, David Champredon and Jonathan Dushoff}
 
@@
 \begin{document}
-\maketitle
+
+\begin{flushleft}{
+	\Large
+	\textbf\newline{
+		Inferring generation-interval distributions from contact tracing data
+	}
+}
+\newline
+\\
+Sang Woo Park\textsuperscript{1,*}
+David Champredon\textsuperscript{2,3}
+Jonathan Dushoff\textsuperscript{2,4}
+\\
+
+\bigskip
+\textbf{1} Department of Mathematics \& Statistics, McMaster University, Hamilton, Ontario, Canada
+\\
+\textbf{2} Department of Biology, McMaster University, Hamilton, Ontario, Canada
+\\
+\textbf{3} Agent-Based Modelling Laboratory, York University, Toronto, Ontario, Canada
+\\
+\textbf{4} Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
+\\
+\bigskip
+
+*parksw3@mcmaster.ca
+\end{flushleft}
 
@@
 As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
-This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
-These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will give us a corrected distribution that contains information about the population structure; using this corrected distribution should allow us to correctly infer $\RR$ from $r$.
+This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
+These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will give us a corrected distribution that contains information about the population structure; using this corrected distribution should allow us to correctly infer the reproductive number from the exponential growth rate.
 
@@
 the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
-\dc{notation $c_{exp}$ is not really consistent with Equation (9) $c_t$}
-\swp{Should I just write it as $c_t$? I can't come up with a better notation.}
-\dc{what about $c_t^{\exp}$ ?}
-\swp{TODO: ask JD}
-\jd{I like $b_0$ for the initial backward. We can then use $c_0$ for the initial censored, I guess.}
 \begin{equation}
-c_t(\tau) = b_t(\tau) \propto g(\tau) \exp(-r\tau).
+c_0(\tau) = b_0(\tau) \propto g(\tau) \exp(-r\tau),
 \label{eq:exp}
 \end{equation}
+where $c_0$ and $b_0$ represent the ``initial'' censored and backward generation-interval distributions (during the exponential growth phase), respectively.
 A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
-\dc{I'm confused... Figure~\ref{fig:censor} does not show they are the same. On the other hand, \autoref{eq:exp} is an equality... which one shall I believe?}
-\swp{This is a notation issue... equality only holds during the exponential period. TODO: talk to Jonathan about notation.}
 Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
@@
 This calculation can be validated by simulating stochastic infection processes on a ``star'' network (i.e, a single infected individual at the center connected to multiple susceptible individuals who are not connected with each other).
-Simulations confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution, while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (\fref{local}).
+Simulations confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution, while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (\fref{local}, middle panel).
 
@@
 When an epidemic is growing exponentially, right-censoring causes the observed generation interval to be discounted by the exponential growth rate \eref{exp};
-\swp{Forgot words. Not sure if ``inverse-exponentially weighted'' is clear, however.}
-\jd{I fiddled the words -- but the equation ref is the key here; I think it's fine.}
 hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
@@
 To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
-\jd{What's wrong with this reference (see rendering)?}
 and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
@@
 It consists of 23133 nodes and 93497 edges.
-\cite{trapman2016inferring} used the same network to simulate epidemics.
+The same network was used by \cite{trapman2016inferring} to study how network structure affects the estimate of the reproductive number.
 

### 2019-06-20 doc/network.tex
tiny change

@@
 As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
-estimates based on the egocentric generation-interval distribution (\eref{conditional}) are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
+estimates based on the egocentric generation-interval distribution \eref{conditional} are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
 Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.

### 2019-06-20 doc/network.tex
minor edits by DC

@@
 \jd{dA is better, and we obviously need to clarify.}
+\dc{So, isn't the standard way to write this integral like that: $ \int_{A} K_a (\tau) da$? Just trying to clarify here, this is confusing... }
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
@@
 \swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.} 
+\dc{mmh, still not super clear, but I guess it's fine for now; I don't really know how to improve this though... also the reviewers will be a good test for the clarity of this part...}
 \begin{equation}
@@
 We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
-Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence spread of disease.
+Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence the spread of disease.
 
@@
 \dc{isn't this based on ``Estimating Initial Epidemic Growth Rates'' from Ma et al BMB 2013?}
-\swp{Yes, but they also have a package called epigrowthfit which is based on Ma et al. Maybe in acknowledgements?}
+\swp{Yes, but they also have a package called epigrowthfit which is based on Ma et al. Maybe in acknowledgements?} \dc{Oh OK, yes}
 

### 2019-06-23 doc/network.tex
Have I pulled? Why still so many comments?

@@
 We show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
+TODO: Something about statistical methods?
 
@@
 
-Generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
+A generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
 Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.
@@
 In fact, censored observations can be thought of as an average of backward generation-interval distributions, over the cohorts which have been infected before the observation time (and weighted by cohort size).
-\jd{Simplified radically.}
 
@@
 These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will give us a corrected distribution that contains information about the population structure; using this corrected distribution should allow us to correctly infer the reproductive number from the exponential growth rate.
+In other words, Trapman's finding can be explained by the gap between the intrinsic generation interval and an ``effective'' (spatially corrected) generation interval \jdnew
 

### 2019-06-23 doc/network.tex
Some edits (went lightly only through Methods).

@@
 
+\newcommand{\jdnew}{\jd{NEW}}
+\newcommand{\jddel}[1]{\jd{DELETE: #1}}
+
 \begin{document}
@@
 	\textbf\newline{
-		Inferring generation-interval distributions from contact tracing data
+		Inferring generation-interval distributions from contact-tracing data
 	}
@@
 We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales.
-We show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
-TODO: Something about statistical methods?
+We develop and test statistical methods for both spatial and temporal corrections of generation intervals,
+and show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
 
@@
 
-Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations across an ``aspect space'' $A$ which describes the population distributions:
+Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations \jdnew:
 \begin{equation}
-K(\tau) = \int K_a (\tau) dA.
+K(\tau) = \int K_a (\tau) f(a) da,
 \end{equation}
-\swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
-\jd{dA is better, and we obviously need to clarify.}
+The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
+where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
 \dc{So, isn't the standard way to write this integral like that: $ \int_{A} K_a (\tau) da$? Just trying to clarify here, this is confusing... }
-The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
+\jd{I haven't thought about this for a long time.
+dA actually represents something pretty confusing (a chunk of the \emph{distribution} across the space).
+Probably easier to change than explain, but we do need a distribution now.}
 
@@
 A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
-Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
+Second, the decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
 As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
@@
 
-Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \citep{trapman2016inferring}.
+Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$. \jddel{, following \citep{trapman2016inferring}}
 In this case, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
@@
 \dc{mmh, still not super clear, but I guess it's fine for now; I don't really know how to improve this though... also the reviewers will be a good test for the clarity of this part...}
+\jd{Not quite sure what Daniel is going for here: this equation depends on having an $r$, which does depend on exponentiality; also it is not how the number is ``defined'' at all.
+Not even sure if it's how it's estimated, except on a star network.
+This should be rewritten and not wait for reviewers.}
 \begin{equation}
@@
 This calculation can be validated by simulating stochastic infection processes on a ``star'' network (i.e, a single infected individual at the center connected to multiple susceptible individuals who are not connected with each other).
-Simulations confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution, while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (\fref{local}, middle panel).
+Simulations (\fref{local}) confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution (left panel), while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (middle panel).
 
@@
 Both the egocentric and local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
-Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
+Finally, global depletion refers to overall depeletion of susceptibility at the population level, and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
 
-\section{Inferring generation-interval distributions from a contact tracing data}
+\section{Inferring generation-interval distributions from contact-tracing data}
 
@@
 
-Since the right-censoring effect is a sampling bias, we want to take it into account explicitly.
+Since the right-censoring effect is a sampling bias, we typically want to correct for it. 
 In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
-We therefore expect that if we correct observed generation intervals for the right-censoring effect, spatial components will be taken into account implicitly. 
+We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread \jdnew.
 When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$.
-Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution $\tsub{g}{eff}(\tau)$.
+We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
 In a large homogeneously mixing population, the effective distribution will be equivalent to the intrinsic distribution.
@@
 Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
-Similarly, using intrinsic and egocentric generation-interval distributions overestimates the reproductive number because they fail to account for local spatial effects.
-Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
+Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially \jdnew.
+Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise \jdnew.
 Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$.
@@
 
-Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
+Here, we investigate two methods for correcting for temporal bias in contact-tracing data (see Methods for details).
 We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
@@
 and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
-As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
-estimates based on the egocentric generation-interval distribution \eref{conditional} are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects.
-Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
+As expected, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
+estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially \jdnew, as they do not account for indirect spatial effects. 
+Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
 While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
@@
 Changes due to temporal censoring reflect observation bias, whereas changes due to spatial or network structure reflect the dynamics of the outbreak.
-Thus correcting the observed distribution for temporal, but not spatial, effects provides the correct link between $r$ and $\RR$.
+Thus, correcting the observed distribution for temporal, but not spatial, effects provides the correct link between $r$ and $\RR$.
 
@@
 
-We classified susceptible depletion into three levels: egocentric, local, and global.
+We think of susceptible depletion as operating on three levels: egocentric, local, and global.
 Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
-Accounting for the egocentric effect allowed us to reinterpret the results by \cite{trapman2016inferring} using arguments based on generation intervals.
+Accounting for the egocentric effect allowed us to link the results by \cite{trapman2016inferring} to established results \jdnew\ based on generation intervals.
 Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
@@
 
-The generation-interval-based approaches to estimating the reproductive number often assumes that an epidemic grows exponentially \citep{wearing2005appropriate, wallinga2007generation, roberts2007model, park2019practical}.
+Generation-interval-based approaches to estimating the reproductive number often assume that an epidemic grows exponentially \citep{wearing2005appropriate, wallinga2007generation, roberts2007model, park2019practical}.
 In practice, heterogeneity in population structure can lead to subexponential growth \citep{szendroi2004polynomial, chowell2015western, chowell2016growing, chowell2016characterizing, kiskowski2016modeling, viboud2016generalized};
@@
 
-\swp{New paragraph:}
 Contact tracing provides an effective way of collecting epidemiological data and controlling an outbreak \citep{clarke1998contact, eames2003contact, donnelly2003epidemiological}.
 In particular, using tracing information allows us to infer real-time estimates of the time-varying reproductive number \citep{cauchemez2006estimating, hens2012robust, jewell2012enhancing, soetens2018real};
-generation-interval distributions, which can be either assumed or estimated, often play a central role in describing these tracing data.
-Our study illustrates that generation intervals measured through contact tracing data contain information about the underlying contact structure, which can be implicitly reflected in the estimates of the reproductive number;
+generation-interval distributions, which can be either assumed or estimated, often play a central role in analyzing tracing data.
+Our study illustrates that generation intervals measured through contact-tracing data contain information about the underlying contact structure, which can be implicitly reflected in the estimates of the reproductive number;
 this perspective can be particularly useful for characterizing an epidemic because detailed information about the contact structure is often unavailable.
 
-Generation-interval distribution is a key, and often under-appreciated, component of disease modeling and forecasting.
+The generation-interval distribution is a key, and often under-appreciated, component of disease modeling and forecasting.
 Different definitions, and different measurement approaches, produce different estimates of these distributions.
@@
 These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
-We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
+We have shown that correcting observed intervals for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population.
 Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence the spread of disease.
@@
 
-\subsection{Population-level method for estimating generation-interval distributions from contact tracing data}
+\subsection{Population-level method for estimating generation-interval distributions from contact-tracing data}
 \label{poplevel}
@@
 
-\subsection{Individual-level method for estimating generation-interval distributions from contact tracing data}
+\subsection{Individual-level method for estimating generation-interval distributions from contact-tracing data}
 \label{indlevel}

### 2019-06-24 doc/network.tex
Messing with minor warnings
have to go help jo

@@
 
-\usepackage{hyperref}
+\usepackage[pdfencoding=auto, psdextra]{hyperref}
+
 \usepackage{natbib}
-\usepackage{hyperref}
 \bibliographystyle{chicago}
@@
 
+\swp{Changed the last two sentences.}
 Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
-Often, generation intervals are measured through contact tracing by identifying who infected whom.
+Generation intervals are often measured through contact tracing by identifying who infected whom.
 We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales.
-We develop and test statistical methods for both spatial and temporal corrections of generation intervals,
-and show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
+Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the censoring and the population structure near the source of disease spread; 
+therefore, we predict that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly links $r$ and $\RR$.
+We develop and test statistical methods for temporal corrections of generation intervals,
+and confirm our prediction using individual-based simulations on an empirical network.
 
@@
 These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will give us a corrected distribution that contains information about the population structure; using this corrected distribution should allow us to correctly infer the reproductive number from the exponential growth rate.
-In other words, Trapman's finding can be explained by the gap between the intrinsic generation interval and an ``effective'' (spatially corrected) generation interval \jdnew
+In other words, Trapman \textit{et al.}'s finding can be explained by the gap between the intrinsic generation interval and an ``effective'' (spatially corrected) generation interval.
 
@@
 
+\swp{New}
 Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence};
 making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
-An individual-level intrinsic infection kernel $K_a(\tau)$ describes the rate at which an infected individual $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
-Individuals may vary in their course of infection (e.g., duration of latent and infectious periods) and their level of infectiousness, which can depend both on biological infectiousness as well contact patterns.
+An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
+Individual `aspects'' may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
 
-Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations \jdnew:
+Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
 \begin{equation}
-K(\tau) = \int K_a (\tau) f(a) da,
+K(\tau) = \int k (\tau; a) f(a) da,
 \end{equation}
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-\dc{So, isn't the standard way to write this integral like that: $ \int_{A} K_a (\tau) da$? Just trying to clarify here, this is confusing... }
-\jd{I haven't thought about this for a long time.
-dA actually represents something pretty confusing (a chunk of the \emph{distribution} across the space).
-Probably easier to change than explain, but we do need a distribution now.}
 
@@
 To explore the effect of multiple contacts on realized generation intervals, we first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
-We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case $a$ in the absence of other infectors:
+We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case with aspect $a$ in the absence of other infectors:
 \begin{equation}
-\hat{K}_a(\tau) = K_a(\tau) \exp \left(- \delta_a \int_0^\tau K_a(s) ds\right),
+\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(s; a) ds\right),
 \end{equation}
-where $K_a(\tau)$ is the individual-level intrinsic kernel and $e^{- \delta_a \int_0^\tau K_a(s) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by individual $a$.
-The dilution term, $\delta_a$, models how contacts are distributed among susceptible acquaintances.
+where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by a particular infected individual.
+The dilution term, $\delta(a)$, models how contacts are distributed among susceptible acquaintances.
 
-Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$. \jddel{, following \citep{trapman2016inferring}}
-In this case, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
+Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$.
+In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) ds$ is the product of the number of acquaintances $N(a)$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
 This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
@@
 \begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
+\hat{K}(\tau) = \int \hat{k}(\tau; a) \hat{f}(a) da,
 \end{equation}
+where $\hat{f}(a)$ represents a probability density over an aspect space, including the contact structure from an egocentric point of view (i.e., the degree distribution across network). \swp{New}
 \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
-Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
-\dc{but that is only during the exponential growth phase, isn't it?}
-\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.} 
-\dc{mmh, still not super clear, but I guess it's fine for now; I don't really know how to improve this though... also the reviewers will be a good test for the clarity of this part...}
-\jd{Not quite sure what Daniel is going for here: this equation depends on having an $r$, which does depend on exponentiality; also it is not how the number is ``defined'' at all.
-Not even sure if it's how it's estimated, except on a star network.
-This should be rewritten and not wait for reviewers.}
+Finally, the link between the growth rate and the egocentric reproductive number can be obtained by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
 \begin{equation}
@@
 this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
+\swp{Simplified.}
 % Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
@@
 In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
-In practice, directly using this distribution to link $r$ and $\RR$ is unrealistic because it requires that we know the per-pair contact rate. 
+In practice, directly using this distribution to link $r$ and $\RR$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
 Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
-we briefly describe this relationship in Methods.
+we briefly describe this relationship in Section~\ref{egosection}.
+\swp{Is it OK to mention this here? Or should I say something about it at the end of the previous paragraph as well? I feel like it's too abstract to say anything there.}
 
@@
 In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
-We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread \jdnew.
+We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread.
 When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$.
 We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
-In a large homogeneously mixing population, the effective distribution will be equivalent to the intrinsic distribution.
+In a large homogeneously mixing population, the effective distribution is expected to be equivalent to the intrinsic distribution \swp{(since the egocentric depletion is negligible and the local depletion does not exist)}.
 
@@
 incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
-For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$.
+For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$ using the Euler-Lotka equation.
 
@@
 As expected, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
-estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially \jdnew, as they do not account for indirect spatial effects. 
+estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially, as they do not account for indirect (local) spatial effects. 
 Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
@@
 Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
-Accounting for the egocentric effect allowed us to link the results by \cite{trapman2016inferring} to established results \jdnew\ based on generation intervals.
+Accounting for the egocentric effect allowed us to link the results by \cite{trapman2016inferring} to established results based on generation intervals.
 Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
@@
 
+\swp{New paragraph; not sure if you already saw this or not}
 Contact tracing provides an effective way of collecting epidemiological data and controlling an outbreak \citep{clarke1998contact, eames2003contact, donnelly2003epidemiological}.
@@
 We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.
-This gives periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
+This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
 \begin{equation}
@@
 We then construct the random infectious period and infectious contact times simultaneously as follows.
-For each of the $n_I$ stages of the infectious period, we draw the number of effective contacts from a geometric distribution with mean $S_i \lambda / n_I\gamma$, where $S_i$ is the number
+For each of the $n_I$ stages of the infectious period, we draw the number of effective contacts (before transitioning to the next compartment) from a geometric distribution with probability $n_I\gamma/(S_i \lambda + n_I \gamma)$, where $S_i$ is the number
 of susceptible acquaintances and $\lambda$ is the per-pair contact rate.
@@
 \subsection{Egocentric relationship between $r$ and $\RR$ (SEIR model)}
+\label{egosection}
 
-Here, we rederive the egocentric relationship between $r$ and $\RR$ as presented by \cite{trapman2016inferring} (see original source for detailed derivations). 
+\swp{Can you check if this subsection is clear?}
+Here, we rederive the egocentric relationship between $r$ and $\RR$ as presented by \cite{trapman2016inferring} (see the original source for detailed derivations). 
 Assume that latent and infectious periods are exponentially distributed with mean $1/\sigma$ and $1/\gamma$, respectively.
@@
 \end{equation}
-which completes the egocentric relationship between $r$ and $\RR$:
+which completes the relationship between the growth rate and the egocentric reproductive number:
 \begin{equation}
@@
 Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.
-\swp{This is using epigrowthfit. How do I cite it? It's not out yet...}
-\dc{isn't this based on ``Estimating Initial Epidemic Growth Rates'' from Ma et al BMB 2013?}
-\swp{Yes, but they also have a package called epigrowthfit which is based on Ma et al. Maybe in acknowledgements?} \dc{Oh OK, yes}
 

### 2019-06-24 doc/network.tex
done for now

@@
 Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the censoring and the population structure near the source of disease spread; 
-therefore, we predict that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly links $r$ and $\RR$.
+therefore, we predict that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly link $r$ and $\RR$.
 We develop and test statistical methods for temporal corrections of generation intervals,
@@
 
-\swp{New}
+\swp{New notation}
 Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence};
@@
 \end{equation}
-where $\hat{f}(a)$ represents a probability density over an aspect space, including the contact structure from an egocentric point of view (i.e., the degree distribution across network). \swp{New}
+where $\hat{f}(a)$ represents a probability density over an aspect space, including the contact structure from an egocentric point of view (i.e., the degree distribution across network).
+\swp{New notation (f hat)}
 \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
-Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially \jdnew.
-Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise \jdnew.
+Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
+Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
 Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$.
@@
 
+\section*{Authors' contributions}
+
+SWP led the literature review, wrote the first draft of the MS, performed analytic calculations and simulations; JD conceived the study and performed analytic calculations; all authors contributed to refining study design, literature review, and final MS writing.
+All authors gave final approval for publication.
+
+\section*{Competing interests}
+
+The authors declare that they have no competing interests.
+
+\section*{Funding}
+
+This work was supported by the Canadian Institutes of Health Research [funding reference number 143486].
+
 \pagebreak

### 2019-06-25 doc/network.tex
minor changes for formatting
Back to Daniel

@@
 \usepackage{amsfonts}
+
+\usepackage{lineno}
+\linenumbers
 
@@
 
-\swp{Changed the last two sentences.}
 Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
@@
 and confirm our prediction using individual-based simulations on an empirical network.
+
+\section*{Keywords}
+
+Infectious disease modeling, generation interval, basic reproduction number, population structure, contact tracing
+
+\pagebreak
 
@@
 
-\swp{New notation}
 Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence};
@@
 An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
-Individual `aspects'' may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
+Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
 
@@
 where $\hat{f}(a)$ represents a probability density over an aspect space, including the contact structure from an egocentric point of view (i.e., the degree distribution across network).
-\swp{New notation (f hat)}
+\swp{New notation (f hat)}\jd{This feels like a mistake.
+We have the same population, and therefore the same distribution over aspect space, even if we are using more information to calculate the egocentric kernel.}
 \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
-\swp{Simplified.}
 % Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
@@
 we briefly describe this relationship in Section~\ref{egosection}.
-\swp{Is it OK to mention this here? Or should I say something about it at the end of the previous paragraph as well? I feel like it's too abstract to say anything there.}
+\swp{Is it OK to mention this here?
+Or should I say something about it at the end of the previous paragraph as well?
+I feel like it's too abstract to say anything there.} \jd{Seems fine this way, but let's discuss your concern.}
 
@@
 In a large homogeneously mixing population, the effective distribution is expected to be equivalent to the intrinsic distribution \swp{(since the egocentric depletion is negligible and the local depletion does not exist)}.
+\jd{This puts a focus on how we talk about global depletion. It's sort of a thing but also sort of accounted for in the equations that contain $S$. Maybe we should adddress that point explicitly.}
 
@@
 
-\swp{New paragraph; not sure if you already saw this or not}
 Contact tracing provides an effective way of collecting epidemiological data and controlling an outbreak \citep{clarke1998contact, eames2003contact, donnelly2003epidemiological}.
@@
 \swp{Can you check if this subsection is clear?}
-Here, we rederive the egocentric relationship between $r$ and $\RR$ as presented by \cite{trapman2016inferring} (see the original source for detailed derivations). 
+\jd{Seems clear enough. See framing below.}
+Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \cite{trapman2016inferring} (see the original source for detailed derivations) can be derived simply by applying the Euler-Lotka equation and using the effective (rather than the intrinsic) generation-interval distribution.
 Assume that latent and infectious periods are exponentially distributed with mean $1/\sigma$ and $1/\gamma$, respectively.
-Assuming a constant per-pair contact rate of $\lambda$ for any pair, the egocentric generation-interval distribution can be written as follows:
+Assuming a constant per-pair contact rate of $\lambda$ for any pair, the egocentric generation-interval distribution can be written:
 \begin{equation}
@@
 \end{equation}
-Substituting this equation into \eref{egorR}, we get
+Substituting into \eref{egorR}, we get
 \begin{equation}
@@
 \begin{equation}
-\hat\RR = \underbrace{\kappa \times \lambda}_{\text{average contact rate}} \times \underbrace{\frac{1}{\gamma + \lambda}}_{\text{mean effective infectious period}},
+\hat\RR = \underbrace{\kappa \lambda}_{\text{average contact rate}} \times \underbrace{\frac{1}{\gamma + \lambda}}_{\text{mean effective infectious period}},
 \end{equation}
-where $\kappa = \sigma^2/\mu + \mu - 1$, referred to as the mean degree excess \citep{newman2003structure}, describes expected number of susceptible individuals that an average infected individual will encounter early in an outbreak.
+where $\kappa = \sigma^2/\mu + \mu - 1$, referred to as the mean degree excess \citep{newman2003structure}, describes the expected number of susceptible individuals that an average infected individual will encounter early in an outbreak.
 Combining the two equations, we get

### 2019-06-25 doc/network.tex
resolving comments

@@
 \begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{k}(\tau; a) \hat{f}(a) da,
+\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da.
 \end{equation}
-where $\hat{f}(a)$ represents a probability density over an aspect space, including the contact structure from an egocentric point of view (i.e., the degree distribution across network).
-\swp{New notation (f hat)}\jd{This feels like a mistake.
-We have the same population, and therefore the same distribution over aspect space, even if we are using more information to calculate the egocentric kernel.}
 \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 
-\swp{Can you check if this subsection is clear?}
-\jd{Seems clear enough. See framing below.}
 Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \cite{trapman2016inferring} (see the original source for detailed derivations) can be derived simply by applying the Euler-Lotka equation and using the effective (rather than the intrinsic) generation-interval distribution.

### 2019-06-25 doc/network.tex
cover letter

@@
 To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
-The SEIR model describes how a disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
+The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
 We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.

### 2019-06-25 doc/network.tex
Ready to submit!!

@@
 The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
-Finally, the link between the growth rate and the egocentric reproductive number can be obtained by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
+Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
 \begin{equation}
@@
 we briefly describe this relationship in Section~\ref{egosection}.
-\swp{Is it OK to mention this here?
-Or should I say something about it at the end of the previous paragraph as well?
-I feel like it's too abstract to say anything there.} \jd{Seems fine this way, but let's discuss your concern.}
 
@@
 We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
-In a large homogeneously mixing population, the effective distribution is expected to be equivalent to the intrinsic distribution \swp{(since the egocentric depletion is negligible and the local depletion does not exist)}.
-\jd{This puts a focus on how we talk about global depletion. It's sort of a thing but also sort of accounted for in the equations that contain $S$. Maybe we should adddress that point explicitly.}
+In a large homogeneously mixing population, the effective distribution is expected to be equivalent to the intrinsic distribution (since the egocentric and local depletion are negligible in the absence of network structure, and global depletion is negligible early in the epidemic).
 

### 2019-07-02 doc/network.tex
Some cutting

@@
 while \emph{observed} generation intervals refer to the time between actual infection events.
-Observed generation intervals in turn can be measured either \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others or \emph{backward} in time by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
+%% Observed generation intervals in turn can be measured either \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others or \emph{backward} in time by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
 % When aggregated generation intervals are observed until a given time in an ongoing epidemic (often via contact tracing), we refer to these as \emph{censored} (or right-censored) intervals.
-
 % Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
@@
 % and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.
+
 While the intrinsic generation-interval distribution is often assumed to be fixed, the shape of the observed generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
-Here, we briefly describe how backward generation-interval distributions change over time because backward measurements better reflect real outbreak scenarios where contact tracing is performed by identifying infectors of infected individuals.
-(see \cite{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic} for discussions on forward generation intervals).
 When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
-A susceptible individual is thus relatively more likely to be infected by a newly infected individual, 
-and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
-When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and backward generation intervals will be longer than the intrinsic one \citep{nishiura2010time, tomba2010some, champredon2015intrinsic}.
+A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
+Thus, ``backward'' generation intervals (which look at a cohort of individuals, and ask when their infectors were infected)  will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
 
@@
 While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
-Thus, there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
-In fact, censored observations can be thought of as an average of backward generation-interval distributions, over the cohorts which have been infected before the observation time (and weighted by cohort size).
+These censored intervals are similar to backward intervals: there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
+%% In fact, censored observations can be thought of as an average of backward generation-interval distributions, over the cohorts which have been infected before the observation time (and weighted by cohort size).
+%% JD: We say this later anyway.
 
 Realized generation intervals are also affected by spatial structure.
-In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population even in the early phase of an epidemic.
+In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population. 
+%% even in the early phase of an epidemic. 
 This means that contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already) 
@@
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
-These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will give us a corrected distribution that contains information about the population structure; using this corrected distribution should allow us to correctly infer the reproductive number from the exponential growth rate.
-In other words, Trapman \textit{et al.}'s finding can be explained by the gap between the intrinsic generation interval and an ``effective'' (spatially corrected) generation interval.
+These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will yield a spatially corrected ``effective'' distribution that contains information about the population structure and will allow us to correctly infer the reproductive number from the exponential growth rate.
+%% In other words, Trapman \textit{et al.}'s finding can be explained by the gap between the intrinsic generation interval and an ``effective'' (spatially corrected) generation interval.
 
@@
 We extend previous frameworks to study how the censored generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
-We classify sources of spatial variation in generation intervals into three levels (egocentric, local, and global) and discuss how they affect observed generation-interval distributions.
+We classify spatial effects on generation intervals into three levels (egocentric, local, and global) and discuss how they affect observed generation-interval distributions.
 Finally, we compare two methods for accounting for temporal bias and test our prediction using simulations.
@@
 where $\RR = S(t) \RR_0$ is the effective reproductive number.
-This model, also referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
-Over a period of time where the proportion of susceptible $S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
+This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
+Over a period of time where the susceptible proportion $S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
 \begin{equation}
@@
 \end{equation}
-As shown in \cite{champredon2015intrinsic}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
-\begin{equation}\label{eq:backward}
-b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-\sigma) g(\sigma) d\sigma}
-\end{equation}
-On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infected at time $s$ before $t$: $\int_0^t i_{s-\tau}(s) ds$.
-Then, the censored generation-interval distribution is given by normalizing
-\begin{equation}\label{eq:obsg}
-\begin{aligned}
-c_t(\tau) 
-&= \frac{\int_0^t i_{s-\tau}(s) ds}{\int_0^t \int_0^t i_{s-\sigma}(s) ds d\sigma}= \frac{\int_0^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t \int_0^t i(s-\sigma) g(x) S(s) ds d\sigma}\\
-\end{aligned}
-\end{equation}
-Substituting \eref{backward} allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
+As shown in \cite{champredon2015intrinsic}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$.
+The censored generation-interval distribution, $c_t(\tau)$, describes \emph{all} cohorts infected at time $s<t$; thus, the censored generation-interval distribution is just the average of the backward distributions of these cohorts, weighted by incidence:
 \begin{equation}
@@
 In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
-the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
+the initial censored (or backward) generation-interval distribution is just the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
 \begin{equation}
-c_0(\tau) = b_0(\tau) \propto g(\tau) \exp(-r\tau),
+c_0(\tau) = b_0(\tau) \propto g(\tau) \exp(-r\tau).
 \label{eq:exp}
 \end{equation}
-where $c_0$ and $b_0$ represent the ``initial'' censored and backward generation-interval distributions (during the exponential growth phase), respectively.
 A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
 Second, the decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
-As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
-Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
-Overall, we expect, naively using the observed generation-interval distribution, to underestimate the reproductive number.
+Overall, we expect naively using the observed generation-interval distribution, to underestimate the reproductive number.
 
@@
 
+\jd{If $\lambda$ is constant, shouldn't that mean that $\delta(a)$ should depend on $\bar N$ and not $N(a)$?}
 Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$.
@@
 
+\jd{I don't love this analytic example. What do we need from this \P?}
 For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
@@
 Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
-(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible (simulated on a star network).
+(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible.
 (Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).

### 2019-07-03 doc/network.tex
Minor fiddling

@@
 Here, we investigate two methods for correcting for temporal bias in contact-tracing data (see Methods for details).
-We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
+We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
 When an epidemic is growing exponentially, right-censoring causes the observed generation interval to be discounted by the exponential growth rate \eref{exp};
@@
 
-We refer to the second method (Section~\ref{indlevel}) as the individual-level method because it relies on individual contact information.
+We refer to the second method as the individual-level method because it relies on individual contact information.
 We model each infection as a non-homogeneous Poisson process arising from the infector \eref{nonhomogeneous}; 
@@
 Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
-While we use the SEIR model for simulations in our main text, we show results based on more realistic distributions in Appendix.
+We show results based on more realistic distributions in Appendix.
 
@@
 
-We develop an algorithm to simulate an individual-based SEIR model on a network to keep track of generation intervals.
-Our algorithm is based on the Gillespie algorithm \citep{gillespie1977exact}.
-We model individuals with nodes on a network and their acquaintanceships with edges; infected individuals can only contact their acquaintances.
+We simulated an individual-based SEIR model on a contact network, using an algorithm based on the ``Gillespie algorithm'' \cite{Bartlett,gillespie1977exact}.
+\jd{Check Gillespie algorithm on wikipedia to find Bartlett cite.}
 
-First, we begin by randomly selecting initially infected individuals; these individuals are assumed to be infected at $t = 0$.
+We begin by randomly selecting individuals assumed to be infected at $t = 0$.
 For each infected individual $i$, we randomly draw the latent period $E_i$ from an Erlang distribution with mean $1/\sigma$ and shape $n_E$.
 We then construct the random infectious period and infectious contact times simultaneously as follows.
-For each of the $n_I$ stages of the infectious period, we draw the number of effective contacts (before transitioning to the next compartment) from a geometric distribution with probability $n_I\gamma/(S_i \lambda + n_I \gamma)$, where $S_i$ is the number
-of susceptible acquaintances and $\lambda$ is the per-pair contact rate.
+For each of the $n_I$ stages of the infectious period, we draw the number of effective contacts (before transitioning to the next compartment) from a geometric distribution with probability $n_I\gamma/(S_i \lambda + n_I \gamma)$, where $S_i$ is the number of susceptible acquaintances and $\lambda$ is the per-pair contact rate.
 We then choose the time between consecutive events (the chosen number of contacts, followed by exit from the given stage of infection) from an exponential distribution with rate $S_i \lambda + n_I\gamma$.
@@
 
-Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \cite{trapman2016inferring} (see the original source for detailed derivations) can be derived simply by applying the Euler-Lotka equation and using the effective (rather than the intrinsic) generation-interval distribution.
+Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \cite{trapman2016inferring} (see the original source for detailed derivations) matches what would be calculated by applying the Euler-Lotka equation to the effective (rather than the intrinsic) generation-interval distribution.
 Assume that latent and infectious periods are exponentially distributed with mean $1/\sigma$ and $1/\gamma$, respectively.
@@
 
-\subsection{Population-level method for estimating generation-interval distributions from contact-tracing data}
-\label{poplevel}
+\subsection{Estimating generation-interval distributions from contact-tracing data}
 
-Population-level method tries to estimate the effective generation-interval distribution by reverting the inverse exponential weighting in the observed generation-interval distribution:
+The \textbf{Population-level method} estimates the effective generation-interval distribution by reversing the inverse exponential weighting in the observed generation-interval distribution:
 \begin{equation}
@@
 
-\subsection{Individual-level method for estimating generation-interval distributions from contact-tracing data}
-\label{indlevel}
-
-We model each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), where the time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
+The \textbf{Individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
 We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
@@
 
-The entire data likelihood of the non-homogeneous poisson process can be written as:
+The likelihood of the non-homogeneous poisson process can be written as:
 \begin{equation}\label{eq:nonhomogeneous}
@@
 
-We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}: 
-\begin{equation}
-c(t) = \frac{K}{1 + \left[(K/c_0) - 1\right] e^{-rt}},
-\end{equation}
-where $K$ is the final size of an epidemic and $c_0$ is the initial number of cases.
-Fitting this curve directly to cumulative incidence can lead to overly confident result \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution.
-This assumption requires us to estimate the dispersion parameter $\theta$ of a negative binomial distribution.
-Parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
+We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}. 
+Fitting directly to cumulative incidence can lead to overly confident results \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution.
+We estimate the logistic parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
 Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.

### 2019-07-03 R/empirical.R
spending so much time on a stupid bst file

@@ -16,18 +16,6 @@ infection.generation <- function(x) {
     generation
 }
 
-##' @param x simulation result
-##' @param reference reference generation (first generation when there are n infected people)
-trapman.R0 <- function(x,
-                       reference=75,
-                       include.initial=FALSE) {
-    gen <- infection.generation(x)
-    if (!include.initial) gen <- gen[-1]
-    
-    k <- head(which(gen > reference), 1)
-    sum(gen[2:(k+1)])/sum(gen[1:k])
-}
-
 empirical.R0 <- function(x,
                          n=75) {
     o <- order(x$t_infected)

### 2019-07-03 doc/network.tex
spending so much time on a stupid bst file
supplementary material

@@
 \usepackage{natbib}
-\bibliographystyle{chicago}
+\setcitestyle{numbers} 
+\bibliographystyle{prsb}
 \date{\today}
@@
 For smaller values of \RR, we expect the differences to become smaller.
-In Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
+In the electronic supplementary material, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
 Overall, our qualitative conclusions do not change.
@@
 
-Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \cite{trapman2016inferring} (see the original source for detailed derivations) matches what would be calculated by applying the Euler-Lotka equation to the effective (rather than the intrinsic) generation-interval distribution.
+Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \cite{trapman2016inferring} (see the original source for detailed derivations) matches what would be calculated by applying the Euler-Lotka equation to the egocentric (rather than the intrinsic) generation-interval distribution.
 Assume that latent and infectious periods are exponentially distributed with mean $1/\sigma$ and $1/\gamma$, respectively.
@@
 \pagebreak
-\appendix
-\renewcommand\thefigure{\thesection.\arabic{figure}}
-\setcounter{figure}{0}    
-\section{Appendix}
 
-\subsection{Comparison of estimates of reproductive number -- smaller $\RR$}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{../fig/cmp_reproductive_small.pdf}
-\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-This figure matches \fref{cmp} in the main text.
-A smaller per-pair contact rate ($\lambda = 0.0026 \textrm{ days}^{-1}$) is used to simulate epidemics.
-All other parameters are the same as in \fref{cmp}.
-}
-\label{fig:cmpsmall}
-\end{figure}
-
-\pagebreak
-
-\subsection{Comparison of estimates of reproductive number -- Erlang distributed latent periods}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{../fig/cmp_reproductive_seminr.pdf}
-\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-This figure matches \fref{cmp} in the main text which used a SEIR model.
-Here, an Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
-All other parameters are the same as in \fref{cmp}.
-}
-\label{fig:cmpseminir}
-\end{figure}
-
-\pagebreak
-
-\subsection{Testing the individual-based method on simulations on a homogeneous network}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{../fig/full_coverage_fig.pdf}
-\caption{\textbf{Wrong distributional assumption may result in biased estimates of the parameters of a generation-interval distribution.}
-We simulate a stochastic SEIR model on a homogenous network with $10^5$ individuals using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
-Then, we apply the individual-based method based on the first 1000 infections.
-(A-B) A boxplot of 100 estimates of the mean generation intervals and their coefficient of variations (CV). 
-Dashed horizontal lines represent the true value.
-(C) Coverage probability of the mean generation intervals and their CVs based on the 95\% confidence interval.
-Dashed horizontal lines represent the 95\% coverage probability.
-(D) Estimated generation-interval distributions based on the gamma approximation and a true generation-interval distribution of an SEIR model.
-(E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.
-Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
-}
-\label{fig:cover}
-\end{figure}
-
-\pagebreak
-
-\bibliography{network}
+\bibliography{networkprsb}
 \end{document}

### 2019-07-10 doc/network.tex
edit

@@
 
-\jd{If $\lambda$ is constant, shouldn't that mean that $\delta(a)$ should depend on $\bar N$ and not $N(a)$?}
 Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$.
@@
 
-\jd{I don't love this analytic example. What do we need from this \P?}
 For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
@@
 
-We simulated an individual-based SEIR model on a contact network, using an algorithm based on the ``Gillespie algorithm'' \cite{Bartlett,gillespie1977exact}.
-\jd{Check Gillespie algorithm on wikipedia to find Bartlett cite.}
-
+We simulated an individual-based SEIR model on a contact network, using an algorithm based on the ``Gillespie algorithm'' \cite{bartlett1953stochastic,gillespie1977exact}.
 We begin by randomly selecting individuals assumed to be infected at $t = 0$.
@@
 Fitting directly to cumulative incidence can lead to overly confident results \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution.
-We estimate the logistic parameters $r$, $K$, $c_0$, and $\theta$ are estimated by maximum likelihood.
+We estimate the logistic parameters $r$, $K$, $c_0$, and $\theta$ by maximum likelihood.
 Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.

### 2019-07-10 doc/supp.tex

\documentclass[12pt]{article}
\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}

\usepackage{graphicx}
%\usepackage{adjustbox}

\newcommand{\eref}[1]{(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\setcitestyle{numbers} 
\bibliographystyle{prsb}
\date{\today}

\usepackage{xspace}
\newcommand*{\ie}{i.e.\@\xspace}

\usepackage{color}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\jdnew}{\jd{NEW}}
\newcommand{\jddel}[1]{\jd{DELETE: #1}}

\begin{document}

\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}    
\section{Appendix}

\subsection{Comparison of estimates of reproductive number -- smaller $\RR$}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive_small.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches Fig. 4 in the main text.
A smaller per-pair contact rate ($\lambda = 0.0026 \textrm{ days}^{-1}$) is used to simulate epidemics.
All other parameters are the same as in Fig. 4 in the main text.
}
\label{fig:cmpsmall}
\end{figure}

\pagebreak

\subsection{Comparison of estimates of reproductive number -- Erlang distributed latent periods}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive_seminr.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
This figure matches Fig. 4 in the main text which used a SEIR model.
Here, an Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
All other parameters are the same as in Fig. 4 in the main text.
}
\label{fig:cmpseminir}
\end{figure}

\pagebreak

\subsection{Testing the individual-based method on simulations on a homogeneous network}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../fig/full_coverage_fig.pdf}
\caption{\textbf{Wrong distributional assumption may result in biased estimates of the parameters of a generation-interval distribution.}
We simulate a stochastic SEIR model on a homogenous network with $10^5$ individuals using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. Then, we apply the individual-based method based on the first 1000 infections.
(A-B) A boxplot of 100 estimates of the mean generation intervals and their coefficient of variations (CV). 
Dashed horizontal lines represent the true value.
(C) Coverage probability of the mean generation intervals and their CVs based on the 95\% confidence interval.
Dashed horizontal lines represent the 95\% coverage probability.
(D) Estimated generation-interval distributions based on the gamma approximation and a true generation-interval distribution of an SEIR model.
(E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.
Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
}
\label{fig:cover}
\end{figure}

\pagebreak

\bibliography{networkprsb}
\end{document}

### 2019-09-19 doc/network.tex
edits

@@
 \setcitestyle{numbers} 
+\setcitestyle{square}
 \bibliographystyle{prsb}
@@
 \newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}
+\newcommand{\rev}[1]{\comment{red}{REV}{#1}}
 
@@
 \\
-Sang Woo Park\textsuperscript{1,*}
-David Champredon\textsuperscript{2,3}
-Jonathan Dushoff\textsuperscript{2,4}
+Sang Woo Park\textsuperscript{1,2,*}
+David Champredon\textsuperscript{3,4}
+Jonathan Dushoff\textsuperscript{3,5}
 \\
@@
 \\
-\textbf{2} Department of Biology, McMaster University, Hamilton, Ontario, Canada
+\textbf{2} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, New Jersey, USA
 \\
-\textbf{3} Agent-Based Modelling Laboratory, York University, Toronto, Ontario, Canada
+\textbf{3} Department of Biology, McMaster University, Hamilton, Ontario, Canada
 \\
-\textbf{4} Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
+\textbf{4} Agent-Based Modelling Laboratory, York University, Toronto, Ontario, Canada
+\\
+\textbf{5} Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
 \\
@@
 
-An epidemic can be characterized by the exponential growth rate, $r$, and the reproductive number, \RR.
+An epidemic can be described by the exponential growth rate, $r$, and the reproductive number, \RR.
 The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
@@
 \caption{\textbf{Comparison of individual- and population-level kernels.}
-(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days. 
-(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively. 
+(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
+this represents an individual realization of a random process.
+(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
+this represents a population average of a random process.
 Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
@@
 Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
+Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
+represent time relative to/from infection.
 
@@
 \end{equation}
+where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
 The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
-where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
 
@@
 \begin{equation}
-i(t) = S(t) \int K(s) i(t-s) ds = \RR \int g(s) i(t-s) ds,
+i(t) = S(t) \int K(s) i(t-s) ds = \RR(t) \int g(s) i(t-s) ds,
 \end{equation}
@@
 \begin{equation}
-i_{t-\tau}(t) = \RR \, i(t-\tau) g(\tau) S(t)
+i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
 \end{equation}
-As shown in \cite{champredon2015intrinsic}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$.
+As shown in \cite{champredon2015intrinsic}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$ with respect to $\tau$.
 The censored generation-interval distribution, $c_t(\tau)$, describes \emph{all} cohorts infected at time $s<t$; thus, the censored generation-interval distribution is just the average of the backward distributions of these cohorts, weighted by incidence:
@@
 
+\rev{
+I have a problem with N(a).
+Does your model allow for different number of acquaintances?
+But then your R_0 is no longer valid.
+Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.
+}
 Infected individuals may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
-Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average, in a limited contact network.
+Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average.
+\swp{
+To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
+Instead, we assume that a disease spreads on a network;
+infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.
+}
 
-To explore the effect of multiple contacts on realized generation intervals, we first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
+We first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
 We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case with aspect $a$ in the absence of other infectors:
@@
 
-Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$.
-In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) ds$ is the product of the number of acquaintances $N(a)$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
+Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
+In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) ds$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
 This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
@@
 
+\rev{I did not follow how you got \hat g. What is a, N(a) and f(a)? Explain more.}
+\swp{Reviewer 2 was not happy with this description. Come back.}
 The population-level egocentric kernel is found by integrating over individual variations:
 \begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da.
+\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
 \end{equation}
-\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
+where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
+Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
 The population-level egocentric generation-interval distribution is:
@@
 \end{equation}
-In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
+In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
 In practice, directly using this distribution to link $r$ and $\RR$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
@@
 
+\rev{
+what is a homogeneous network? Complete network, alla have same degree or ...? Explain better why your networks of size 5 matter in a large community (local aspects?)
+}
 The egocentric generation interval \eref{conditional} only explains some of the reduction in generation times that occurs on most networks, however.
 Generation intervals are also shortened by indirect connections: a susceptible individual can be infected through another route before the focal individual makes infectious contacts.
-Simulations on a small homogeneous network confirm this additional effect (\fref{local}, right panel).
+Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (\fref{local}, right panel). 
+\swp{
+We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
+we expect realistic local network structures (and their effects on the realized generation intervals) to lay between a star network and a complete network. 
+}
 
@@
 \section{Inferring generation-interval distributions from contact-tracing data}
+\rev{
+some of the 4 effects described also affect R whereas others don't.
+Be more clear to the reader.
+For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation.
+I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.
+}
 
@@
 \frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
-\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} I_n) && \text{for } n = 2, 3, \dots, n_I\\
+\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} - I_n) && \text{for } n = 2, 3, \dots, n_I\\
 \end{aligned}
@@
 After repeating the contact process for all initially infected individuals, all contacts are put into a sorted queue.
-The first person in the queue becomes infected, and the current time is updated to infection time of this individual. 
+The first person in the queue becomes infected (and therefore $S_i$ is decreased by 1), and the current time is updated to infection time of this individual. 
 Any subsequent contacts made to this individual are removed from the queue because they will no longer be effective.
@@
 Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
-Since the estimate of $\RR$ is sensitive to under-reporting, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate (using the Euler-Lotka equation).
+Note that the estimate of $\RR$ is sensitive to under-reporting;
+instead, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate (using the Euler-Lotka equation).
 \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

### 2019-09-23 doc/network.tex
error

@@
 \rev{
-I have a problem with N(a).
+I have a problem with $N(a)$.
 Does your model allow for different number of acquaintances?
-But then your R_0 is no longer valid.
+But then your $R_0$ is no longer valid.
 Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.
@@
 
-\rev{I did not follow how you got \hat g. What is a, N(a) and f(a)? Explain more.}
+\rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
 \swp{Reviewer 2 was not happy with this description. Come back.}
@@
 }
+\swp{They all affect R. I guess we weren't clear enough. Also, the point of our paper is that we can use 
+Euler-Lotka equations as long as we know which distribution to use. I tried to clarify these points.}
 
 When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
-We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect.
+We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
+\swp{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number.}
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
-both local and global depletion effects reduce number of infections that occur and shorten generation intervals.
+both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.
 
@@
 We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread.
-When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$.
+When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ \swp{through the Euler-Lotka equation}.
 We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
@@
 Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
+\swp{Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.}
 Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$.
@@
 Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
-Note that the estimate of $\RR$ is sensitive to under-reporting;
+We expect estimates of $\RR$ to be sensitive to under-reporting;
 instead, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate (using the Euler-Lotka equation).
-\cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
+White and Pagano \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
 

### 2019-09-23 doc/response.tex
stuff

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\begin{document}

\noindent Dear Editor:

Thank you for the chance to transfer this submission. 
Below please find our responses from the previous round.

\rev{Reviewer 1}

\revtext{This paper considers estimation of generation interval distributions from data, paying particular attention to the questions of censoring. These are important questions, and the paper clearly lays out the theory of generation times and its implications.}

\revtext{My main concern about the paper is audience; for a typical Proceedings B / applied audience, there is no work with real epidemiological data (although there is real network data); but for a more theoretical audience, the new analytic results are not quite as general as would be expected.}

\revtext{Given that the questions considered are important and worthwhile, my suggestion is that the paper should be revised with a substantial additional piece of work added involving work with a real epidemic dataset (not necessarily novel data) to demonstrate the theoretical developments, with R code in a tidier / more user-friendly format than is currently on the associated GitHub site.
}

As per editor's suggestion, we have transferred our submission to Interface journal, which better suits our manuscript in its current form. Due to time constraints, we will not be able to make substantial changes to our manuscript.

Generation-interval-based approaches (Euler-Lotak equation) are still widely used in outbreak analyses to estimate the reproduction number. While some of the analytical results that we present here (e.g., egocentric kernel) may not be as general as would be expected, we expect biological intuitions (the effects of censoring and spatial reduction on generation intervals and reproduction number) that we get from our results to be fairly general.

\revtext{Some minor comments:}

\revtext{Introduction. “Can be characterized by”. These are properties but don’t uniquely determine the epidemic.}

We have followed the reviewer's comment in changing the text:

``An epidemic can be described by the exponential growth rate, $r$, and the reproductive number, \RR.''

\revtext{Fig 1 -> two things change between each figure. I would describe the left hand as an individual realisation of the random process, and the right hand as a population-level behaviour emerging from the random process.}

We have followed the reviewer's comment in changing the figure caption:

``(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
this represents a population average of a random process.''

\revtext{Eq(15) -> typo on last line of the equations.}

Fixed.

\rev{Reviewer 2}

\revtext{The manuscript addresses very important questions. Some comments:}

\revtext{The same notation, e.g. (3) is used for equation three and reference 3 (and also some effects mention in beginning of Sec 5). Use different notation.}

Fixed.

\revtext{p5, eq 3: perhaps $\RR \rightarrow \RR(t)$?}

We made minor changes to this equation for clarity:

\begin{equation*}
i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
\end{equation*}

\revtext{p5, eq 5: I think S(t) should be removed as it is contained in R?}

This was a typo. We changed $\RR$ to $\RR_0$:

\begin{equation*}
i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
\end{equation*}

\revtext{p5-6: perhaps mention that $t, s$ is for calendar time and $\tau$ time relative to/from infection.}

We added the following sentence in the beginning of section 2 to clarify this:

``Hereafter, we use $t, s$ to represent calendar time and $\tau$ to represent time relative to/from infection.''

\revtext{p6, top: when you say and write "proportional" there are 3 time parameters. Be explicit and write: with respect to $\tau$}

Fixed.

\revtext{p8, top and later: I have a problem with $N(a)$. Does your model allow for different number of acquaintances? But then your $\RR_0$ is no longer valid. Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.}

We added the following sentences in the beginning of section 3 to make this clear:

``To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic.''

\revtext{p8, bottom: I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}

We tried to clarify this.

``... number of acquaintances $N(a)$ ...''

``The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.
Trapman \textit{et al.} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair ...''

\revtext{p9, bottom: what is a homogeneous network? Complete network, alla have same degree or ...? Explain better why your networks of size 5 matter in a large community (local aspects?)}

We tried to clarify this.

``
Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (Fig. 3, right panel). We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
we expect realistic local network structures (and their effects on the realized generation intervals) to lay between a star network and a complete network...''

\revtext{Sec 5: some of the 4 effects described also affect R whereas others don't. Be more clear to the reader. For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation. I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.}

All 4 effects affect $\RR$. Our simulations show that Euler-Lotka equation can work even in network situation, as long as we can find the proper generation-interval distribution (by using temporal correction early in an epidemic). We tried to make this point clear.

``When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number.
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.''

``When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ through the Euler-Lotka equation.''

\revtext{In your analysis you assume that you observe the sampled contact traced generation times without error. It is known (e.g. your ref 18) that onset of symptoms rather than infection time are observed, and that often there are several potential infectors. You don't have to include this in your analysis, but at least it should be acknowledged.}

We added the following sentences:

``For brevity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number.
We do not pursue these directions in this study.''

\revtext{p14, sec 7.2: the $S_i$ should vary over time. When is it measured in the algorithm?}

We added the following sentence:

``... (and therefore decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual) ...''

\revtext{p17, middle. Suddenly you talk about under-reporting. This is another very important area which would affect your analysis in many ways. Mention that you neglect this elsewhere.
}

We added the following sentence:

``Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error''

\end{document}

### 2019-09-23 doc/network.tex
stuff

@@
 \begin{equation}
-i(t) = S(t) \int K(s) i(t-s) ds = \RR(t) \int g(s) i(t-s) ds,
+i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
+\end{equation}
+This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
+Over a period of time where the susceptible proportion $S(t) = S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
+\begin{equation}
+\frac{1}{\RR} = \int g(s) \exp(-r s) ds,
 \end{equation}
 where $\RR = S(t) \RR_0$ is the effective reproductive number.
-This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
-Over a period of time where the susceptible proportion $S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
-\begin{equation}
-\frac{1}{\RR} = \int g(s) \exp(-r s) ds.
-\end{equation}
 
@@
 \rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
-\swp{Reviewer 2 was not happy with this description. Come back.}
-The population-level egocentric kernel is found by integrating over individual variations:
+The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
 \begin{equation}\label{eq:ego}
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
+\swp{Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.}.
 Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 \end{equation}
-The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
+The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
 Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
@@
 }
-\swp{They all affect R. I guess we weren't clear enough. Also, the point of our paper is that we can use 
-Euler-Lotka equations as long as we know which distribution to use. I tried to clarify these points.}
 
@@
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
-\swp{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number.}
+\swp{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number \cite{wallinga2007generation, park2019practical}.}
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
@@
 Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
-\swp{Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.}
 Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$.
@@
 and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
+\swp{
+For brevity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
+In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
+We do not pursue these direction in this study.
+}
+
 As expected, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
@@
 After repeating the contact process for all initially infected individuals, all contacts are put into a sorted queue.
-The first person in the queue becomes infected (and therefore $S_i$ is decreased by 1), and the current time is updated to infection time of this individual. 
+The first person in the queue becomes infected \swp{(and therefore decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual)}, and the current time is updated to infection time of this individual.
 Any subsequent contacts made to this individual are removed from the queue because they will no longer be effective.
@@
 
-The \textbf{Population-level method} estimates the effective generation-interval distribution by reversing the inverse exponential weighting in the observed generation-interval distribution:
+The \textbf{population-level method} estimates the effective generation-interval distribution by reversing the inverse exponential weighting in the observed generation-interval distribution:
 \begin{equation}
@@
 
-The \textbf{Individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
+The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
 We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.

### 2019-09-29 doc/response.tex
Some minor thoughts about the response and changes. One thing we should talk about.

@@
 
-As per editor's suggestion, we have transferred our submission to Interface journal, which better suits our manuscript in its current form.
-Due to time constraints, we will not be able to make substantial changes to our manuscript.
+Based on the comments of the transmitting editor, Dr.~Heesterbeek, we have decided that it makes more sense to transfer to \emph{Interface} than to engage at this time with an epidemic dataset.
+We have made an effort to make our GitHub site more user-friendly.
 
-Generation-interval-based approaches (Euler-Lotak equation) are still widely used in outbreak analyses to estimate the reproduction number.
+\jd{Not sure what this \P\ is for.
+Drop it?} Generation-interval-based approaches (Euler-Lotka equation) are still widely used in outbreak analyses to estimate the reproduction number.
 While some of the analytical results that we present here (e.g., egocentric kernel) may not be as general as would be expected, we expect biological intuitions (the effects of censoring and spatial reduction on generation intervals and reproduction number) that we get from our results to be fairly general.
@@
 
-\revtext{Fig 1 -> two things change between each figure.
+\revtext{Fig 1: two things change between each figure.
 I would describe the left hand as an individual realisation of the random process, and the right hand as a population-level behaviour emerging from the random process.}
@@
 
-\revtext{Eq(15) -> typo on last line of the equations.}
+\revtext{Eq(15): typo on last line of the equations.}
 
@@
 
-``For brevity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
+``For simplicity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
 In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number.

### 2019-09-29 doc/network.tex
Some minor thoughts about the response and changes. One thing we should talk about.

@@
 \newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
-\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
 \newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
 \newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}
-\newcommand{\rev}[1]{\comment{red}{REV}{#1}}
+
+%% \newcommand{\rev}[1]{\comment{red}{REV}{#1}}
+\newcommand{\rev}[1]{}
+
+\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
+\newcommand{\new}[1]{\textcolor{blue}{#1}}
 
@@
 Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average.
-\swp{
+\new{
 To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
@@
 }
+\jd{I feel that this requires more thought. If we're no longer assuming homogeneity, how much of what we said before still works, and how do we know we can rely on what we're relying on?}
 
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-\swp{Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.}.
+\new{Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.}.
 Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (\fref{local}, right panel). 
-\swp{
+\new{
 We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
@@
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
-\swp{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number \cite{wallinga2007generation, park2019practical}.}
+\new{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number \cite{wallinga2007generation, park2019practical}.}
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
@@
 We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread.
-When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ \swp{through the Euler-Lotka equation}.
+When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ \new{through the Euler-Lotka equation}.
 We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
@@
 and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
-\swp{
-For brevity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
+\new{
+For simplicity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
 In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
@@
 
-As expected, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
+As expected, estimating the reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
 estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially, as they do not account for indirect (local) spatial effects. 
@@
 After repeating the contact process for all initially infected individuals, all contacts are put into a sorted queue.
-The first person in the queue becomes infected \swp{(and therefore decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual)}, and the current time is updated to infection time of this individual.
+The first person in the queue becomes infected \new{(thus decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual)}, and the current time is updated to infection time of this individual.
 Any subsequent contacts made to this individual are removed from the queue because they will no longer be effective.

### 2019-10-02 doc/network.tex
Comments and changes during meeting.

@@
 Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
-represent time relative to/from infection.
+represent time relative to/from infection. \jd{Is this just time since infection?}
 
@@
 \new{
-To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
+To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous.
 Instead, we assume that a disease spreads on a network;
@@
 }
-\jd{I feel that this requires more thought. If we're no longer assuming homogeneity, how much of what we said before still works, and how do we know we can rely on what we're relying on?}
 
@@
 for brevity, we do not pursue these directions here.
+\jd{Confused now about $\delta$.}
 
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-\new{Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.}.
+\new{Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual been infected by the focal individual.}
 Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 We do not pursue these direction in this study.
-}
+} \jd{We should disentangle under-reporting from g-i estimation.}
 

### 2019-10-09 doc/response.tex
stuff

@@
 We have made an effort to make our GitHub site more user-friendly.
-
-\jd{Not sure what this \P\ is for.
-Drop it?} Generation-interval-based approaches (Euler-Lotka equation) are still widely used in outbreak analyses to estimate the reproduction number.
-While some of the analytical results that we present here (e.g., egocentric kernel) may not be as general as would be expected, we expect biological intuitions (the effects of censoring and spatial reduction on generation intervals and reproduction number) that we get from our results to be fairly general.
 
@@
 
-``Hereafter, we use $t, s$ to represent calendar time and $\tau$ to represent time relative to/from infection.''
+``Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
+represent time since infection.''
 
@@
 
-``To explore the effect of multiple contacts on realized generation intervals, we no longer assume that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
+``To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
 Instead, we assume that a disease spreads on a network;
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-Essentially, the population-level egocentric kernel further accounts for the probability that a susceptible individual has not been infected within an infected-susceptible pair.
+Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
 Trapman \textit{et al.} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 
-``For simplicity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
-In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number.
-We do not pursue these directions in this study.''
+``For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting. 
+We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
+On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
+We do not pursue these direction in this study.''
 
@@
 
-``... (and therefore decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual) ...''
+``... (thus decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual) ...''
 

### 2019-10-09 doc/network.tex
Autocommit (doc)
stuff

@@
 Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
-represent time relative to/from infection. \jd{Is this just time since infection?}
+represent time since infection.
 
@@
 for brevity, we do not pursue these directions here.
-\jd{Confused now about $\delta$.}
 
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-\new{Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual been infected by the focal individual.}
+\new{Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.}
 Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 \new{
-For simplicity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
-In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
+For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting. 
+We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
+On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
 We do not pursue these direction in this study.
-} \jd{We should disentangle under-reporting from g-i estimation.}
+}
 

### 2019-10-09 R/empirical.R
tiny comment

@@ -16,6 +16,7 @@ infection.generation <- function(x) {
     generation
 }
 
+## estimating empirical R0
 empirical.R0 <- function(x,
                          n=75) {
     o <- order(x$t_infected)

### 2019-10-09 doc/response.tex
update readme

@@
 We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
-On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
+On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number.
 We do not pursue these direction in this study.''

### 2020-05-10 doc/response.tex
don't remember these changes...

@@
 We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
-On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number.
+On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproduction number.
 We do not pursue these direction in this study.''
@@
 
-``Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error''
+``For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting.''
 

### 2020-05-10 doc/network.tex
don't remember these changes...

@@
 
-*parksw3@mcmaster.ca
+*swp2@princeton.edu
 \end{flushleft}
@@
 \end{equation}
-This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
+This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
 Over a period of time where the susceptible proportion $S(t) = S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
@@
 Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average.
-\new{
 To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous.
@@
 infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.
-}
 
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-\new{Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.}
+Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
 Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
@@
 \begin{equation}
-g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma t} - e^{-\sigma t}\right),
+g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma \tau} - e^{-\sigma \tau}\right),
 \end{equation}
@@
 Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (\fref{local}, right panel). 
-\new{
 We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
 we expect realistic local network structures (and their effects on the realized generation intervals) to lay between a star network and a complete network.
-}
 
@@
 Both the egocentric and local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
-Finally, global depletion refers to overall depeletion of susceptibility at the population level, and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
+Finally, global depletion refers to overall depletion of susceptibility at the population level, and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).
 
@@
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
-\new{we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number \cite{wallinga2007generation, park2019practical}.}
+we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number \cite{wallinga2007generation, park2019practical}.
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
@@
 We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread.
-When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ \new{through the Euler-Lotka equation}.
+When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ through the Euler-Lotka equation.
 We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
@@
 and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
-\new{
 For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting. 
 We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
-On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
+On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproduction number \cite{britton2019estimation}.
 We do not pursue these direction in this study.
-}
 
@@
 To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
-The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infeceted individuals become infectious after a latent period.
+The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infected individuals become infectious after a latent period.
 We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.
@@
 After repeating the contact process for all initially infected individuals, all contacts are put into a sorted queue.
-The first person in the queue becomes infected \new{(thus decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual)}, and the current time is updated to infection time of this individual.
+The first person in the queue becomes infected (thus decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual), and the current time is updated to infection time of this individual.
 Any subsequent contacts made to this individual are removed from the queue because they will no longer be effective.
@@
 
-The likelihood of the non-homogeneous poisson process can be written as:
+The likelihood of the non-homogeneous Poisson process can be written as:
 \begin{equation}\label{eq:nonhomogeneous}

### 2020-05-11 doc/response.tex
edits

--- 2019-10-09 doc/response.tex
@@
 \usepackage{color}
+
+\usepackage{natbib}
+\setcitestyle{numbers} 
+\setcitestyle{square}
+\bibliographystyle{prsb}
 
@@
 
-Thank you for the chance to transfer this submission. 
-Below please find our responses from the previous round.
+Thank you for the chance to revise and resubmit.
+We apologize for the delay.
+Reviewer 1 suggested that we should divide the paper into two: theory and application.
+Given the relevance of this manuscript to the current COVID-19 outbreak, we think it is best to write an application-based one first (this one) and then write a theory-based on later.
+We believe that incorporating some theory with application will help readers understand the material better at the moment.
+Otherwise, we have tried our best to incorporate the reviewers' suggestions.
+See below for detailed response.
 
-\rev{Reviewer 1}
+\rev{Reviewer \#1}
 
-\revtext{This paper considers estimation of generation interval distributions from data, paying particular attention to the questions of censoring.
-These are important questions, and the paper clearly lays out the theory of generation times and its implications.}
+\revtext{The main objective of this paper is to correct selection bias in observational
+(contact tracing) data to yield unbiased estimate of the "true" distribution
+corresponding to generation intervals defined in Lines 42-43 (if I understand
+properly).}
 
-\revtext{My main concern about the paper is audience; for a typical Proceedings B / applied audience, there is no work with real epidemiological data (although there is real network data); but for a more theoretical audience, the new analytic results are not quite as general as would be expected.}
+The main objective is to estimate the effective generation-interval distribution, which implicitly accounts for spatial structure. We tried to make this clear now.
 
-\revtext{Given that the questions considered are important and worthwhile, my suggestion is that the paper should be revised with a substantial additional piece of work added involving work with a real epidemic dataset (not necessarily novel data) to demonstrate the theoretical developments, with R code in a tidier / more user-friendly format than is currently on the associated GitHub site.
+\revtext{Lines 42-43: "A generation interval is defined as the time between when a
+person becomes infected and when that person infects another person."
+(with reference [10]). Indeed, most of the cited literature in this paper
+start with the definition exactly the same way. I feel that this definition
+is imprecise and is not as obvious as it reads, especially in its relationship
+with the intrinsic generation interval (Line 46) and its distribution (Lines
+49,101). It is not proven that they are equivalent.}
+
+We agree that this definition may be imprecise but we feel that it may be clearer for the readers to begin with a familiar definition. 
+Instead, we now point to imprecision of this definition.
+We also make clear that this definition corresponds to the definition at the individual-level (i.e., a single interval) and distinguish it from the population-level distributions:
+
+``At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
+while this definition is widely used in the literature, it is not connected directly to a population-level distribution.''
+
+\revtext{Line 46: "...intrinsic generation intervals..." This is not yet defined. Is it
+equivalent to the definition in Lines 42-43?}
+
+We now define intrinsic generation interval properly:
+
+``\emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population...''
+
+\revtext{Line 46: "......intrinsic generation intervals measure the infectiousness of an
+infected individual." This statement is vague. I will get back to this phrase
+a few pages later in the discussion on "equilibrium distributions" and
+suggest ways to make it more precise.}
+
+\revtext{Lines 49,101: "...the intrinsic generation interval distribution..." While the
+interval itself is not defined in Line 46 (in my opinion), its distribution is
+clarified in Line 101 in relation to equation (2) under homogeneous mixing.
+I hereby pose the following questions:}
+
+\revtext{
+1. Although $g(\tau) > 0$ satisfies all the properties of a probability density
+function (p.d.f.) of a non-negative continues random variable, does that
+automatically correspond to a specific random variable relevant to any of
+the observable (or conceptually observable) "generation intervals"?
 }
 
-Based on the comments of the transmitting editor, Dr.~Heesterbeek, we have decided that it makes more sense to transfer to \emph{Interface} than to engage at this time with an epidemic dataset.
-We have made an effort to make our GitHub site more user-friendly.
+We now define $g(\tau)$ as follows: ``Expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population''
 
-\revtext{Some minor comments:}
-
-\revtext{Introduction. “Can be characterized by”. These are properties but don’t uniquely determine the epidemic.}
-
-We have followed the reviewer's comment in changing the text:
-
-``An epidemic can be described by the exponential growth rate, $r$, and the reproductive number, \RR.''
-
-\revtext{Fig 1: two things change between each figure.
-I would describe the left hand as an individual realisation of the random process, and the right hand as a population-level behaviour emerging from the random process.}
-
-We have followed the reviewer's comment in changing the figure caption:
-
-``(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
-this represents an individual realization of a random process.
-(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
-this represents a population average of a random process.''
-
-\revtext{Eq(15): typo on last line of the equations.}
-
-Fixed.
-
-\rev{Reviewer 2}
-
-\revtext{The manuscript addresses very important questions. Some comments:}
-
-\revtext{The same notation, e.g. (3) is used for equation three and reference 3 (and also some effects mention in beginning of Sec 5). Use different notation.}
-
-Fixed.
-
-\revtext{p5, eq 3: perhaps $\RR \rightarrow \RR(t)$?}
-
-We made minor changes to this equation for clarity:
-
-\begin{equation*}
-i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
-\end{equation*}
-
-\revtext{p5, eq 5: I think S(t) should be removed as it is contained in R?}
-
-This was a typo. We changed $\RR$ to $\RR_0$:
-
-\begin{equation*}
-i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
-\end{equation*}
-
-\revtext{p5-6: perhaps mention that $t, s$ is for calendar time and $\tau$ time relative to/from infection.}
-
-We added the following sentence in the beginning of section 2 to clarify this:
-
-``Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
-represent time since infection.''
-
-\revtext{p6, top: when you say and write "proportional" there are 3 time parameters. Be explicit and write: with respect to $\tau$}
-
-Fixed.
-
-\revtext{p8, top and later: I have a problem with $N(a)$.
-Does your model allow for different number of acquaintances?
-But then your $\RR_0$ is no longer valid.
-Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.}
-
-We added the following sentences in the beginning of section 3 to make this clear:
-
-``To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
-Instead, we assume that a disease spreads on a network;
-infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic.''
-
-\revtext{p8, bottom: I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
-
-We tried to clarify this.
-
-``... number of acquaintances $N(a)$ ...''
-
-``The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
-\begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
-\end{equation}
-where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
-Trapman \textit{et al.} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
-The population-level egocentric generation-interval distribution is:
-\begin{equation}
-\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
-\label{eq:conditional}
-\end{equation}
-The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair ...''
-
-\revtext{p9, bottom: what is a homogeneous network? Complete network, alla have same degree or ...? Explain better why your networks of size 5 matter in a large community (local aspects?)}
-
-We tried to clarify this.
-
-``
-Simulations on a small homogeneous network (i.e., complete network) confirm this additional effect (Fig.
-3, right panel).
-We can think of simulations on this network as an approximation of (local) infection process in a small household, consisting of 5 individuals;
-we expect realistic local network structures (and their effects on the realized generation intervals) to lay between a star network and a complete network...''
-
-\revtext{Sec 5: some of the 4 effects described also affect R whereas others don't.
-Be more clear to the reader.
-For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation.
-I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.}
-
-All 4 effects affect $\RR$.
-Our simulations show that Euler-Lotka equation can work even in network situation, as long as we can find the proper generation-interval distribution (by using temporal correction early in an epidemic).
-We tried to make this point clear.
-
-``When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
-We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
-we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number.
-While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
-both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.''
-
-``When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ through the Euler-Lotka equation.''
-
-\revtext{In your analysis you assume that you observe the sampled contact traced generation times without error.
-It is known (e.g.
-your ref 18) that onset of symptoms rather than infection time are observed, and that often there are several potential infectors.
-You don't have to include this in your analysis, but at least it should be acknowledged.}
-
-We added the following sentences:
-
-``For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting. 
-We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
-On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number.
-We do not pursue these direction in this study.''
-
-\revtext{p14, sec 7.2: the $S_i$ should vary over time. When is it measured in the algorithm?}
-
-We added the following sentence:
-
-``... (thus decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual) ...''
-
-\revtext{p17, middle. Suddenly you talk about under-reporting. This is another very important area which would affect your analysis in many ways. Mention that you neglect this elsewhere.
+\revtext{
+2. Is it the p.d.f. corresponding to the definition in Lines 42-43? Proof?
 }
 
-We added the following sentence:
+Definition in Lines 42-43 does not correspond to the definition of the intrinsic generation-interval distribution. We have tried to make this clearer in the text.
 
-``Here, we assume that \emph{all} generation intervals (no under-reporting) are observed without error''
+\revtext{
+Lines 44-48: "... the observed generation-interval distribution can change...
+...observed generation intervals refer to the time between actual infection
+events." The actual infection events correspond to the exact moments of
+passing the infection between a pair of infected-susceptible individuals.
+Are they observable? In most cases, including contact tracing, observable
+events are clinical (e.g. onset of symptoms). The relationship between
+the interval deÖned by the time between actual infection events and the
+interval deÖned by the time between clinical onsets is unclear. At least the
+latter depends on the incubation period (of the infectee), in addition to the
+latent period and the infectious period (of the infector). The incubation
+period is defined as the time from infection to symptom onset, and is
+different from the latent period such as in the SEIR model. This paper
+does not make such distinction. It assumes that the time between actual
+infection events "were observable" in a hypothetical way.
+}
 
+\revtext{
+Suggestion: Re-phrase "observed generation intervals" or make some further
+clarification on this. This is important because naive readers may tend
+to take the face value and collect different kinds of observed data and
+make inferences, such as that based on clinical events. This is potentially
+misleading.
+}
+
+We agree that calling it ``observed'' can be misleading. We now refer to them as \emph{realized} intervals.
+
+``There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population,
+while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
+Due to individual variation in infection time, the realized generation-interval distribution (at the population-level) can change depending on when and how it is defined \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.''
+
+\revtext{Lines 54-55: ``the backward generation interval...'' This is a key measurement in this paper along with the length-bias in relation to data observation scheme.
+Conceptually, there may also be a "forward" generation
+interval. Both concepts were initially raised in Svensson (2007, ref.[21]
+in the paper), who pointed out the difference in distribution between the
+primary generation time as measured prospectively from the time of the
+infection of the infector to the transmission to the infectee, and the secondary generation time as measured retrospectively from the time of the
+transmission to the infectee to the infection of the infector.}
+
+We now introduce forward.
+
+\revtext{1. does that mean that your objective is to make unbiased inference on the
+"forward", or the primary, generation time?}
+
+Thank you for pointing this out. This is what we want. We have tried to clarify this throughout the manuscript. We no longer use the term ``effective'' generation-interval distribution.
+
+\revtext{2. does the definition in Lines 42-43 correspond to the forward generation
+interval?}
+
+No.
+
+\revtext{3. does the definition in Lines 42-43 correspond to the intrinsic generation
+interval (Line 46)?}
+
+No.
+
+\revtext{4. does the definition in Lines 42-43 correspond to a random variable with
+p.d.f. given by $g(\tau)$?}
+
+No.
+
+\revtext{5. do you imply that the intrinsic generation interval (not defined in Line 46
+but distributed according to Line 101) and the forward measure, and the
+objective of the unbiased inference, are all equivalent?}
+
+No.
+
+\bibliography{networkprsb}
 \end{document}

### 2020-05-11 doc/network.tex
edits

@@
 
-A generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
-Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.
-There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals measure the infectiousness of an infected individual,
-while \emph{observed} generation intervals refer to the time between actual infection events.
-%% Observed generation intervals in turn can be measured either \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others or \emph{backward} in time by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
-% When aggregated generation intervals are observed until a given time in an ongoing epidemic (often via contact tracing), we refer to these as \emph{censored} (or right-censored) intervals.
-% Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
-% As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
-% and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.
+At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
+While this definition is widely used in the literature, it is not connected directly to a population-level distribution.
+There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case,
+while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
+Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.
 
-While the intrinsic generation-interval distribution is often assumed to be fixed, the shape of the observed generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+While the shape of intrinsic generation-interval distribution is often assumed to be invariant across time, the shape of the realized generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
 When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
 A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
-Thus, ``backward'' generation intervals (which look at a cohort of individuals, and ask when their infectors were infected)  will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
+Thus, ``backward'' generation intervals (which look at a cohort of infectees, and ask when their infectors were infected) will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
+Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
+Forward generation intervals can contract over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.
 
-In practice, generation intervals are often measured by aggregating available information from contact tracing. 
-%% Whereas backward generation intervals represent infections that are one generation apart from a cohort of infectees, these aggregated generation intervals represent infections across multiple cohorts (i.e., the time of infection of the infectees can range anywhere from the beginning of an epidemic to the most recent infection).
+Forward generation intervals are also affected by spatial structure.
+In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population. 
+This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
+As a result, forward generation intervals will have shorter mean than the intrinsic generation-interval distribution.
+This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
+
+In practice, realized generation intervals are often measured by aggregating available information from contact tracing.
 While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
 These censored intervals are similar to backward intervals: there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
-%% In fact, censored observations can be thought of as an average of backward generation-interval distributions, over the cohorts which have been infected before the observation time (and weighted by cohort size).
-%% JD: We say this later anyway.
-
-Realized generation intervals are also affected by spatial structure.
-In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population. 
-%% even in the early phase of an epidemic. 
-This means that contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already) 
-As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
-This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
 These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will yield a spatially corrected ``effective'' distribution that contains information about the population structure and will allow us to correctly infer the reproductive number from the exponential growth rate.
-%% In other words, Trapman \textit{et al.}'s finding can be explained by the gap between the intrinsic generation interval and an ``effective'' (spatially corrected) generation interval.
+We further predict the effective generation-interval distribution to be equivalent to the \emph{initial} forward generation-interval distribution, which describes the renewal process of infection during the exponential growth rate.
 

### 2020-05-20 doc/response.tex
working on it...

@@
 \newcommand{\Tc}{\ensuremath{C}}
+
+\newcommand{\Rini}{\Rx{\textrm{\tiny initial}}}
 
@@
 
-The main objective is to estimate the effective generation-interval distribution, which implicitly accounts for spatial structure. We tried to make this clear now.
+Lines 42-43 refers to a single measurement, rather than a distribution. We have made this clear (see next comment).
 
@@
 
-We agree that this definition may be imprecise but we feel that it may be clearer for the readers to begin with a familiar definition. 
-Instead, we now point to imprecision of this definition.
-We also make clear that this definition corresponds to the definition at the individual-level (i.e., a single interval) and distinguish it from the population-level distributions:
+We make clear that this definition corresponds to the definition at the individual-level (i.e., a single interval), which is unambiguous, and distinguish it from the population-level distributions:
 
-``At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
-while this definition is widely used in the literature, it is not connected directly to a population-level distribution.''
+``At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
+While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.''
 
@@
 
-``\emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population...''
+``The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.''
 
@@
 suggest ways to make it more precise.}
+
+We have now made this definition more precise. In Section 2, we also introduce the idea that the intrinsic distributions can be thought of as equilibrium distributions:
+
+``$g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
+Since all infectious contacts lead to infection in a disease-free equilibrium of a homogeneously mixing population, $g(\tau)$ also describes the expected time distribution of secondary cases caused by an average primary case in a fully susceptible population and can be treated as an ``equilibrium'' distribution of the corresponding homogeneous mixing model.''
 
@@
 
-We now define $g(\tau)$ as follows: ``Expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population''
+We now define $g(\tau)$ as follows: ``expected time distribution of infectious contacts by a primary case''.
 
@@
 
-Definition in Lines 42-43 does not correspond to the definition of the intrinsic generation-interval distribution. We have tried to make this clearer in the text.
+Definition in Lines 42-43 does not correspond to the definition of the intrinsic generation-interval distribution. 
+It is a single measurement --- we tried to make it clear that it is not directly connected to any p.d.f.
+We have tried to make this clearer in the text.
 
@@
 events are clinical (e.g. onset of symptoms). The relationship between
-the interval deÖned by the time between actual infection events and the
-interval deÖned by the time between clinical onsets is unclear. At least the
+the interval defined by the time between actual infection events and the
+interval defined by the time between clinical onsets is unclear. At least the
 latter depends on the incubation period (of the infectee), in addition to the
@@
 
+We have also added the following sentence as per suggestions in previous comments:
+
+``While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
+we typically want to try to make inference based on all available data.''
+
 \revtext{Lines 54-55: ``the backward generation interval...'' This is a key measurement in this paper along with the length-bias in relation to data observation scheme.
@@
 
-We now introduce forward.
+We now introduce the forward distribution as well.
 
@@
 
-Thank you for pointing this out. This is what we want. We have tried to clarify this throughout the manuscript. We no longer use the term ``effective'' generation-interval distribution.
+Yes, we want to make inference on the \emph{initial} forward generation-interval distribution, rather than the intrinsic generation-interval distribution.
+We have tried to make this clear and no longer use the term ``effective'' generation-interval distribution.
 
@@
 
-No.
+No. Lines 42-43 is simply a single measurement and does not correspond to any specific generation interval.
 
@@
 
-No.
+No. See above.
 
@@
 
-No.
+No. The intrinsic distributions does not account for any spatial effect and therefore are different from the forward distributions.
+We have tried to make this clear throughout the manuscript.
+In particular, we provide the following sentence in Introduction to distinguish intrinsic from the realized distributions:
+
+``Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.''
+
+\revtext{
+The "the backward generation interval (or the secondary generation time)" is statistically sound. An infectee, before being infected, has been exposed to many infectors in a prevalent cohort and
+could be independently infected by any of them. Each infectee is only associated with a single backward generation time whose distribution arises
+from an independent competing risk framework. At the population level,
+all backward measured intervals are independent random variables (but
+not necessarily identically distributed). On the other hand, the "forward
+generation interval" is problematic. A typical infected individual may
+infect several contacts, but not independently if the infectious period is
+random. Therefore, it is not appropriate to consider that the times to
+transmission of the pathogen to its infectees as an iid. sample. This can
+be shown in Exercise 3.5 (f) of Yan and Chowell (2019). At the population level, should such data can be observed, there might exist correlated
+intervals arising from the same infector, something to ponder about.
+}
+
+\swp{I don't know.}
+
+\revtext{This paper addresses length-bias in the "backward generation interval" and
+different ways of adjusting this bias, assuming the intervals were observable
+especially in the contact tracing setting. This is scientifically important, not
+only under the context of the current paper, but also applicable in many other
+contexts in science. Several issues need to be discussed.}
+
+Thank you. We have tried to address the issues outlined in the review.
+
+\revtext{
+Line 43: "...individual variation in infection time..." I presume the authors
+meant individuals infected at different (calendar) times in the course of
+an outbreak. If this is the case, I suggest re-wording to avoid confusion
+}
+
+We have removed this sentence.
+
+\revtext{
+Lines 59, 114-115, 129, etc.: There is a distinction between "truncation"
+and "censoring" in survival analysis. In the current context (e.g. contact tracing), when the inclusion criteria is conditioning on the infection
+event of the infectee and looking backward to ascertain the infection time
+of its infector, it is right-truncation. It leads to over-sampling short durations (as stated in Line 129). I suggest to use "right-truncation" instead
+of "censoring" throughout the paper, wherever it appears.
+}
+
+We now use truncation instead of censoring.
+
+\revtext{
+Lines 121 and 123: The backward generation-interval distribution $b_t(\tau)$ and
+the "censored" (right-truncated?) generation-interval distribution $c_t(\tau)$
+should be clearly defined and expressed (e.g. in terms of conditional distribution) before expressing the relationship (6).
+}
+
+Done.
+
+\revtext{
+Lines 233-241: Once more, the sampling bias corresponds to right-truncation,
+not censoring. A great deal of words is involved leading to gef $g_{eff}(\tau)$. The
+mathematical language is more powerful and perhaps gef $g_{eff}(\tau)$ could be
+explained more concisely (and take less space?)
+}
+
+We have removed this paragraph and tried to make the explanation more concise.
+We no longer use the term effective generation-interval distribution.
+
+\rev{Lines 241-243: "In a large homogeneously mixing population, the e§ective
+distribution is expected to be equivalent to the intrinsic distribution..."
+Define both $g(\tau)$ and gef $g_{eff}(\tau)$ mathematically at the first place and prove
+their equivalence in a large homogeneously mixing population.}
+
+Done.
+
+\rev{However, I suggest the authors to compare the "individual-based
+method" in the paragraph starting from Line 252 to Line 260 with the statistical methods for analyzing right-truncated survival data as mentioned
+above, and explicitly write out the likelihood function (Line 258) to see if
+they are the same or different.}
+
+\rev{The authors has used the notations $\RR$ (Line 35) and $\Ro$ (Lines 98-99) for reproduction numbers and
+linked them via $\RR = \Ro S(t)$:}
+
+\rev{Shouldnít $\RR$ be a function of time $t$ via $S(t)$?}
+
+We no longer use $\RR$ to avoid confusion. We now use $\Rini = \Ro S(0)$ to reflect the possibility that $S(0)$ may not be 1.
+
+\rev{Is $S(t)$ a number or a proportion? In deterministic models, this matters
+less as one can argue that all deterministic models are describing the
+change of proportions (by re-scaling on population size, assuming it is
+constant); but it matters a lot in stochastic models because the population
+size is an essential parameter.}
+
+Proportion. Made this clear.
+
+\rev{
+When stating that ``$\RR$'' transcends to the final size of the epidemic (Line
+37), do the authors actually mean $\Ro$; which does not depend on time?
+}
+
+Yes.
+
+\rev{
+When talking about the final size, one considers that the system is at a
+disease-free equilibrium. When using the final size equation to calculate a
+value for the reproduction number, the calculated value does not depend
+on time. If there are behavior changes and public health interventions
+throughout the outbreak, the calculated value $\RR_{final}$ is usually smaller
+than $\Ro$ (defined at another disease-free equilibrium at the very moment
+when an infector was seeded).
+So now we have three reproduction numbers, $\Ro$, a time-dependent $\RR = \Ro S(t)$ and $\RR_{final}$: Need to clarify the differences and reveal the hidden assumptions when discussing (and using)
+each of them.
+}
+
+Discussion!
+
+\rev{
+The authors stated that $\RR$ transcends to the final size of the epidemic
+(with cited references). I boldly assume that the authors actually meant
+by $\Ro$: This statement is in the context of a large outbreak that ends
+when either there is no infectious individuals or there is no susceptible
+individuals left in the population. Under a different context, the outbreak
+may reach an endemic steady state with sustained on-going transmission
+persisting for a very long period of time due to re-infection or demography.
+It could be further said that $\Ro$ can transcend to the long term endemicity
+level, such as $S(\infty) = \RR$: I leave to the authors to search the literature
+for the latter
+}
+
+Yes, we meant $\Ro$. We have now clarified this in the text.
+We now mention in Introduction that $\Ro$ can also affect the endemicity level and added a few appropriate citations.
+
+\rev{$g(\tau)$ as a p.d.f. given by (2) and Line 101 is also an equilibrium
+distribution.}
+
+We disagree with the reviewer that the intrinsic generation-interval is an equilibrium distribution.
+It is only an equilibrium distribution for a model that assumes a large, homogeneously mixing population.
+As we show in this paper, when there is spatial structure, the realized generation-interval distributions differ from the intrinsic one. 
+In particular, even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egeocentric effect.
+
+\rev{ Under suitable assumptions on homogeneous mixing, the intrinsic
+generation interval distribution $g(\tau)$ measures the infectiousness within
+an infected individual without involving another individual. (In SEIS,
+SEIR, SEIRS settings, it is a function of the latent period distribution
+and the infectious period distribution.)}
+
+We agree that $g(\tau)$ measures the infectiousness within an infected individual without involving another individual.
+We defined the intrinsic generation-interval distribution as follows:
+
+``The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.''
+
+\revtext{The distribution $g(\tau)$ links the two intrinsic parameters R0 and r
+(homogeneous mixing). Both $\Ro$ and $g(\tau)$ are intrinsic in the context of the equilibrium conditions. Therefore the parameter $r$ is a
+theoretical value and is very di§erent from any observed exponential
+or sub-exponential growth rate as the system is not at equilibrium.}
+
+\rev{Observed exponential or sub-exponential growth events are different from the infection events. The current paper only addresses the infection events (mostly unobservable). This was
+already addressed in the comments with respect to Line 47-48.
+Any attempt to link the estimated $r$ based on data collected
+on observed events using the Euler-Lotka equation could be extremely misleading}
+
+We point out in the Discussion that estimated $r$ can be different from the true $r$. 
+However, it is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
+See:
+
+``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''
+
+\rev{
+The intrinsic parameters R0 and r are parameters in branching
+process, which is used to approximate the "initial phase" of the
+outbreak and that approximation is poor. By the way, the condition in Lines 107-108 $S(t) = S$ is also inaccurate. The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
+$n \to \infty$; where $n$ is the population size. The so-called initial
+phase (when the depletion of the susceptible population is negligible) totally ignores the difference between the initial disease free equilibrium and the highly explosive dynamic growth
+}
+
+``The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
+$n \to \infty$'' 
+This is inaccurate.
+It is still possible to observe exponential growth when the population is not fully susceptible ($S(t) \approx S(0) \neq 1$) as long as the depletion of the susceptible population is negligible.
+To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.
+
+\rev{
+ It is too strong a statement that "the Erlang distributions are
+more realistic than the exponentially distributed periods in the standard
+SEIR model." The advantage of using the Erlang distributions is that the
+system can be analyzed using ordinary differential equations. They are
+only suitable for latent and infectious periods with very small variances
+}
+
+The exponential distribution is a special case of the Erlang distribution (when the shape parameter is equal to 1).
+We disagree the statement that it is ``too strong'' to say that a more general distribution is more realistic than a special case.
+To avoid confusion, we now say refer to Erlang distributions as ``more realistic approximations''.
+
+\rev{
+The Erlang distribution is not only a subset of the gamma distribution with integer valued shape parameters, but also it is limited to distributions with variances smaller than the corresponding exponential distribution with the same mean value.
+}
+
+The Erlang distribution includes the exponential distribution.
+
+\rev{
+One can not even claim the gamma distributions are more realistic
+without supported by data. For one thing: all gamma distributions
+have exponential tails which may be contested in real data.
+}
+
+One can at least claim that the gamma distribution is more realistic than its subset, the exponential distribution.
+
+\rev{
+It is more convenient to use the Laplace transforms as a tool.
+}
+
+The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial components.
+The Laplace tranforms are only more useful for the intrinsic generation-interval distributions. 
+The forward generation-interval distribution is analytically intractable and therefore, the Laplace transforms no longer hold.
+Given that our goal is to infer the initial forward generation-interval distribution, rather than the intrinsic distribution, we feel that the Laplace transforms are not necessarily helpful for the discussion.
+In particular, the reviewer asks us to prove special cases of the Euler-Lotka equation, which has been proven numerous times as suggested by the references, using Laplace transforms; 
+we do not think this is any relevant to the discussion of our paper.
+We cite all Yan et al papers that were mentioned in the review in our main text.
+
+\rev{
+How to assess if the gamma distribution is a good choice as a
+model? Run some kind of goodness-of-fit tests? How does the likelihood
+function look like? Since the "observed" generation-intervals (assuming
+observable) do not arise from a designed random experiment, but observational data, I expect the likelihood function should reflect the observation
+scheme (e.g. retrospective)? Please be explicit.
+}
+
+\rev{
+Elaborate how the estimation of $\RR$ is expected to be sensitive to under-reporting (of which data)
+}
+
+\rev{
+How exactly is $\RR$ (and, is it \Ro or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref. [54] in the paper)?
+}
+
+``the inferred generation-interval distribution is then used to estimate the reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.''
+
+\rev{
+How close is the "related approach" in White and Pagano
+(2007, ref. [54] in the paper) related to the current paper? Did they use the
+Euler-Lotka equation at all? Did they define and measure the generation intervals that were consistent with the generation-intervals discussed in
+the current paper?
+}
+
+\rev{
+The authors use a logistic function with a scale parameter $r$,
+the carrier capacity parameter $K$ and the initial condition $c_0$ to model the
+cumulative incidence $c(t)$: Assuming, (theoretically), the infection events
+are observable. It is inaccurate to claim that r is the exponential growth
+rate
+}
+
+\rev{
+The paper focuses on estimating the generating-interval distribution. There are
+several generating-intervals and distributions mentioned in this paper, such as
+$g(\tau)$; $b_t(\tau)$; $c_t(\tau)$; $g_eff(\tau)$; etc. To the end, it seems that the primary focus is the
+intrinsic distribution $g(\tau)$; and relating this distribution with two other intrinsic
+parameters: the reproduction number $\Ro$ and the Malthusian number $r$
+}
+
+The intrinsic distribution is not our primary focus. 
+As we show in Figure 4, the intrinsic distribution overesitmates $\Rini$ as it does not account for spatial effects.
+We tried to make clear that our goal is the initial forward distribution.
+
+\rev{
+The authors have well demonstrated different sources of sampling bias
+arising from contact tracing data, more noticeably, (i) length-bias due
+to right-truncation because of retrospective measurement together with
+different infection times during the course of the outbreak; (ii) due to the
+network structure (e.g. the degree distribution and the degree excess (Line
+408). Both sources, the right-truncation and the egocentric point of view
+(Line 174), tend to give shorter on average observed intervals. These are
+correct and important.
+}
+
+\rev{
+Any notion of bias is in the context against some "truth". The authors
+proposed, and demonstrated by simulation, ways to adjust for the bias in
+order to estimate this "truth". However, the authors equate the "true"
+distribution to the intrinsic distribution $g(\tau)$ as if "obvious". It is not
+obvious and this needs to be proven.
+}
+
+We had never equated the true distribution to the intrinsic distribution.
+We make clear that we do not want the intrinsic distribution.
+
+\rev{
+The intrinsic distribution $g(\tau)$ is an equilibrium distribution that
+involves the equilibrium distribution of the infectious period (in SEIS,
+SEIR, SEIRS settings). It is a renewal process concept and is only
+valid when the system is under the equilibrium condition. It is not
+obvious how it is related to measurements of "generation intervals"
+when the system is not at equilibrium.
+}
+
+\rev{
+ Under suitable assumptions on homogeneous mixing, the intrinsic
+distribution $g(\tau)$ measures the infectiousness within an infected individual without involving another individual. It is not obvious how it
+is related to measures involving two individuals, no matter whether
+they are prospectively or retrospectively measured.
+}
+
+\rev{
+Potentially observable events do not take place at the exact moment
+when transmission occurs.
+}
+
+\rev{
+I rate this paper very high in scientific importance and in general interest.
+In the last 20 years, especially since SARS outbreak in 2003, there has
+been a huge volumn of related work, more than those referenced in this
+paper. These "intervals" have been widely cited, measured and applied to
+outbreak investigations, especially during the early stage. However, there
+is a lot of ambiguity. There are occasions that the same terminology is
+associated with two different definitions, whereas there are other occasions
+that the same definition is assigned to different terminologies by different
+researchers.
+}
+
+\rev{
+I suggest major revision of the paper and hope my comments are
+constructive for the revision(s) for better clarification. It will be a
+great contribution to this field.
+}
+
+\rev{
+The immediate objective of this paper is to construct some "unbiased" estimate of the generation interval distribution (presumably, the intrinsic distribution $g(\tau)$).
+Its important goes beyond this immediate objective, that is, to attempt using quantities (assuming observable, at least conceptually) measured during a the very dynamic
+phase of the system to make inferences on these intrinsic quantities
+$\Ro$ and $r$: I feel the authors have fallen short of achieving these objectives. However, I would like to share some of my opinions and ideas
+and hope that, after revision(s), some of these goals will be achieved.
+}
+
+\rev{
+I also recommend, for the authors to consider, breaking this paper
+into two papers: one on mathematical theory and models to better
+clarify concepts which I believe are still ambiguous in currently published literature; and another one on statistical models and methods
+to link data to models.
+}
 

### 2020-05-20 doc/network.tex
working on it...

@@
 
-\newcommand{\eref}[1]{(\ref{eq:#1})}
+\newcommand{\eref}[1]{Eq.~\ref{eq:#1}}
 \newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
@@
 
-\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
+\usepackage{xspace}
+
+\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
 \newcommand{\Ro}{\Rx{0}}
 \newcommand{\RR}{\ensuremath{{\mathcal R}}}
+\newcommand{\Rini}{\Rx{\textrm{\tiny initial}}}
 \newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
@@
 	\textbf\newline{
-		Inferring generation-interval distributions from contact-tracing data
+		Inferring the generation-interval distribution that correctly links speed and strength of an epidemic
 	}
@@
 Generation intervals are often measured through contact tracing by identifying who infected whom.
-We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales.
-Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the censoring and the population structure near the source of disease spread; 
-therefore, we predict that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly link $r$ and $\RR$.
+We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales.
+Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
+therefore, we predict that correcting observed intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly link $r$ and $\RR$.
 We develop and test statistical methods for temporal corrections of generation intervals,
@@
 
-Infectious disease modeling, generation interval, basic reproduction number, population structure, contact tracing
+Infectious disease modeling, generation interval, basic reproductive number, population structure, contact tracing
 
@@
 An epidemic can be described by the exponential growth rate, $r$, and the reproductive number, \RR.
-The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
-However, estimating the reproductive number directly from disease natural history requires detailed knowledge which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
-Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
+The reproductive number is defined as the average number of secondary cases arising from a primary case;
+its value in a fully susceptible population, also known as the basic reproductive number \Ro, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition} as well as the endemicity level \citep{kribs2000simple, van2002reproduction, smith2012ross}.
+However, estimating the reproductive number directly from disease life history requires detailed knowledge, which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
+Instead, the reproductive number can be indirectly estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
 These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.
@@
 At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
-While this definition is widely used in the literature, it is not connected directly to a population-level distribution.
-There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case,
-while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
+While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.
+There are important distinctions to be made when defining generation-interval distributions at the population-level.
+The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.
+On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
 Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.
 
-While the shape of intrinsic generation-interval distribution is often assumed to be invariant across time, the shape of the realized generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+The shape of the realized generation-interval distribution depends on the reference time and perspective \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
 When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
@@
 Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
-Forward generation intervals can contract over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.
+Mean forward generation intervals can decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.
 
-Forward generation intervals are also affected by spatial structure.
+Realized generation intervals are also affected by spatial structure.
 In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population.
 This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
-As a result, forward generation intervals will have shorter mean than the intrinsic generation-interval distribution.
+As a result, realized generation intervals (from the perspective of an infector) will typically have shorter mean than the intrinsic generation-interval distribution in a non-homogenous population.
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
 
-In practice, realized generation intervals are often measured by aggregating available information from contact tracing.
-While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
-These censored intervals are similar to backward intervals: there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
-These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will yield a spatially corrected ``effective'' distribution that contains information about the population structure and will allow us to correctly infer the reproductive number from the exponential growth rate.
-We further predict the effective generation-interval distribution to be equivalent to the \emph{initial} forward generation-interval distribution, which describes the renewal process of infection during the exponential growth rate.
+While an epidemic is ongoing, realized generation intervals until the current epidemic time can be aggregated to form a single distribution.
+While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
+we typically want to try to make inference based on all available data.
+These aggregated measurements are ``truncated'' because we do not know what happens after the current epidemic time.
+These truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
+These observations allow us to make the following prediction: removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which is ``spatially corrected'' (i.e., contains information about the population structure) and will allow us to correctly infer the initial reproductive number from the exponential growth rate.
 
-In this study, we explore spatiotemporal variation in generation intervals measured through contact tracing.
-We extend previous frameworks to study how the censored generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
-We classify spatial effects on generation intervals into three levels (egocentric, local, and global) and discuss how they affect observed generation-interval distributions.
+In this study, we explore spatiotemporal variation in realized (particularly forward) generation intervals.
+We extend previous frameworks to study how the aggregated generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
+We classify spatial effects on realized generation intervals into three levels (egocentric, local, and global) and discuss how they affect realize generation-interval distributions.
 Finally, we compare two methods for accounting for temporal bias and test our prediction using simulations.
@@
 Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence};
-making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
-An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
+making this distinction clear will be particularly useful when we discuss spatial components later (see \fref{indpop}).
+An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``infectious contacts'' (contacts which will cause infection if the contactee is susceptible).
 Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
@@
 where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
-The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
+The population-level kernel describes the rate at which infectious contacts are made by an an infected individual, on \emph{average}.
 
@@
 \end{equation}
-where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population), 
-and $g(\tau)$ is the expected time distribution of those cases (the intrinsic generation-interval distribution).
+where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population \citep{anderson1991infectious}),
+and $g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
+Since all infectious contacts lead to infection in a disease-free equilibrium of a homogeneously mixing population, $g(\tau)$ also describes the expected time distribution of secondary cases caused by an average primary case in a fully susceptible population and can be treated as an ``equilibrium'' distribution of the corresponding homogeneous mixing model.
 
@@
 This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
-Over a period of time where the susceptible proportion $S(t) = S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
+Over a period of time where the susceptible proportion $S(t) = S(0)$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
 \begin{equation}
-\frac{1}{\RR} = \int g(s) \exp(-r s) ds,
+\frac{1}{\Rini} = \int g(s) \exp(-r s) ds.
+\label{eq:euler}
 \end{equation}
-where $\RR = S(t) \RR_0$ is the effective reproductive number.
+Under the homogeneous mixing assumption, the intrinsic generation-interval distribution $g(\tau)$ provides the correct link between $r$ and $\Rini$.
 
-\section{Generation-interval distributions across time}
+\section{Realized generation-interval distributions across time}
 
-When generation intervals are estimated through contact tracing during an outbreak, infection events that have not happened yet are not observed. 
-This effect is called ``right-censoring''.
-We can understand the effect of right-censoring using backward generation-interval distributions \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}:
-the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions (weighted by incidence) up until the observation time.
+Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
+The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by infectors who were infected at time $t$.
+Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} of infectees who were infected at time $t$.
+For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
+Therefore, the total number of infection that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
+\begin{equation}
+i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
+\label{eq:equivalence}
+\end{equation}
 
-The density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
+In a homogeneously mixing population, the forward and backward generation-interval distributions can be calculated exactly.
+The density of new infections occurring at time $t+\tau$ caused by infectors who were infected at time $t$ is given by
+\begin{equation}
+i_{t}(t+\tau) = \RR_0 i(t) g(\tau) S(t+\tau)
+\end{equation}
+As shown in \cite{champredon2015intrinsic}, the forward generation-interval distribution, $f_t(\tau)$, is proportional to $i_{t+\tau}(t)$:
+\begin{equation}
+f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(\tau) S(t+\tau) d\tau}.
+\label{eq:forward}
+\end{equation}
+In this case, the initial foward generation-interval distribution during the exponential growth phase (when $S(t) \approx S(0)$) is equivalent to the intrinsic generation-interval distribution and, therefore, provides the correct link between $r$ and \Rini.
+Likewise, the density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
 \begin{equation}
@@
 \end{equation}
-As shown in \cite{champredon2015intrinsic}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$ with respect to $\tau$.
-The censored generation-interval distribution, $c_t(\tau)$, describes \emph{all} cohorts infected at time $s<t$; thus, the censored generation-interval distribution is just the average of the backward distributions of these cohorts, weighted by incidence:
+The backward generation-interval distribution, $b_t(\tau)$, is proportional to $i_{t-\tau}(t)$:
 \begin{equation}
-c_t(\tau) \propto \int_0^t b_s(\tau) i(s) ds.
+b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-\tau) g(\tau) d\tau}.
 \end{equation}
+
+During an outbreak, generation intervals cannot be measured for infection events that have not happened yet.
+This effect is called ``right-truncation''.
+Therefore, even if we aggregate all realized generation intervals over the course of an epidemic, their mean will be shorter than the intrinsic generation intervals.
+The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until the epidemic time $t$:
+\begin{equation}
+a_t(\tau) \propto \int_{-\infty}^t b_s(\tau) i(s) ds.
+\end{equation}
+The aggregated generation-interval distribution can be also equivalently expressed in terms of the forward generation-interval distributions:
+\begin{equation}
+a_t(\tau) \propto \int_{-\infty}^{t-\tau} f_s(\tau) i(s) ds.
+\end{equation}
+\eref{equivalence} confirms that both expressions are identical.
+
+For a single outbreak, the mean aggregated generation interval will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
+There are two reasons for this phenomenon.
+First, longer generation intervals are more likely to be missed due to right truncation (and short generation intervals are more likely to be realized).
+In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$),
+the initial aggregated (or backward) generation-interval distribution is just the initial forward generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
+\begin{equation}
+a_0(\tau) = b_0(\tau) \propto f_0(\tau) \exp(-r\tau).
+\label{eq:exp}
+\end{equation}
+A deterministic simulation confirms that the aggregated generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
+Second, the decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
+Overall, we expect naively using the aggregated generation-interval distribution, to underestimate the reproductive number.
 
@@
 A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
-The backward and the censored mean generation interval were calculated over the course of an epidemic.
+The backward and the aggregated mean generation interval were calculated over the course of an epidemic.
 The dotted horizontal line represents the mean intrinsic generation interval.
@@
 
-For a single outbreak, the observed mean  generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
-There are two reasons for this phenomenon.
-First, longer generation intervals are more likely to be missed due to right censoring (and short generation intervals are more likely to be observed).
-In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
-the initial censored (or backward) generation-interval distribution is just the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
-\begin{equation}
-c_0(\tau) = b_0(\tau) \propto g(\tau) \exp(-r\tau).
-\label{eq:exp}
-\end{equation}
-A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
-Second, the decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
-Overall, we expect naively using the observed generation-interval distribution, to underestimate the reproductive number.
+\section{Realized generation-interval distributions across space}
 
-% When a disease is at (or near) endemic equilibrium, the number of susceptibles remains (approximately) constant over time, and we expect the observed generation-interval distribution to be similar to the intrinsic generation-interval distribution.
-
-\section{Generation-interval distributions across space}
-
-\rev{
-I have a problem with $N(a)$.
-Does your model allow for different number of acquaintances?
-But then your $R_0$ is no longer valid.
-Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.
-}
 Infected individuals may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
@@
 
-\rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
 The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
@@
 This calculation can be validated by simulating stochastic infection processes on a ``star'' network (i.e, a single infected individual at the center connected to multiple susceptible individuals who are not connected with each other).
-Simulations (\fref{local}) confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution (left panel), while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (middle panel).
+Simulations (\fref{local}) confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution (left panel), while the distribution of realized generation intervals (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (middle panel).
 
@@
 
-\rev{
-what is a homogeneous network? Complete network, alla have same degree or ...? Explain better why your networks of size 5 matter in a large community (local aspects?)
-}
-The egocentric generation interval \eref{conditional} only explains some of the reduction in generation times that occurs on most networks, however.
+The egocentric generation interval (\eref{conditional}) only explains some of the reduction in generation intervals that occurs on most networks, however.
 Generation intervals are also shortened by indirect connections: a susceptible individual can be infected through another route before the focal individual makes infectious contacts.
@@
 
-In general, spatial reduction in the mean generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
+In general, spatial reduction in the mean realized generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
 Egocentric depletion, as discussed previously, is caused by an infected individual making multiple contacts to the same individual.
@@
 
-\section{Inferring generation-interval distributions from contact-tracing data}
-\rev{
-some of the 4 effects described also affect R whereas others don't.
-Be more clear to the reader.
-For instance, the ordinare Euler-Lotka equations is not fulfilled in network situation.
-I think Ball, Pellis and Trapman have investigated different reproduction numbers for network epidemics and related.
-}
+\section{Inferring the initial forward generation-interval distribution}
 
-When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
-We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-censoring effect;
-we showed that these effects shorten the mean observed generation intervals, which in turn will reduce the estimate of the reproduction number \cite{wallinga2007generation, park2019practical}.
+In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
+In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution;
+since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
+
+When realized generation intervals are aggregated over the course of an epidemic, there will be four effects present in the sample: (1) right-truncation effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
+We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-truncation effect;
+these effects shorten the mean realized generation intervals, which in turn will reduce the estimate of the reproductive number \cite{wallinga2007generation, park2019practical}.
 While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
 both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.
-
-Since the right-censoring effect is a sampling bias, we typically want to correct for it. 
-In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
-We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread.
-When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ through the Euler-Lotka equation.
-We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
-In a large homogeneously mixing population, the effective distribution is expected to be equivalent to the intrinsic distribution (since the egocentric and local depletion are negligible in the absence of network structure, and global depletion is negligible early in the epidemic).
+If we can correct for the temporal bias early in an outbreak, during the exponential growth phase, we should be able to infer the initial forward generation-interval distribution, which incorporates egocentric and local spatial effects but not the global effects, from the aggregated distribution.
 
@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
+Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-truncation severely underestimates the reproductive number.
 Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
@@
 
-Here, we investigate two methods for correcting for temporal bias in contact-tracing data (see Methods for details).
-We refer to the first method as the population-level method as it relies on the observed distribution aggregated across the entire population.
-When an epidemic is growing exponentially, right-censoring causes the observed generation interval to be discounted by the exponential growth rate \eref{exp};
-hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
+Here, we investigate two methods for correcting for temporal bias in aggregated generation-interval data (see Methods for details).
+We refer to the first method as the population-level method as it relies on realized generation intervals aggregated across the entire population.
+When an epidemic is growing exponentially, right-truncation causes the aggregated generation interval to be discounted by the exponential growth rate (\eref{exp});
+hence, we can ``undo'' the truncation by exponentially weighting the aggregated generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
 \begin{equation}
-\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau),
+f_0(\tau) \propto a_0(\tau) \exp(r\tau),
 \label{eq:geff}
@@
 We refer to the second method as the individual-level method because it relies on individual contact information.
-We model each infection as a non-homogeneous Poisson process arising from the infector \eref{nonhomogeneous}; 
-incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
-For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$ using the Euler-Lotka equation.
+We model each infection as a non-homogeneous Poisson process arising from the infector (\eref{nonhomogeneous}); 
+incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the truncation process in the realized generation intervals.
+For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.
 
@@
 We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
-On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproduction number \cite{britton2019estimation}.
+On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproductive number \cite{britton2019estimation}.
 We do not pursue these direction in this study.
@@
 As expected, estimating the reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
-estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially, as they do not account for indirect (local) spatial effects. 
+estimates based on the egocentric generation-interval distribution (\eref{conditional}) address this problem only partially, as they do not account for indirect (local) spatial effects. 
 Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
@@
 \section{Discussion}
+\swp{something about little r}
+\swp{something about big R; and time dependent R}
 
-The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\RR$) of an epidemic \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
-However, observed generation-interval distributions can vary depending on how and when they are measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
-determining which distribution correctly links $r$ and $\RR$ can be challenging.
-Here, we analyze how observed generation intervals, measured through contact tracing, differ from intrinsic generation intervals.
-Changes due to temporal censoring reflect observation bias, whereas changes due to spatial or network structure reflect the dynamics of the outbreak.
-Thus, correcting the observed distribution for temporal, but not spatial, effects provides the correct link between $r$ and $\RR$.
+The intrinsic generation-interval distribution, which describes the expected time distribution of infectous contacts, provides a direct link between speed (exponential growth rate, $r$) and strength (initial reproductive number, $\Rini$) of an epidemic in a homogeneously mixing population \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
+However, realized generation-interval distributions can vary depending on how and when they are measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
+determining which distribution correctly links $r$ and $\Rini$ can be challenging.
+Here, we analyze how realized generation intervals differ from intrinsic generation intervals.
+Changes due to right-truncation reflect observation bias, whereas changes due to spatial or network structure reflect the dynamics of the outbreak.
+Thus, correcting the aggregated distribution for temporal, but not spatial, effects provides the correct link between $r$ and $\Rini$.
 
-Observed generation intervals are subject to right-censoring -- it is not possible to trace individuals that have not been infected yet.
-These right-censored distributions can be thought of as averages of ``backward'' generation intervals (measured by tracing the infectors of a cohort of infected individuals) \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
-During an ongoing outbreak, the observed generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-censoring.
-Early in the outbreak, censored intervals are expected to match the early-outbreak backward intervals.
-Near the end of an outbreak, the effect of right-censoring becomes negligible but the observed generation intervals are still shorter on average than intrinsic generation intervals, because of depletion of the susceptible population.
+Realized generation intervals that have been aggregated over the course of an epidemic are subject to right-truncation -- it is not possible to trace individuals that have not been infected yet.
+The aggregated distributions can be thought of as averages of ``backward'' generation intervals (measured by looking at infectors of a cohort ofindividuals infected at the same time) \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+During an ongoing outbreak, the aggregated generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-truncation.
+Early in the outbreak, aggregated intervals are expected to match the early-outbreak backward intervals.
+Near the end of an outbreak, the effect of right-truncation becomes negligible but the aggregated generation intervals are still shorter on average than intrinsic generation intervals, because of depletion of the susceptible population.
 
@@
 Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
-Accounting for the egocentric effect allowed us to link the results by \cite{trapman2016inferring} to established results based on generation intervals.
+Accounting for the egocentric effect allows us to link the results by \cite{trapman2016inferring} to established results based on generation intervals.
 Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
@@
 Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
-Therefore, we predicted the observed generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the \emph{effective} generation-interval distribution by simply accounting for the right-censoring.
-Simulation studies confirmed our prediction: using the effective generation-interval distribution provides the correct link between $r$ and $\RR$.
+Therefore, we predicted the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right-truncation.
+Simulation studies confirmed our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\RR$.
 
-We compare two methods for estimating the effective generation-interval distribution and assume that the effective generation-interval distribution follows a gamma distribution.
+We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
 The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
@@
 Contact tracing provides an effective way of collecting epidemiological data and controlling an outbreak \citep{clarke1998contact, eames2003contact, donnelly2003epidemiological}.
-In particular, using tracing information allows us to infer real-time estimates of the time-varying reproductive number \citep{cauchemez2006estimating, hens2012robust, jewell2012enhancing, soetens2018real};
-generation-interval distributions, which can be either assumed or estimated, often play a central role in analyzing tracing data.
-Our study illustrates that generation intervals measured through contact-tracing data contain information about the underlying contact structure, which can be implicitly reflected in the estimates of the reproductive number;
+In particular, using tracing information allows us to infer real-time estimates of the time-varying reproductive number \citep{cauchemez2006estimating, hens2012robust, jewell2012enhancing, soetens2018real}.
+Generation-interval distributions, which can be either assumed or estimated, often play a central role in analyzing tracing data.
+Our study illustrates that realized generation intervals over the course of an epidemic contain information about the underlying contact structure, which can be implicitly reflected in the estimates of the reproductive number;
 this perspective can be particularly useful for characterizing an epidemic because detailed information about the contact structure is often unavailable.
@@
 Different definitions, and different measurement approaches, produce different estimates of these distributions.
-We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
+We have shown that estimates based on aggregated generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
 These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
-We have shown that correcting observed intervals for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population.
+We have shown that correcting aggregated intervals for temporal effects allows us to estimate a spatially informed initial forward distribution, which accurately describes how disease spreads in a population.
 Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence the spread of disease.
@@
 
-To study the effects on right-censoring on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
+To study the effects on right-truncation on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
 The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infected individuals become infectious after a latent period.
 We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.
-This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
+This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more realistic approximations than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
 \begin{equation}
@@
 
-The \textbf{population-level method} estimates the effective generation-interval distribution by reversing the inverse exponential weighting in the observed generation-interval distribution:
+The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution:
 \begin{equation}
-\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau).
+f_0(\tau) \propto a_0(\tau) \exp(r\tau).
 \end{equation}
-In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
+In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; 
+specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
 Then, the effective generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
 
-The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the censorship time ($\tsub{t}{censor}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
+The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the truncation time ($\tsub{t}{truncate}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
 We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
-Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{censor}$ is equal to
+Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ is equal to
 \begin{equation}
-\frac{\RR^{n_j} \exp \big(- \RR\, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
+\frac{\RR^{n_j} \exp \big(- \RR\, \tsub{G}{eff}(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
 \end{equation}
@@
 \begin{aligned}
-&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{censor})\\
-&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \, \tsub{G}{eff}(\tsub{t}{censor} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
+&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{truncate})\\
+&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \, \tsub{G}{eff}(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
 \end{aligned}
@@
 
-SWP led the literature review, wrote the first draft of the MS, performed analytic calculations and simulations; JD conceived the study and performed analytic calculations; all authors contributed to refining study design, literature review, and final MS writing.
+SWP led the literature review, performed analytic calculations and simulations, and wrote the first draft of the MS; JD conceived the study, performed analytic calculations and wrote the first draft of the MS; all authors contributed to refining study design, literature review, and final MS writing.
 All authors gave final approval for publication.

### 2020-05-20 doc/response.tex
changes

@@
 
-Done.
+We now define both backward and forward generation-interval distributions more appropriately in the text.
+We no longer use the term ``censored'' generation-interval distribution. 
+We refer to it as the aggregated generation-interval distribution, which is a distribution of all realized generation intervals until certain time point.
+We think thie definition provides a clearer definition.
 
@@
 We have removed this paragraph and tried to make the explanation more concise.
-We no longer use the term effective generation-interval distribution.
+We no longer use the term effective generation-interval distribution but rather refer to it as the initial forward generation-interval distribution.
 
-\rev{Lines 241-243: "In a large homogeneously mixing population, the e§ective
+\revtext{Lines 241-243: "In a large homogeneously mixing population, the effective
 distribution is expected to be equivalent to the intrinsic distribution..."
@@
 
-Done.
+We now show that the initial forward generation-interval distribution is equivalent to the intrinsic distribution.
+We also point to the analytical intractability of the spatial effects on the realized generation-interval distributions (and therefore the initial forward distribution) and therefore cannot be expressed mathematically (for non-homogeneous cases):
 
-\rev{However, I suggest the authors to compare the "individual-based
+``Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.''
+
+\revtext{However, I suggest the authors to compare the "individual-based
 method" in the paragraph starting from Line 252 to Line 260 with the statistical methods for analyzing right-truncated survival data as mentioned
@@
 
-\rev{The authors has used the notations $\RR$ (Line 35) and $\Ro$ (Lines 98-99) for reproduction numbers and
+The likelihood function is provided in the Methods section.
+\swp{TODO. It seems like they're different... Ugh. I'll get back to this.}
+
+\revtext{The authors has used the notations $\RR$ (Line 35) and $\Ro$ (Lines 98-99) for reproduction numbers and
 linked them via $\RR = \Ro S(t)$:}
 
-\rev{Shouldnít $\RR$ be a function of time $t$ via $S(t)$?}
+\revtext{Shouldn't $\RR$ be a function of time $t$ via $S(t)$?}
 
-We no longer use $\RR$ to avoid confusion. We now use $\Rini = \Ro S(0)$ to reflect the possibility that $S(0)$ may not be 1.
+We no longer use $\RR$ to avoid confusion.
+We now use $\Rini = \Ro S(0)$ to reflect the possibility that $S(0)$ may not be 1 but clearly convey the idea that we're not estimating time dependent reproduction number.
 
-\rev{Is $S(t)$ a number or a proportion? In deterministic models, this matters
+\revtext{Is $S(t)$ a number or a proportion? In deterministic models, this matters
 less as one can argue that all deterministic models are describing the
@@
 
-Proportion. Made this clear.
+We make these distinctions clear in the main text.
 
-\rev{
+\revtext{
 When stating that ``$\RR$'' transcends to the final size of the epidemic (Line
@@
 
-Yes.
+Yes. We now make a clear distinction between $\RR$ and $\Ro$.
 
-\rev{
+\revtext{
 When talking about the final size, one considers that the system is at a
@@
 
-Discussion!
+\swp{TODO? Just say we are interesting in the initial phase? maybe write something in the discussion? We're only using $\Rini$}
 
-\rev{
+\revtext{
 The authors stated that $\RR$ transcends to the final size of the epidemic
@@
 Yes, we meant $\Ro$. We have now clarified this in the text.
-We now mention in Introduction that $\Ro$ can also affect the endemicity level and added a few appropriate citations.
+We now mention in Introduction that $\Ro$ can also affect the endemicity level and added a few citations.
 
-\rev{$g(\tau)$ as a p.d.f. given by (2) and Line 101 is also an equilibrium
+\revtext{$g(\tau)$ as a p.d.f. given by (2) and Line 101 is also an equilibrium
 distribution.}
 
-We disagree with the reviewer that the intrinsic generation-interval is an equilibrium distribution.
-It is only an equilibrium distribution for a model that assumes a large, homogeneously mixing population.
+While we agree with the reviewer that the intrinsic generation-interval can be interpreted an equilibrium distribution, this only holds under homogeneous mixing assumption. 
 As we show in this paper, when there is spatial structure, the realized generation-interval distributions differ from the intrinsic one. 
 In particular, even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egeocentric effect.
+Therefore, the realized generation intervals of a single, primary infector in an otherwise susceptible population will not follow the intrinsic generation-interval distribution.
+We mention in the manuscript that the intrinsic distribution can be interpreted as an equilibrium distribution but do not want to emphasize on using the term ``equilibrium'' particularly given that we're interested in realized distributions on a spatially heterogeneous network.
+The reviewer also presented a series of formulas on the mean generation interval for special cases of the SEIR model without any suggestions;
+given that our goal is not to explore all special cases of the SEIR model, we do not think presenting all those formulas is relevant to the discussion of the paper.
 
-\rev{ Under suitable assumptions on homogeneous mixing, the intrinsic
+\revtext{
+Expression (13) is the convolution (the sum of) the exponentially
+distributed latent period with mean $1/\sigma$ and another exponential distribution, with mean $1/(\gamma+\lambda)$; as written in Lines 188-189. It is also equivalent
+to the homogeneous mixing situation, in which, individuals during their
+infectious periods are isolated (with rate $\lambda$) so that the "effective" duration
+of the infectious period is exponentially distributed with mean $1/(\gamma + \lambda)$:
+Any comments from the authors on this coincidence?
+}
+
+We explain this in the main text:
+
+``In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
+Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
+Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.''
+
+\revtext{Under suitable assumptions on homogeneous mixing, the intrinsic
 generation interval distribution $g(\tau)$ measures the infectiousness within
@@
 
-We agree that $g(\tau)$ measures the infectiousness within an infected individual without involving another individual.
-We defined the intrinsic generation-interval distribution as follows:
+We now define the intrinsic generation-interval distribution as follows as we think it is more concrete; it is also easier to distinguish intrinsic vs. realized distributiosn under this definition:
 
@@
 \revtext{The distribution $g(\tau)$ links the two intrinsic parameters R0 and r
-(homogeneous mixing). Both $\Ro$ and $g(\tau)$ are intrinsic in the context of the equilibrium conditions. Therefore the parameter $r$ is a
-theoretical value and is very di§erent from any observed exponential
+(homogeneous mixing).
+Both $\Ro$ and $g(\tau)$ are intrinsic in the context of the equilibrium conditions.
+Therefore the parameter $r$ is a theoretical value and is very different from any observed exponential
 or sub-exponential growth rate as the system is not at equilibrium.}
 
-\rev{Observed exponential or sub-exponential growth events are different from the infection events. The current paper only addresses the infection events (mostly unobservable). This was
+\swp{???}
+
+\revtext{Observed exponential or sub-exponential growth events are different from the infection events. The current paper only addresses the infection events (mostly unobservable). This was
 already addressed in the comments with respect to Line 47-48.
@@
 
-We point out in the Discussion that estimated $r$ can be different from the true $r$. 
-However, it is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
-See:
+It is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
+See \swp{How do I respond to this, actually?}:
 
@@
 
-\rev{
+\revtext{
 The intrinsic parameters R0 and r are parameters in branching
@@
 This is inaccurate.
-It is still possible to observe exponential growth when the population is not fully susceptible ($S(t) \approx S(0) \neq 1$) as long as the depletion of the susceptible population is negligible.
+It is still possible for exponential growth to occur when the population is not fully susceptible ($S(t) \approx S(0) \neq 1$) as long as the depletion of the susceptible population is negligible.
 To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.
 
-\rev{
+\revtext{
  It is too strong a statement that "the Erlang distributions are
@@
 The exponential distribution is a special case of the Erlang distribution (when the shape parameter is equal to 1).
-We disagree the statement that it is ``too strong'' to say that a more general distribution is more realistic than a special case.
-To avoid confusion, we now say refer to Erlang distributions as ``more realistic approximations''.
+We disagree the statement that it is ``too strong'' to say that a more general distribution is more realistic than its special case.
+To avoid confusion, we now say refer to Erlang distributions as ``more flexible''.
 
-\rev{
+\revtext{
 The Erlang distribution is not only a subset of the gamma distribution with integer valued shape parameters, but also it is limited to distributions with variances smaller than the corresponding exponential distribution with the same mean value.
@@
 
-\rev{
+\revtext{
 One can not even claim the gamma distributions are more realistic
@@
 
-\rev{
+\revtext{
+"... based on more realistic distributions in Appendix." I do
+not see "more realistic distributions" more than the Erlang distributions.
+How about heavy tailed distributions and distributions with large variances? In fact, many of the discussions in this paper can be generalized
+to arbitrarily distributed latent and infectious periods.
+}
+
+We had omitted a sentence previous --- we have now clarified this sentence as follows:
+
+``In the main text, we present results based on exponentially distributed latent and infectious periods;
+we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.''
+
+\revtext{
 It is more convenient to use the Laplace transforms as a tool.
@@
 
-The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial components.
-The Laplace tranforms are only more useful for the intrinsic generation-interval distributions. 
-The forward generation-interval distribution is analytically intractable and therefore, the Laplace transforms no longer hold.
+The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial effects.
+The Laplace tranforms are only more useful for the intrinsic generation-interval distributions under the homogeneous assumption. 
+The forward generation-interval distribution in a spatially hetergeneous environment is analytically intractable and therefore, the Laplace transforms no longer hold.
 Given that our goal is to infer the initial forward generation-interval distribution, rather than the intrinsic distribution, we feel that the Laplace transforms are not necessarily helpful for the discussion.
-In particular, the reviewer asks us to prove special cases of the Euler-Lotka equation, which has been proven numerous times as suggested by the references, using Laplace transforms; 
-we do not think this is any relevant to the discussion of our paper.
-We cite all Yan et al papers that were mentioned in the review in our main text.
+In particular, the reviewer asks us to prove special cases of the Euler-Lotka equation, which has been proven previosuly as suggested by the references, using Laplace transforms; 
+we do not think this is relevant to the discussion of our paper.
 
-\rev{
-How to assess if the gamma distribution is a good choice as a
-model? Run some kind of goodness-of-fit tests? How does the likelihood
+\revtext{How to assess if the gamma distribution is a good choice as a
+model? Run some kind of goodness-of-fit tests?}
+
+The gamma distribution has been previously tested for many diseases. 
+In our case, Figure 4 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
+We discuss potential caveats in assuming a shape of the generation-interval distribution in Discussion.
+
+``We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
+The gamma approximation of the generation interval distribution has been widely used due to its simplicity;
+we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$.
+However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
+These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).''
+
+\revtext{How does the likelihood
 function look like? Since the "observed" generation-intervals (assuming
@@
 
+The population-level method does not take into account the observation scheme or the infection process. 
+We simply fit the gamma distribution to all realized generation intervals (i.e., the aggregated distribution);
+we feel that writing down the likelihood function is an overkill for a basic gamma fit.
+We have clarified the text:
+
+``The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom):
+\begin{equation}
+f_0(\tau) \propto a_0(\tau) \exp(r\tau).
+\end{equation}
+In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
+specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
+Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
+We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).''
+
 \rev{
@@
 }
+
+That was a bad notation on our side --- we should not have used $\RR$ in the likelihood. 
+We have changed the notation for the Poisson process and provide a better explanation.
 
@@
 
-``the inferred generation-interval distribution is then used to estimate the reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.''
+We estimate $\Rini$ via Euler-Lotka. We have tried to make this clear throughout the main text.
 
@@
 
+\swp{TODO.}
+
 \rev{
@@
 }
+
+\swp{TODO.}
 
@@
 We had never equated the true distribution to the intrinsic distribution.
-We make clear that we do not want the intrinsic distribution.
+We make clear that the intrinsic distributions are not our goal.
 
@@
 
+Thank you!
+
 \rev{
@@
 }
+
+We have made major revisions to the manuscript as per suggested comments.
+We hope that our manuscript is clearer.
 

### 2020-05-20 doc/network.tex
changes

@@
 \newcommand{\Rini}{\Rx{\textrm{\tiny initial}}}
+\newcommand{\Rego}{\Rx{\textrm{\tiny egocentric}}}
 \newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
@@
 \begin{equation}
-\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
+\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
 \label{eq:egorR}
 \end{equation}
-As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will always be smaller than $\Rini$ estimated from the intrinsic distribution;
 this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
@@
 \end{equation}
-In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
-In practice, directly using this distribution to link $r$ and $\RR$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
+In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
+Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
+Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
+In practice, directly using the egocentric distribution to link $r$ and $\Rini$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
 Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
@@
 
-The egocentric generation interval (\eref{conditional}) only explains some of the reduction in generation intervals that occurs on most networks, however.
+The egocentric generation interval (\eref{conditional}) only explains some of the reduction in realized generation intervals that occurs on most networks, however.
 Generation intervals are also shortened by indirect connections: a susceptible individual can be infected through another route before the focal individual makes infectious contacts.
@@
 In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution;
-since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
+since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we predict the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
+Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.
 
@@
 Therefore, we predicted the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right-truncation.
-Simulation studies confirmed our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\RR$.
+Simulation studies confirmed our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\Rini$.
 
@@
 The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
-we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\RR$ \citep{park2019practical}.
+we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ \citep{park2019practical}.
 However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
@@
 We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.
-This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more realistic approximations than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
+This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more flexible than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
 \begin{equation}
@@
 Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
-We show results based on more realistic distributions in Appendix.
+In the main text, we present results based on exponentially distributed latent and infectious periods;
+we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.
 
@@
 
-\subsection{Estimating generation-interval distributions from contact-tracing data}
+\subsection{Estimating the initial forward generation-interval distribution}
 
-The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution:
+The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom) \citep{tomba2010some, nishiura2010time, britton2019estimation}:
 \begin{equation}
@@
 \end{equation}
-In order to do so, we first fit a gamma distribution to the observed generation-interval distributions; 
+In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
 specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
-Then, the effective generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
+Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
+We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
 
-The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the truncation time ($\tsub{t}{truncate}$), with time-varying Poisson rate at time $t$ is equal to $\RR \tsub{g}{eff}(t - t_j)$, where $\tsub{g}{eff}(t)$ is the effective generation-interval distribution \citep{daley2007introduction}.
-We use a gamma distribution (parameterized by its mean and shape) to model the effective generation-interval distribution.
+The \textbf{individual-level method} models each infection $i$ from an infected individual $j$ as a non-homogeneous Poisson process between the time at which infector $j$ was infected ($t_j$) and the truncation time ($\tsub{t}{truncate}$), with time-varying Poisson rate at time $t$ is equal to $\Lambda f_0(t - t_j)$, where $f_0(t)$ is the initial forward generation-interval distribution \citep{daley2007introduction}.
+We use a gamma distribution (parameterized by its mean and shape) to model the initial forward generation-interval distribution.
 Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ is equal to
 \begin{equation}
-\frac{\RR^{n_j} \exp \big(- \RR\, \tsub{G}{eff}(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta)}{n_j!},
+\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!},
 \end{equation}
-where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the generation-interval distribution $g$ (and the corresponding cumulative distribution function $\tsub{G}{eff}$).
+where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the initial forward generation-interval distribution $f_0$ (and the corresponding cumulative distribution function $F_0$).
 
@@
 \begin{aligned}
-&\mathcal{L}(\RR, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{truncate})\\
-&\quad=\prod_{j=1}^N \left(\RR^{n_j} \exp \big(- \RR \, \tsub{G}{eff}(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} \tsub{g}{eff}(s_{i, j}; \theta) \right),
+&\mathcal{L}(\Lambda, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{truncate})\\
+&\quad=\prod_{j=1}^N \left(\Lambda^{n_j} \exp \big(- \Lambda \, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta) \right),
 \end{aligned}
@@
 where $N$ is the total number of infected individuals.
-Here, we estimate parameters $\RR$ and $\theta$ by maximum likelihood.
-We expect estimates of $\RR$ to be sensitive to under-reporting;
-instead, we use the estimated generation-interval distribution to infer the effective reproductive number from the estimated growth rate (using the Euler-Lotka equation).
+Here, we estimate parameters $\Lambda$ and $\theta$ by maximum likelihood.
+We then use the estimated distribution $f_0$ to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
+We expect estimates of $\Lambda$ to be sensitive to under-reporting.
 White and Pagano \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

### 2020-05-20 doc/supp.tex
changes

@@
 This figure matches Fig. 4 in the main text which used a SEIR model.
-Here, an Erland-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
+Here, an Erlang-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
 All other parameters are the same as in Fig. 4 in the main text.

### 2020-05-20 doc/response.tex
middle of meeting!

@@
 
-\newcommand{\rev}{\subsection*}
+\newcommand{\reviewer}{\subsection*}
 \newcommand{\revtext}{\textsf}
@@
 
-\rev{Reviewer \#1}
+\reviewer{Reviewer \#1}
 
@@
 
-\rev{
+\revtext{
 Elaborate how the estimation of $\RR$ is expected to be sensitive to under-reporting (of which data)
@@
 
-\rev{
+\revtext{
 How exactly is $\RR$ (and, is it \Ro or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref. [54] in the paper)?
@@
 
-\rev{
+\revtext{
 How close is the "related approach" in White and Pagano
@@
 
-\rev{
+\revtext{
 The authors use a logistic function with a scale parameter $r$,
@@
 
-\rev{
+\revtext{
 The paper focuses on estimating the generating-interval distribution. There are
@@
 
-\rev{
+\revtext{
 The authors have well demonstrated different sources of sampling bias
@@
 
-\rev{
+\revtext{
 Any notion of bias is in the context against some "truth". The authors
@@
 
-\rev{
+\revtext{
 The intrinsic distribution $g(\tau)$ is an equilibrium distribution that
@@
 
-\rev{
+\revtext{
  Under suitable assumptions on homogeneous mixing, the intrinsic
@@
 
-\rev{
+\revtext{
 Potentially observable events do not take place at the exact moment
@@
 
-\rev{
+\revtext{
 I rate this paper very high in scientific importance and in general interest.
@@
 
-\rev{
+\revtext{
 I suggest major revision of the paper and hope my comments are
@@
 
-\rev{
+\revtext{
 The immediate objective of this paper is to construct some "unbiased" estimate of the generation interval distribution (presumably, the intrinsic distribution $g(\tau)$).
@@
 
-\rev{
+\revtext{
 I also recommend, for the authors to consider, breaking this paper

### 2020-05-20 doc/response.tex
what is this

--- 2020-05-20 doc/response.tex
@@
-\documentclass[12pt]{article}
 \usepackage[utf8]{inputenc}
@@
 properly).}
-
-Lines 42-43 refers to a single measurement, rather than a distribution. We have made this clear (see next comment).
 
@@
 
-It is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
-See \swp{How do I respond to this, actually?}:
-
-``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''
+\swp{Incubation, reporting delays are concerns; we're not worried about that here.}
 
@@
 To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.
+
+\swp{Leave it to JD.}
 

### 2020-05-20 doc/response.tex
Sectioning
done
oops!

--- 2020-05-20 doc/response.tex
@@
+\documentclass[12pt]{article}
 \usepackage[utf8]{inputenc}
@@
 properly).}
+
+Lines 42-43 refers to a single measurement, rather than a distribution. We have made this clear (see next comment).
 
@@
 
-\swp{Incubation, reporting delays are concerns; we're not worried about that here.}
+It is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
+See \swp{How do I respond to this, actually?}:
+
+``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''
 
@@
 To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.
-
-\swp{Leave it to JD.}
 

### 2020-05-21 doc/response.tex
final meeting merge
Take a break, hope to return at least once more tonight.

--- 2020-05-20 doc/response.tex
@@
 \newcommand{\revtext}{\textsf}
+\newcommand{\newtext}[1]{\textsl{``#1''}}
 \setlength{\parskip}{\baselineskip}
@@
 
+There are many ways to measure a generation interval.
+We are interested here in exploring these different definitions, and in finding how these definition affect the \rR\ link.
+We have tried to clarify this throughout the revised MS.
+
 \revtext{Lines 42-43: "A generation interval is defined as the time between when a
@@
 
-We make clear that this definition corresponds to the definition at the individual-level (i.e., a single interval), which is unambiguous, and distinguish it from the population-level distributions:
+We start with the standard conceptual definition. There is no way to define GI that is both general and precise. We develop the theme of different definitions of GI throught the paper.
 
-``At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
-While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.''
+We now start earlier to discuss these issues. At this point we added text:
 
-\revtext{Line 46: "...intrinsic generation intervals..." This is not yet defined. Is it
-equivalent to the definition in Lines 42-43?}
+\newtext{At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
+While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.}
 
-We now define intrinsic generation interval properly:
-
-``The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.''
+\revtext{Line 46: "...intrinsic generation intervals..." This is not yet defined. Is it equivalent to the definition in Lines 42-43?}
 
@@
 
-We have now made this definition more precise. In Section 2, we also introduce the idea that the intrinsic distributions can be thought of as equilibrium distributions:
+We now define intrinsic generation interval properly:
 
-``$g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
-Since all infectious contacts lead to infection in a disease-free equilibrium of a homogeneously mixing population, $g(\tau)$ also describes the expected time distribution of secondary cases caused by an average primary case in a fully susceptible population and can be treated as an ``equilibrium'' distribution of the corresponding homogeneous mixing model.''
+\newtext{The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.}
+
+In Section 2, we also introduce the idea that the intrinsic distributions can be thought of as equilibrium distributions:
+
+\newtext{$g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
+If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed at the endemic equilibrium or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized generaton interval.}
 
@@
 
-We now define $g(\tau)$ as follows: ``expected time distribution of infectious contacts by a primary case''.
-
 \revtext{
@@
 
-Definition in Lines 42-43 does not correspond to the definition of the intrinsic generation-interval distribution. 
-It is a single measurement --- we tried to make it clear that it is not directly connected to any p.d.f.
-We have tried to make this clearer in the text.
+Our new definition of the generation interval and explanation in S2 (just above) address these questions to some extent. We prefer not to go deeper into the mathematical details for this paper.
 
@@
 
-``There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population,
+\newtext{There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population,
 while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
-Due to individual variation in infection time, the realized generation-interval distribution (at the population-level) can change depending on when and how it is defined \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.''
+Due to individual variation in infection time, the realized generation-interval distribution (at the population-level) can change depending on when and how it is defined \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.}
 
@@
 
-``While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
-we typically want to try to make inference based on all available data.''
+\newtext{While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
+we typically want to try to make inference based on all available data.}
 
@@
 We have tried to make this clear and no longer use the term ``effective'' generation-interval distribution.
+
+\jd{I liked effective, meaning that one that links correctly. I think it's still in the abstract.}
 
@@
 
-``Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.''
+\newtext{Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.}
 
@@
 
-``Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.''
+\newtext{Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.}
 
@@
 
-``In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
+\newtext{In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
 Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
-Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.''
+Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.}
 
@@
 
-``The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.''
+\newtext{The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.}
 
@@
 
-``In the main text, we present results based on exponentially distributed latent and infectious periods;
-we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.''
+\newtext{In the main text, we present results based on exponentially distributed latent and infectious periods;
+we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.}
 
@@
 
-``We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
+\newtext{We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
 The gamma approximation of the generation interval distribution has been widely used due to its simplicity;
@@
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).''
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).}
 
@@
 
-``The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom):
+\newtext{The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom):
 \begin{equation}
@@
 Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
-We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).''
+We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).}
 
@@
 
+\reviewer{Reviewer \#2}
+
+Thank you for your superlative review!!!!!!!
+
 \bibliography{networkprsb}

### 2020-05-22 doc/response.tex
not much progress, sorry
fixed minor comments from rev 2

@@
 Yes, we want to make inference on the \emph{initial} forward generation-interval distribution, rather than the intrinsic generation-interval distribution.
+This is the distribution that correctly links the initial values of $r$ and \Reff.
 We have tried to make this clear and no longer use the term ``effective'' generation-interval distribution.
@@
 
-Thank you for your superlative review!!!!!!!
+\revtext{
+This manuscript extends previous work to examine how observed generation intervals (estimated through contract tracing) are shaped by both temporal and spatial effects during the course of an epidemic and leverages this to construct two methods that correct for the time-censoring bias.
+These methods are then applied to simulated data with complete reporting and shown to accurately infer the reproductive number from the exponential growth rate.
+}
+
+\revtext{
+The manuscript is well-written and presents novel methods to correct observed generation intervals to obtain estimates of R0.
+When reading sections 3 and 4, I was wondering if there was a conceptual diagram that would help the reader visualize the different spatio-temporal processes that impact the observed generation-interval distributions (summarized on lines 224-226).
+Finally, while the methods are tested on simulated data with complete reporting, it's unclear how they would perform with incomplete reporting, and ultimately, with empirical data.
+Of course, the latter is beyond the scope of the current manuscript but it would be useful to include a straightforward treatment of simulated data with incomplete reporting to support the expectations discussed on lines 265-267.
+}
+
+\swp{TODO:}
+
+``For simplicity, we assume that generation264intervals are observed without error and assume that there is no under-reporting.
+We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process.''
+
+\revtext{Minor points:}
+
+\revtext{Figure 1: the average latent period is 11.4 days and the average infectious period is 5 days. However, these seem to be switched in Figures 2, 3 & 4. Why?}
+
+Thank you for pointing this out. We made typographical errors in figure captions. We have corrected this.
+
+\revtext{Line 145: "acquantanceships" -> "acquaintanceships"}
+
+Done.
+
+\revtext{Figure 3 caption, last sentence: has recovered -> have recovered}
+
+Done.
+
+\revtext{Line 270: direction -> directions }
+
+Done.
 

### 2020-05-22 doc/network.tex
fixed minor comments from rev 2

@@
 \caption{\textbf{Temporal variation in the mean observed generation interval.}
-A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
 The backward and the aggregated mean generation interval were calculated over the course of an epidemic.
@@
 Instead, we assume that a disease spreads on a network;
-infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.
+infected individuals contact their ``acquaintances'' at random, but ``acquaintanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.
 
@@
 All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
-mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
+mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
 Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
-Each simulation is run until all individuals are either susceptible or has recovered.
+Each simulation is run until all individuals are either susceptible or have recovered.
 }
@@
 Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
-Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$.
+Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$.
 Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
@@
 On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproductive number \cite{britton2019estimation}.
-We do not pursue these direction in this study.
+We do not pursue these directions in this study.
 

### 2020-05-23 analysis/cmp_compare_small.R
refactoring some code

@@ -118,9 +118,42 @@ intrinsic.est <- lapply(growth$r, function(x) 1/mean(exp(-x*gen)))
 
 network.est <- lapply(growth$r, R0fun.network)
 
-trapman.est <- lapply(reslist, trapman.R0, 100)
-
 empirical.est <- lapply(reslist, empirical.R0, 100)
+
+count <- 0
+
+forward.est <- sapply(reslist, function(x) {
+  count <<- count + 1
+  n <- 75
+  
+  o <- order(x$t_infected)
+  tmax <- tail(x$data$time, 1)
+  
+  n <- min(n, sum(x$t_recovered[o] < tmax, na.rm=TRUE))
+  o <- o[1:n]
+  
+  forwardgen <- unlist(lapply(o, function(v) {
+    x$t_infected[which(x$infected_by==v)]-x$t_infected[v]
+  }))
+  
+  1/mean(exp(-growth$r[count] * forwardgen))
+})
+
+forward.gen <- sapply(reslist, function(x) {
+  n <- 75
+  
+  o <- order(x$t_infected)
+  tmax <- tail(x$data$time, 1)
+  
+  n <- min(n, sum(x$t_recovered[o] < tmax, na.rm=TRUE))
+  o <- o[1:n]
+  
+  forwardgen <- unlist(lapply(o, function(v) {
+    x$t_infected[which(x$infected_by==v)]-x$t_infected[v]
+  }))
+  
+  mean(forwardgen)
+})
 
 RRdata <- data.frame(
 	observed=observed.est$estimate[observed.est$type=="RR"],
@@ -128,14 +161,15 @@ RRdata <- data.frame(
 	population=population.est$estimate[population.est$type=="RR"],
 	intrinsic=unlist(intrinsic.est),
 	network=unlist(network.est),
-	trapman=unlist(trapman.est),
+	forward=forward.est,
 	empirical=unlist(empirical.est)
 )
 
 gendata <- data.frame(
 	observed=observed.est$estimate[observed.est$type=="mean"],
 	individual=individual.est$estimate[individual.est$type=="mean"],
-	population=population.est$estimate[population.est$type=="mean"]
+	population=population.est$estimate[population.est$type=="mean"],
+	forward=forward.gen
 )
 
 save("RRdata", "gendata", file="cmp_compare_small.rda")

### 2020-05-23 fig/cmp.R

@@ -11,7 +11,7 @@ scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palet
 load("../analysis/cmp_compare.rda")
 
 RRgroup <- data.frame(
-	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "egocentric", "intrinsic"),
+	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "initial\nforward", "empirical", "egocentric", "intrinsic"),
 	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
 				 levels=c("tracing based", "empirical", "individual based"))
 )
@@ -19,11 +19,10 @@ RRgroup <- data.frame(
 RRdata2 <- RRdata %>%
 	gather(key, value) %>%
 	mutate(key=factor(key, 
-					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
+					  levels=c("observed", "population", "individual", "forward", "empirical", "network", "intrinsic"),
 					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
-					  		 "tree\nbased", "empirical", "egocentric", "intrinsic"))) %>%
-	merge(RRgroup) %>%
-	filter(key != "tree\nbased")
+					  		 "initial\nforward", "empirical", "egocentric", "intrinsic"))) %>%
+	merge(RRgroup)
 
 ggR <- ggplot(RRdata2) +
 	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
@@ -36,4 +35,34 @@ ggR <- ggplot(RRdata2) +
 		axis.title.x=element_blank()
 	)
 
+g2 <- ggplot(RRdata) +
+  geom_point(aes(forward, individual), shape=1) +
+  geom_abline(lty=2) +
+  scale_x_continuous("initial forward") +
+  scale_y_continuous("individual correction") +
+  theme(
+    panel.grid = element_blank()
+  )
+
+g3 <- ggplot(RRdata) +
+  geom_point(aes(forward, population), shape=1) +
+  geom_abline(lty=2) +
+  scale_x_continuous("initial forward") +
+  scale_y_continuous("population correction") +
+  theme(
+    panel.grid = element_blank()
+  )
+
[…]

### 2020-05-23 fig/cmp_seminr.R

@@ -11,19 +11,18 @@ scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palet
 load("../analysis/cmp_compare_seminr.rda")
 
 RRgroup <- data.frame(
-	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "egocentric", "intrinsic"),
-	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
-				 levels=c("tracing based", "empirical", "individual based"))
+  key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "initial\nforward", "empirical", "egocentric", "intrinsic"),
+  group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
+               levels=c("tracing based", "empirical", "individual based"))
 )
 
 RRdata2 <- RRdata %>%
-	gather(key, value) %>%
-	mutate(key=factor(key, 
-					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
-					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
-					  		 "tree\nbased", "empirical", "egocentric", "intrinsic"))) %>%
-	merge(RRgroup) %>%
-	filter(key != "tree\nbased")
+  gather(key, value) %>%
+  mutate(key=factor(key, 
+                    levels=c("observed", "population", "individual", "forward", "empirical", "network", "intrinsic"),
+                    labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
+                             "initial\nforward", "empirical", "egocentric", "intrinsic"))) %>%
+  merge(RRgroup)
 
 ggR <- ggplot(RRdata2) +
 	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +

### 2020-05-23 fig/cmp_small.R

@@ -11,23 +11,22 @@ scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palet
 load("../analysis/cmp_compare_small.rda")
 
 RRgroup <- data.frame(
-	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "tree\nbased", "empirical", "egocentric", "intrinsic"),
-	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
-				 levels=c("tracing based", "empirical", "individual based"))
+  key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "initial\nforward", "empirical", "egocentric", "intrinsic"),
+  group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
+               levels=c("tracing based", "empirical", "individual based"))
 )
 
 RRdata2 <- RRdata %>%
-	gather(key, value) %>%
-	mutate(key=factor(key, 
-					  levels=c("observed", "population", "individual", "trapman", "empirical", "network", "intrinsic"),
-					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
-					  		 "tree\nbased", "empirical", "egocentric", "intrinsic"))) %>%
-	merge(RRgroup) %>%
-	filter(key != "tree\nbased")
+  gather(key, value) %>%
+  mutate(key=factor(key, 
+                    levels=c("observed", "population", "individual", "forward", "empirical", "network", "intrinsic"),
+                    labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
+                             "initial\nforward", "empirical", "egocentric", "intrinsic"))) %>%
+  merge(RRgroup)
 
 ggR <- ggplot(RRdata2) +
 	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
-	scale_y_log10("Reproductive number", breaks=c(1, 2, 3, 4, 8, 16)) +
+	scale_y_log10("Reproductive number", breaks=c(1, 2, 4, 8, 16)) +
 	facet_grid(~group, scale="free", space="free_x") +
 	theme(
 		panel.spacing.x = unit(0, "lines"),

### 2020-05-23 doc/response.tex
one more analysis

@@
 The manuscript is well-written and presents novel methods to correct observed generation intervals to obtain estimates of R0.
-When reading sections 3 and 4, I was wondering if there was a conceptual diagram that would help the reader visualize the different spatio-temporal processes that impact the observed generation-interval distributions (summarized on lines 224-226).
-Finally, while the methods are tested on simulated data with complete reporting, it's unclear how they would perform with incomplete reporting, and ultimately, with empirical data.
-Of course, the latter is beyond the scope of the current manuscript but it would be useful to include a straightforward treatment of simulated data with incomplete reporting to support the expectations discussed on lines 265-267.
-}
+When reading sections 3 and 4, I was wondering if there was a conceptual diagram that would help the reader visualize the different spatio-temporal processes that impact the observed generation-interval distributions (summarized on lines 224-226).}
 
-\swp{TODO:}
+\revtext{Finally, while the methods are tested on simulated data with complete reporting, it's unclear how they would perform with incomplete reporting, and ultimately, with empirical data.
+Of course, the latter is beyond the scope of the current manuscript but it would be useful to include a straightforward treatment of simulated data with incomplete reporting to support the expectations discussed on lines 265-267.}
 
-``For simplicity, we assume that generation264intervals are observed without error and assume that there is no under-reporting.
-We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process.''
+We have added a new figure to the supplementary figures to account for incomplete reporting.
 

### 2020-05-23 analysis/cmp_compare_underreport.R

@@ -1,0 +1,184 @@
+library(igraph)
+library(dplyr)
+library(bbmle)
+library(epigrowthfit)
+source("../sim/ebola_param_seminr.R")
+source("../R/generation.R")
+source("../R/estimate.R")
+source("../R/empirical.R")
+
+rprob <- 0.3
+
+load("../sim/cmp_seir_sim.rda")
+
+intrinsic_fun <- function(tau) {
+	sigma*gamma/(gamma-sigma) * (exp(-sigma*tau)-exp(-gamma*tau))
+}
+
+conditional_fun <- function(tau) {
+	sigma*(gamma+lambda)/(gamma+lambda-sigma) * (exp(-sigma*tau)-exp(-(gamma+lambda)*tau))
+}
+
+true.var <- integrate(function(x) (x-true.mean)^2 * intrinsic_fun(x), 0, Inf)[[1]]
+
+true.shape <- true.mean^2/true.var
+
+cmpGraph <- read.table("../data/CA-CondMat.txt")
+cmpGraph <- graph.data.frame(cmpGraph)
+cmpGraph <- as.undirected(cmpGraph)
+cmpGraph <- delete_vertex_attr(cmpGraph, "name")
+
+d <- degree(cmpGraph)
+
+kappa <- var(d)/mean(d) + mean(d) -1
+
+## This is SEIR
+R0fun.network <- function(r) (gamma+r)/(gamma*sigma/(sigma+r)+r/kappa)
+
+## This is SE2IR
+## R0fun.network <- function(r) {
+##	lambda <- (2 * sigma+r)^2 * (r + gamma)/((kappa-1) * 4 * sigma^2 - 4 * sigma * r - r^2)
+##	
+##	kappa * lambda/(lambda + gamma)
+## }
+
+censor.gi <- lapply(
+	reslist
+	, network.generation
+	, plot=FALSE
+	, interval=c(10, 1010)
+	, interval.type="cases"
+)
+
+count <- 0
+
+growth <- lapply(reslist, function(x){
+	count <<- count + 1
+	print(count)
+	day <- floor(x$data$time) + 1
+	
+	fitdata <- data.frame(
+		day=unique(day),
+		cases=as.vector(table(day))
+	)
+	
+	ee <- epigrowthfit(time=fitdata$day,
+				 deaths=fitdata$cases,
+				 model="logistic")
+	
+	dd <- data.frame(r=coef(ee)[1])
+	
+	dd
+}) %>%
+	bind_rows(.id="sim")
+
+set.seed(101)
[…]

### 2020-05-23 fig/cmp_underreport.R

@@ -1,0 +1,38 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+source("../sim/ebola_param_seminr.R")
+
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+load("../analysis/cmp_compare_underreport.rda")
+
+RRgroup <- data.frame(
+	key=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", "initial\nforward", "empirical", "egocentric", "intrinsic"),
+	group=factor(c("tracing based", "tracing based", "tracing based", "empirical", "empirical", "individual based", "individual based"),
+				 levels=c("tracing based", "empirical", "individual based"))
+)
+
+RRdata2 <- RRdata %>%
+	gather(key, value) %>%
+	mutate(key=factor(key, 
+					  levels=c("observed", "population", "individual", "forward", "empirical", "network", "intrinsic"),
+					  labels=c("contact\ntracing", "population\ncorrection", "individual\ncorrection", 
+					  		 "initial\nforward", "empirical", "egocentric", "intrinsic"))) %>%
+	merge(RRgroup)
+
+ggR <- ggplot(RRdata2) +
+	geom_boxplot(aes(key, value, fill=key), alpha=0.7) +
+	scale_y_log10("Reproductive number", breaks=c(1, 2, 4, 8, 16)) +
+	facet_grid(~group, scale="free", space="free_x") +
+	theme(
+		panel.spacing.x = unit(0, "lines"),
+		strip.background = element_blank(),
+		legend.position = "none",
+		axis.title.x=element_blank()
+	)
+
+ggsave("cmp_reproductive_underreport.pdf", ggR, width=6, height=3)

### 2020-05-23 doc/network.tex
tweaks

@@
 
-\begin{figure}[!pbth]
+\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
@@
 
-\begin{figure}[!pbth]
+\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
-\caption{\textbf{Temporal variation in the mean observed generation interval.}
-A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
-The backward and the aggregated mean generation interval were calculated over the course of an epidemic.
+\caption{\textbf{Temporal variation in the mean backward and aggregated generation interval.}
+A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model is simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
+The backward and the aggregated mean generation interval are calculated over the course of an epidemic.
 The dotted horizontal line represents the mean intrinsic generation interval.
@@
 
-\begin{figure}[!pbth]
+\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{../fig/local_effect.pdf}
@@
 
-When realized generation intervals are aggregated over the course of an epidemic, there will be four effects present in the sample: (1) right-truncation effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../fig/diagram.pdf}
+\caption{
+\textbf{Spatiotemporal effects on generation intervals.}
+}
+\label{fig:diagram}
+\end{figure}
+
+When realized generation intervals are aggregated over the course of an epidemic, there will be four effects present in the sample (\fref{diagram}): (1) right-truncation effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-truncation effect;
@@
 
-\begin{figure}[!pbth]
+\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}

### 2020-05-23 fig/diagram.tex
tweaks

\documentclass{article}

\usepackage{graphics}
\usepackage[vcentering]{geometry}

\geometry{papersize={9.25in,2.4in}, top=0.05in, bottom=0in, left=0in, right=0in}

\input{tikzlib.tex}
\input{diagram_geom.tex}

% \definecolor{c1}{RGB}{217, 95, 2}
% \definecolor{c2}{RGB}{117, 112, 179}
% \definecolor{c3}{RGB}{231, 41, 138}
% \definecolor{c4}{RGB}{102, 166, 30}

\definecolor{localc}{RGB}{236, 175, 100}
\definecolor{effc}{RGB}{186, 184, 217}
\definecolor{obsc}{RGB}{243, 148, 197}
\definecolor{intc}{RGB}{153, 209, 89}

\usetikzlibrary{arrows.meta}

\begin{document}

\begin{tikzpicture}
\node(int)[bigcompartment, text width=5cm, fill opacity=0.7, text opacity=1]{
    \textbf{Intrinsic GI}
    \footnotesize{
        \begin{itemize}
        \item expected time distribution of infectious contacts made by a primary case
        \end{itemize}
    }
};

\node(local)[bigcompartment, right=3.4cm of int, text width=5cm, fill opacity=0.7, text opacity=1]{
    \textbf{Realized GI}
    \footnotesize{
        \begin{itemize}
        	\item time distribution between actual infection events
        	\item egocentric depletion, local depletion, global depletion
        \end{itemize}
    }
};

\node(forward)[bigcompartment, below=1.61cm of local, text width=5cm, fill opacity=0.7, text opacity=1]{
    \textbf{Initial forward GI}
    \footnotesize{
        \begin{itemize}
        	\item reflects network structure, but corrects for time truncation
        	\item \textbf{gives the correct link} between $r$ and $\mathcal{R}_{\textrm{\tiny initial}}$
        \end{itemize}
    }
};

\node(agg)[bigcompartment, right=3.4cm of local, text width=5cm, fill opacity=0.7, text opacity=1]{
    \textbf{Aggregated GI}
    \footnotesize{
        \begin{itemize}
        	\item analogous to contact tracing data
        	\item truncated at epidemic time
        \end{itemize}
    }
};

\node(iniagg)[bigcompartment, below=1.61cm of agg, text width=5cm, fill opacity=0.7, text opacity=1]{
    \textbf{Initial aggregated GI}
    \footnotesize{
        \begin{itemize}
        	\item aggregated GI during the exponential growth phase
        	\item truncation bias is understood exactly
        \end{itemize}
    }
};
\draw[->, very thick, dashed] (int) -- node [midway, left, text width=2cm, align=right] {} (local);
\draw[->, very thick, dashed] (local) -- node [midway, right, text width=2cm] {exponential growth \\ phase} (forward);
\draw[->, very thick, dashed] (local) -- node [midway, right, text width=2cm] {} (agg);
\draw[->, very thick, dashed] (agg) -- node [midway, right, text width=2cm] {exponential growth \\ phase} (iniagg);
\draw[->, very thick] (iniagg) -- node [midway, above, text width=1.5cm] {temporal \\ correction} (forward);

\end{tikzpicture}

\end{document}

### 2020-05-23 fig/diagram_geom.tex
tweaks

\tikzset{
  auto,
  compartment/.style={
    rectangle, minimum size=9mm, rounded corners=2mm,
    thick, draw=black!15, top color=white,bottom color=black!30
  },
  %
  bigcompartment/.style={
    rectangle, minimum size=20mm, rounded corners=2mm,
    thick, draw=black!30, top color=white,bottom color=black!20
  },
  %
  point/.style={
    circle, inner sep=2pt, fill=black!5
  },
  %
  mytextbox/.style={
    rectangle, text=black!50, thin, 
    draw=white, top color=white,bottom color=white, fill=white
  }
}

### 2020-05-23 fig/temporal_effect.R

@@ -72,7 +72,7 @@ sumfun <- function(out,
 	
 	censor.df <- data.frame(
 		t=tau
-		, censored=cm
+		, aggregated=cm
 		, backward=bm
 	)
 }
@@ -117,8 +117,8 @@ g2 <- ggplot(gendata, aes(t, value, lty=key)) +
     geom_line(lwd=1.2) +
     geom_hline(yintercept = 1/sigma + 1/gamma, lty=3) +
     scale_x_continuous("time (days)", expand=c(0,0), limits=c(0, 300), breaks=seq(0, 250, by=50)) +
-    scale_y_continuous("mean observed GI (days)", expand=c(0,0), limits=c(0, 30)) +
-	scale_linetype_manual(values=c(2, 1)) +
+    scale_y_continuous("mean GI (days)", expand=c(0,0), limits=c(0, 30)) +
+	scale_linetype_manual(values=c(1, 2)) +
 	# facet_wrap(~type) +
 	theme(
         panel.grid=element_blank(),
@@ -128,10 +128,10 @@ g2 <- ggplot(gendata, aes(t, value, lty=key)) +
         strip.text = element_blank(),
         panel.spacing = unit(0, "cm"),
         legend.title = element_blank(),
-        legend.position = c(0.15, 0.87),
+        legend.position = c(0.16, 0.8),
         legend.key.width = grid::unit(2, "cm")
     )
 
 gg_temporal <- arrangeGrob(g1, g2, nrow=2, heights=c(0.4, 0.6))
 
-ggsave("temporal_effect.pdf", gg_temporal, width=6, height=6)
+ggsave("temporal_effect.pdf", gg_temporal, width=6, height=4)

### 2020-05-23 doc/network.tex
supp

@@
 For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting. 
-We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
+We do not expect under-reporting to affect the inference of generation-interval distributions (Appendix A.3) unless there are systematic biases in the observation process.
 On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproductive number \cite{britton2019estimation}.
@@
 For smaller values of \RR, we expect the differences to become smaller.
-In the electronic supplementary material, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
+In the Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
 Overall, our qualitative conclusions do not change.
@@
 we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ \citep{park2019practical}.
-However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
+However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.4).
 

### 2020-05-23 doc/supp.tex
supp

@@
 
+\subsection{Comparison of estimates of reproductive number -- under-reporting of generation intervals}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{../fig/cmp_reproductive_underreport.pdf}
+\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
+This figure matches Fig. 4 in the main text.
+We assume that infected indidividuals have 30\% binomial probability of reporting their generation intervals (time between their infection and the infection of their infectors) when we apply population- and individual-based methods to estimate \Rini;
+under-reporting of generation intervals still give unbiased estimates of \Rini.
+All other parameters and estimates are the same as in Fig. 4 in the main text.
+}
+\label{fig:cmpsmall}
+\end{figure}
+
+\pagebreak
+
 \subsection{Testing the individual-based method on simulations on a homogeneous network}
@@
 \caption{\textbf{Wrong distributional assumption may result in biased estimates of the parameters of a generation-interval distribution.}
-We simulate a stochastic SEIR model on a homogenous network with $10^5$ individuals using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$, mean infectious period $1/\gamma = 11.4 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
+We simulate a stochastic SEIR model on a homogenous network with $10^5$ individuals using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
 Then, we apply the individual-based method based on the first 1000 infections.
@@
 Dashed horizontal lines represent the 95\% coverage probability.
-(D) Estimated generation-interval distributions based on the gamma approximation and a true generation-interval distribution of an SEIR model.
+(D) Estimated generation-interval distributions based on the gamma approximation and the true generation-interval distribution of an SEIR model.
 (E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.

### 2020-05-23 doc/network.tex
some tweaks

@@
 While an epidemic is ongoing, realized generation intervals until the current epidemic time can be aggregated to form a single distribution.
-While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
+While realized generation intervals can be difficult to measure in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
 we typically want to try to make inference based on all available data.

### 2020-05-23 doc/network.tex
one more

@@
 We then use the estimated distribution $f_0$ to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
-We expect estimates of $\Lambda$ to be sensitive to under-reporting.
+When the entire transmission process is known, we expect $\Lambda$ to match $\Rini$;
+otherwise, $\Lambda$ will be sensitive to under-reporting of number of infections caused by each infected individual.
+As we show in Appendix A.3, the estimates of $\Rini$ using the Euler-Lotka equation from $f_0$ remains unbiased even in the presence of under-reporting.
 White and Pagano \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.

### 2020-05-24 doc/response.tex
tweaks
still not done with the crazy reviewer!

@@
 Thank you for the chance to revise and resubmit.
-We apologize for the delay.
+%% We apologize for the delay. (They don't care.)
 Reviewer 1 suggested that we should divide the paper into two: theory and application.
@@
 
-We now start earlier to discuss these issues. At this point we added text:
+We now discuss these issues. At this point we added text:
 
@@
 
-We agree that calling it ``observed'' can be misleading. We now refer to them as \emph{realized} intervals.
+We appreciate this point. We now refer to them as \emph{realized} intervals.
 
@@
 
-\swp{I don't know.}
+We appreciate this point and have tried not to imply strong assumptions, and also taken care to empirically validate our statistical approaches.
 
@@
 
-We now use truncation instead of censoring.
+Done.
 
@@
 We now define both backward and forward generation-interval distributions more appropriately in the text.
-We no longer use the term ``censored'' generation-interval distribution. 
-We refer to it as the aggregated generation-interval distribution, which is a distribution of all realized generation intervals until certain time point.
-We think thie definition provides a clearer definition.
+We now refer to $c$ as the ``aggregated'' generation-interval distribution (all realized generation intervals until certain time point).
 
@@
 
-We have removed this paragraph and tried to make the explanation more concise.
-We no longer use the term effective generation-interval distribution but rather refer to it as the initial forward generation-interval distribution.
+Thank you so much for telling us to remove all examples of censoring and then continuing to tell us to remove specific examples of censoring!
+Also, censoring is very commonly used in this context, so you were being pedantic in the first place.
+But we removed it, and can you shut up now?
+\jd{Maybe try to tone this text down a little bit?}
+
+We have shortened and tried to improve this explanation.
 
@@
 We now show that the initial forward generation-interval distribution is equivalent to the intrinsic distribution.
-We also point to the analytical intractability of the spatial effects on the realized generation-interval distributions (and therefore the initial forward distribution) and therefore cannot be expressed mathematically (for non-homogeneous cases):
+We also point to the analytical intractability of the spatial effects on the realized generation-interval distributions (and therefore the initial forward distribution):
 
@@
 We no longer use $\RR$ to avoid confusion.
-We now use $\Rini = \Ro S(0)$ to reflect the possibility that $S(0)$ may not be 1 but clearly convey the idea that we're not estimating time dependent reproduction number.
+We now use $\Rini = \Ro S(0)$ to convey the idea that we consider the possibility that $S(0)$ may not be 1 but are not explicitly considering how $S$ changes through time.
 
@@
 
-We make these distinctions clear in the main text.
+We have tried to clarify these points. $S$ is a proportion, but we do attempt to address issues of individuality appropriately.
+
+\jd{Is the reviewer getting paid by the word, or what?}
 
@@
 As we show in this paper, when there is spatial structure, the realized generation-interval distributions differ from the intrinsic one. 
-In particular, even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egeocentric effect.
+Even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egocentric effect.
 Therefore, the realized generation intervals of a single, primary infector in an otherwise susceptible population will not follow the intrinsic generation-interval distribution.
-We mention in the manuscript that the intrinsic distribution can be interpreted as an equilibrium distribution but do not want to emphasize on using the term ``equilibrium'' particularly given that we're interested in realized distributions on a spatially heterogeneous network.
-The reviewer also presented a series of formulas on the mean generation interval for special cases of the SEIR model without any suggestions;
-given that our goal is not to explore all special cases of the SEIR model, we do not think presenting all those formulas is relevant to the discussion of the paper.
+We do mention in the manuscript that the intrinsic distribution can be interpreted as an equilibrium distribution but do not want to over-emphasize this perspective, since we're interested in realized distributions on a spatial network.
+The reviewer also presented a series of formulas on the mean generation interval for special cases of the SEIR model; these are interesting, but not a direction that we want to go with this version of the paper.
 
@@
 
-We now define the intrinsic generation-interval distribution as follows as we think it is more concrete; it is also easier to distinguish intrinsic vs. realized distributiosn under this definition:
+We now define the intrinsic generation-interval distribution as follows as we think it is more concrete; it is also easier to distinguish intrinsic vs. realized distributions under this definition:
 
@@
 
-\swp{???}
-
 \revtext{Observed exponential or sub-exponential growth events are different from the infection events. The current paper only addresses the infection events (mostly unobservable). This was
@@
 
-It is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
-See \swp{How do I respond to this, actually?}:
+We have tried to clarify points related to this objection in the MS. 
+Briefly, $r$ can be defined as the logarithmic rate of change in incidence $\dot i/i$.
+It changes through time, but we are interested in estimating the value near the beginning (after transients, but before appreciable depletion of susceptibles).
+To do this with (simulated or real) data, we use a method that estimates the initial value of $r$, while making a first-order correction for the fact that it might be changing.
 
 ``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''
+
+We do expect the initial $r$, $R$ and forward generation time to obey Euler-Lotka.
 
@@
 
-``The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
-$n \to \infty$'' 
-This is inaccurate.
-It is still possible for exponential growth to occur when the population is not fully susceptible ($S(t) \approx S(0) \neq 1$) as long as the depletion of the susceptible population is negligible.
+The reviewer is apparently conflating the general case, which would in fact be problematic, with the case we consider: where the \emph{initial} population is not fully susceptible.
+In the latter case, there is a period where incidence is low and the depletion of the susceptible population is negligible.
 To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.
@@
 only suitable for latent and infectious periods with very small variances
-}
-
-The exponential distribution is a special case of the Erlang distribution (when the shape parameter is equal to 1).
-We disagree the statement that it is ``too strong'' to say that a more general distribution is more realistic than its special case.
-To avoid confusion, we now say refer to Erlang distributions as ``more flexible''.
-
-\revtext{
 The Erlang distribution is not only a subset of the gamma distribution with integer valued shape parameters, but also it is limited to distributions with variances smaller than the corresponding exponential distribution with the same mean value.
@@
 
-The Erlang distribution includes the exponential distribution.
+The exponential distribution is a special case of the Erlang distribution (when the shape parameter is equal to 1). So our statement is at least technically correct.
+We agree that we should not over-emphasize the flexibility of the Erlang distribution.
+We now say ``often'' more realistic.
 
@@
 }
-
-One can at least claim that the gamma distribution is more realistic than its subset, the exponential distribution.
 
@@
 
+The reviewers discussion of and suggestions about Laplace transformations are interesting, but we think beyond the scope of this paper.
+Our experience has been that Laplace transformations and moment generating functions are as likely to obscure the key concepts underlying disease dynamics as to clarify them.
+
+\jd{I would skip the following text.}
+
 The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial effects.
@@
 we do not think this is relevant to the discussion of our paper.
+
+\jd{Had to stop here, sorry.}
 
@@
 
+We have added a diagram to summarize these effects.
+
 \revtext{Finally, while the methods are tested on simulated data with complete reporting, it's unclear how they would perform with incomplete reporting, and ultimately, with empirical data.
@@
 
-\revtext{Figure 1: the average latent period is 11.4 days and the average infectious period is 5 days. However, these seem to be switched in Figures 2, 3 & 4. Why?}
+\revtext{Figure 1: the average latent period is 11.4 days and the average infectious period is 5 days. However, these seem to be switched in Figures 2, 3 \& 4. Why?}
 

### 2020-05-25 doc/response.tex
holy moses!

@@
 
-\jd{Had to stop here, sorry.}
-
 \revtext{How to assess if the gamma distribution is a good choice as a
@@
 
+These are hard issues, but beyond the scope of this paper. Using exponentials is routine in many applications. Using gammas (or Erlangs) generalizes this, and is also commonly done. 
 The gamma distribution has been previously tested for many diseases. 
 In our case, Figure 4 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
-We discuss potential caveats in assuming a shape of the generation-interval distribution in Discussion.
+We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as cuss potential caveats in assuming a shape of the generation-interval distribution in Discussion.
 
@@
 
+\jd{I am not sure what the last part could mean, except that some biases don't turn out as bad as others -- they can't cancel exactly, so we shouldn't say unbiased estimates...
+Am I missing something?}
+
 \revtext{How does the likelihood
@@
 The population-level method does not take into account the observation scheme or the infection process. 
-We simply fit the gamma distribution to all realized generation intervals (i.e., the aggregated distribution);
-we feel that writing down the likelihood function is an overkill for a basic gamma fit.
+We simply fit the gamma distribution to all realized generation intervals (i.e., the aggregated distribution).
 We have clarified the text:
@@
 In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
-specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
-Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
+specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a by fitting a gamma distribution to the observed values by maximum likelihood.
+The inferred initial forward generation-interval distribution also follows a gamma distribution with the same shape, but with mean $\alpha/(\alpha/\bar G - r)$.
 We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).}
@@
 obvious and this needs to be proven.
-}
-
-We had never equated the true distribution to the intrinsic distribution.
-We make clear that the intrinsic distributions are not our goal.
-
-\revtext{
 The intrinsic distribution $g(\tau)$ is an equilibrium distribution that
@@
 }
+
+We now discuss this point in much more detail (and have discussed these points above).
+Briefly: we are interested in the initial forward distribution because this provides the \rR\ link.
+This approaches the intrinsic distribution in the case where mixing is homogeneous and the initial number infectious is small.
+We hope we have fixed all of the places where we failed to explain ourselves clearly.
 
@@
 }
+
+This is an important concern, and we have discussed this issue a bit more.
+We note that this is not always the case: for example, transmission is one of the most observable features in rabies dynamics.
+In Ebola, dates of transmission are often estimated.
+We are currently working to extend our framework to cases that explicitly treat other events as the observables.
 
@@
 
+See above for our discussion of these issues
+
 \reviewer{Reviewer \#2}

### 2020-05-25 doc/response.tex
tweaks

@@
 }
-
-Thank you so much for telling us to remove all examples of censoring and then continuing to tell us to remove specific examples of censoring!
-Also, censoring is very commonly used in this context, so you were being pedantic in the first place.
-But we removed it, and can you shut up now?
-\jd{Maybe try to tone this text down a little bit?}
 

### 2020-05-26 doc/response.tex
changes

@@
 Thank you for the chance to revise and resubmit.
-%% We apologize for the delay. (They don't care.)
 Reviewer 1 suggested that we should divide the paper into two: theory and application.
@@
 
-\newtext{There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population,
-while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
-Due to individual variation in infection time, the realized generation-interval distribution (at the population-level) can change depending on when and how it is defined \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.}
+\newtext{There are important distinctions to be made when defining generation-interval distributions at the population-level.
+The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.
+On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
+Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.}
 
@@
 
-\newtext{While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
-we typically want to try to make inference based on all available data.}
+\newtext{While an epidemic is ongoing, realized generation intervals can be measured by identifying who infected whom and can be aggregated to form a single distribution.
+In practice, realized generation intervals is difficult to measure because it is difficult to observe the exact moment at which individuals become infected;
+in most cases, observable events are clinical (e.g. onset of symptoms).
+However, it is still possible to directly measure generation intervals through contact tracing in some cases (e.g., a rabid dog biting another dog \citep{hampson2009transmission}),.
+Thus, the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic ---
+we typically want to try to make inference based on all available data that have been gathered since the beginning of an epidemic.}
 
@@
 Yes, we want to make inference on the \emph{initial} forward generation-interval distribution, rather than the intrinsic generation-interval distribution.
-This is the distribution that correctly links the initial values of $r$ and \Reff.
+This is the distribution that correctly links the initial values of $r$ and \Rini.
 We have tried to make this clear and no longer use the term ``effective'' generation-interval distribution.
-
-\jd{I liked effective, meaning that one that links correctly. I think it's still in the abstract.}
 
@@
 
+\swp{??? What do you mean by ``also taken care to empirically validate our statistical approaches''}
+
 \revtext{This paper addresses length-bias in the "backward generation interval" and
@@
 We now define both backward and forward generation-interval distributions more appropriately in the text.
-We now refer to $c$ as the ``aggregated'' generation-interval distribution (all realized generation intervals until certain time point).
+We now use $a_t$ to refer to the ``aggregated'' generation-interval distribution (all realized generation intervals until certain time point) and do not use $c_t$.
 
@@
 
-\jd{Is the reviewer getting paid by the word, or what?}
-
 \revtext{
@@
 
-\swp{TODO? Just say we are interesting in the initial phase? maybe write something in the discussion? We're only using $\Rini$}
+We are interested in the initial value of $\RR$, which we now refer to as $\Rini = \Ro S(0)$. We have tried to make this clear throughout the manuscript.
 
@@
 
-\newtext{In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
+\newtext{In this case, with constant transmission rate during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
 Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
@@
 To do this with (simulated or real) data, we use a method that estimates the initial value of $r$, while making a first-order correction for the fact that it might be changing.
-
-``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''
-
-We do expect the initial $r$, $R$ and forward generation time to obey Euler-Lotka.
 
@@
 
-\jd{I would skip the following text.}
-
-The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial effects.
-The Laplace tranforms are only more useful for the intrinsic generation-interval distributions under the homogeneous assumption. 
-The forward generation-interval distribution in a spatially hetergeneous environment is analytically intractable and therefore, the Laplace transforms no longer hold.
-Given that our goal is to infer the initial forward generation-interval distribution, rather than the intrinsic distribution, we feel that the Laplace transforms are not necessarily helpful for the discussion.
-In particular, the reviewer asks us to prove special cases of the Euler-Lotka equation, which has been proven previosuly as suggested by the references, using Laplace transforms; 
-we do not think this is relevant to the discussion of our paper.
-
 \revtext{How to assess if the gamma distribution is a good choice as a
@@
 
-These are hard issues, but beyond the scope of this paper. Using exponentials is routine in many applications. Using gammas (or Erlangs) generalizes this, and is also commonly done. 
+These are hard issues, but beyond the scope of this paper. 
+Using exponentials is routine in many applications. 
+Using gammas (or Erlangs) generalizes this, and is also commonly done. 
 The gamma distribution has been previously tested for many diseases. 
 In our case, Figure 4 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
-We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as cuss potential caveats in assuming a shape of the generation-interval distribution in Discussion.
+We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as potential caveats in assuming a shape of the generation-interval distribution in Discussion.
 
@@
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.3).}
-
-\jd{I am not sure what the last part could mean, except that some biases don't turn out as bad as others -- they can't cancel exactly, so we shouldn't say unbiased estimates...
-Am I missing something?}
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estiamte of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of \Rini (Appendix A.4).}
 
@@
 
-\newtext{The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom):
+\newtext{The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom) \citep{tomba2010some, nishiura2010time, britton2019estimation}:
 \begin{equation}
@@
 In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
-specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a by fitting a gamma distribution to the observed values by maximum likelihood.
-The inferred initial forward generation-interval distribution also follows a gamma distribution with the same shape, but with mean $\alpha/(\alpha/\bar G - r)$.
+specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
+Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
 We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).}
@@
 
-\swp{TODO.}
+We had initially thought their approach is similar due to their use of Poisson likelihood.
+We realized that their approach is quite different from the goal of our paper and have removed the sentence.
+They do not use Euler-Lotka or measure generation intervals.
 
@@
 
-\swp{TODO.}
+We have added the following justification to the text:
+
+\newtext{
+We estimate the initial exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}:
+\begin{equation}
+c(t) = \frac{K}{1 + \left[(K/c_0)-1\right]e^{-rt}}.
+\end{equation}
+While the exponential growth rate (i.e., the rate of change in log incidence) of the logistic function changes throughout an epidemic,
+we focus strictly on estimating the \emph{initial} exponential growth rate (when $t \to -\infty$).
+The method of estimating the initial exponential growth rate by fitting a logistic curve has been previously validated against simulations of stochastic comparmental models \citep{ma2014estimating}.
+}
 
@@
 
-\revtext{Line 145: "acquantanceships" -> "acquaintanceships"}
+\revtext{Line 145: "acquantanceships" $\rightarrow$ "acquaintanceships"}
 
@@
 
-\revtext{Line 270: direction -> directions }
+\revtext{Line 270: direction $\rightarrow$ directions }
 

### 2020-05-26 doc/network.tex
changes

@@
 	\textbf\newline{
-		Inferring the generation-interval distribution that correctly links speed and strength of an epidemic
+		Inferring generation-interval distributions from contact-tracing data
 	}
@@
 David Champredon\textsuperscript{3,4}
-Jonathan Dushoff\textsuperscript{3,5}
+Jonathan Dushoff\textsuperscript{3,5,$\dagger$}
 \\
@@
 \bigskip
-\textbf{1} Department of Mathematics \& Statistics, McMaster University, Hamilton, Ontario, Canada
+\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, New Jersey, USA
 \\
-\textbf{2} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, New Jersey, USA
+\textbf{2} Department of Mathematics \& Statistics, McMaster University, Hamilton, Ontario, Canada
 \\
@@
 
-*swp2@princeton.edu
+*swp2@princeton.edu\\
+$\dagger$dushoff@mcmaster.ca
 \end{flushleft}
@@
 
-Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
-Generation intervals are often measured through contact tracing by identifying who infected whom.
-We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales.
-Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
-therefore, we predict that correcting observed intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly link $r$ and $\RR$.
+Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the initial reproductive number, $\Rini$, and the initial rate of exponential growth, $r$.
+Generation intervals can be measured through contact tracing by identifying who infected whom.
+We study how realized intervals differ from ``intrinsic'' intervals that could be estimated by measuring individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales.
+Early in an epidemic, we expect the variation in the realized generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
+therefore, we predict that correcting realized intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide the initial forward generation-interval distribution, which is spatially informed and correctly links $r$ and $\Rini$.
 We develop and test statistical methods for temporal corrections of generation intervals,
@@
 This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
-As a result, realized generation intervals (from the perspective of an infector) will typically have shorter mean than the intrinsic generation-interval distribution in a non-homogenous population.
+As a result, realized generation intervals (from the perspective of an infector) will typically have shorter mean than the intrinsic generation-interval distribution in a non-homogeneous population.
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
 
-While an epidemic is ongoing, realized generation intervals until the current epidemic time can be aggregated to form a single distribution.
-While realized generation intervals can be difficult to measure in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
-we typically want to try to make inference based on all available data.
+While an epidemic is ongoing, realized generation intervals can be measured by identifying who infected whom and can be aggregated to form a single distribution.
+In practice, realized generation intervals is difficult to measure because it is difficult to observe the exact moment at which individuals become infected;
+in most cases, observable events are clinical (e.g. onset of symptoms).
+However, it is still possible to directly measure generation intervals through contact tracing in some cases (e.g., a rabid dog biting another dog \citep{hampson2009transmission}),.
+Thus, the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic ---
+we typically want to try to make inference based on all available data that have been gathered since the beginning of an epidemic.
+
 These aggregated measurements are ``truncated'' because we do not know what happens after the current epidemic time.
 These truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
-These observations allow us to make the following prediction: removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which is ``spatially corrected'' (i.e., contains information about the population structure) and will allow us to correctly infer the initial reproductive number from the exponential growth rate.
+These observations allow us to make the following prediction: removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.
 
@@
 We classify spatial effects on realized generation intervals into three levels (egocentric, local, and global) and discuss how they affect realize generation-interval distributions.
-Finally, we compare two methods for accounting for temporal bias and test our prediction using simulations.
+Finally, we compare two methods for accounting for temporal bias and test our prediction using individual-based simulations.
 
@@
 \begin{equation}
-K(\tau) = \int k (\tau; a) f(a) da,
+K(\tau) = \int k (\tau; a) f(a) \,\mathrm{d}a,
 \end{equation}
@@
 \end{equation}
-where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population \citep{anderson1991infectious}),
+where $\RR_0 = \int K(\tau)\,\mathrm{d}\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population \citep{anderson1991infectious}),
 and $g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
-Since all infectious contacts lead to infection in a disease-free equilibrium of a homogeneously mixing population, $g(\tau)$ also describes the expected time distribution of secondary cases caused by an average primary case in a fully susceptible population and can be treated as an ``equilibrium'' distribution of the corresponding homogeneous mixing model.
+If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed at the endemic equilibrium or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized generation intervals.
 
@@
 \begin{equation}
-i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
+i(t) = S(t) \int K(s) i(t-s)\,\mathrm{d}s = \RR_0 S(t) \int g(s) i(t-s)\,\mathrm{d}s.
 \end{equation}
 This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
-Over a period of time where the susceptible proportion $S(t) = S(0)$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
+Over a period of time where the susceptible proportion $S(t) = S(0)$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
 \begin{equation}
-\frac{1}{\Rini} = \int g(s) \exp(-r s) ds.
+\frac{1}{\Rini} = \int g(s) \exp(-r s)\,\mathrm{d}s.
 \label{eq:euler}
@@
 Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
-The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by infectors who were infected at time $t$.
-Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} of infectees who were infected at time $t$.
+The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
+Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
 For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
@@
 \begin{equation}
-f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(\tau) S(t+\tau) d\tau}.
+f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(\tau) S(t+\tau) \,\mathrm{d}\tau}.
 \label{eq:forward}
 \end{equation}
-In this case, the initial foward generation-interval distribution during the exponential growth phase (when $S(t) \approx S(0)$) is equivalent to the intrinsic generation-interval distribution and, therefore, provides the correct link between $r$ and \Rini.
+In this case, the initial forward generation-interval distribution $f_0(\tau)$ during the exponential growth phase (when $S(t) \approx S(0)$) is equivalent to the intrinsic generation-interval distribution $g(\tau)$ and, therefore, provides the correct link between $r$ and \Rini.
 Likewise, the density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
@@
 \begin{equation}
-b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-\tau) g(\tau) d\tau}.
+b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-\tau) g(\tau)\,\mathrm{d}\tau}.
 \end{equation}
 
-During an outbreak, generation intervals cannot be measured for infection events that have not happened yet.
+During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
 This effect is called ``right-truncation''.
-Therefore, even if we aggregate all realized generation intervals over the course of an epidemic, their mean will be shorter than the intrinsic generation intervals.
+Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the intrinsic generation intervals.
 The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until the epidemic time $t$:
 \begin{equation}
-a_t(\tau) \propto \int_{-\infty}^t b_s(\tau) i(s) ds.
+a_t(\tau) \propto \int_{-\infty}^t b_s(\tau) i(s) \,\mathrm{d}s.
 \end{equation}
@@
 \begin{equation}
-a_t(\tau) \propto \int_{-\infty}^{t-\tau} f_s(\tau) i(s) ds.
+a_t(\tau) \propto \int_{-\infty}^{t-\tau} f_s(\tau) i(s) \,\mathrm{d}s.
 \end{equation}
@@
 A deterministic simulation confirms that the aggregated generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
-Second, the decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
-Overall, we expect naively using the aggregated generation-interval distribution, to underestimate the reproductive number.
+Second, the decreasing number of susceptibles over the course of an epidemic makes long infectious contacts less likely to result in infection \citep{champredon2015intrinsic}.
+Overall, we expect naively using the aggregated generation-interval distribution, to underestimate the initial reproductive number.
 
@@
 A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model is simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
-The backward and the aggregated mean generation interval are calculated over the course of an epidemic.
+The mean backward and aggregated generation interval are calculated over the course of an epidemic.
 The dotted horizontal line represents the mean intrinsic generation interval.
@@
 
-Infected individuals may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
-Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average.
+Infected individuals may contact the same susceptible individual multiple times, but only the first infectious contact gives rise to infection in a given individual (after this, they are no longer susceptible).
+Therefore, we expect realized generation intervals from an individual to have a smaller mean than their mean intrinsic generation interval.
 To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous.
@@
 \begin{equation}
-\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(s; a) ds\right),
+\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(s; a) \,\mathrm{d}s\right),
 \end{equation}
 where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by a particular infected individual.
-The dilution term, $\delta(a)$, models how contacts are distributed among susceptible acquaintances.
+The dilution term, $\delta(a)$, models how contacts are distributed among the acquaintances.
 
 Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
-In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) ds$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
+In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
 This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
-for brevity, we do not pursue these directions here.
+for simplicity, we do not pursue these directions here.
 
@@
 \begin{equation}\label{eq:ego}
-\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
+\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) \,\mathrm{d}a,
 \end{equation}
@@
 \begin{equation}
-\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
+\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s)\,\mathrm{d}s}.
 \label{eq:conditional}
@@
 The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
-Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
+Finally, the link between the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
 \begin{equation}
@@
 \end{equation}
-As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will always be smaller than $\Rini$ estimated from the intrinsic distribution;
-this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be typically smaller than $\Rini$ estimated from the intrinsic distribution;
+this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
 % Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.
@@
 \end{equation}
-In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
+In this case, with constant transmission rate during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$).
 Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
@@
 Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
-(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible.
+(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the contact results in infection.
 (Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
@@
 since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we predict the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
-Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.
+Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to validate our prediction.
 
@@
 \caption{
-\textbf{Spatiotemporal effects on generation intervals.}
+\textbf{A summary of spatiotemporal effects on generation intervals.}
+The intrinsic generation-interval distribution describe the expected time distribution of infectious contacts made by a primary case.
+The realized (forward) generation-interval distribution describe the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
+The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
+Our goal is to estimate the initial forward generation-interval distribution from the initial aggregated generation-interval distribution, measured during the exponential growth period.
 }
@@
 
-When realized generation intervals are aggregated over the course of an epidemic, there will be four effects present in the sample (\fref{diagram}): (1) right-truncation effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
+When realized generation intervals are aggregated over the course of an epidemic, there will be four effects present in the data (\fref{diagram}): (1) right-truncation effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
 We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-truncation effect;
@@
 both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.
-If we can correct for the temporal bias early in an outbreak, during the exponential growth phase, we should be able to infer the initial forward generation-interval distribution, which incorporates egocentric and local spatial effects but not the global effects, from the aggregated distribution.
+If we can correct for the truncation bias early in an outbreak, during the exponential growth phase, we should be able to infer the initial forward generation-interval distribution, which incorporates egocentric and local spatial effects but not the global effects, from the aggregated distribution.
 
@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-truncation severely underestimates the reproductive number.
+Using the aggregated generation-interval distributions (based on the first 1000 realized generation intervals) without correcting for right-truncation (labeled as `contact tracing') severely underestimates the reproductive number.
 Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
-Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
+Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 75 infected individuals, as well as the estimates based on the initial forward generation intervals, which are calculated by applying the Euler-Lotka equation to the realized generation intervals of all infections caused by the first 75 infected individuals.
 Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$.
@@
 To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
-and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
+and compare the estimates of initial reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 75 infected individuals,
+as well as the estimates of the initial reproductive number calculated from the initial forward generation intervals, which we define as the generation intervals for all infections caused by the first 75 infected individuals (\fref{cmp}).
 For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting.
@@
 Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
-While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
+While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the initial forward generation-interval distribution on average,
 population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
@@
 \section{Discussion}
-\swp{something about little r}
-\swp{something about big R; and time dependent R}
 
-The intrinsic generation-interval distribution, which describes the expected time distribution of infectous contacts, provides a direct link between speed (exponential growth rate, $r$) and strength (initial reproductive number, $\Rini$) of an epidemic in a homogeneously mixing population \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
+The intrinsic generation-interval distribution, which describes the expected time distribution of infectious contacts, provides a direct link between speed (initial exponential growth rate, $r$) and strength (initial reproductive number, $\Rini$) of an epidemic in a homogeneously mixing population \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
 However, realized generation-interval distributions can vary depending on how and when they are measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
 determining which distribution correctly links $r$ and $\Rini$ can be challenging.
-Here, we analyze how realized generation intervals differ from intrinsic generation intervals.
+Here, we analyze how realized generation intervals aggregated over the course of an epidemic, possibly through contact tracing, differ from intrinsic generation intervals.
 Changes due to right-truncation reflect observation bias, whereas changes due to spatial or network structure reflect the dynamics of the outbreak.
@@
 Realized generation intervals that have been aggregated over the course of an epidemic are subject to right-truncation -- it is not possible to trace individuals that have not been infected yet.
-The aggregated distributions can be thought of as averages of ``backward'' generation intervals (measured by looking at infectors of a cohort ofindividuals infected at the same time) \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+The aggregated distributions can be thought of as averages of ``backward'' generation intervals (measured by looking at infectors of a cohort of individuals infected at the same time) \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
 During an ongoing outbreak, the aggregated generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-truncation.
@@
 Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
-Therefore, we predicted the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right-truncation.
-Simulation studies confirmed our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\Rini$.
+Therefore, we predict the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right-truncation.
+Simulation studies confirm our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\Rini$.
 
@@
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may cancel out and give unbiased estimate of the reproductive number (Appendix A.4).
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of \Rini (Appendix A.4).
 
@@
 Different definitions, and different measurement approaches, produce different estimates of these distributions.
-We have shown that estimates based on aggregated generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
+We have shown that estimates based on aggregated generation intervals (e.g., measured through contact tracing) differ in predictable ways from intrinsic estimates based on underlying measures of infectiousness (e.g., from shedding studies).
 These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
-We have shown that correcting aggregated intervals for temporal effects allows us to estimate a spatially informed initial forward distribution, which accurately describes how disease spreads in a population.
+Correcting aggregated intervals for temporal effects allows us to estimate a spatially informed initial forward distribution, which accurately describes how disease spreads in a population.
 Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence the spread of disease.
@@
 We use a \seminar\ model, which extends the SEIR model to have multiple equivalent stages in the latent and infectious periods.
-This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more flexible than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
+This gives latent and infectious periods with Erlang distributions (gamma distributions with integer shape parameters, including the exponential distribution), which are often more realistic than the exponentially distributed periods in the standard SEIR model \citep{anderson1980spread, bailey1964some}:
 \begin{equation}
 \begin{aligned}
-\frac{dS}{dt} &= - \beta S \frac{\sum_{k=1}^{n_I} I_k}{N}\\
-\frac{dE_1}{dt} &= \beta S \frac{\sum_{k=1}^{n_I} I_k}{N} - n_E \sigma E_1\\
+\frac{dS}{dt} &= - \beta S \sum_{k=1}^{n_I} I_k\\
+\frac{dE_1}{dt} &= \beta S \sum_{k=1}^{n_I} I_k - n_E \sigma E_1\\
 \frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
@@
 \end{equation}
-where $S$ is the number of susceptible individuals, $E_m$ is the number of exposed individuals in the $m$-th compartment, $I_n$ is the number of infectious individuals in the $n$-th compartment, and $R$ is the number of recovered individuals.
-Parameters of the model are specified as follows: $N$ is the total population size, $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
+where $S$ is the proportion of susceptible individuals, $E_m$ is the proportion of exposed individuals in the $m$-th compartment, $I_n$ is the proportion of infectious individuals in the $n$-th compartment, and $R$ is the proportion of recovered individuals.
+Parameters of the model are specified as follows:  $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
+We scale the proportions of individuals in each compartment are scaled by the total population size $N$.
 In the main text, we present results based on exponentially distributed latent and infectious periods;
@@
 
-We simulated an individual-based SEIR model on a contact network, using an algorithm based on the ``Gillespie algorithm'' \cite{bartlett1953stochastic,gillespie1977exact}.
+We simulate an individual-based SEIR model on a contact network, using an algorithm based on the ``Gillespie algorithm'' \cite{bartlett1953stochastic,gillespie1977exact}.
 We begin by randomly selecting individuals assumed to be infected at $t = 0$.
@@
 
-Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \cite{trapman2016inferring} (see the original source for detailed derivations) matches what would be calculated by applying the Euler-Lotka equation to the egocentric (rather than the intrinsic) generation-interval distribution.
+Here, we show that the egocentric relationship between $r$ and $\RR$ derived by Trapman \textit{et al.} \cite{trapman2016inferring} (see the original source for detailed derivations) matches what would be calculated by applying the Euler-Lotka equation to the egocentric (rather than the intrinsic) generation-interval distribution.
 Assume that latent and infectious periods are exponentially distributed with mean $1/\sigma$ and $1/\gamma$, respectively.
@@
 \begin{equation}
-\hat\RR = \left(1 + \frac{r}{\sigma}\right) \left(1 + \frac{r}{\gamma + \lambda}\right),
+\Rego = \left(1 + \frac{r}{\sigma}\right) \left(1 + \frac{r}{\gamma + \lambda}\right),
 \end{equation}
 where $r$ is the exponential growth rate.
-Alternatively, the reproductive number can be expressed based on the degree distribution (mean $\mu$ and variance $\sigma^2$) of a network:
+Alternatively, the reproductive number can be expressed based on the degree distribution (mean $\mu$ and variance $v$) of a network:
 \begin{equation}
-\hat\RR = \underbrace{\kappa \lambda}_{\text{average contact rate}} \times \underbrace{\frac{1}{\gamma + \lambda}}_{\text{mean effective infectious period}},
+\Rego = \underbrace{\kappa \lambda}_{\text{average contact rate}} \times \underbrace{\frac{1}{\gamma + \lambda}}_{\text{mean effective infectious period}},
 \end{equation}
-where $\kappa = \sigma^2/\mu + \mu - 1$, referred to as the mean degree excess \citep{newman2003structure}, describes the expected number of susceptible individuals that an average infected individual will encounter early in an outbreak.
+where $\kappa = v/\mu + \mu - 1$, referred to as the mean degree excess \citep{newman2003structure}, describes the expected number of susceptible individuals that an average infected individual will encounter early in an outbreak.
 Combining the two equations, we get
@@
 \end{equation}
-which completes the relationship between the growth rate and the egocentric reproductive number:
+which completes the relationship between the growth rate and the egocentric reproductive number \citep{trapman2016inferring}:
 \begin{equation}
-\hat\RR = \frac{\gamma + r}{\gamma \sigma/(\sigma + r) + r/\kappa}.
+\Rego = \frac{\gamma + r}{\gamma \sigma/(\sigma + r) + r/\kappa}.
 \end{equation}
@@
 \begin{equation}
-f_0(\tau) \propto a_0(\tau) \exp(r\tau).
+f_0(\tau) \propto a_0(\tau) \exp(r\tau),
 \end{equation}
-In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that aggregated realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
+where $r$ is the initial exponential growth rate.
+In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
 specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
@@
 As we show in Appendix A.3, the estimates of $\Rini$ using the Euler-Lotka equation from $f_0$ remains unbiased even in the presence of under-reporting.
-White and Pagano \cite{forsberg2008likelihood} proposed a related approach based on discretized incidence reports.
 
@@
 
-We estimate the exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}. 
-Fitting directly to cumulative incidence can lead to overly confident results \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution.
-We estimate the logistic parameters $r$, $K$, $c_0$, and $\theta$ by maximum likelihood.
+We estimate the initial exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}:
+\begin{equation}
+c(t) = \frac{K}{1 + \left[(K/c_0)-1\right]e^{-rt}}.
+\end{equation}
+While the exponential growth rate (i.e., the rate of change in log incidence) of the logistic function changes throughout an epidemic,
+we focus strictly on estimating the \emph{initial} exponential growth rate (when $t \to -\infty$).
+The method of estimating the initial exponential growth rate by fitting a logistic curve has been previously validated against simulations of stochastic compartmental models \citep{ma2014estimating}.
+
+Fitting directly to cumulative incidence can lead to overly confident results \citep{king2015avoidable}; instead we fit interval incidence $x(t) = c(t + \Delta t) - c(t)$, where $\Delta t$ is 1 day, to daily incidence, assuming that daily incidence follows a negative binomial distribution with overdispersion parameter $\theta$.
+We estimate parameters $r$, $K$, $c_0$, and $\theta$ by maximum likelihood.
 Fitting time window is defined from the last trough before the peak of an epidemic to the first day after the peak of an epidemic.

### 2020-05-26 fig/cmp.R

@@ -35,34 +35,4 @@ ggR <- ggplot(RRdata2) +
 		axis.title.x=element_blank()
 	)
 
-g2 <- ggplot(RRdata) +
-  geom_point(aes(forward, individual), shape=1) +
-  geom_abline(lty=2) +
-  scale_x_continuous("initial forward") +
-  scale_y_continuous("individual correction") +
-  theme(
-    panel.grid = element_blank()
-  )
-
-g3 <- ggplot(RRdata) +
-  geom_point(aes(forward, population), shape=1) +
-  geom_abline(lty=2) +
-  scale_x_continuous("initial forward") +
-  scale_y_continuous("population correction") +
-  theme(
-    panel.grid = element_blank()
-  )
-
-g4 <- ggplot(RRdata) +
-  geom_point(aes(forward, empirical), shape=1) +
-  geom_abline(lty=2) +
-  scale_x_continuous("initial forward") +
-  scale_y_continuous("empirical") +
-  theme(
-    panel.grid = element_blank()
-  )
-
-gtot <- arrangeGrob(ggR, g2, g3, g4, layout_matrix = matrix(c(1, 1, 1, 2, 3, 4), nrow=3), widths=c(2, 1))
-
 ggsave("cmp_reproductive.pdf", ggR, width=6, height=3)
-ggsave("cmp_reproductive2.pdf", gtot, width=10, height=6)

### 2020-05-26 fig/individual_and_population.R

@@ -27,7 +27,7 @@ dd_population_poly <- data.frame(
 g1 <- ggplot(dd_individual) +
 	geom_line(aes(time, density), lwd=1.1) +
 	scale_x_continuous("Time (days)", expand=c(0, 0)) +
-	scale_y_continuous(expression(K[a](tau)), expand=c(0,0), limit=c(0, 0.45)) +
+	scale_y_continuous(expression(k(tau*";"~a)), expand=c(0,0), limit=c(0, 0.45)) +
 	facet_wrap(~type) +
 	theme(
 		strip.background = element_blank(),
@@ -45,5 +45,3 @@ gtot <- arrangeGrob(
 )
 
 ggsave("individual_and_population.pdf", gtot, width=6, height=3)
-
-

### 2020-05-26 doc/response.tex
done???

@@
 
-Our new definition of the generation interval and explanation in S2 (just above) address these questions to some extent. We prefer not to go deeper into the mathematical details for this paper.
+Our new definition of the generation interval and explanation in Section 2 (just above) address these questions to some extent. 
+The definition in Lines 42-43 we had earlier does not correspond to any p.d.f.
+We prefer not to go deeper into the mathematical details for this paper.
 
@@
 
+Thank you for pointing this out. 
+We realized that we were implicitly combining two likelihoods, one of which is based on the non-homogeneous Poisson process and the other is based on truncated distributions (described in suggested references).
+We now provide a more detailed explanation of the method with appropriate references.
+We also allude to the fact that our proposed likelihood is a special case of the likelihood derived in Kalbfleisch and Lawless (1989).
 The likelihood function is provided in the Methods section.
-\swp{TODO. It seems like they're different... Ugh. I'll get back to this.}
 
@@
 We have tried to clarify these points. $S$ is a proportion, but we do attempt to address issues of individuality appropriately.
+
+\swp{What do you mean we have tried to address issues of individuality? Can we just get rid of that phrase and say S is a proportion?}
 
@@
 
-The reviewers discussion of and suggestions about Laplace transformations are interesting, but we think beyond the scope of this paper.
+The reviewer's discussion of and suggestions about Laplace transformations are interesting, but we think beyond the scope of this paper.
 Our experience has been that Laplace transformations and moment generating functions are as likely to obscure the key concepts underlying disease dynamics as to clarify them.
@@
 The gamma distribution has been previously tested for many diseases. 
-In our case, Figure 4 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
+In our case, Figure 5 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
 We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as potential caveats in assuming a shape of the generation-interval distribution in Discussion.
@@
 \newtext{We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
-The gamma approximation of the generation interval distribution has been widely used due to its simplicity;
-we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$.
-However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.3), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
+The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
+we previously showed that summarizing the entire distribution with two moments (mean and variance) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
+However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estiamte of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of \Rini (Appendix A.4).}
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).}
 
@@
 
-We had initially thought their approach is similar due to their use of Poisson likelihood.
+We had included this reference in earlier versions of this manuscript a few years ago, thinking that their approach is similar due to their use of Poisson likelihood.
 We realized that their approach is quite different from the goal of our paper and have removed the sentence.
-They do not use Euler-Lotka or measure generation intervals.
+They do not use Euler-Lotka or rely on ``observed'' generation intervals;
+instead they try to infer a generation-interval distribution from incidence reports.
 
@@
 several generating-intervals and distributions mentioned in this paper, such as
-$g(\tau)$; $b_t(\tau)$; $c_t(\tau)$; $g_eff(\tau)$; etc. To the end, it seems that the primary focus is the
+$g(\tau)$; $b_t(\tau)$; $c_t(\tau)$; $g_{eff}(\tau)$; etc. To the end, it seems that the primary focus is the
 intrinsic distribution $g(\tau)$; and relating this distribution with two other intrinsic
@@
 We now discuss this point in much more detail (and have discussed these points above).
-Briefly: we are interested in the initial forward distribution because this provides the \rR\ link.
+Briefly: we are interested in the initial forward distribution because this provides the link between $r$ and \Rini.
 This approaches the intrinsic distribution in the case where mixing is homogeneous and the initial number infectious is small.

### 2020-05-26 doc/network.tex
done???

@@
 There are important distinctions to be made when defining generation-interval distributions at the population-level.
-The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.
+The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
 On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
@@
 \begin{equation}
-\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
+\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
 \label{eq:egorR}
@@
 The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
-we previously thought that summarizing the entire distribution with two moments (mean and variance) is sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ \citep{park2019practical}.
-However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the gamma distribution ``looks'' indistinguishable from the shape of the intrinsic generation-interval distribution (derived from the SEIR model).
+we previously showed that summarizing the entire distribution with two moments (mean and variance) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
+However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of \Rini (Appendix A.4).
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).
 
@@
 \begin{aligned}
-\frac{dS}{dt} &= - \beta S \sum_{k=1}^{n_I} I_k\\
-\frac{dE_1}{dt} &= \beta S \sum_{k=1}^{n_I} I_k - n_E \sigma E_1\\
-\frac{dE_m}{dt} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
-\frac{dI_1}{dt} &= n_E \sigma E_m - n_I \gamma I_1\\
-\frac{dI_n}{dt} &= n_I \gamma (I_{n-1} - I_n) && \text{for } n = 2, 3, \dots, n_I\\
+\frac{\mathrm{d}S}{\mathrm{d}t} &= - \beta S \sum_{k=1}^{n_I} I_k\\
+\frac{\mathrm{d}E_1}{\mathrm{d}t} &= \beta S \sum_{k=1}^{n_I} I_k - n_E \sigma E_1\\
+\frac{\mathrm{d}E_m}{\mathrm{d}t} &= n_E \sigma (E_{m-1} - E_m) && \text{for } m = 2, 3, \dots, n_E\\
+\frac{\mathrm{d}I_1}{\mathrm{d}t} &= n_E \sigma E_m - n_I \gamma I_1\\
+\frac{\mathrm{d}I_n}{\mathrm{d}t} &= n_I \gamma (I_{n-1} - I_n) && \text{for } n = 2, 3, \dots, n_I\\
 \end{aligned}
@@
 We use a gamma distribution (parameterized by its mean and shape) to model the initial forward generation-interval distribution.
-Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ is equal to
+Then, the probability that an individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ is equal to:
 \begin{equation}
-\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!},
+\frac{\Lambda^{n_j} F_0(\tsub{t}{truncate} - t_j ;\theta)^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big)}{n_j!},
+\label{eq:nPoisson}
 \end{equation}
-where $s_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is a (vector) parameter of the initial forward generation-interval distribution $f_0$ (and the corresponding cumulative distribution function $F_0$).
+where $\theta$ is a (vector) parameter of the initial forward generation-interval distribution $f_0$ (and the corresponding cumulative distribution function $F_0$).
+On the other hand, the probability density that the realized generation-interval between infector $j$ and infectee $i$ is equal to $0 \leq s_{i,j} \leq \tsub{t}{truncate} - t_j$ can be expressed using a truncated distribution \citep{lagakos1988nonparametric, kalbfleisch1991regression}:
+\begin{equation}
+\frac{f_0(s_{i, j}; \theta)}{F_0(\tsub{t}{truncate} - t_j ;\theta)}.
+\label{eq:truncdist}
+\end{equation}
+Therefore, the probability density that individual $j$ infects $n_j$ individuals between $t_j$ and $\tsub{t}{truncate}$ with realized generation intervals $s_{i,j}$ for $i=1,\dots,n_j$ is a product of \eref{nPoisson} and \eref{truncdist}:
+\begin{equation}
+\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!}.
+\end{equation}
 
-The likelihood of the non-homogeneous Poisson process can be written as:
+The full likelihood of contact tracing data, which include aggregated generation intervals as well as information about who infected whom, until time $\tsub{t}{truncate}$ can be written as:
 \begin{equation}\label{eq:nonhomogeneous}
@@
 &\mathcal{L}(\Lambda, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{truncate})\\
-&\quad=\prod_{j=1}^N \left(\Lambda^{n_j} \exp \big(- \Lambda \, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta) \right),
+&\quad=\prod_{j=1}^N \left(\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!} \right),
 \end{aligned}
@@
 where $N$ is the total number of infected individuals.
+This likelihood is a special case of the likelihood suggested in \citep{kalbfleisch1989inference} which assumes that the second event (observation of infection) occurs simultaneously with the corresponding initiating event (infection of an individual).
 Here, we estimate parameters $\Lambda$ and $\theta$ by maximum likelihood.

### 2020-05-27 doc/response.tex
Quickly through R2

@@
 That was a bad notation on our side --- we should not have used $\RR$ in the likelihood. 
-We have changed the notation for the Poisson process and provide a better explanation.
+We have changed the notation for the Poisson process and provided a better explanation.
 
 \revtext{
-How exactly is $\RR$ (and, is it \Ro or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref. [54] in the paper)?
+How exactly is $\RR$ (and, is it \Ro\ or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref.
+[54] in the paper)?
 }
@@
 
-We have added a diagram to summarize these effects.
+We have added a diagram as suggested.
 
@@
 
-We have added a new figure to the supplementary figures to account for incomplete reporting.
+We have added a new supplementary figure to address this question.
 

### 2020-05-27 doc/network.tex
done????
Light comments on intro

@@
 \\
-\textbf{4} Agent-Based Modelling Laboratory, York University, Toronto, Ontario, Canada
+\textbf{4} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
 \\
@@
 Generation intervals can be measured through contact tracing by identifying who infected whom.
-We study how realized intervals differ from ``intrinsic'' intervals that could be estimated by measuring individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales.
+We study how realized intervals differ from ``intrinsic'' intervals that describe individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales.
 Early in an epidemic, we expect the variation in the realized generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
@@
 However, estimating the reproductive number directly from disease life history requires detailed knowledge, which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
-Instead, the reproductive number can be indirectly estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
+Instead, the reproductive number is often indirectly estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
 These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.
@@
 On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
-Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.
+Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.
 
@@
 Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
-Mean forward generation intervals can decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.
+Mean forward generation intervals tend to decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.
 
@@
 
-While an epidemic is ongoing, realized generation intervals can be measured by identifying who infected whom and can be aggregated to form a single distribution.
-In practice, realized generation intervals is difficult to measure because it is difficult to observe the exact moment at which individuals become infected;
+In practice, realized generation intervals are often difficult to measure for many diseases, because it is difficult to observe when individuals become infected;
 in most cases, observable events are clinical (e.g. onset of symptoms).
-However, it is still possible to directly measure generation intervals through contact tracing in some cases (e.g., a rabid dog biting another dog \citep{hampson2009transmission}),.
-Thus, the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic ---
-we typically want to try to make inference based on all available data that have been gathered since the beginning of an epidemic.
+There are some exceptions: for example, generation intervals are commonly measured directly through contact tracing for canine rabies, where infection events are bites \citep{hampson2009transmission}.
+Intervals between observed disease progression events (commonly, onset of signs or symptoms) are called serial intervals.
+Serial intervals are in many ways similar to generation intervals, but there are also complexities in their use.
+We will not address these complexities here.
 
+While an epidemic is ongoing, realized generation intervals can at least in theory can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
+We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
 These aggregated measurements are ``truncated'' because we do not know what happens after the current epidemic time.
-These truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
-These observations allow us to make the following prediction: removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.
+The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
+We therefore predict that removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.
 
-In this study, we explore spatiotemporal variation in realized (particularly forward) generation intervals.
-We extend previous frameworks to study how the aggregated generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
-We classify spatial effects on realized generation intervals into three levels (egocentric, local, and global) and discuss how they affect realize generation-interval distributions.
+In this study, we explore spatiotemporal variation in realized generation intervals.
+We extend previous frameworks to investigate how aggregated generation-interval distributions change over time.
+We classify spatial effects on realized generation intervals into three levels (egocentric, local, and global) and discuss how these affect realized generation-interval distributions.
 Finally, we compare two methods for accounting for temporal bias and test our prediction using individual-based simulations.
+
+\jd{I don't love the prediction framework even after I toned it down: I feel we show that it should work in theory, not just predict that it should work in practice.}
 

### 2020-05-30 doc/network.tex
discussion

@@
 We study how realized intervals differ from ``intrinsic'' intervals that describe individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales.
-Early in an epidemic, we expect the variation in the realized generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
-therefore, we predict that correcting realized intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide the initial forward generation-interval distribution, which is spatially informed and correctly links $r$ and $\Rini$.
+Early in an epidemic, we expect the variation in the realized generation intervals to be mainly driven by truncation and by the population structure near the source of disease spread; 
+we predict that correcting realized intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide the initial forward generation-interval distribution, which is spatially informed and correctly links $r$ and $\Rini$.
 We develop and test statistical methods for temporal corrections of generation intervals,
 and confirm our prediction using individual-based simulations on an empirical network.
+\jd{I did not understand the therefore at all (and have deleted it).}
 
@@
 There are important distinctions to be made when defining generation-interval distributions at the population-level.
-The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
-On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
+The \emph{intrinsic} generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
+On the other hand, \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
 Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.
@@
 We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
-These aggregated measurements are ``truncated'' because we do not know what happens after the current epidemic time.
-The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
+These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observaton.
+The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded in time to be observed.
 We therefore predict that removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.
@@
 and $g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
-If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed at the endemic equilibrium or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized generation intervals.
+If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed population at the endemic equilibrium, or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized generation intervals.
 
@@
 This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
-Over a period of time where the susceptible proportion $S(t) = S(0)$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
+Over a period of time where the susceptible proportion $S(t) = S(0)$ remains approximately constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
 \begin{equation}
@@
 
+\jd{Deleted ``epidemic time'' everywhere, it does not seem clear to me.}
 During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
@@
 Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the intrinsic generation intervals.
-The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until the epidemic time $t$:
+The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until calendar time $t$:
 \begin{equation}
@@
 \end{equation}
-The aggregated generation-interval distribution can be also equivalently expressed in terms of the forward generation-interval distributions:
+The aggregated generation-interval distribution can be also equivalently expressed in terms of truncated forward generation-interval distributions:
 \begin{equation}
@@
 There are two reasons for this phenomenon.
-First, longer generation intervals are more likely to be missed due to right truncation (and short generation intervals are more likely to be realized).
+First, longer generation intervals are more likely than short intervals to be missed due to right truncation.
 In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$),
@@
 Second, the decreasing number of susceptibles over the course of an epidemic makes long infectious contacts less likely to result in infection \citep{champredon2015intrinsic}.
-Overall, we expect naively using the aggregated generation-interval distribution, to underestimate the initial reproductive number.
+Overall, we therefore expect naively using the aggregated generation-interval distribution to underestimate the initial reproductive number.
 
@@
 
+\jd{How do you feel about rewriting this paragraph?
+It sounds like we introduce space to understand effects of multiple contacts; in fact, the opposite is true; we got into the effects of multiple contacts because we were interested in space.}
 Infected individuals may contact the same susceptible individual multiple times, but only the first infectious contact gives rise to infection in a given individual (after this, they are no longer susceptible).
@@
 \end{equation}
-where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by a particular infected individual.
+where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by the focal individual.
 The dilution term, $\delta(a)$, models how contacts are distributed among the acquaintances.
@@
 
-The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
+The population-level egocentric kernel is found by integrating the individual-level kernel over individual variations:
 \begin{equation}\label{eq:ego}
@@
 The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
-Finally, the link between the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
+Finally, the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
 \begin{equation}
@@
 \end{equation}
-As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be typically smaller than $\Rini$ estimated from the intrinsic distribution;
+As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be smaller than $\Rini$ estimated from the intrinsic distribution;
 this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
@@
 In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution;
-since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we predict the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
-Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to validate our prediction.
+since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
+Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations to validate this prediction.
 
@@
 Accounting for the egocentric effect allows us to link the results by \cite{trapman2016inferring} to established results based on generation intervals.
-Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
-Global susceptible depletion refers to the overall decrease in the susceptible pool in the population.
+Local susceptible depletion refers to the effect of multiple ``linked'' individuals (e.g., in the same household or neighborhood) making infectious contacts to the same susceptible individual.
+Global susceptible depletion refers to the decrease in the susceptible proportion of the whole population.
 
@@
 Egocentric and local depletion effects are present from the beginning of an epidemic, even when depletion in the global susceptible population is negligible, and can strongly affect the initial spread of an epidemic.
-Therefore, we predict the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right-truncation.
+Therefore, we predict the realized generation intervals during an exponential growth phase to contain information about the contact structure, allowing us to estimate the initial forward generation-interval distribution by simply accounting for the right truncation.
 Simulation studies confirm our prediction: using the initial forward generation-interval distribution provides the correct link between $r$ and $\Rini$.
@@
 The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
-we previously showed that summarizing the entire distribution with two moments (mean and variance) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
+we previously showed that in many cases a gamma approximation (requiring estimation of only two parameters) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
 However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).

### 2020-05-31 doc/network.tex
double checking calculations

@@
 and confirm our prediction using individual-based simulations on an empirical network.
-\jd{I did not understand the therefore at all (and have deleted it).}
 
@@
 This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
-As a result, realized generation intervals (from the perspective of an infector) will typically have shorter mean than the intrinsic generation-interval distribution in a non-homogeneous population.
+As a result, realized generation intervals (from the perspective of an infector) will typically have a shorter mean than the intrinsic generation-interval distribution in a non-homogeneous population.
 This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
@@
 There are some exceptions: for example, generation intervals are commonly measured directly through contact tracing for canine rabies, where infection events are bites \citep{hampson2009transmission}.
-Intervals between observed disease progression events (commonly, onset of signs or symptoms) are called serial intervals.
-Serial intervals are in many ways similar to generation intervals, but there are also complexities in their use.
+Intervals between observed disease progression events (commonly, onset of signs or symptoms) are called serial intervals \citep{svensson2007note}.
+Serial intervals are in many ways similar to generation intervals, but there are also complexities in their use  \citep{britton2019estimation}.
 We will not address these complexities here.
 
-While an epidemic is ongoing, realized generation intervals can at least in theory can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
+While an epidemic is ongoing, realized generation intervals, at least in theory, can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
 We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
@@
 Finally, we compare two methods for accounting for temporal bias and test our prediction using individual-based simulations.
-
-\jd{I don't love the prediction framework even after I toned it down: I feel we show that it should work in theory, not just predict that it should work in practice.}
 
@@
 and $g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
-If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed population at the endemic equilibrium, or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized generation intervals.
+If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed population at the endemic equilibrium, or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized (forward) generation intervals.
 
@@
 This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
-Over a period of time where the susceptible proportion $S(t) = S(0)$ remains approximately constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
+Over a period of time where the susceptible proportion remains approximately constant ($S(t) \approx S(0)$), we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
 \begin{equation}
@@
 \begin{equation}
-i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
+\mathcal R_{\rm c}(t) i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
 \label{eq:equivalence}
 \end{equation}
+Here, $\mathcal R_{\rm c}(t)$ represents the case reproduction number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.
 
@@
 \begin{equation}
-f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(\tau) S(t+\tau) \,\mathrm{d}\tau}.
+f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(x) S(t+x) \,\mathrm{d}x}.
 \label{eq:forward}
@@
 \begin{equation}
-b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-\tau) g(\tau)\,\mathrm{d}\tau}.
+b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-x) g(x)\,\mathrm{d}x}.
 \end{equation}
+Substituting $\mathcal R_{\rm c}(t) = \mathcal R_0 \int_0^\infty g(x) S(t+x) \,\mathrm{d}x$ confirms that \eref{equivalence} holds for this model.
 
-\jd{Deleted ``epidemic time'' everywhere, it does not seem clear to me.}
 During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
@@
 \begin{equation}
-a_t(\tau) \propto \int_{-\infty}^t b_s(\tau) i(s) \,\mathrm{d}s.
+a_t(\tau) \propto \int_{-\infty}^t i(s) b_s(\tau) \,\mathrm{d}s.
 \end{equation}
-The aggregated generation-interval distribution can be also equivalently expressed in terms of truncated forward generation-interval distributions:
+The aggregated generation-interval distribution can be also equivalently expressed in terms of forward generation-interval distributions (weighted by incidence and case reproduction number):
 \begin{equation}
-a_t(\tau) \propto \int_{-\infty}^{t-\tau} f_s(\tau) i(s) \,\mathrm{d}s.
+a_t(\tau) \propto \int_{-\infty}^{t-\tau} \mathcal R_{\rm c}(s) i(s) f_s(\tau) \,\mathrm{d}s.
 \end{equation}

### 2020-05-31 fig/temporal_effect.R

@@ -34,7 +34,7 @@ sumfun <- function(out,
 		
 		i <- which(t==tau)
 		
-		beta*sum(incidence[1:(m-i+1)] * out$S[i:m]/N)/total*genden[i]
+		beta/gamma*sum(incidence[1:(m-i+1)] * out$S[i:m]/N)/total*genden[i]
 	}
 	
 	backward <- function(tau, tmax) {
@@ -88,14 +88,10 @@ names(out1)[1] <- "t"
 
 gensum <- sumfun(out1, tmax=tmax, t=t, tauby=5)
 
-# gensum_long <- sumfun(out4, tmax=10000, t=0:10000, tauby=1000)
-
 outdata <- out1 %>%
 	mutate(
 		incidence=beta*S*I/N
 	)
-
-## We probably don't need to show the endemic case.. seems trivial...
 
 g1 <- ggplot(outdata, aes(t, incidence)) +
     geom_line(lwd=1.2) +

### 2020-05-31 doc/response.tex
ready to submit

@@
 
-\newtext{The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.}
+\newtext{The \emph{intrinsic} generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.}
 
@@
 \newtext{$g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
-If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed at the endemic equilibrium or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized generaton interval.}
+If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed population at the endemic equilibrium, or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized (forward) generation intervals.}
 
@@
 \newtext{There are important distinctions to be made when defining generation-interval distributions at the population-level.
-The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.
-On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
-Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.}
+The \emph{intrinsic} generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
+On the other hand, \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
+Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.}
 
@@
 
-\newtext{While an epidemic is ongoing, realized generation intervals can be measured by identifying who infected whom and can be aggregated to form a single distribution.
-In practice, realized generation intervals is difficult to measure because it is difficult to observe the exact moment at which individuals become infected;
+\newtext{In practice, realized generation intervals are often difficult to measure for many diseases, because it is difficult to observe when individuals become infected;
 in most cases, observable events are clinical (e.g. onset of symptoms).
-However, it is still possible to directly measure generation intervals through contact tracing in some cases (e.g., a rabid dog biting another dog \citep{hampson2009transmission}),.
-Thus, the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic ---
-we typically want to try to make inference based on all available data that have been gathered since the beginning of an epidemic.}
+There are some exceptions: for example, generation intervals are commonly measured directly through contact tracing for canine rabies, where infection events are bites \citep{hampson2009transmission}.
+Intervals between observed disease progression events (commonly, onset of signs or symptoms) are called serial intervals \citep{svensson2007note}.
+Serial intervals are in many ways similar to generation intervals, but there are also complexities in their use  \citep{britton2019estimation}.
+We will not address these complexities here.}
 
@@
 
-\newtext{Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.}
+\newtext{Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.}
 
@@
 We appreciate this point and have tried not to imply strong assumptions, and also taken care to empirically validate our statistical approaches.
-
-\swp{??? What do you mean by ``also taken care to empirically validate our statistical approaches''}
 
@@
 
-\newtext{Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations hereafter to vadliate our prediction.}
+\newtext{Spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles); therefore, we rely on simulations to validate this prediction.}
 
@@
 We have tried to clarify these points. $S$ is a proportion, but we do attempt to address issues of individuality appropriately.
-
-\swp{What do you mean we have tried to address issues of individuality? Can we just get rid of that phrase and say S is a proportion?}
 
@@
 \newtext{In the main text, we present results based on exponentially distributed latent and infectious periods;
-we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.}
+we show results based on Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.}
 
@@
 The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
-we previously showed that summarizing the entire distribution with two moments (mean and variance) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
+we previously showed that in many cases a gamma approximation (requiring estimation of only two parameters) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
 However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).}
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Rini whereas narrower generation-interval distribution leads to higher \Rini) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).}
 
@@
 we focus strictly on estimating the \emph{initial} exponential growth rate (when $t \to -\infty$).
-The method of estimating the initial exponential growth rate by fitting a logistic curve has been previously validated against simulations of stochastic comparmental models \citep{ma2014estimating}.
+The method of estimating the initial exponential growth rate by fitting a logistic curve has been previously validated against simulations of stochastic compartmental models \citep{ma2014estimating}.
 }
@@
 
-See above for our discussion of these issues
+See above for our discussion of these issues.
 
@@
 
-\revtext{Figure 3 caption, last sentence: has recovered -> have recovered}
+\revtext{Figure 3 caption, last sentence: has recovered $\rightarrow$ have recovered}
 

### 2020-05-31 doc/network.tex
ready to submit

@@
 We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
-These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observaton.
+These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observation.
 The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded in time to be observed.
@@
 For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
-Therefore, the total number of infection that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
+Therefore, the total number of infections that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
 \begin{equation}
@@
 
-\jd{How do you feel about rewriting this paragraph?
-It sounds like we introduce space to understand effects of multiple contacts; in fact, the opposite is true; we got into the effects of multiple contacts because we were interested in space.}
+The effects of spatial structure on realized generation intervals can be understood in terms of effect of multiple contacts.
 Infected individuals may contact the same susceptible individual multiple times, but only the first infectious contact gives rise to infection in a given individual (after this, they are no longer susceptible).
-Therefore, we expect realized generation intervals from an individual to have a smaller mean than their mean intrinsic generation interval.
-To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous.
+Therefore, we expect realized generation intervals from an individual in a spatially structured population to have a smaller mean than their mean intrinsic generation interval.
+To explore the effects of spatial structure on realized generation intervals, we relax our assumption that the population is homogeneous.
 Instead, we assume that a disease spreads on a network;
@@
 Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
-In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
+In this case, the intrinsic infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
 This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
@@
 Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
-Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
+Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of the basic reproductive number.
 The population-level egocentric generation-interval distribution is:
@@
 For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
-The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{Champredon2018equivalence}:
+The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{yan2008separate, Champredon2018equivalence}:
 \begin{equation}
@@
 Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
-(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the contact results in infection.
+(Left) the intrinsic generation-interval distribution corresponds to all infectious contacts by a focal individual, regardless of whether the contact results in infection.
 (Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
-(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
+(Right) realized generation-interval distributions have a shorter mean than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
 All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
@@
 In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
-In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution;
-since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
-Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations to validate this prediction.
+In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution.
+Since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
+Spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles); therefore, we rely on simulations to validate this prediction.
 
@@
 \textbf{A summary of spatiotemporal effects on generation intervals.}
-The intrinsic generation-interval distribution describe the expected time distribution of infectious contacts made by a primary case.
-The realized (forward) generation-interval distribution describe the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
+The intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case.
+The realized (forward) generation-interval distribution describes the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
 The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
@@
 When an epidemic is growing exponentially, right-truncation causes the aggregated generation interval to be discounted by the exponential growth rate (\eref{exp});
-hence, we can ``undo'' the truncation by exponentially weighting the aggregated generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
+hence, we can ``undo'' the truncation by exponentially weighting the aggregated generation-interval distribution \citep{nishiura2010time, tomba2010some, britton2019estimation}:
 \begin{equation}
@@
 incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the truncation process in the realized generation intervals.
-For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.
+For both methods, mean and coefficient of variation (CV) of the initial forward generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distributions are then used to estimate the initial reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.
 
@@
 and compare the estimates of initial reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 75 infected individuals,
-as well as the estimates of the initial reproductive number calculated from the initial forward generation intervals, which we define as the generation intervals for all infections caused by the first 75 infected individuals (\fref{cmp}).
-For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting.
+as well as the estimates of the initial reproductive number calculated from the empirical initial forward generation intervals, which we define as the generation intervals for all infections caused by the first 75 infected individuals (\fref{cmp}).
+For simplicity, we assume that realized generation intervals are observed without error and assume that there is no under-reporting of generation intervals.
 We do not expect under-reporting to affect the inference of generation-interval distributions (Appendix A.3) unless there are systematic biases in the observation process.
-On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproductive number \cite{britton2019estimation}.
+On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the initial reproductive number \cite{britton2019estimation}.
 We do not pursue these directions in this study.
@@
 estimates based on the egocentric generation-interval distribution (\eref{conditional}) address this problem only partially, as they do not account for indirect (local) spatial effects. 
-Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
-While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the initial forward generation-interval distribution on average,
+Direct estimates based on the aggregated generation intervals from contact tracing (via Euler-Lotka) severely underestimate the empirical estimates.
+While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the empirical initial forward generation-interval distribution on average,
 population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
@@
 Realized generation intervals that have been aggregated over the course of an epidemic are subject to right-truncation -- it is not possible to trace individuals that have not been infected yet.
-The aggregated distributions can be thought of as averages of ``backward'' generation intervals (measured by looking at infectors of a cohort of individuals infected at the same time) \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+The aggregated distributions can be thought of as averages of ``backward'' generation intervals (measured by looking at infectors of a cohort of individuals infected at the same time) \citep{champredon2015intrinsic, kenah2008generation, nishiura2010time, tomba2010some, britton2019estimation}.
 During an ongoing outbreak, the aggregated generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-truncation.
@@
 These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
-Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Ro whereas narrower generation-interval distribution leads to higher \Ro) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).
+Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Rini whereas narrower generation-interval distribution leads to higher \Rini) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).
 
@@
 
-To study the effects on right-truncation on the observed generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
+To study the effects on right-truncation on the realized generation intervals, we use the deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model.
 The SEIR model describes how disease spreads in a homogeneously mixing population; it assumes that infected individuals become infectious after a latent period.
@@
 Parameters of the model are specified as follows:  $\beta$ is the transmission rate, $1/\sigma$ is the mean latent period, $n_E$ is the number of latent compartments, $1/\gamma$ is the mean infectious period, and $n_I$ is the number of infectious compartments.
-We scale the proportions of individuals in each compartment are scaled by the total population size $N$.
+We scale the proportions of individuals in each compartment by the total population size $N$.
 In the main text, we present results based on exponentially distributed latent and infectious periods;
-we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.
+we show results based on Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.
 
@@
 We then construct the random infectious period and infectious contact times simultaneously as follows.
-For each of the $n_I$ stages of the infectious period, we draw the number of effective contacts (before transitioning to the next compartment) from a geometric distribution with probability $n_I\gamma/(S_i \lambda + n_I \gamma)$, where $S_i$ is the number of susceptible acquaintances and $\lambda$ is the per-pair contact rate.
+For each of the $n_I$ stages of the infectious period, we draw the number of infectious contacts (before transitioning to the next compartment) from a geometric distribution with probability $n_I\gamma/(S_i \lambda + n_I \gamma)$, where $S_i$ is the number of susceptible acquaintances and $\lambda$ is the per-pair contact rate.
 We then choose the time between consecutive events (the chosen number of contacts, followed by exit from the given stage of infection) from an exponential distribution with rate $S_i \lambda + n_I\gamma$.
@@
 Here, we estimate parameters $\Lambda$ and $\theta$ by maximum likelihood.
+In theory, the forward generation intervals arising from the same infector may be correlated due to non-independence in the contact process \citep{yan2008separate}; although we do not account for this potential correlation in our likelihood, our simulations (\fref{cmp}) suggest that approximating the initial forward generation-interval distribution with a single distribution provides a viable approach for estimating the reproductive number.
+
 We then use the estimated distribution $f_0$ to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
@@
 It consists of 23133 nodes and 93497 edges.
-The same network was used by \cite{trapman2016inferring} to study how network structure affects the estimate of the reproductive number.
+The same network was used by \cite{trapman2016inferring} to study how network structure affects the estimate of the basic reproductive number.
 

### 2020-05-31 fig/diagram.tex
changes
ready to submit

@@
         	\item time distribution between actual infection events
-        	\item egocentric depletion, local depletion, global depletion
+        	\item subject to egocentric, local, and global depletion
         \end{itemize}
@@
         \begin{itemize}
-        	\item analogous to contact tracing data
-        	\item truncated at epidemic time
+        	\item contact tracing data
+        	\item truncated at observation time
         \end{itemize}
@@
         \begin{itemize}
-        	\item aggregated GI during the exponential growth phase
-        	\item truncation bias is understood exactly
+        	\item measured during the exponential growth phase
         \end{itemize}
@@
 \draw[->, very thick, dashed] (local) -- node [midway, right, text width=2cm] {exponential growth \\ phase} (forward);
-\draw[->, very thick, dashed] (local) -- node [midway, right, text width=2cm] {} (agg);
+\draw[->, very thick, dashed] (local) -- node [midway, above, text width=2cm] {truncation \\ bias} (agg);
 \draw[->, very thick, dashed] (agg) -- node [midway, right, text width=2cm] {exponential growth \\ phase} (iniagg);

### 2020-05-31 doc/network.tex
updating directory
ugh more minor tweaks

@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
+\includegraphics[width=\textwidth]{individual_and_population.pdf}
 \caption{\textbf{Comparison of individual- and population-level kernels.}
@@
 \end{equation}
-Here, $\mathcal R_{\rm c}(t)$ represents the case reproduction number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.
+Here, $\mathcal R_{\rm c}(t)$ represents the case reproductive number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.
 
@@
 \end{equation}
-The aggregated generation-interval distribution can be also equivalently expressed in terms of forward generation-interval distributions (weighted by incidence and case reproduction number):
+The aggregated generation-interval distribution can be also equivalently expressed in terms of forward generation-interval distributions (weighted by incidence and case reproductive number):
 \begin{equation}
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
+\includegraphics[width=\textwidth]{temporal_effect.pdf}
 \caption{\textbf{Temporal variation in the mean backward and aggregated generation interval.}
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
+\includegraphics[width=\textwidth]{local_effect.pdf}
 \caption{
@@
 
-In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
+In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproductive number \Rini (see \eref{forward}).
 In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution.
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../fig/diagram.pdf}
+\includegraphics[width=\textwidth]{diagram.pdf}
 \caption{
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
+\includegraphics[width=\textwidth]{cmp_reproductive.pdf}
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}

### 2020-06-03 doc/network.tex
clean u[

@@
 At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
-While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.
+While this definition is widely used in the literature, it is not directly related to a population-level \emph{distribution}.
 There are important distinctions to be made when defining generation-interval distributions at the population-level.
@@
 
-The shape of the realized generation-interval distribution depends on the reference time and perspective \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
+The shape of the realized generation-interval distribution depends on the reference time and perspective \citep{champredon2015intrinsic,kenah2008generation, nishiura2010time, tomba2010some, britton2019estimation}.
 When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
 A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
-Thus, ``backward'' generation intervals (which look at a cohort of infectees, and ask when their infectors were infected) will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
+Thus, ``backward'' generation intervals, which look at a cohort of infectees and ask when their infectors were infected, will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
 Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
-Mean forward generation intervals tend to decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.
+Mean forward generation intervals tends to decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.
 
@@
 As a result, realized generation intervals (from the perspective of an infector) will typically have a shorter mean than the intrinsic generation-interval distribution in a non-homogeneous population.
-This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
+This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller than would be predicted from homogeneous mixing.
 
 In practice, realized generation intervals are often difficult to measure for many diseases, because it is difficult to observe when individuals become infected;
-in most cases, observable events are clinical (e.g. onset of symptoms).
+in most cases, observable events are clinical (e.g., onset of symptoms).
 There are some exceptions: for example, generation intervals are commonly measured directly through contact tracing for canine rabies, where infection events are bites \citep{hampson2009transmission}.
@@
 While an epidemic is ongoing, realized generation intervals, at least in theory, can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
-We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
+We typically want to try to make inference based on this aggregated distribution --- that is, on all available data that have been gathered since the beginning of an epidemic.
 These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observation.
@@
 this represents an individual realization of a random process.
-(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
+(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 days and 5 days, respectively;
 this represents a population average of a random process.
@@
 Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
-Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
+Hereafter, we use $t, s$ to represent calendar time and $\tau, x$ to
 represent time since infection.
@@
 \begin{equation}
-i(t) = S(t) \int K(s) i(t-s)\,\mathrm{d}s = \RR_0 S(t) \int g(s) i(t-s)\,\mathrm{d}s.
+i(t) = S(t) \int K(\tau) i(t-\tau)\,\mathrm{d}\tau = \RR_0 S(t) \int g(\tau) i(t-\tau)\,\mathrm{d}\tau.
 \end{equation}
-This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
+This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{roberts2007model, wallinga2007generation, heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, Champredon2018equivalence}.
 Over a period of time where the susceptible proportion remains approximately constant ($S(t) \approx S(0)$), we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
 \begin{equation}
-\frac{1}{\Rini} = \int g(s) \exp(-r s)\,\mathrm{d}s.
+\frac{1}{\Rini} = \int g(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
 \label{eq:euler}
@@
 For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
-Therefore, the total number of infections that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
+Therefore, the density of new infections occurring at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
 \begin{equation}
@@
 \end{equation}
-As shown in \cite{champredon2015intrinsic}, the forward generation-interval distribution, $f_t(\tau)$, is proportional to $i_{t+\tau}(t)$:
+As shown in \cite{champredon2015intrinsic}, the forward generation-interval distribution, $f_t(\tau)$, is proportional to $i_{t}(t+\tau)$:
 \begin{equation}
@@
 \end{equation}
-Substituting $\mathcal R_{\rm c}(t) = \mathcal R_0 \int_0^\infty g(x) S(t+x) \,\mathrm{d}x$ confirms that \eref{equivalence} holds for this model.
+Substituting $\mathcal R_{\rm c}(t) = \mathcal R_0 \int_0^\infty g(\tau) S(t+\tau) \,\mathrm{d}\tau$ confirms that \eref{equivalence} holds for this model.
 
@@
 This effect is called ``right-truncation''.
-Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the intrinsic generation intervals.
+Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the mean intrinsic generation interval.
 The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until calendar time $t$:
@@
 \end{equation}
-The aggregated generation-interval distribution can be also equivalently expressed in terms of forward generation-interval distributions (weighted by incidence and case reproductive number):
+The aggregated generation-interval distribution can be also expressed equivalently in terms of forward generation-interval distributions (weighted by incidence and case reproductive number):
 \begin{equation}
@@
 First, longer generation intervals are more likely than short intervals to be missed due to right truncation.
-In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$),
+In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) \approx i(0) \exp(rt)$ and $\mathcal R_{\rm c}(t) \approx \mathcal R_{\rm c}(0)$),
 the initial aggregated (or backward) generation-interval distribution is just the initial forward generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
@@
 \begin{equation}
-\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(s; a) \,\mathrm{d}s\right),
+\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(x; a) \,\mathrm{d}x\right),
 \end{equation}
-where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by the focal individual.
+where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(x; a)  \,\mathrm{d}x}$ is the probability that a susceptible acquaintance has not yet been contacted by the focal individual.
 The dilution term, $\delta(a)$, models how contacts are distributed among the acquaintances.
@@
 Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
-In this case, the intrinsic infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
+In this case, the intrinsic infectiousness of an individual $\RR(a) = \int k(\tau; a) \,\mathrm{d}\tau$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, the contact rate $\lambda$, and the duration of infectious period; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
 This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
@@
 \begin{equation}
-\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s)\,\mathrm{d}s}.
+\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(x)\,\mathrm{d}x}.
 \label{eq:conditional}
@@
 The intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case.
-The realized (forward) generation-interval distribution describes the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
+The realized (forward) generation-interval distribution describes the time between actual infection events --- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
 The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
@@
 While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the empirical initial forward generation-interval distribution on average,
-population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
-For smaller values of \RR, we expect the differences to become smaller.
-In the Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
+population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimates of the initial exponential growth rate.
+For smaller values of $\Rini$, we expect the differences to become smaller.
+In the Appendix, we present the same figure using smaller $\Rini$ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
 Overall, our qualitative conclusions do not change.
@@
 During an ongoing outbreak, the aggregated generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-truncation.
-Early in the outbreak, aggregated intervals are expected to match the early-outbreak backward intervals.
+Early in the outbreak, the initial aggregated intervals are expected to match the initial backward intervals.
 Near the end of an outbreak, the effect of right-truncation becomes negligible but the aggregated generation intervals are still shorter on average than intrinsic generation intervals, because of depletion of the susceptible population.
@@
 The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
-we previously showed that in many cases a gamma approximation (requiring estimation of only two parameters) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
+we previously showed that a gamma approximation (requiring estimation of only two parameters) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
 However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
@@
 &\mathcal{L}(\Lambda, \theta\,|\, \mathbf{s}, \mathbf{t}, \mathbf{n}, \tsub{t}{truncate})\\
-&\quad=\prod_{j=1}^N \left(\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!} \right),
+&\quad=\prod_{j=1}^{N_I} \left(\frac{\Lambda^{n_j} \exp \big(- \Lambda\, F_0(\tsub{t}{truncate} - t_j ;\theta) \big) \prod_{i=1}^{n_j} f_0(s_{i, j}; \theta)}{n_j!} \right),
 \end{aligned}
 \end{equation}
-where $N$ is the total number of infected individuals.
+where $N_I$ is the total number of infected individuals (in the data).
 This likelihood is a special case of the likelihood suggested in \citep{kalbfleisch1989inference} which assumes that the second event (observation of infection) occurs simultaneously with the corresponding initiating event (infection of an individual).
@@
 
-We then use the estimated distribution $f_0$ to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
+We use the estimated distribution $f_0$ to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).
 When the entire transmission process is known, we expect $\Lambda$ to match $\Rini$;
@@
 This graph describes co-authorship among anyone who submitted a paper to Condensed Matter category in the arXiv between January 1993 and April 2003 \citep{leskovec2007graph}.
-It consists of 23133 nodes and 93497 edges.
+It consists of 23,133 nodes and 93,497 edges.
 The same network was used by \cite{trapman2016inferring} to study how network structure affects the estimate of the basic reproductive number.

### 2020-06-03 doc/network.tex
using bbl instead

@@
 
-\bibliography{networkprsb}
+\begin{thebibliography}{10}
+
+\bibitem{anderson1991infectious}
+Anderson RM and May RM.
+\newblock 1991 {\em Infectious diseases of humans: dynamics and control}.
+\newblock Oxford university press.
+
+\bibitem{diekmann1990definition}
+Diekmann O, Heesterbeek JAP, and Metz JA.
+\newblock 1990 On the definition and the computation of the basic reproduction
+  ratio {R0} in models for infectious diseases in heterogeneous populations.
+\newblock {\em J. Math. Biol}.
+\newblock {\bfseries 28}, 365--382.
+
+\bibitem{kribs2000simple}
+Kribs-Zaleta CM and Velasco-Hern{\'a}ndez JX.
+\newblock 2000 A simple vaccination model with multiple endemic states.
+\newblock {\em Math. Biosci}.
+\newblock {\bfseries 164}, 183--201.
+
+\bibitem{van2002reproduction}
+Van~den Driessche P and Watmough J.
+\newblock 2002 Reproduction numbers and sub-threshold endemic equilibria for
+  compartmental models of disease transmission.
+\newblock {\em Math. Biosci}.
+\newblock {\bfseries 180}, 29--48.
+
+\bibitem{smith2012ross}
+Smith DL, Battle KE, Hay SI, Barker CM, Scott TW, and McKenzie FE.
+\newblock 2012 {Ross, Macdonald, and a theory for the dynamics and control of
+  mosquito-transmitted pathogens}.
+\newblock {\em PLoS Pathog}.
+\newblock {\bfseries 8}.
+
+\bibitem{dietz1993estimation}
+Dietz K.
+\newblock 1993 The estimation of the basic reproduction number for infectious
+  diseases.
+\newblock {\em Stat. Methods Med. Res}.
+\newblock {\bfseries 2}, 23--41.
+
+\bibitem{chowell2003sars}
+Chowell G, Fenimore PW, Castillo-Garsow MA, and Castillo-Chavez C.
+\newblock 2003 {SARS outbreaks in Ontario, Hong Kong and Singapore: the role of
+  diagnosis and isolation as a control mechanism}.
+\newblock {\em J. Theor. Biol}.
+\newblock {\bfseries 224}, 1--8.
+
+\bibitem{mills2004transmissibility}
+Mills CE, Robins JM, and Lipsitch M.
+\newblock 2004 Transmissibility of 1918 pandemic influenza.
+\newblock {\em Nature}.
+\newblock {\bfseries 432}, 904.
+
+\bibitem{nishiura2009transmission}
+Nishiura H, Castillo-Chavez C, Safan M, and Chowell G.
+\newblock 2009 {Transmission potential of the new influenza A (H1N1) virus and
+  its age-specificity in Japan}.
+\newblock {\em Euro Surveill}.
+\newblock {\bfseries 14}, 19227.
+
+\bibitem{nishiura2010pros}
+Nishiura H, Chowell G, Safan M, and Castillo-Chavez C.
+\newblock 2010 {Pros and cons of estimating the reproduction number from early
+  epidemic growth rate of influenza A (H1N1) 2009}.
+\newblock {\em Theor. Biol. Med. Model}.
+\newblock {\bfseries 7}, 1.
+
+\bibitem{ma2014estimating}
+Ma~J, Dushoff J, Bolker BM, and Earn DJ.
+\newblock 2014 Estimating initial epidemic growth rates.
+\newblock {\em Bull. Math. Biol}.
+\newblock {\bfseries 76}, 245--260.
+
+\bibitem{wearing2005appropriate}
+Wearing HJ, Rohani P, and Keeling MJ.
+\newblock 2005 Appropriate models for the management of infectious diseases.
+\newblock {\em PLoS Med}.
+\newblock {\bfseries 2}, e174.
+
+\bibitem{svensson2007note}
+Svensson {\AA}.
+\newblock 2007 A note on generation times in epidemic models.
+\newblock {\em Math. Biosci}.
+\newblock {\bfseries 208}, 300--311.
+
+\bibitem{roberts2007model}
+Roberts M and Heesterbeek J.
+\newblock 2007 Model-consistent estimation of the basic reproduction number
+  from the incidence of an emerging infection.
+\newblock {\em J. Math. Biol}.
+\newblock {\bfseries 55}, 803.
+
+\bibitem{wallinga2007generation}
+Wallinga J and Lipsitch M.
+\newblock 2007 How generation intervals shape the relationship between growth
+  rates and reproductive numbers.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 274}, 599--604.
+
+\bibitem{park2019practical}
+Park SW, Champredon D, Weitz JS, and Dushoff J.
+\newblock 2019 A practical generation-interval-based approach to inferring the
+  strength of epidemics from their speed.
+\newblock {\em Epidemics}.
+\newblock {\bfseries 27}, 12--18.
+
+\bibitem{champredon2015intrinsic}
+Champredon D and Dushoff J.
+\newblock 2015 Intrinsic and realized generation intervals in
+  infectious-disease transmission.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 282}.
+
+\bibitem{kenah2008generation}
+Kenah E, Lipsitch M, and Robins JM.
+\newblock 2008 Generation interval contraction and epidemic data analysis.
+\newblock {\em Math. Biosci}.
+\newblock {\bfseries 213}, 71--79.
+
+\bibitem{nishiura2010time}
+Nishiura H.
+\newblock 2010 Time variations in the generation time of an infectious disease:
+  implications for sampling to appropriately quantify transmission potential.
+\newblock {\em Math. Biosci. Eng}.
+\newblock {\bfseries 7}, 851--869.
+
+\bibitem{tomba2010some}
+Scalia~Tomba G, Svensson {\AA}, Asikainen T, and Giesecke J.
+\newblock 2010 Some model based considerations on observing generation times
+  for communicable diseases.
+\newblock {\em Math. Biosci}.
+\newblock {\bfseries 223}, 24--31.
+
+\bibitem{britton2019estimation}
+Britton T and Scalia~Tomba G.
+\newblock 2019 Estimation in emerging epidemics: {Biases} and remedies.
+\newblock {\em J. Royal Soc. Interface}.
+\newblock {\bfseries 16}, 20180670.
+
+\bibitem{trapman2016inferring}
+Trapman P, Ball F, Dhersin JS, Tran VC, Wallinga J, and Britton T.
+\newblock 2016 Inferring {R0} in emerging epidemics--the effect of common
+  population structure is small.
+\newblock {\em J. Royal Soc. Interface}.
+\newblock {\bfseries 13}.
+
+\bibitem{hampson2009transmission}
+Hampson K, Dushoff J, Cleaveland S, Haydon DT, Kaare M, Packer C, and Dobson A.
+\newblock 2009 Transmission dynamics and prospects for the elimination of
+  canine rabies.
+\newblock {\em PLoS Biol}.
+\newblock {\bfseries 7}.
+
+\bibitem{who2014ebola}
+{WHO Ebola Response Team} .
+\newblock 2014 {Ebola virus disease in West Africa--the first 9 months of the
+  epidemic and forward projections}.
+\newblock {\em N. Engl. J. Med}.
+\newblock {\bfseries 371}, 1481--1495.
+
+\bibitem{svensson2015influence}
+Svensson {\AA}.
+\newblock 2015 The influence of assumptions on generation time distributions in
+  epidemic models.
+\newblock {\em Math. Biosci}.
+\newblock 270, 81--89.
+
+\bibitem{heesterbeek1996concept}
+Heesterbeek J and Dietz K.
+\newblock 1996 The concept of {R0} in epidemic theory.
+\newblock {\em Stat. Neerl}.
+\newblock {\bfseries 50}, 89--110.
+
+\bibitem{diekmann2000mathematical}
+Diekmann O and Heesterbeek JAP.
+\newblock 2000 {\em Mathematical epidemiology of infectious diseases: model
+  building, analysis and interpretation}, volume~5.
+\newblock John Wiley \& Sons.
+
+\bibitem{roberts2004modelling}
+Roberts M.
+\newblock 2004 Modelling strategies for minimizing the impact of an imported
+  exotic infection.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 271}, 2411--2415.
+
+\bibitem{aldis2005integral}
+Aldis G and Roberts M.
+\newblock 2005 An integral equation model for the control of a smallpox
+  outbreak.
+\newblock {\em Math. Biosci}.
+\newblock {\bfseries 195}, 1--22.
+
+\bibitem{Champredon2018equivalence}
+Champredon D, Dushoff J, and Earn DJD.
+\newblock 2018 {Equivalence of the Erlang-distributed SEIR epidemic model and
+  the renewal equation}.
+\newblock {\em SIAM J. Appl. Math}.
+\newblock 78, 3258--3278.
+
+\bibitem{lotka1907relation}
+Lotka AJ.
+\newblock 1907 Relation between birth rates and death rates.
+\newblock {\em Science}.
+\newblock {\bfseries 26}, 21--22.
+
+\bibitem{fraser2007estimating}
+Fraser C.
+\newblock 2007 Estimating individual and household reproduction numbers in an
+  emerging epidemic.
+\newblock {\em PloS One}.
+\newblock {\bfseries 2}.
+
+\bibitem{ball1997epidemics}
+Ball F, Mollison D, and Scalia-Tomba G.
+\newblock 1997 Epidemics with two levels of mixing.
+\newblock {\em Ann. Appl. Probab}.
+\newblock pages 46--89.
+
+\bibitem{ball2002general}
+Ball F and Neal P.
+\newblock 2002 A general model for stochastic {SIR} epidemics with two levels
+  of mixing.
+\newblock {\em Math. Biosci}.
+\newblock {\bfseries 180}, 73--102.
+
+\bibitem{yan2008separate}
+Yan P.
+\newblock 2008 Separate roles of the latent and infectious periods in shaping
+  the relation between the basic reproduction number and the intrinsic growth
+  rate of infectious disease outbreaks.
+\newblock {\em J. Theor. Biol}.
+\newblock {\bfseries 251}, 238--252.
+
+\bibitem{leskovec2016snap}
+Leskovec J and Krevl A.
+\newblock 2014 {SNAP} {Datasets}: {Stanford Large Network Dataset Collection}.
+\newblock http://snap.stanford.edu/data.
+
+\bibitem{mcbryde2009early}
+McBryde E, Bergeri I, van Gemert C, Rotty J, Headley E, Simpson K, Lester R,
+  Hellard M, and Fielding JE.
+\newblock 2009 {Early transmission characteristics of influenza A (H1N1) v in
+  Australia: Victorian state, 16 May--3 June 2009}.
+\newblock {\em Euro Surveill}.
+\newblock {\bfseries 14}, 19363.
+
+\bibitem{roberts2011early}
+Roberts MG and Nishiura H.
+\newblock 2011 {Early estimation of the reproduction number in the presence of
+  imported cases: pandemic influenza H1N1-2009 in New Zealand}.
+\newblock {\em PLoS One}.
+\newblock {\bfseries 6}, e17835.
+
+\bibitem{trichereau2012estimation}
+Trichereau J, Verret C, Mayet A, Manet G, Decam C, Meynard JB, Deparis X, and
+  Migliani R.
+\newblock 2012 {Estimation of the reproductive number for A (H1N1) pdm09
+  influenza among the French armed forces, September 2009--March 2010}.
+\newblock {\em J. Infect}.
+\newblock {\bfseries 64}, 628--630.
+
+\bibitem{nishiura2015theoretical}
+Nishiura H and Chowell G.
+\newblock 2015 Theoretical perspectives on the infectiousness of {Ebola} virus
+  disease.
+\newblock {\em Theor. Biol. Med. Model}.
+\newblock {\bfseries 12}, 1.
+
+\bibitem{szendroi2004polynomial}
+Szendroi B and Cs{\'a}nyi G.
+\newblock 2004 Polynomial epidemics and clustering in contact networks.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 271}, S364--S366.
+
+\bibitem{chowell2015western}
+Chowell G, Viboud C, Hyman JM, and Simonsen L.
+\newblock 2015 {The Western Africa} {Ebola} virus disease epidemic exhibits
+  both global exponential and local polynomial growth rates.
+\newblock {\em PLoS Curr}.
+\newblock 7.
+
+\bibitem{chowell2016growing}
+Chowell G and Viboud C.
+\newblock 2016 Is it growing exponentially fast?--impact of assuming
+  exponential growth for characterizing and forecasting epidemics with initial
+  near-exponential growth dynamics.
+\newblock {\em Infect. Dis. Model}.
+\newblock {\bfseries 1}, 71--78.
+
+\bibitem{chowell2016characterizing}
+Chowell G, Viboud C, Simonsen L, and Moghadas SM.
+\newblock 2016 Characterizing the reproduction number of epidemics with early
+  subexponential growth dynamics.
+\newblock {\em J. Royal Soc. Interface}.
+\newblock {\bfseries 13}, 20160659.
+
+\bibitem{kiskowski2016modeling}
+Kiskowski M and Chowell G.
+\newblock 2016 Modeling household and community transmission of {Ebola} virus
+  disease: epidemic growth, spatial dynamics and insights for epidemic control.
+\newblock {\em Virulence}.
+\newblock {\bfseries 7}, 163--173.
+
+\bibitem{viboud2016generalized}
+Viboud C, Simonsen L, and Chowell G.
+\newblock 2016 A generalized-growth model to characterize the early ascending
+  phase of infectious disease outbreaks.
+\newblock {\em Epidemics}.
+\newblock 15, 27--37.
+
+\bibitem{clarke1998contact}
+Clarke J.
+\newblock 1998 Contact tracing for chlamydia: data on effectiveness.
+\newblock {\em Int. J. STD. AIDS}.
+\newblock {\bfseries 9}, 187--191.
+
+\bibitem{eames2003contact}
+Eames KT and Keeling MJ.
+\newblock 2003 Contact tracing and disease control.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 270}, 2565--2571.
+
+\bibitem{donnelly2003epidemiological}
+Donnelly CA, Ghani AC, Leung GM, Hedley AJ, Fraser C, Riley S, Abu-Raddad LJ,
+  Ho~LM, Thach TQ, Chau P, and others .
+\newblock 2003 Epidemiological determinants of spread of causal agent of severe
+  acute respiratory syndrome in {Hong Kong}.
+\newblock {\em Lancet}.
+\newblock {\bfseries 361}, 1761--1766.
+
+\bibitem{cauchemez2006estimating}
+Cauchemez S, Bo{\"e}lle PY, Thomas G, and Valleron AJ.
+\newblock 2006 Estimating in real time the efficacy of measures to control
+  emerging communicable diseases.
+\newblock {\em Am. J. Epidemiol}.
+\newblock {\bfseries 164}, 591--597.
+
+\bibitem{hens2012robust}
+Hens N, Calatayud L, Kurkela S, Tamme T, and Wallinga J.
+\newblock 2012 Robust reconstruction and analysis of outbreak data: influenza
+  {A (H1N1)} v transmission in a school-based population.
+\newblock {\em Am. J. Epidemiol}.
+\newblock {\bfseries 176}, 196--203.
+
+\bibitem{jewell2012enhancing}
+Jewell CP and Roberts GO.
+\newblock 2012 Enhancing {Bayesian} risk prediction for epidemics using contact
+  tracing.
+\newblock {\em Biostatistics}.
+\newblock {\bfseries 13}, 567--579.
+
+\bibitem{soetens2018real}
+Soetens L, Klinkenberg D, Swaan C, Hahn{\'e} S, and Wallinga J.
+\newblock 2018 Real-time estimation of epidemiologic parameters from contact
+  tracing data during an emerging infectious disease outbreak.
+\newblock {\em Epidemiology}.
+\newblock {\bfseries 29}, 230--236.
+
+\bibitem{anderson1980spread}
+Anderson D and Watson R.
+\newblock 1980 On the spread of a disease with gamma distributed latent and
+  infectious periods.
+\newblock {\em Biometrika}.
+\newblock {\bfseries 67}, 191--198.
+
+\bibitem{bailey1964some}
+Bailey NT.
+\newblock 1964 Some stochastic models for small epidemics in large populations.
+\newblock {\em J. R. Stat. Soc. Ser. C. Appl. Stat}.
+\newblock 13, 9--19.
+
+\bibitem{bartlett1953stochastic}
+Bartlett MS.
+\newblock 1953 Stochastic processes or the statistics of change.
+\newblock {\em J. R. Stat. Soc. Ser. C. Appl. Stat}.
+\newblock {\bfseries 2}, 44--64.
+
+\bibitem{gillespie1977exact}
+Gillespie DT.
+\newblock 1977 Exact stochastic simulation of coupled chemical reactions.
+\newblock {\em J. Phys. Chem}.
+\newblock {\bfseries 81}, 2340--2361.
+
+\bibitem{newman2003structure}
+Newman ME.
+\newblock 2003 The structure and function of complex networks.
+\newblock {\em SIAM Rev}.
+\newblock {\bfseries 45}, 167--256.
+
+\bibitem{daley2007introduction}
+Daley DJ and Vere-Jones D.
+\newblock 2007 {\em {An introduction to the theory of point processes: volume
+  II: general theory and structure}}.
+\newblock Springer Science \& Business Media.
+
+\bibitem{lagakos1988nonparametric}
+Lagakos SW, Barraj LM, and Gruttola Vd.
+\newblock 1988 {Nonparametric analysis of truncated survival data, with
+  application to AIDS}.
+\newblock {\em Biometrika}.
+\newblock {\bfseries 75}, 515--523.
+
+\bibitem{kalbfleisch1991regression}
+Kalbfleisch J and Lawless J.
+\newblock 1991 {Regression models for right truncated data with applications to
+  AIDS incubation times and reporting lags}.
+\newblock {\em Statistica Sinica}.
+\newblock pages 19--32.
+
+\bibitem{kalbfleisch1989inference}
+Kalbfleisch J and Lawless JF.
+\newblock 1989 {Inference based on retrospective ascertainment: an analysis of
+  the data on transfusion-related AIDS}.
+\newblock {\em J. Am. Stat. Assoc}.
+\newblock {\bfseries 84}, 360--372.
+
+\bibitem{king2015avoidable}
+King AA, de~Celles MD, Magpantay FM, and Rohani P.
+\newblock 2015 Avoidable errors in the modelling of outbreaks of emerging
+  pathogens, with special reference to {Ebola}.
+\newblock {\em Proc. R. Soc. B}.
+\newblock {\bfseries 282}, 20150347.
+
+\bibitem{leskovec2007graph}
+Leskovec J, Kleinberg J, and Faloutsos C.
+\newblock 2007 Graph evolution: {Densification} and shrinking diameters.
+\newblock {\em ACM T. Knowl. Discov. D.}
+\newblock {\bfseries 1}, 2.
+
+\end{thebibliography}
+
 \end{document}

### 2020-06-03 doc/supp.tex
clean u[
typo in the supp

@@
 This figure matches Fig. 4 in the main text.
-A smaller per-pair contact rate ($\lambda = 0.0026 \textrm{ days}^{-1}$) is used to simulate epidemics.
+A smaller per-pair contact rate ($\lambda = 0.026 \textrm{ days}^{-1}$) is used to simulate epidemics.
 All other parameters are the same as in Fig. 4 in the main text.
@@
 (D) Estimated generation-interval distributions based on the gamma approximation and the true generation-interval distribution of an SEIR model.
-(E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$-$\RR$ relationship of an SEIR model.
+(E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$--$\RR$ relationship of an SEIR model.
 Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.

### 2020-06-04 doc/supp.tex
final final

@@
 This figure matches Fig. 4 in the main text.
-We assume that infected indidividuals have 30\% binomial probability of reporting their generation intervals (time between their infection and the infection of their infectors) when we apply population- and individual-based methods to estimate \Rini;
-under-reporting of generation intervals still give unbiased estimates of \Rini.
-All other parameters and estimates are the same as in Fig. 4 in the main text.
+We assume that infected indidividuals have 30\% binomial probability of reporting their generation intervals (time between their infection and the infection of their infectors) when we apply population- and individual-based methods to estimate \Rini.
+We still obtain unbiased estimates of \Rini when generation intervals are under-reported.
+All other parameters and \Rini estimates (boxplots) are the same as in Fig. 4 in the main text.
 }

### 2020-06-04 analysis/cmp_compare_underreport.R

@@ -77,7 +77,7 @@ individual.est <- lapply(reslist, function(x){
 	tmax <- x$data[1001,1]
 	gendata <- generation.data(x, tmax=tmax)
 	
-	ww <- rbinom(1:nrow(gendata), 1, rprob)==1
+	ww <- rbinom(1:nrow(gendata), 1, rprob)==0
 	
 	gendata$infected_by[ww] <- NA
 	gendata$generation[ww] <- NA

### 2020-06-15 doc/supp.tex
Small changes to supp

@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-This figure matches Fig. 4 in the main text.
-A smaller per-pair contact rate ($\lambda = 0.026 \textrm{ days}^{-1}$) is used to simulate epidemics.
-All other parameters are the same as in Fig. 4 in the main text.
+This figure matches Fig.
+5 in the main text, but with a smaller per-pair contact rate ($\lambda = 0.026 \textrm{ days}^{-1}$) used to simulate epidemics.
+All other parameters are the same as in Fig.
+5.
 }
@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-This figure matches Fig. 4 in the main text which used a SEIR model.
-Here, an Erlang-distributed latent period ($n_E=2$) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease than the exponential assumption \citep{who2014ebola}.
-All other parameters are the same as in Fig. 4 in the main text.
+This figure matches Fig.
+5 in the main text, but an Erlang-distributed latent period ($n_E=2$ instead of 1) is used to simulate epidemics; this assumption better matches the incubation period distribution of the Ebola virus disease \citep{who2014ebola}.
+All other parameters are the same as in Fig. 5 in the main text.
 }
@@
 \caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
-This figure matches Fig. 4 in the main text.
-We assume that infected indidividuals have 30\% binomial probability of reporting their generation intervals (time between their infection and the infection of their infectors) when we apply population- and individual-based methods to estimate \Rini.
+This figure matches Fig. 5 in the main text, except that we assume that not all generation intervals are observed. 
+Instead, we assume that each generation interval independently has a 30\% probability of being reported when we apply population- and individual-based methods to estimate \Rini.
+Independence is a simplifying assumption that should not affect the conclusions.
 We still obtain unbiased estimates of \Rini when generation intervals are under-reported.
-All other parameters and \Rini estimates (boxplots) are the same as in Fig. 4 in the main text.
+All parameters (and boxplots not based on these methods) are the same as in Fig. 5 in the main text.
 }
@@
 \includegraphics[width=\textwidth]{../fig/full_coverage_fig.pdf}
-\caption{\textbf{Wrong distributional assumption may result in biased estimates of the parameters of a generation-interval distribution.}
+\caption{\textbf{Wrong distributional assumptions may result in biased estimates of the parameters of a generation-interval distribution.}
 We simulate a stochastic SEIR model on a homogenous network with $10^5$ individuals using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$.
@@
 (E) Estimated $r$--$\RR$ relationships based on the gamma approximation and a true $r$--$\RR$ relationship of an SEIR model.
-Even though there are biases in the estimates of the parameters of the generation-interval distribution, we can still get an unbiased estimate of the $r$--$\RR$ relationship: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
+Even though there are biases in the estimates of the parameters of the generation-interval distribution, the $r$--$\RR$ relationship is much less biased: a shorter mean generation interval decreases $\RR$ whereas a tighter generation-interval distribution increases $\RR$ \citep{wallinga2007generation, park2019practical}.
 }
